---
title: "Factor Foundations"
---

```{r setup, include=FALSE}
box::use(
    ../modules/data[load_market, load_factors, filter_dates, merge_with_factors],
    ../modules/stats[sharpe_ratio, annualised_return, annualised_vol, max_drawdown],
    ../modules/viz[theme_trading, trading_colors]
)

box::use(
    data.table[...],
    ggplot2[...]
)

tc <- unlist(trading_colors)

knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 10,
    fig.height = 6,
    fig.path = "../figures/09-1_"
)

set.seed(42)
```

# Factor Foundations

Factor investing represents one of the most significant intellectual achievements in quantitative finance. Rather than picking individual stocks, factor strategies systematically harvest *characteristics* that have historically predicted returns.

The approach is simple: sort stocks by a characteristic (like book-to-market ratio), go long the top decile, short the bottom decile, and collect the premium. The challenge: understanding *why* these premia exist and whether they'll persist.

---

## 9.1 What Are Factors?

### 9.1.1 Prose/Intuition

A **factor** is a measurable characteristic that explains cross-sectional differences in expected returns. The idea is foundational: if you can identify what makes some stocks outperform others, you can build portfolios that systematically capture that outperformance.

**The Fama-French Revolution:**

Eugene Fama and Kenneth French transformed asset pricing in 1992-1993. Their key insight: the CAPM's single-factor model (market beta) couldn't explain why small stocks and value stocks consistently outperformed. They introduced two additional factors:
- **SMB (Small Minus Big):** Small-cap stocks outperform large-caps
- **HML (High Minus Low):** High book-to-market (value) stocks outperform low book-to-market (growth) stocks

**Risk vs Behavioural Explanations:**

Why do factors earn premia? Two camps:
1. **Risk-based:** Factor premia compensate for systematic risk. Value stocks are distressed and risky; investors demand higher returns.
2. **Behavioural:** Factor premia arise from investor mistakes. Overconfidence, extrapolation, and slow reaction to information create mispricings.

Neither explanation is definitively proven. Both probably contribute.

**The Factor Zoo:**

By 2020, researchers had documented over 400 "significant" factors. Most are spurious—products of data mining. The robust factors that survive scrutiny:
- Market (Mkt-RF)
- Size (SMB)
- Value (HML)
- Momentum (UMD/WML)
- Profitability (RMW)
- Investment (CMA)
- Low volatility/beta (BAB)

### 9.1.2 Visual Evidence

```{r factor-cumulative-returns, fig.cap="Cumulative returns of Fama-French factors since 1963. Market dominates in magnitude; other factors show distinct risk/return profiles."}
# Load Fama-French 5-factor data
ff5 <- load_factors("ff5", freq = "daily")
ff5 <- filter_dates(ff5, as.Date("1963-07-01"), as.Date("2023-12-31"))

# Calculate cumulative returns
ff5[, cum_mkt := cumprod(1 + Mkt_RF) - 1]
ff5[, cum_smb := cumprod(1 + SMB) - 1]
ff5[, cum_hml := cumprod(1 + HML) - 1]
ff5[, cum_rmw := cumprod(1 + RMW) - 1]
ff5[, cum_cma := cumprod(1 + CMA) - 1]

# Prepare for plotting
factors_long <- melt(
    ff5[, .(date, `Market (Mkt-RF)` = cum_mkt, `Size (SMB)` = cum_smb,
            `Value (HML)` = cum_hml, `Profitability (RMW)` = cum_rmw,
            `Investment (CMA)` = cum_cma)],
    id.vars = "date", variable.name = "Factor", value.name = "Cumulative_Return"
)

# Plot all factors
ggplot(factors_long, aes(x = date, y = Cumulative_Return, colour = Factor)) +
    geom_line(linewidth = 0.5) +
    scale_colour_manual(values = c(tc[1], tc[2], tc[3], tc[4], tc[5])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Fama-French Factor Cumulative Returns (1963-2023)",
         subtitle = "60 years of factor data reveals distinct patterns",
         x = NULL, y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")
```

```{r factor-annual-returns, fig.cap="Annual returns by factor. Factors have good and bad years; diversification across factors reduces volatility."}
# Calculate annual returns
ff5[, year := year(date)]

annual_returns <- ff5[, .(
    Mkt_RF = prod(1 + Mkt_RF) - 1,
    SMB = prod(1 + SMB) - 1,
    HML = prod(1 + HML) - 1,
    RMW = prod(1 + RMW) - 1,
    CMA = prod(1 + CMA) - 1
), by = year]

# Filter to recent decades for readability
annual_returns <- annual_returns[year >= 2000]

annual_long <- melt(annual_returns, id.vars = "year",
                    variable.name = "Factor", value.name = "Return")

ggplot(annual_long, aes(x = year, y = Return, fill = Factor)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = c(tc[1], tc[2], tc[3], tc[4], tc[5])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Annual Factor Returns (2000-2023)",
         subtitle = "Each factor has distinct performance cycles",
         x = NULL, y = "Annual Return") +
    theme_trading() +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 45, hjust = 1))
```

### 9.1.3 Mathematical Derivation

**The Factor Model:**

Asset returns are explained by exposure to common factors:

$$
r_i = \alpha_i + \sum_{k=1}^{K} \beta_{ik} f_k + \varepsilon_i
$$

where:
- $r_i$: Return of asset $i$
- $\alpha_i$: Alpha (unexplained return)
- $\beta_{ik}$: Loading of asset $i$ on factor $k$
- $f_k$: Return of factor $k$
- $\varepsilon_i$: Idiosyncratic (asset-specific) return

**Expected Return:**

Taking expectations:
$$
E[r_i] = \alpha_i + \sum_{k=1}^{K} \beta_{ik} E[f_k]
$$

If the model is well-specified, $\alpha_i = 0$ and expected returns are entirely determined by factor exposures.

**Factor Premium:**

The factor premium is the expected return of the factor:
$$
\lambda_k = E[f_k]
$$

Estimated from the time-series average:
$$
\hat{\lambda}_k = \frac{1}{T} \sum_{t=1}^{T} f_{k,t}
$$

**Alpha as Risk-Adjusted Return:**

Alpha measures performance after accounting for factor exposures:
$$
\alpha_i = \bar{r}_i - \sum_{k=1}^{K} \hat{\beta}_{ik} \bar{f}_k
$$

A positive alpha means the asset outperformed what its factor exposures would predict.

**Pricing vs Characteristic Factors:**

- **Pricing factors:** Explain cross-section of expected returns (CAPM, Fama-French)
- **Characteristic factors:** Observable firm characteristics used to construct portfolios (B/M, momentum)

The distinction matters: characteristics predict returns, but factors explain returns.

```{r factor-statistics, fig.cap="Factor summary statistics. Annualised returns, volatility, and Sharpe ratios for each factor."}
# Compute factor statistics
factor_stats <- data.table(
    Factor = c("Market (Mkt-RF)", "Size (SMB)", "Value (HML)",
               "Profitability (RMW)", "Investment (CMA)"),
    `Ann. Return` = sprintf("%.1f%%", c(
        annualised_return(ff5$Mkt_RF) * 100,
        annualised_return(ff5$SMB) * 100,
        annualised_return(ff5$HML) * 100,
        annualised_return(ff5$RMW) * 100,
        annualised_return(ff5$CMA) * 100)),
    `Ann. Vol` = sprintf("%.1f%%", c(
        sd(ff5$Mkt_RF, na.rm = TRUE) * sqrt(252) * 100,
        sd(ff5$SMB, na.rm = TRUE) * sqrt(252) * 100,
        sd(ff5$HML, na.rm = TRUE) * sqrt(252) * 100,
        sd(ff5$RMW, na.rm = TRUE) * sqrt(252) * 100,
        sd(ff5$CMA, na.rm = TRUE) * sqrt(252) * 100)),
    Sharpe = sprintf("%.2f", c(
        sharpe_ratio(ff5$Mkt_RF),
        sharpe_ratio(ff5$SMB),
        sharpe_ratio(ff5$HML),
        sharpe_ratio(ff5$RMW),
        sharpe_ratio(ff5$CMA)))
)

cat("=== Fama-French Factor Statistics (1963-2023) ===\n\n")
print(factor_stats)
```

### 9.1.4 Implementation & Application

```{r factor-regression, fig.cap="Factor regression example. Regressing SPY returns on Fama-French factors reveals its factor exposures."}
# Load SPY and merge with factors
spy <- load_market("SPY")
spy <- filter_dates(spy, as.Date("2000-01-01"), as.Date("2023-12-31"))
spy[, returns := c(NA, diff(log(adjusted)))]
spy <- spy[!is.na(returns)]

# Merge with factors
spy_factors <- merge(spy[, .(date, spy_ret = returns)], ff5, by = "date")
spy_factors <- spy_factors[!is.na(spy_ret)]

# Calculate excess return
spy_factors[, spy_excess := spy_ret - RF]

# Factor regression
fit_ff5 <- lm(spy_excess ~ Mkt_RF + SMB + HML + RMW + CMA, data = spy_factors)

cat("=== SPY Factor Regression (Fama-French 5-Factor) ===\n\n")
cat("Model: SPY_excess = α + β_Mkt*Mkt + β_SMB*SMB + β_HML*HML + β_RMW*RMW + β_CMA*CMA + ε\n\n")

# Extract coefficients
coefs <- summary(fit_ff5)$coefficients
coef_dt <- data.table(
    Factor = c("Alpha (annualised)", "Market", "SMB", "HML", "RMW", "CMA"),
    Estimate = c(coefs[1, 1] * 252, coefs[2:6, 1]),  # Annualise alpha
    `Std Error` = c(coefs[1, 2] * sqrt(252), coefs[2:6, 2]),
    `t-stat` = coefs[, 3],
    `p-value` = sprintf("%.4f", coefs[, 4])
)
print(coef_dt)

cat(sprintf("\nR-squared: %.1f%%\n", summary(fit_ff5)$r.squared * 100))
cat(sprintf("Annualised Alpha: %.2f%%\n", coefs[1, 1] * 252 * 100))

# Visualise betas
beta_dt <- data.table(
    Factor = c("Market", "SMB", "HML", "RMW", "CMA"),
    Beta = coefs[2:6, 1],
    SE = coefs[2:6, 2]
)

ggplot(beta_dt, aes(x = Factor, y = Beta)) +
    geom_col(fill = tc[1], alpha = 0.7) +
    geom_errorbar(aes(ymin = Beta - 1.96 * SE, ymax = Beta + 1.96 * SE),
                  width = 0.2, colour = tc[2]) +
    geom_hline(yintercept = c(0, 1), linetype = c("solid", "dashed"),
               colour = c("grey50", tc[3])) +
    labs(title = "SPY Factor Exposures (Betas)",
         subtitle = "Market beta ≈ 1; other exposures reveal style tilts",
         x = NULL, y = "Beta") +
    theme_trading()
```

---

## 9.2 Value Factor

### 9.2.1 Prose/Intuition

The value factor buys cheap stocks (high book-to-market) and sells expensive stocks (low book-to-market). Graham and Dodd pioneered this approach in the 1930s; Fama and French documented it statistically in 1992.

**Why Value Might Work:**

Risk-based explanations:
- Value stocks are often distressed companies facing bankruptcy risk
- They have high operating leverage and are sensitive to economic cycles
- Investors demand compensation for holding these risks

Behavioural explanations:
- Investors extrapolate past performance; they overpay for glamour stocks
- Overconfidence in growth forecasts leads to disappointment
- Mean reversion in profitability surprises investors

**The Value Crisis (2010s):**

Value experienced its worst decade on record from 2010-2020. Growth stocks (especially tech) dominated. Many questioned whether value was dead.

Possible explanations:
- Low interest rates made growth stocks more valuable (longer duration)
- Intangible assets (not captured by book value) became more important
- Sector composition shifted toward tech

Value has historically recovered from such periods, but past is not prologue.

### 9.2.2 Visual Evidence

```{r value-factor-performance, fig.cap="Value factor (HML) cumulative returns. The factor earned substantial premia historically but suffered during 2010-2020."}
# HML performance over time
ff5[, cum_hml := cumprod(1 + HML) - 1]

# Highlight periods
ggplot(ff5, aes(x = date, y = cum_hml)) +
    geom_line(colour = tc[1], linewidth = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    annotate("rect", xmin = as.Date("2010-01-01"), xmax = as.Date("2020-03-01"),
             ymin = -Inf, ymax = Inf, fill = tc[4], alpha = 0.1) +
    annotate("text", x = as.Date("2015-01-01"), y = max(ff5$cum_hml, na.rm = TRUE) * 0.8,
             label = "Value Drought\n(2010-2020)", colour = tc[4], size = 3) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Value Factor (HML) Cumulative Returns",
         subtitle = "The 2010s were historically poor for value investors",
         x = NULL, y = "Cumulative Return") +
    theme_trading()
```

```{r value-by-decade, fig.cap="Value factor returns by decade. The factor has been consistently positive except for 2010-2020."}
# Decade analysis
ff5[, decade := floor(year(date) / 10) * 10]

decade_returns <- ff5[decade >= 1970 & decade <= 2020, .(
    HML_Return = prod(1 + HML) - 1,
    Years = .N / 252
), by = decade]

decade_returns[, Ann_Return := (1 + HML_Return)^(1/Years) - 1]

ggplot(decade_returns, aes(x = factor(decade), y = Ann_Return)) +
    geom_col(fill = ifelse(decade_returns$Ann_Return > 0, tc[3], tc[4]), alpha = 0.7) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Value Factor Annualised Returns by Decade",
         subtitle = "Value worked for decades, then struggled in 2010s",
         x = "Decade", y = "Annualised Return") +
    theme_trading()
```

### 9.2.3 Mathematical Derivation

**Book-to-Market Ratio:**

$$
\text{B/M} = \frac{\text{Book Equity}}{\text{Market Cap}}
$$

Book equity = Total assets - Total liabilities (shareholders' equity).

High B/M = "cheap" (value); Low B/M = "expensive" (growth).

**Earnings Yield:**

$$
\text{E/P} = \frac{\text{EPS}}{\text{Price}} = \frac{1}{\text{P/E}}
$$

Alternative value measure; more responsive to current profitability.

**Composite Value:**

Combining multiple metrics reduces noise:

$$
\text{Value}_{i} = \frac{1}{3}\left(\text{rank}(\text{B/M}_i) + \text{rank}(\text{E/P}_i) + \text{rank}(\text{CF/P}_i)\right)
$$

where ranks are cross-sectional percentiles.

**HML Construction (Fama-French):**

1. Sort NYSE stocks by book-to-market into thirds: Low, Medium, High
2. Sort NYSE stocks by size into halves: Small, Big
3. Form 6 portfolios at intersection
4. HML = (Small High + Big High)/2 - (Small Low + Big Low)/2

This controls for size effects when measuring value.

```{r value-metrics-correlation, fig.cap="Value metrics are correlated but not identical. Using multiple metrics provides diversification."}
# Simulate value metrics correlation (illustrative)
n <- 1000
bm <- rnorm(n)
ep <- 0.6 * bm + sqrt(1 - 0.6^2) * rnorm(n)  # Correlated with B/M
cfp <- 0.5 * bm + 0.3 * ep + sqrt(1 - 0.5^2 - 0.3^2) * rnorm(n)

metrics_dt <- data.table(
    `Book/Market` = bm,
    `Earnings/Price` = ep,
    `CashFlow/Price` = cfp
)

# Correlation matrix
cor_matrix <- cor(metrics_dt)

cat("=== Value Metrics Correlation ===\n\n")
print(round(cor_matrix, 3))

# Visualise
metrics_long <- melt(metrics_dt[1:200], measure.vars = names(metrics_dt),
                     variable.name = "Metric", value.name = "Value")
metrics_long[, Stock := rep(1:200, 3)]

ggplot(metrics_dt[1:200], aes(x = `Book/Market`, y = `Earnings/Price`)) +
    geom_point(colour = tc[1], alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, colour = tc[2]) +
    labs(title = "Value Metrics: B/M vs E/P",
         subtitle = sprintf("Correlation = %.2f", cor_matrix[1, 2]),
         x = "Book/Market (rank)", y = "Earnings/Price (rank)") +
    theme_trading()
```

### 9.2.4 Implementation & Application

```{r value-factor-construction, fig.cap="Constructing a value factor from scratch using simulated data. The process: rank, sort, form portfolios, compute returns."}
# Simulate cross-section of stocks for illustration
n_stocks <- 100
n_months <- 120

# Generate characteristics and returns
sim_data <- data.table()

for (t in 1:n_months) {
    # Cross-section at time t
    bm <- rnorm(n_stocks)  # Book-to-market
    size <- rnorm(n_stocks)  # Size (market cap)

    # True factor premia (for simulation)
    value_premium <- 0.003  # 30 bps/month for high B/M
    size_premium <- 0.002   # 20 bps/month for small size

    # Generate returns with factor structure
    returns <- 0.01 + value_premium * (bm > median(bm)) -
               size_premium * (size > median(size)) +
               rnorm(n_stocks, 0, 0.05)

    sim_data <- rbind(sim_data, data.table(
        month = t,
        stock = 1:n_stocks,
        bm = bm,
        size = size,
        returns = returns
    ))
}

# Construct HML factor
sim_data[, bm_rank := frank(bm, ties.method = "average") / .N, by = month]
sim_data[, size_rank := frank(size, ties.method = "average") / .N, by = month]

# Classify
sim_data[, value_tercile := cut(bm_rank, breaks = c(0, 1/3, 2/3, 1),
                                 labels = c("Low", "Med", "High"), include.lowest = TRUE)]
sim_data[, size_half := cut(size_rank, breaks = c(0, 0.5, 1),
                             labels = c("Small", "Big"), include.lowest = TRUE)]

# Portfolio returns (equal-weighted)
portfolio_returns <- sim_data[, .(
    ret = mean(returns)
), by = .(month, value_tercile, size_half)]

# HML calculation
hml_data <- dcast(portfolio_returns, month ~ value_tercile + size_half, value.var = "ret")

# Handle potential NA column names
hml_cols <- names(hml_data)
if ("High_Small" %in% hml_cols && "High_Big" %in% hml_cols &&
    "Low_Small" %in% hml_cols && "Low_Big" %in% hml_cols) {

    hml_data[, HML := (High_Small + High_Big) / 2 - (Low_Small + Low_Big) / 2]
    hml_data[, SMB := (High_Small + Low_Small) / 2 - (High_Big + Low_Big) / 2]

    # Cumulative returns
    hml_data[, cum_HML := cumprod(1 + HML) - 1]

    # Plot
    ggplot(hml_data, aes(x = month, y = cum_HML)) +
        geom_line(colour = tc[1], linewidth = 0.7) +
        geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
        labs(title = "Constructed HML Factor (Simulated Data)",
             subtitle = "Long high B/M, short low B/M",
             x = "Month", y = "Cumulative Return") +
        theme_trading()

    cat("=== Simulated HML Factor Statistics ===\n")
    cat(sprintf("Mean monthly return: %.2f%%\n", mean(hml_data$HML, na.rm = TRUE) * 100))
    cat(sprintf("Monthly volatility: %.2f%%\n", sd(hml_data$HML, na.rm = TRUE) * 100))
    cat(sprintf("Annualised Sharpe: %.2f\n", mean(hml_data$HML, na.rm = TRUE) /
                sd(hml_data$HML, na.rm = TRUE) * sqrt(12)))
}
```

---

## 9.3 Momentum Factor

### 9.3.1 Prose/Intuition

Momentum is the tendency for recent winners to continue winning and recent losers to continue losing. Jegadeesh and Titman (1993) documented that buying stocks with high 12-month returns (skipping the most recent month) and selling stocks with low 12-month returns generated significant profits.

**Why Momentum Might Work:**

Behavioural explanations dominate:
- **Underreaction:** Investors respond slowly to news; prices adjust gradually
- **Herding:** Trend-followers amplify initial price moves
- **Confirmation bias:** Winners attract attention and more buying
- **Disposition effect:** Investors sell winners too early, hold losers too long

Risk-based explanations are harder to justify—momentum crashes are severe, but crashes happen *after* losers become winners (2009), not during systematic stress.

**Momentum Crashes:**

Momentum has the worst tail risk of any major factor. In March 2009, momentum lost ~40% in one month as beaten-down stocks (losers) rebounded violently. Similar crashes occurred in 1932 and 2020.

The pattern: after severe market downturns, when the market rebounds, former losers become big winners, and momentum portfolios (which are short these stocks) collapse.

### 9.3.2 Visual Evidence

```{r momentum-performance, fig.cap="Momentum factor (approximated via factor spread). The factor has high returns but severe crash risk."}
# Use FF5 data - we'll simulate a momentum proxy from market data
# In practice, you'd load UMD (Up Minus Down) directly

# For this example, we'll use a rolling past-return strategy on SPY
# This is illustrative; real momentum uses cross-sectional ranking

spy <- load_market("SPY")
spy <- filter_dates(spy, as.Date("1995-01-01"), as.Date("2023-12-31"))
spy[, returns := c(NA, diff(log(adjusted)))]

# 12-month (252-day) momentum signal, skip last month (21 days)
spy[, mom_signal := frollapply(returns, 252, function(x) {
    if (length(x) < 252) return(NA)
    prod(1 + x[1:231]) - 1  # First 11 months
})]

spy <- spy[!is.na(mom_signal)]

# Momentum strategy: long when momentum positive, flat/short when negative
spy[, mom_position := ifelse(mom_signal > 0, 1, -1)]
spy[, mom_ret := mom_position * returns]

# Cumulative
spy[, cum_bh := cumprod(1 + returns) - 1]
spy[, cum_mom := cumprod(1 + mom_ret) - 1]

# Plot
mom_long <- melt(spy[, .(date, `Buy & Hold` = cum_bh, `Momentum` = cum_mom)],
                 id.vars = "date", variable.name = "Strategy", value.name = "Return")

ggplot(mom_long, aes(x = date, y = Return, colour = Strategy)) +
    geom_line(linewidth = 0.6) +
    annotate("rect", xmin = as.Date("2009-01-01"), xmax = as.Date("2009-06-01"),
             ymin = -Inf, ymax = Inf, fill = tc[4], alpha = 0.2) +
    annotate("text", x = as.Date("2009-03-01"), y = max(mom_long$Return) * 0.9,
             label = "2009\nCrash", colour = tc[4], size = 3) +
    scale_colour_manual(values = c(tc[1], tc[2])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Time-Series Momentum Strategy (SPY)",
         subtitle = "Momentum profits but crashes hard in reversals",
         x = NULL, y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")
```

```{r momentum-crashes, fig.cap="Momentum's worst months. The factor has extreme negative skewness."}
# Worst momentum months (using SPY momentum returns)
spy[, month := format(date, "%Y-%m")]
monthly_mom <- spy[, .(
    mom_ret = prod(1 + mom_ret) - 1
), by = month]

# Histogram
ggplot(monthly_mom, aes(x = mom_ret)) +
    geom_histogram(bins = 50, fill = tc[1], alpha = 0.7) +
    geom_vline(xintercept = quantile(monthly_mom$mom_ret, 0.01),
               colour = tc[4], linetype = "dashed") +
    scale_x_continuous(labels = scales::percent_format()) +
    labs(title = "Distribution of Monthly Momentum Returns",
         subtitle = "Left tail shows crash risk; 1% quantile marked",
         x = "Monthly Return", y = "Count") +
    theme_trading()

# Worst months
cat("=== Worst Momentum Months ===\n\n")
print(monthly_mom[order(mom_ret)][1:10])
```

### 9.3.3 Mathematical Derivation

**Momentum Signal:**

12-1 month momentum (skip most recent month due to short-term reversal):

$$
\text{Mom}_{i,t} = \prod_{s=t-12}^{t-2}(1 + r_{i,s}) - 1
$$

Equivalently using log returns:
$$
\text{Mom}_{i,t} = \sum_{s=t-12}^{t-2} r_{i,s}
$$

**Cross-Sectional Ranking:**

Rank all stocks by momentum signal:
$$
\text{Rank}_i = \frac{\text{Position of } \text{Mom}_i \text{ among all stocks}}{N}
$$

**Portfolio Construction:**

- **Winners:** Top decile (rank > 0.9)
- **Losers:** Bottom decile (rank < 0.1)
- **UMD (WML):** Winners - Losers (Up Minus Down, or Winners Minus Losers)

**Expected Return Under Momentum:**

If momentum persists:
$$
E[r_{t+1} | r_{t-12:t-2}] = \alpha + \beta \cdot \text{Mom}_{t}
$$

with $\beta > 0$.

**Momentum Crash Condition:**

Momentum crashes when:
1. Market has crashed (large negative returns)
2. Market starts to rebound
3. Past losers (now in short leg) surge
4. Past winners (now in long leg) lag

Technically: $\text{Corr}(\text{Mom}_{t-1}, r_t) < 0$ during reversals.

```{r momentum-decay, fig.cap="Momentum signal decay. Returns are predictable for 3-12 months; reversal occurs after 3-5 years."}
# Simulate momentum decay pattern (stylised)
horizons <- 1:60  # months

# Expected return given past 12-month momentum
# Positive initially, reverses after ~36 months
expected_return <- sapply(horizons, function(h) {
    if (h <= 12) {
        0.5 * (1 - h/12)  # Decay over 12 months
    } else if (h <= 36) {
        0  # Flat
    } else {
        -0.3 * (1 - exp(-(h - 36)/12))  # Reversal
    }
})

decay_dt <- data.table(
    Horizon = horizons,
    `Expected Return` = expected_return
)

ggplot(decay_dt, aes(x = Horizon, y = `Expected Return`)) +
    geom_line(colour = tc[1], linewidth = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = c(12, 36), linetype = "dotted", colour = tc[2]) +
    annotate("text", x = 6, y = 0.4, label = "Momentum\nProfitable", size = 3) +
    annotate("text", x = 48, y = -0.2, label = "Long-term\nReversal", size = 3) +
    labs(title = "Momentum Signal Decay (Stylised)",
         subtitle = "Momentum works for ~12 months, then reverses",
         x = "Holding Period (months)", y = "Expected Return (normalised)") +
    theme_trading()
```

### 9.3.4 Implementation & Application

```{r cross-sectional-momentum, fig.cap="Cross-sectional momentum: ranking stocks by past returns. Winners and losers diverge over time."}
# Simulate cross-sectional momentum
n_stocks <- 50
n_months <- 60

# Generate returns with momentum structure
set.seed(42)
cs_data <- data.table()

# Initial momentum signals
signals <- rnorm(n_stocks)

for (t in 1:n_months) {
    # Returns partially depend on lagged signal
    returns <- 0.01 + 0.05 * signals + rnorm(n_stocks, 0, 0.1)

    cs_data <- rbind(cs_data, data.table(
        month = t,
        stock = 1:n_stocks,
        signal = signals,
        returns = returns
    ))

    # Update signals with autocorrelation
    signals <- 0.8 * signals + 0.2 * (returns - 0.01) / 0.1 + rnorm(n_stocks, 0, 0.3)
}

# Cross-sectional momentum strategy
cs_data[, signal_rank := frank(signal, ties.method = "average") / .N, by = month]

# Form quintiles
cs_data[, quintile := cut(signal_rank, breaks = seq(0, 1, 0.2),
                           labels = c("Q1 (Losers)", "Q2", "Q3", "Q4", "Q5 (Winners)"),
                           include.lowest = TRUE)]

# Quintile returns
quintile_returns <- cs_data[, .(ret = mean(returns)), by = .(month, quintile)]
quintile_wide <- dcast(quintile_returns, month ~ quintile, value.var = "ret")

# Calculate WML
quintile_wide[, WML := `Q5 (Winners)` - `Q1 (Losers)`]

# Cumulative returns by quintile
quintile_cum <- quintile_returns[, .(cum_ret = cumprod(1 + ret) - 1), by = .(quintile, month)]

ggplot(quintile_cum, aes(x = month, y = cum_ret, colour = quintile)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[4], tc[5], "grey50", tc[2], tc[3])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Cross-Sectional Momentum: Quintile Performance",
         subtitle = "Winners outperform losers persistently",
         x = "Month", y = "Cumulative Return", colour = "Quintile") +
    theme_trading() +
    theme(legend.position = "right")
```

---

## 9.4 Size, Quality, and Low Volatility

### 9.4.1 Prose/Intuition

Beyond value and momentum, three other factors have robust empirical support:

**Size (SMB):**
Small-cap stocks have historically outperformed large-caps. Explanations include:
- Higher risk and illiquidity require premium
- Less analyst coverage creates mispricings
- But: the size premium has weakened since its discovery (1981)

**Quality (QMJ - Quality Minus Junk):**
Profitable, stable, growing companies with low leverage outperform unprofitable, volatile companies with high leverage. Quality factors include:
- Profitability (ROE, gross profit/assets)
- Earnings stability (low earnings volatility)
- Growth (earnings and sales growth)
- Low leverage (debt/equity)

Quality is "flight to safety"—it outperforms during downturns.

**Low Volatility (BAB - Betting Against Beta):**
Low-volatility and low-beta stocks have historically delivered higher risk-adjusted returns than predicted by CAPM. This contradicts theory: higher risk should mean higher return.

Explanations:
- Leverage constraints: investors who can't lever up buy high-beta stocks, overpricing them
- Lottery preferences: investors overpay for high-volatility "lottery tickets"
- Benchmarking: fund managers who are benchmarked to an index prefer high-beta stocks

### 9.4.2 Visual Evidence

```{r size-factor, fig.cap="Size factor (SMB) cumulative returns. The small-cap premium has weakened in recent decades."}
# SMB performance
ff5[, cum_smb := cumprod(1 + SMB) - 1]

ggplot(ff5, aes(x = date, y = cum_smb)) +
    geom_line(colour = tc[2], linewidth = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    annotate("rect", xmin = as.Date("1980-01-01"), xmax = as.Date("2000-01-01"),
             ymin = -Inf, ymax = Inf, fill = tc[3], alpha = 0.1) +
    annotate("text", x = as.Date("1990-01-01"), y = max(ff5$cum_smb, na.rm = TRUE) * 0.8,
             label = "Size premium\nstrong (1980-2000)", colour = tc[3], size = 3) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Size Factor (SMB) Cumulative Returns",
         subtitle = "Small minus Big; premium has weakened since discovery",
         x = NULL, y = "Cumulative Return") +
    theme_trading()
```

```{r factor-correlations, fig.cap="Factor correlation matrix. Factors have low correlations, providing diversification benefits."}
# Factor correlations
factor_cols <- c("Mkt_RF", "SMB", "HML", "RMW", "CMA")
cor_matrix <- cor(ff5[, ..factor_cols], use = "complete.obs")

# Convert to long format for plotting
cor_dt <- as.data.table(cor_matrix, keep.rownames = "Factor1")
cor_long <- melt(cor_dt, id.vars = "Factor1", variable.name = "Factor2", value.name = "Correlation")

ggplot(cor_long, aes(x = Factor1, y = Factor2, fill = Correlation)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", Correlation)), colour = "white", size = 3) +
    scale_fill_gradient2(low = tc[4], mid = "white", high = tc[3],
                         midpoint = 0, limits = c(-1, 1)) +
    labs(title = "Fama-French Factor Correlation Matrix",
         subtitle = "Low correlations enable diversification",
         x = NULL, y = NULL) +
    theme_trading() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 9.4.3 Mathematical Derivation

**Size Factor (SMB):**

$$
\text{SMB} = \frac{1}{3}\left(\text{S/L} + \text{S/M} + \text{S/H}\right) - \frac{1}{3}\left(\text{B/L} + \text{B/M} + \text{B/H}\right)
$$

where S = Small, B = Big, and L/M/H = Low/Medium/High value.

**Quality Metrics:**

Profitability:
$$
\text{ROE} = \frac{\text{Net Income}}{\text{Shareholders' Equity}}
$$

Gross profitability (Novy-Marx):
$$
\text{GP/A} = \frac{\text{Revenue} - \text{COGS}}{\text{Total Assets}}
$$

**Betting Against Beta (BAB):**

Frazzini and Pedersen (2014):

1. Estimate beta for each stock
2. Rank by beta: Low-beta portfolio (L), High-beta portfolio (H)
3. Lever low-beta to beta = 1: $w_L = 1/\beta_L$
4. Delever high-beta to beta = 1: $w_H = 1/\beta_H$

$$
\text{BAB} = \frac{r_L - r_f}{\beta_L} - \frac{r_H - r_f}{\beta_H}
$$

**RMW (Robust Minus Weak - Profitability):**

$$
\text{RMW} = \frac{1}{2}(\text{Small Robust} + \text{Big Robust}) - \frac{1}{2}(\text{Small Weak} + \text{Big Weak})
$$

**CMA (Conservative Minus Aggressive - Investment):**

$$
\text{CMA} = \frac{1}{2}(\text{Small Conservative} + \text{Big Conservative}) - \frac{1}{2}(\text{Small Aggressive} + \text{Big Aggressive})
$$

Conservative = low asset growth; Aggressive = high asset growth.

### 9.4.4 Implementation & Application

```{r quality-factor-proxy, fig.cap="Quality factor proxy: profitability (RMW). Quality tends to outperform during market stress."}
# RMW performance
ff5[, cum_rmw := cumprod(1 + RMW) - 1]

# Identify stress periods (market drops > 10% in 60 days)
ff5[, mkt_60d := frollapply(Mkt_RF, 60, function(x) prod(1 + x) - 1)]
ff5[, stress := mkt_60d < -0.10]

ggplot(ff5, aes(x = date, y = cum_rmw)) +
    geom_line(colour = tc[3], linewidth = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Profitability Factor (RMW) Cumulative Returns",
         subtitle = "Robust (profitable) minus Weak; defensive characteristics",
         x = NULL, y = "Cumulative Return") +
    theme_trading()
```

```{r multi-factor-performance, fig.cap="Combining factors improves risk-adjusted returns. Equal-weight multi-factor outperforms individual factors."}
# Simple equal-weight multi-factor
ff5[, multi_factor := (Mkt_RF + SMB + HML + RMW + CMA) / 5]
ff5[, cum_multi := cumprod(1 + multi_factor) - 1]

# Compare
compare_dt <- melt(
    ff5[, .(date, Market = cum_mkt, `Multi-Factor` = cum_multi)],
    id.vars = "date", variable.name = "Strategy", value.name = "Return"
)

ggplot(compare_dt, aes(x = date, y = Return, colour = Strategy)) +
    geom_line(linewidth = 0.6) +
    scale_colour_manual(values = c(tc[1], tc[2])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Market vs Equal-Weight Multi-Factor Portfolio",
         subtitle = "Diversification across factors smooths returns",
         x = NULL, y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")

# Statistics comparison
cat("=== Factor Combination Statistics ===\n\n")
combo_stats <- data.table(
    Strategy = c("Market", "Multi-Factor (EW)"),
    `Ann. Return` = sprintf("%.1f%%", c(
        annualised_return(ff5$Mkt_RF) * 100,
        annualised_return(ff5$multi_factor) * 100)),
    `Ann. Vol` = sprintf("%.1f%%", c(
        sd(ff5$Mkt_RF, na.rm = TRUE) * sqrt(252) * 100,
        sd(ff5$multi_factor, na.rm = TRUE) * sqrt(252) * 100)),
    Sharpe = sprintf("%.2f", c(
        sharpe_ratio(ff5$Mkt_RF),
        sharpe_ratio(ff5$multi_factor)))
)
print(combo_stats)
```

---

## Quick Reference

### Factor Definitions

| Factor | Long | Short | Premium |
|--------|------|-------|---------|
| Market (Mkt-RF) | Market portfolio | Risk-free rate | ~6% p.a. |
| Size (SMB) | Small caps | Large caps | ~2% p.a. |
| Value (HML) | High B/M | Low B/M | ~3% p.a. |
| Momentum (UMD) | Past winners | Past losers | ~8% p.a. |
| Profitability (RMW) | Profitable | Unprofitable | ~3% p.a. |
| Investment (CMA) | Conservative | Aggressive | ~2% p.a. |

### Factor Model Regression

```r
# Fama-French 5-factor regression
fit <- lm(excess_return ~ Mkt_RF + SMB + HML + RMW + CMA, data = merged_data)
summary(fit)
```

### Key Formulas

| Metric | Formula |
|--------|---------|
| Alpha | $\alpha = \bar{r} - \sum \beta_k \bar{f}_k$ |
| Factor premium | $\lambda_k = \bar{f}_k$ |
| Book-to-market | B/M = Book Equity / Market Cap |
| Momentum | $\prod_{t-12}^{t-2}(1+r_t) - 1$ |

### Data Sources

- Fama-French: [mba.tuck.dartmouth.edu/pages/faculty/ken.french/](https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html)
- AQR: [aqr.com/Insights/Datasets](https://www.aqr.com/Insights/Datasets)
- WRDS: Academic data access
