---
title: "Factor Portfolio Construction"
---

```{r setup, include=FALSE}
box::use(
    ../modules/data[load_market, load_factors, filter_dates],
    ../modules/stats[sharpe_ratio, annualised_return, annualised_vol, max_drawdown],
    ../modules/viz[theme_trading, trading_colors]
)

box::use(
    data.table[...],
    ggplot2[...]
)

tc <- unlist(trading_colors)

knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 10,
    fig.height = 6,
    fig.path = "../figures/09-2_"
)

set.seed(42)
```

# Factor Portfolio Construction

Knowing that factors exist is one thing; constructing factor portfolios is another. The details matter: how you sort, weight, and rebalance determines whether you capture the premium or lose it to costs and noise.

This chapter covers the mechanics: from raw signals to tradeable portfolios.

---

## 9.5 Sorting and Ranking

### 9.5.1 Prose/Intuition

The classic factor construction method is **portfolio sorting**:

1. Rank all stocks by the factor characteristic (e.g., book-to-market)
2. Divide into quantiles (deciles, quintiles, or terciles)
3. Go long the top quantile, short the bottom quantile
4. Hold for a period (typically monthly), then repeat

This non-parametric approach has advantages:
- No assumptions about functional form
- Robust to outliers
- Easy to interpret and implement

The Fama-French methodology uses 2×3 or 5×5 sorts to control for multiple characteristics simultaneously.

### 9.5.2 Visual Evidence

```{r univariate-sort, fig.cap="Univariate sort: ranking stocks by a single characteristic. The spread between top and bottom deciles captures the factor premium."}
# Simulate cross-section for sorting demonstration
n_stocks <- 500
n_periods <- 60

# Generate characteristics and returns
sim_data <- data.table()

for (t in 1:n_periods) {
    # Characteristic (e.g., B/M)
    characteristic <- rnorm(n_stocks)

    # True premium: characteristic predicts returns
    true_beta <- 0.03  # 3% per period spread

    returns <- 0.01 + true_beta * characteristic + rnorm(n_stocks, 0, 0.08)

    sim_data <- rbind(sim_data, data.table(
        period = t,
        stock = 1:n_stocks,
        characteristic = characteristic,
        returns = returns
    ))
}

# Cross-sectional ranking
sim_data[, char_rank := frank(characteristic, ties.method = "average") / .N, by = period]

# Form deciles
sim_data[, decile := cut(char_rank, breaks = seq(0, 1, 0.1),
                          labels = paste0("D", 1:10), include.lowest = TRUE)]

# Decile returns
decile_returns <- sim_data[, .(
    mean_ret = mean(returns),
    n = .N
), by = .(period, decile)]

# Average across periods
avg_decile <- decile_returns[, .(
    avg_return = mean(mean_ret),
    se = sd(mean_ret) / sqrt(.N)
), by = decile]

# Ensure proper ordering
avg_decile[, decile_num := as.numeric(gsub("D", "", decile))]
setorder(avg_decile, decile_num)

ggplot(avg_decile, aes(x = decile, y = avg_return)) +
    geom_col(fill = tc[1], alpha = 0.7) +
    geom_errorbar(aes(ymin = avg_return - 1.96 * se, ymax = avg_return + 1.96 * se),
                  width = 0.2, colour = tc[2]) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Univariate Sort: Average Returns by Decile",
         subtitle = "Monotonic pattern indicates factor premium",
         x = "Decile (D1 = Low, D10 = High)", y = "Average Return") +
    theme_trading()
```

```{r monotonicity-test, fig.cap="Testing for monotonicity. A well-specified factor shows generally increasing returns across deciles."}
# Cumulative returns by decile
decile_wide <- dcast(decile_returns, period ~ decile, value.var = "mean_ret")

# Calculate long-short
decile_wide[, `Long-Short (D10-D1)` := D10 - D1]
decile_wide[, cum_ls := cumprod(1 + `Long-Short (D10-D1)`) - 1]
decile_wide[, cum_d1 := cumprod(1 + D1) - 1]
decile_wide[, cum_d10 := cumprod(1 + D10) - 1]

# Plot
ls_long <- melt(decile_wide[, .(period, `D1 (Low)` = cum_d1, `D10 (High)` = cum_d10,
                                 `Long-Short` = cum_ls)],
                id.vars = "period", variable.name = "Portfolio", value.name = "Return")

ggplot(ls_long, aes(x = period, y = Return, colour = Portfolio)) +
    geom_line(linewidth = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    scale_colour_manual(values = c(tc[4], tc[3], tc[1])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Factor Portfolio Performance: Long vs Short vs Long-Short",
         subtitle = "Long-short captures premium regardless of market direction",
         x = "Period", y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")
```

### 9.5.3 Mathematical Derivation

**Cross-Sectional Ranking:**

For characteristic $C_i$ of stock $i$:
$$
\text{Rank}_i = \frac{\#\{j : C_j \le C_i\}}{N}
$$

This maps any distribution to uniform [0, 1].

**Quantile Breakpoints:**

For $K$ quantiles, breakpoints are:
$$
q_k = F^{-1}\left(\frac{k}{K}\right), \quad k = 1, \ldots, K-1
$$

where $F$ is the CDF. Stocks fall into quantile $k$ if $q_{k-1} < C_i \le q_k$.

**Univariate Sort:**

Assign stock $i$ to portfolio $P_k$ if:
$$
\frac{k-1}{K} < \text{Rank}_i \le \frac{k}{K}
$$

**Bivariate Sort (Independent):**

Sort independently on two characteristics, creating $K_1 \times K_2$ portfolios.

Example (2×3 Fama-French):
- Size: Small (S), Big (B) based on median NYSE market cap
- Value: Low (L), Medium (M), High (H) based on 30th/70th percentiles of NYSE B/M

6 portfolios: S/L, S/M, S/H, B/L, B/M, B/H

**Bivariate Sort (Dependent):**

Sort first on one characteristic, then within each group sort on the second.

Example:
1. Sort into size quintiles
2. Within each size quintile, sort into value quintiles
3. Creates 25 portfolios

Dependent sorts control more precisely but create smaller portfolios.

**Long-Short Factor Return:**

$$
f_t = \bar{r}_{t, \text{High}} - \bar{r}_{t, \text{Low}}
$$

where $\bar{r}$ is the (weighted) average return of stocks in each group.

```{r bivariate-sort, fig.cap="Bivariate sort: 5×5 portfolios sorted independently on two characteristics. Allows examining interaction effects."}
# Simulate two characteristics
n_stocks <- 500
n_periods <- 60

bi_data <- data.table()

for (t in 1:n_periods) {
    char1 <- rnorm(n_stocks)  # e.g., Size
    char2 <- rnorm(n_stocks)  # e.g., Value

    # Both characteristics predict returns
    returns <- 0.01 + 0.02 * char1 + 0.03 * char2 + rnorm(n_stocks, 0, 0.08)

    bi_data <- rbind(bi_data, data.table(
        period = t,
        stock = 1:n_stocks,
        char1 = char1,
        char2 = char2,
        returns = returns
    ))
}

# Bivariate ranking (5×5)
bi_data[, char1_quintile := cut(frank(char1) / .N, breaks = seq(0, 1, 0.2),
                                 labels = paste0("C1_Q", 1:5), include.lowest = TRUE), by = period]
bi_data[, char2_quintile := cut(frank(char2) / .N, breaks = seq(0, 1, 0.2),
                                 labels = paste0("C2_Q", 1:5), include.lowest = TRUE), by = period]

# 25 portfolio returns
bi_port_returns <- bi_data[, .(mean_ret = mean(returns)),
                            by = .(period, char1_quintile, char2_quintile)]

# Average across time
bi_avg <- bi_port_returns[, .(avg_ret = mean(mean_ret)),
                           by = .(char1_quintile, char2_quintile)]

# Extract quintile numbers
bi_avg[, c1_num := as.numeric(gsub("C1_Q", "", char1_quintile))]
bi_avg[, c2_num := as.numeric(gsub("C2_Q", "", char2_quintile))]

# Heatmap
ggplot(bi_avg, aes(x = factor(c1_num), y = factor(c2_num), fill = avg_ret)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.1f%%", avg_ret * 100)), colour = "white", size = 3) +
    scale_fill_gradient2(low = tc[4], mid = "grey90", high = tc[3],
                         midpoint = median(bi_avg$avg_ret),
                         labels = scales::percent_format()) +
    labs(title = "Bivariate Sort: 5×5 Portfolio Returns",
         subtitle = "Both characteristics contribute to returns",
         x = "Characteristic 1 Quintile", y = "Characteristic 2 Quintile",
         fill = "Avg Return") +
    theme_trading()
```

### 9.5.4 Implementation & Application

```{r sorting-functions, fig.cap="Sorting implementation in R. Functions for univariate and bivariate portfolio sorts."}
# Sorting function
create_quantile_portfolios <- function(data, char_col, n_quantiles = 10,
                                        weight_col = NULL) {
    # Cross-sectional ranking
    data[, rank := frank(get(char_col), ties.method = "average") / .N, by = period]

    # Assign to quantiles
    breaks <- seq(0, 1, 1/n_quantiles)
    labels <- paste0("Q", 1:n_quantiles)
    data[, portfolio := cut(rank, breaks = breaks, labels = labels, include.lowest = TRUE)]

    # Calculate portfolio returns
    if (is.null(weight_col)) {
        # Equal weight
        port_ret <- data[, .(port_ret = mean(returns)), by = .(period, portfolio)]
    } else {
        # Value weight
        data[, weight_norm := get(weight_col) / sum(get(weight_col)), by = .(period, portfolio)]
        port_ret <- data[, .(port_ret = sum(returns * weight_norm)), by = .(period, portfolio)]
    }

    return(port_ret)
}

# Apply to simulated data
port_returns <- create_quantile_portfolios(sim_data, "characteristic", n_quantiles = 5)

# Wide format
port_wide <- dcast(port_returns, period ~ portfolio, value.var = "port_ret")
port_wide[, `Long-Short` := Q5 - Q1]

# Statistics
cat("=== Quintile Portfolio Statistics ===\n\n")
quintile_stats <- port_returns[, .(
    `Ann. Return` = sprintf("%.1f%%", mean(port_ret) * 12 * 100),
    `Ann. Vol` = sprintf("%.1f%%", sd(port_ret) * sqrt(12) * 100),
    Sharpe = sprintf("%.2f", mean(port_ret) / sd(port_ret) * sqrt(12))
), by = portfolio]
print(quintile_stats)

cat("\nLong-Short (Q5 - Q1):\n")
cat(sprintf("Ann. Return: %.1f%%\n", mean(port_wide$`Long-Short`) * 12 * 100))
cat(sprintf("Ann. Vol: %.1f%%\n", sd(port_wide$`Long-Short`) * sqrt(12) * 100))
cat(sprintf("Sharpe: %.2f\n", mean(port_wide$`Long-Short`) / sd(port_wide$`Long-Short`) * sqrt(12)))
```

---

## 9.6 Factor Weighting Schemes

### 9.6.1 Prose/Intuition

How you weight positions within a factor portfolio significantly affects performance:

**Equal Weight:**
- Each stock gets the same weight: $w_i = 1/N$
- Pros: Simple, diversified, tilts toward small caps
- Cons: High turnover, capacity constrained

**Value Weight:**
- Weight proportional to market cap: $w_i = \text{MCap}_i / \sum \text{MCap}$
- Pros: Investable at scale, matches market structure
- Cons: Concentrated in large stocks, dilutes factor exposure

**Signal Weight (z-score):**
- Weight proportional to signal strength: $w_i = z_i / \sum|z_i|$
- Pros: Stronger factor exposure, exploits signal intensity
- Cons: Concentrated in extreme stocks, higher turnover

**Risk Parity:**
- Weight inversely proportional to volatility: $w_i \propto 1/\sigma_i$
- Pros: Balanced risk contribution
- Cons: Complex, requires volatility estimates

### 9.6.2 Visual Evidence

```{r weighting-comparison, fig.cap="Equal vs value-weighted factor portfolios. Equal weight gives stronger factor exposure but lower capacity."}
# Simulate with size (market cap) variation
n_stocks <- 200
n_periods <- 60

weight_data <- data.table()

for (t in 1:n_periods) {
    # Characteristics
    signal <- rnorm(n_stocks)
    size <- rlnorm(n_stocks, meanlog = 5, sdlog = 1)  # Log-normal market cap

    # Small stocks have higher signal-return relationship (size anomaly interaction)
    returns <- 0.01 + 0.04 * signal * (1 / sqrt(size/mean(size))) + rnorm(n_stocks, 0, 0.08)

    weight_data <- rbind(weight_data, data.table(
        period = t,
        stock = 1:n_stocks,
        signal = signal,
        size = size,
        returns = returns
    ))
}

# Rank by signal
weight_data[, signal_rank := frank(signal) / .N, by = period]
weight_data[, quintile := cut(signal_rank, breaks = seq(0, 1, 0.2),
                               labels = paste0("Q", 1:5), include.lowest = TRUE), by = period]

# Equal-weighted returns
ew_returns <- weight_data[quintile %in% c("Q1", "Q5"), .(
    ew_ret = mean(returns)
), by = .(period, quintile)]
ew_wide <- dcast(ew_returns, period ~ quintile, value.var = "ew_ret")
ew_wide[, ew_ls := Q5 - Q1]

# Value-weighted returns
vw_returns <- weight_data[quintile %in% c("Q1", "Q5"), {
    w <- size / sum(size)
    .(vw_ret = sum(returns * w))
}, by = .(period, quintile)]
vw_wide <- dcast(vw_returns, period ~ quintile, value.var = "vw_ret")
vw_wide[, vw_ls := Q5 - Q1]

# Combine
compare_wide <- merge(ew_wide[, .(period, EW = ew_ls)],
                      vw_wide[, .(period, VW = vw_ls)], by = "period")

compare_wide[, cum_ew := cumprod(1 + EW) - 1]
compare_wide[, cum_vw := cumprod(1 + VW) - 1]

# Plot
compare_long <- melt(compare_wide[, .(period, `Equal Weight` = cum_ew, `Value Weight` = cum_vw)],
                     id.vars = "period", variable.name = "Weighting", value.name = "Return")

ggplot(compare_long, aes(x = period, y = Return, colour = Weighting)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[1], tc[2])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Equal vs Value-Weighted Factor Portfolio",
         subtitle = "Equal weight often captures more premium but has less capacity",
         x = "Period", y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")

# Statistics comparison
cat("=== Weighting Scheme Comparison ===\n\n")
weight_stats <- data.table(
    Scheme = c("Equal Weight", "Value Weight"),
    `Ann. Return` = sprintf("%.1f%%", c(
        mean(compare_wide$EW) * 12 * 100,
        mean(compare_wide$VW) * 12 * 100)),
    `Ann. Vol` = sprintf("%.1f%%", c(
        sd(compare_wide$EW) * sqrt(12) * 100,
        sd(compare_wide$VW) * sqrt(12) * 100)),
    Sharpe = sprintf("%.2f", c(
        mean(compare_wide$EW) / sd(compare_wide$EW) * sqrt(12),
        mean(compare_wide$VW) / sd(compare_wide$VW) * sqrt(12)))
)
print(weight_stats)
```

### 9.6.3 Mathematical Derivation

**Equal Weight:**
$$
w_i = \frac{1}{N}
$$

Portfolio return: $r_p = \frac{1}{N} \sum_{i=1}^{N} r_i$

**Value Weight:**
$$
w_i = \frac{\text{MCap}_i}{\sum_{j} \text{MCap}_j}
$$

Portfolio return: $r_p = \sum_{i} w_i r_i$

**Signal Weight (z-score normalised):**

First standardise the signal:
$$
z_i = \frac{C_i - \bar{C}}{\sigma_C}
$$

Then normalise to sum to 1 (for long) and -1 (for short):
$$
w_i = \frac{z_i}{\sum_{j: z_j > 0} z_j} \text{ for longs}, \quad w_i = \frac{z_i}{\sum_{j: z_j < 0} |z_j|} \text{ for shorts}
$$

**Capacity Constraint:**

Maximum investable position:
$$
w_i \le c \times \frac{\text{ADV}_i}{AUM}
$$

where ADV = Average Daily Volume, AUM = Assets Under Management, and $c$ is a participation rate (typically 5-10%).

**Turnover:**

One-way turnover:
$$
\text{TO} = \frac{1}{2} \sum_{i} |w_{i,t} - w_{i,t-1}|
$$

Cost-adjusted return:
$$
r_{\text{net}} = r_{\text{gross}} - c \times \text{TO}
$$

### 9.6.4 Implementation & Application

```{r signal-weighting, fig.cap="Signal-weighted (z-score) portfolio. Positions scale with signal strength for stronger factor exposure."}
# Signal-weighted implementation
weight_data[, z_score := (signal - mean(signal)) / sd(signal), by = period]

# Calculate signal-weighted returns
sw_returns <- weight_data[, {
    # Long positions (z > 0)
    long <- .SD[z_score > 0]
    if (nrow(long) > 0) {
        long_w <- long$z_score / sum(long$z_score)
        long_ret <- sum(long$returns * long_w)
    } else {
        long_ret <- 0
    }

    # Short positions (z < 0)
    short <- .SD[z_score < 0]
    if (nrow(short) > 0) {
        short_w <- abs(short$z_score) / sum(abs(short$z_score))
        short_ret <- sum(short$returns * short_w)
    } else {
        short_ret <- 0
    }

    .(sw_ret = long_ret - short_ret)
}, by = period]

# Compare all three
all_compare <- merge(
    merge(ew_wide[, .(period, EW = ew_ls)], vw_wide[, .(period, VW = vw_ls)], by = "period"),
    sw_returns[, .(period, SW = sw_ret)], by = "period"
)

all_compare[, cum_ew := cumprod(1 + EW) - 1]
all_compare[, cum_vw := cumprod(1 + VW) - 1]
all_compare[, cum_sw := cumprod(1 + SW) - 1]

all_long <- melt(all_compare[, .(period, `Equal Weight` = cum_ew,
                                  `Value Weight` = cum_vw, `Signal Weight` = cum_sw)],
                 id.vars = "period", variable.name = "Weighting", value.name = "Return")

ggplot(all_long, aes(x = period, y = Return, colour = Weighting)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[1], tc[2], tc[3])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Factor Portfolio: Three Weighting Schemes",
         subtitle = "Signal weight maximises factor exposure but may lack capacity",
         x = "Period", y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")
```

---

## 9.7 Multi-Factor Portfolios

### 9.7.1 Prose/Intuition

Single factors are volatile. Value had a terrible 2010s; momentum crashes periodically. Combining multiple factors provides diversification.

**Why Combine Factors?**

1. **Lower correlation:** Factors have low correlations with each other (often near zero)
2. **Smoother returns:** When one factor underperforms, others may compensate
3. **Higher Sharpe:** Combined portfolio has better risk-adjusted returns than individuals

**How to Combine:**

1. **Portfolio mixing:** Hold separate factor portfolios, combine at portfolio level
2. **Signal mixing:** Average signals, sort on composite score
3. **Integrated:** Optimise weights on individual stocks subject to multiple factor constraints

**Trade-offs:**

- Signal mixing creates more balanced factor exposure per stock
- Portfolio mixing is simpler and preserves factor definitions
- Integrated approaches can have highest exposure but are complex

### 9.7.2 Visual Evidence

```{r multi-factor-diversification, fig.cap="Multi-factor diversification. Combining uncorrelated factors reduces volatility without sacrificing returns."}
# Load actual factor data
ff5 <- load_factors("ff5", freq = "daily")
ff5 <- filter_dates(ff5, as.Date("2000-01-01"), as.Date("2023-12-31"))

# Individual factors
factors <- c("SMB", "HML", "RMW", "CMA")

# Calculate correlations
factor_cor <- cor(ff5[, ..factors], use = "complete.obs")

cat("=== Factor Correlation Matrix ===\n\n")
print(round(factor_cor, 3))

# Equal-weight multi-factor
ff5[, multi_factor := rowMeans(.SD), .SDcols = factors]

# Cumulative returns
ff5[, cum_smb := cumprod(1 + SMB) - 1]
ff5[, cum_hml := cumprod(1 + HML) - 1]
ff5[, cum_rmw := cumprod(1 + RMW) - 1]
ff5[, cum_cma := cumprod(1 + CMA) - 1]
ff5[, cum_multi := cumprod(1 + multi_factor) - 1]

# Plot
multi_long <- melt(ff5[, .(date, SMB = cum_smb, HML = cum_hml, RMW = cum_rmw,
                            CMA = cum_cma, `Multi-Factor` = cum_multi)],
                   id.vars = "date", variable.name = "Factor", value.name = "Return")

ggplot(multi_long, aes(x = date, y = Return, colour = Factor)) +
    geom_line(linewidth = 0.5, alpha = 0.7) +
    scale_colour_manual(values = c(tc[1], tc[2], tc[3], tc[4], "black")) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Individual Factors vs Multi-Factor Portfolio",
         subtitle = "Combining factors smooths the return path",
         x = NULL, y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")
```

```{r composite-signal, fig.cap="Composite signal approach: averaging z-scores across factors creates stocks with balanced factor exposure."}
# Simulate composite signal approach
n_stocks <- 200
n_periods <- 60

multi_data <- data.table()

for (t in 1:n_periods) {
    # Multiple characteristics
    value_signal <- rnorm(n_stocks)
    momentum_signal <- rnorm(n_stocks)
    quality_signal <- rnorm(n_stocks)

    # All predict returns (different premia)
    returns <- 0.01 +
        0.025 * value_signal +
        0.030 * momentum_signal +
        0.020 * quality_signal +
        rnorm(n_stocks, 0, 0.08)

    multi_data <- rbind(multi_data, data.table(
        period = t,
        stock = 1:n_stocks,
        value = value_signal,
        momentum = momentum_signal,
        quality = quality_signal,
        returns = returns
    ))
}

# Standardise each signal
multi_data[, value_z := (value - mean(value)) / sd(value), by = period]
multi_data[, momentum_z := (momentum - mean(momentum)) / sd(momentum), by = period]
multi_data[, quality_z := (quality - mean(quality)) / sd(quality), by = period]

# Composite signal (equal weight average of z-scores)
multi_data[, composite := (value_z + momentum_z + quality_z) / 3]

# Single-factor portfolios (quintile long-short)
calc_factor_return <- function(data, signal_col) {
    data[, rank := frank(get(signal_col)) / .N, by = period]
    data[, quintile := cut(rank, breaks = seq(0, 1, 0.2),
                            labels = paste0("Q", 1:5), include.lowest = TRUE)]

    returns <- data[quintile %in% c("Q1", "Q5"), .(ret = mean(returns)), by = .(period, quintile)]
    wide <- dcast(returns, period ~ quintile, value.var = "ret")
    wide[, ls := Q5 - Q1]
    return(wide$ls)
}

value_ret <- calc_factor_return(copy(multi_data), "value")
momentum_ret <- calc_factor_return(copy(multi_data), "momentum")
quality_ret <- calc_factor_return(copy(multi_data), "quality")
composite_ret <- calc_factor_return(copy(multi_data), "composite")

# Combine
compare_methods <- data.table(
    period = 1:n_periods,
    Value = value_ret,
    Momentum = momentum_ret,
    Quality = quality_ret,
    Composite = composite_ret
)

# Also create "portfolio mixing" (average of three factor returns)
compare_methods[, `Portfolio Mix` := (Value + Momentum + Quality) / 3]

# Cumulative
compare_methods[, cum_value := cumprod(1 + Value) - 1]
compare_methods[, cum_momentum := cumprod(1 + Momentum) - 1]
compare_methods[, cum_quality := cumprod(1 + Quality) - 1]
compare_methods[, cum_composite := cumprod(1 + Composite) - 1]
compare_methods[, cum_mix := cumprod(1 + `Portfolio Mix`) - 1]

# Plot composite vs mix
mix_long <- melt(compare_methods[, .(period, `Composite Signal` = cum_composite,
                                      `Portfolio Mix` = cum_mix)],
                 id.vars = "period", variable.name = "Method", value.name = "Return")

ggplot(mix_long, aes(x = period, y = Return, colour = Method)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[1], tc[2])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Multi-Factor: Composite Signal vs Portfolio Mixing",
         subtitle = "Both methods capture diversification benefits",
         x = "Period", y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")
```

### 9.7.3 Mathematical Derivation

**Composite Score (Signal Mixing):**

$$
S_i = \sum_{k=1}^{K} w_k \times \text{rank}(C_{ik})
$$

where $C_{ik}$ is stock $i$'s value of characteristic $k$, and $w_k$ are weights (often $1/K$).

**Factor Orthogonalisation:**

To isolate one factor's contribution, regress out other factors:

$$
\tilde{C}_{ik} = C_{ik} - \sum_{j \neq k} \beta_j C_{ij}
$$

This creates a "pure" factor signal uncorrelated with others.

**Multi-Factor Alpha:**

$$
\alpha_{MF} = \bar{r} - \sum_{k=1}^{K} \hat{\beta}_k \bar{f}_k
$$

A positive multi-factor alpha means the strategy outperforms after controlling for factor exposures.

**Optimal Factor Weights:**

Mean-variance optimal weights:
$$
w^* = \frac{1}{\gamma} \Sigma^{-1} \mu
$$

where $\mu$ = expected factor returns, $\Sigma$ = factor covariance matrix, $\gamma$ = risk aversion.

In practice, naive equal weights often perform as well due to estimation error.

### 9.7.4 Implementation & Application

```{r multi-factor-stats, fig.cap="Multi-factor portfolio statistics. Combining factors improves risk-adjusted returns."}
# Statistics comparison
factor_stats <- data.table(
    Factor = c("Value", "Momentum", "Quality", "Composite", "Portfolio Mix"),
    `Ann. Return` = sprintf("%.1f%%", c(
        mean(compare_methods$Value) * 12 * 100,
        mean(compare_methods$Momentum) * 12 * 100,
        mean(compare_methods$Quality) * 12 * 100,
        mean(compare_methods$Composite) * 12 * 100,
        mean(compare_methods$`Portfolio Mix`) * 12 * 100)),
    `Ann. Vol` = sprintf("%.1f%%", c(
        sd(compare_methods$Value) * sqrt(12) * 100,
        sd(compare_methods$Momentum) * sqrt(12) * 100,
        sd(compare_methods$Quality) * sqrt(12) * 100,
        sd(compare_methods$Composite) * sqrt(12) * 100,
        sd(compare_methods$`Portfolio Mix`) * sqrt(12) * 100)),
    Sharpe = sprintf("%.2f", c(
        mean(compare_methods$Value) / sd(compare_methods$Value) * sqrt(12),
        mean(compare_methods$Momentum) / sd(compare_methods$Momentum) * sqrt(12),
        mean(compare_methods$Quality) / sd(compare_methods$Quality) * sqrt(12),
        mean(compare_methods$Composite) / sd(compare_methods$Composite) * sqrt(12),
        mean(compare_methods$`Portfolio Mix`) / sd(compare_methods$`Portfolio Mix`) * sqrt(12)))
)

cat("=== Multi-Factor Strategy Comparison ===\n\n")
print(factor_stats)
```

```{r factor-contribution, fig.cap="Factor contribution analysis. Decomposing multi-factor returns by source shows diversification at work."}
# Contribution analysis
contrib_long <- melt(compare_methods[, .(period, Value, Momentum, Quality)],
                     id.vars = "period", variable.name = "Factor", value.name = "Contribution")

# Rolling 12-period returns
contrib_long[, roll_ret := frollsum(Contribution, 12), by = Factor]

# Area chart
ggplot(contrib_long[!is.na(roll_ret)], aes(x = period, y = roll_ret, fill = Factor)) +
    geom_area(position = "stack", alpha = 0.7) +
    scale_fill_manual(values = c(tc[1], tc[2], tc[3])) +
    geom_hline(yintercept = 0, colour = "grey50") +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Rolling 12-Period Factor Contributions",
         subtitle = "Different factors contribute at different times",
         x = "Period", y = "12-Period Return Contribution") +
    theme_trading() +
    theme(legend.position = "bottom")
```

---

## Quick Reference

### Portfolio Sorting

| Type | Method | Portfolios |
|------|--------|------------|
| Univariate | Rank by one characteristic | N quantiles |
| Bivariate (independent) | Rank each characteristic separately | N₁ × N₂ |
| Bivariate (dependent) | Rank one within groups of other | N₁ × N₂ |

### Weighting Schemes

| Scheme | Formula | Pros | Cons |
|--------|---------|------|------|
| Equal | $w_i = 1/N$ | Simple, diversified | High turnover |
| Value | $w_i = \text{MCap}_i / \Sigma$ | Scalable | Diluted exposure |
| Signal | $w_i = z_i / \Sigma\|z\|$ | Strong exposure | Concentrated |

### Multi-Factor Construction

| Approach | Description | Use Case |
|----------|-------------|----------|
| Portfolio mix | Average factor portfolio returns | Transparency |
| Signal mix | Composite z-score ranking | Balanced exposure |
| Integrated | Optimise with factor constraints | Maximum efficiency |

### Key Formulas

| Metric | Formula |
|--------|---------|
| Cross-sectional rank | $\text{Rank}_i = \#\{j: C_j \le C_i\}/N$ |
| Long-short return | $f_t = \bar{r}_{High} - \bar{r}_{Low}$ |
| Composite score | $S_i = \sum w_k \text{rank}(C_{ik})$ |
| Turnover | $\text{TO} = \frac{1}{2}\sum\|w_{i,t} - w_{i,t-1}\|$ |
