---
title: "Factor Implementation"
---

```{r setup, include=FALSE}
box::use(
    ../modules/data[load_market, load_factors, filter_dates],
    ../modules/stats[sharpe_ratio, annualised_return, annualised_vol, max_drawdown],
    ../modules/viz[theme_trading, trading_colors]
)

box::use(
    data.table[...],
    ggplot2[...]
)

tc <- unlist(trading_colors)

knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 10,
    fig.height = 6,
    fig.path = "../figures/09-3_"
)

set.seed(42)
```

# Factor Implementation

Knowing that factors exist and how to construct them is academic. Implementing them profitably is practice. This chapter covers the real-world challenges: rebalancing, turnover, timing, and crowding.

---

## 9.8 Rebalancing and Turnover

### 9.8.1 Prose/Intuition

Factor portfolios require periodic rebalancing. As prices move, portfolio weights drift from targets. As characteristics change, stocks move between quantiles.

**The Trade-off:**

- **Frequent rebalancing:** Maintains factor exposure but generates turnover and costs
- **Infrequent rebalancing:** Reduces costs but allows factor exposure to decay

Most factor strategies rebalance monthly or quarterly. Weekly is too frequent (excessive costs); annually is too slow (signal decay).

**Turnover Costs:**

Turnover isn't just commissions. It includes:
- Bid-ask spreads (often the largest cost)
- Market impact (moving prices against you)
- Opportunity cost (waiting for execution)

A long-short factor portfolio with 100% annual turnover per side and 50 bps total cost loses 1% annually to friction.

### 9.8.2 Visual Evidence

```{r rebalancing-frequency, fig.cap="Rebalancing frequency trade-off. More frequent rebalancing maintains exposure but increases costs."}
# Simulate factor strategy with different rebalancing frequencies
n_stocks <- 200
n_days <- 252 * 5  # 5 years

# Generate daily data
daily_data <- data.table()

# Signal changes slowly (monthly characteristics)
monthly_signals <- matrix(rnorm(n_stocks * 60), nrow = 60, ncol = n_stocks)

for (d in 1:n_days) {
    month <- ceiling(d / 21)
    signal <- monthly_signals[min(month, 60), ]

    # Signal predicts returns but with noise
    returns <- 0.0003 + 0.001 * signal + rnorm(n_stocks, 0, 0.02)

    daily_data <- rbind(daily_data, data.table(
        day = d,
        stock = 1:n_stocks,
        signal = signal,
        returns = returns
    ))
}

# Function to calculate returns with given rebalancing frequency
calc_rebal_returns <- function(data, rebal_days, cost_bps = 10) {
    n_periods <- n_days / rebal_days
    period_returns <- numeric(n_periods)
    prev_weights <- rep(0, n_stocks)

    for (p in 1:n_periods) {
        start_day <- (p - 1) * rebal_days + 1
        end_day <- min(p * rebal_days, n_days)

        # Get signal at rebalancing date
        rebal_signal <- data[day == start_day, signal]

        # Rank and assign weights (top/bottom 20%)
        ranks <- rank(rebal_signal) / n_stocks
        weights <- ifelse(ranks > 0.8, 1, ifelse(ranks <= 0.2, -1, 0))
        weights <- weights / sum(abs(weights)) * 2  # Normalise

        # Calculate turnover
        turnover <- sum(abs(weights - prev_weights)) / 2
        prev_weights <- weights

        # Get period returns
        period_data <- data[day >= start_day & day <= end_day]
        stock_returns <- period_data[, .(ret = prod(1 + returns) - 1), by = stock]

        # Portfolio return (gross and net)
        gross_ret <- sum(weights * stock_returns$ret)
        cost <- turnover * cost_bps / 10000
        net_ret <- gross_ret - cost

        period_returns[p] <- net_ret
    }

    return(period_returns)
}

# Different frequencies
daily_ret <- calc_rebal_returns(daily_data, 1, cost_bps = 10)
weekly_ret <- calc_rebal_returns(daily_data, 5, cost_bps = 10)
monthly_ret <- calc_rebal_returns(daily_data, 21, cost_bps = 10)
quarterly_ret <- calc_rebal_returns(daily_data, 63, cost_bps = 10)

# Align lengths
min_len <- min(length(daily_ret), length(weekly_ret) * 5,
               length(monthly_ret) * 21, length(quarterly_ret) * 63)

# Convert to monthly for comparison
monthly_equiv <- data.table(
    period = 1:length(monthly_ret),
    Monthly = monthly_ret,
    Weekly = sapply(1:length(monthly_ret), function(m) {
        prod(1 + weekly_ret[((m-1)*4+1):min(m*4, length(weekly_ret))]) - 1
    }),
    Quarterly = sapply(1:length(monthly_ret), function(m) {
        q <- ceiling(m / 3)
        if (m %% 3 == 0 && q <= length(quarterly_ret)) {
            quarterly_ret[q]^(1/3) - 1 + rnorm(1, 0, 0.001)
        } else {
            quarterly_ret[min(q, length(quarterly_ret))]^(1/3) - 1 + rnorm(1, 0, 0.001)
        }
    })
)

# Cumulative
monthly_equiv[, cum_monthly := cumprod(1 + Monthly) - 1]
monthly_equiv[, cum_weekly := cumprod(1 + Weekly) - 1]
monthly_equiv[, cum_quarterly := cumprod(1 + Quarterly) - 1]

# Plot
rebal_long <- melt(monthly_equiv[, .(period, `Monthly` = cum_monthly,
                                      `Weekly` = cum_weekly, `Quarterly` = cum_quarterly)],
                   id.vars = "period", variable.name = "Frequency", value.name = "Return")

ggplot(rebal_long, aes(x = period, y = Return, colour = Frequency)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[1], tc[2], tc[3])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Factor Returns by Rebalancing Frequency (Net of Costs)",
         subtitle = "Monthly often balances exposure maintenance vs costs",
         x = "Month", y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")
```

```{r turnover-costs, fig.cap="Impact of transaction costs on factor returns. Higher turnover strategies are more sensitive to costs."}
# Show impact of different cost assumptions
cost_levels <- c(5, 10, 20, 50)  # basis points

cost_impact <- data.table()

for (cost in cost_levels) {
    ret <- calc_rebal_returns(daily_data, 21, cost_bps = cost)
    cost_impact <- rbind(cost_impact, data.table(
        period = 1:length(ret),
        cost = paste0(cost, " bps"),
        returns = ret
    ))
}

cost_impact[, cum_ret := cumprod(1 + returns) - 1, by = cost]

ggplot(cost_impact, aes(x = period, y = cum_ret, colour = cost)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[3], tc[2], tc[4], tc[1])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Factor Returns Under Different Cost Assumptions",
         subtitle = "Transaction costs significantly impact net returns",
         x = "Month", y = "Cumulative Return", colour = "Cost/Trade") +
    theme_trading() +
    theme(legend.position = "bottom")
```

### 9.8.3 Mathematical Derivation

**Turnover:**

One-way turnover at time $t$:
$$
\text{TO}_t = \frac{1}{2} \sum_{i=1}^{N} |w_{i,t} - w_{i,t-1}^*|
$$

where $w_{i,t-1}^* = w_{i,t-1} \times (1 + r_{i,t-1}) / \sum_j w_{j,t-1}(1+r_{j,t-1})$ is the drifted weight.

**Cost-Adjusted Return:**

$$
r_{\text{net},t} = r_{\text{gross},t} - c \times \text{TO}_t
$$

where $c$ = round-trip cost (bid-ask + impact + commission).

**Optimal Rebalancing Frequency:**

Trade-off function:
$$
\max_{\Delta t} \left[ \text{Factor Exposure}(\Delta t) - \text{Cost}(\text{TO}(\Delta t)) \right]
$$

With signal half-life $h$ and cost $c$:
$$
\Delta t^* \approx \sqrt{\frac{h \cdot \sigma_{\text{signal}}}{c}}
$$

Faster-decaying signals require more frequent rebalancing; higher costs favour less frequent.

**Rebalancing Bands:**

Instead of fixed frequency, rebalance when weights drift beyond threshold:
$$
\text{Rebalance if } |w_i^* - w_i^{\text{target}}| > \delta
$$

Reduces unnecessary trades while maintaining exposure.

### 9.8.4 Implementation & Application

```{r turnover-reduction, fig.cap="Turnover reduction techniques. Buffer zones reduce unnecessary trading."}
# Implement buffer zone approach
calc_buffered_returns <- function(data, buffer = 0.05, cost_bps = 10) {
    n_days <- max(data$day)
    daily_returns <- numeric(n_days)
    prev_weights <- rep(0, n_stocks)
    target_weights <- rep(0, n_stocks)
    total_turnover <- 0

    for (d in 1:n_days) {
        day_data <- data[day == d]

        # Update target based on signal
        ranks <- rank(day_data$signal) / n_stocks

        # New targets
        new_targets <- ifelse(ranks > 0.8, 1, ifelse(ranks <= 0.2, -1, 0))
        new_targets <- new_targets / sum(abs(new_targets)) * 2

        # Check if rebalancing needed (buffer zone)
        drift <- abs(prev_weights - new_targets)
        if (max(drift) > buffer || d == 1) {
            turnover <- sum(abs(new_targets - prev_weights)) / 2
            total_turnover <- total_turnover + turnover
            prev_weights <- new_targets
        }

        # Calculate return
        daily_returns[d] <- sum(prev_weights * day_data$returns)

        # Drift weights
        prev_weights <- prev_weights * (1 + day_data$returns)
        prev_weights <- prev_weights / sum(abs(prev_weights)) * 2
    }

    # Apply cost
    daily_cost <- total_turnover * cost_bps / 10000 / n_days
    net_returns <- daily_returns - daily_cost

    return(list(returns = net_returns, turnover = total_turnover))
}

# Compare different buffer sizes
buffers <- c(0.01, 0.05, 0.10, 0.20)
buffer_results <- data.table()

for (b in buffers) {
    result <- calc_buffered_returns(daily_data, buffer = b, cost_bps = 10)
    buffer_results <- rbind(buffer_results, data.table(
        day = 1:length(result$returns),
        buffer = paste0(b * 100, "% buffer"),
        returns = result$returns,
        turnover = result$turnover
    ))
}

buffer_results[, cum_ret := cumprod(1 + returns) - 1, by = buffer]

# Monthly aggregation for cleaner plot
buffer_results[, month := ceiling(day / 21)]
buffer_monthly <- buffer_results[, .(cum_ret = tail(cum_ret, 1)), by = .(buffer, month)]

ggplot(buffer_monthly, aes(x = month, y = cum_ret, colour = buffer)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[1], tc[2], tc[3], tc[4])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Factor Returns with Different Buffer Zones",
         subtitle = "Larger buffers reduce turnover but allow more drift",
         x = "Month", y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")

# Turnover comparison
turnover_summary <- buffer_results[, .(total_turnover = turnover[1]), by = buffer]
cat("=== Annual Turnover by Buffer Size ===\n\n")
turnover_summary[, annual_turnover := sprintf("%.0f%%", total_turnover / 5 * 100)]
print(turnover_summary[, .(buffer, annual_turnover)])
```

---

## 9.9 Factor Timing

### 9.9.1 Prose/Intuition

Factor premia vary over time. Value had a terrible decade (2010-2020). Momentum crashes periodically. Can we time factor exposure?

**Arguments For Timing:**

- Factors have predictable cycles (value spreads predict value returns)
- Macro conditions affect factors differently
- Avoiding crashes improves returns

**Arguments Against Timing:**

- Premia are hard to predict out-of-sample
- Timing adds turnover and costs
- Model error compounds
- Missing a few good periods destroys performance

The evidence is mixed. Most factor timing strategies don't reliably beat buy-and-hold after costs.

**Simple Timing Signals:**

1. **Valuation spread:** When value spread is wide (cheap stocks very cheap), value may outperform
2. **Momentum after crashes:** Avoid momentum after market drawdowns
3. **Volatility regime:** Some factors perform better in low-vol environments

### 9.9.2 Visual Evidence

```{r factor-timing-value, fig.cap="Value factor timing using valuation spread. When spread is wide, future value returns tend to be higher."}
# Load factor data
ff5 <- load_factors("ff5", freq = "daily")
ff5 <- filter_dates(ff5, as.Date("1990-01-01"), as.Date("2023-12-31"))

# Calculate rolling spread indicator (proxy for value spread)
# Use rolling z-score of HML as timing signal
ff5[, hml_12m := frollsum(HML, 252)]
ff5[, hml_zscore := (hml_12m - frollmean(hml_12m, 252 * 5)) /
                    frollapply(hml_12m, 252 * 5, sd)]

ff5 <- ff5[!is.na(hml_zscore)]

# Forward returns
ff5[, hml_fwd_12m := frollapply(HML, 252, function(x) prod(1 + x) - 1, align = "left")]
ff5[, hml_fwd_12m := shift(hml_fwd_12m, -252)]

# Scatter plot
ff5_timing <- ff5[!is.na(hml_fwd_12m)]

ggplot(ff5_timing[seq(1, .N, 21)], aes(x = hml_zscore, y = hml_fwd_12m)) +
    geom_point(colour = tc[1], alpha = 0.3, size = 1) +
    geom_smooth(method = "lm", se = TRUE, colour = tc[2]) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Value Factor Timing: Past Performance vs Future Returns",
         subtitle = "Weak relationship; timing is difficult",
         x = "Past 12M HML Z-Score", y = "Future 12M HML Return") +
    theme_trading()

# Calculate correlation
cat("=== Value Timing Signal ===\n")
cat(sprintf("Correlation (past 12M vs future 12M): %.3f\n",
            cor(ff5_timing$hml_zscore, ff5_timing$hml_fwd_12m, use = "complete.obs")))
```

```{r momentum-crash-timing, fig.cap="Momentum crash timing. Avoiding momentum after market crashes can reduce drawdowns."}
# Momentum timing: avoid after large market drawdowns
ff5[, mkt_drawdown := {
    cum_ret <- cumprod(1 + Mkt_RF)
    rolling_max <- cummax(cum_ret)
    (cum_ret / rolling_max) - 1
}]

# Create momentum proxy (using market as proxy since we don't have UMD)
# Simulate momentum returns correlated with market reversals
ff5[, momentum := rnorm(.N, 0.0003, 0.01) -
                  0.5 * shift(Mkt_RF, 1) * (Mkt_RF > 0.02)]  # Crashes on reversals

ff5[, cum_momentum := cumprod(1 + momentum) - 1]

# Timed momentum: reduce exposure when market in drawdown
ff5[, timing_signal := ifelse(mkt_drawdown < -0.10, 0, 1)]
ff5[, timed_momentum := momentum * timing_signal]
ff5[, cum_timed := cumprod(1 + timed_momentum) - 1]

# Plot
timing_long <- melt(ff5[, .(date, `Momentum (untimed)` = cum_momentum,
                             `Momentum (timed)` = cum_timed)],
                    id.vars = "date", variable.name = "Strategy", value.name = "Return")

ggplot(timing_long, aes(x = date, y = Return, colour = Strategy)) +
    geom_line(linewidth = 0.6) +
    scale_colour_manual(values = c(tc[4], tc[3])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Momentum Timing: Reduce Exposure During Market Drawdowns",
         subtitle = "Timing can reduce crash risk but may sacrifice returns",
         x = NULL, y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")
```

### 9.9.3 Mathematical Derivation

**Valuation Spread as Timing Signal:**

Define the characteristic spread:
$$
\text{Spread}_t = \bar{C}_{t,\text{Long}} - \bar{C}_{t,\text{Short}}
$$

High spread suggests potential for mean reversion → higher factor return.

Predictive regression:
$$
f_{t+h} = \alpha + \beta \times \text{Spread}_t + \varepsilon
$$

If $\beta > 0$, wide spreads predict higher returns.

**Momentum Crash Indicator:**

Momentum crashes after market rebounds. Indicator:
$$
I_{\text{crash}} = \mathbf{1}\left[\text{Drawdown}_{t-1} < -20\% \text{ and } r_{m,t-1:t} > 10\%\right]
$$

Timing rule:
$$
w_{\text{mom},t} = \begin{cases} 1 & \text{if } I_{\text{crash}} = 0 \\ 0 & \text{if } I_{\text{crash}} = 1 \end{cases}
$$

**Conditional Factor Premium:**

$$
E[f_t | \text{Signal}_t] = \alpha + \beta \times \text{Signal}_t
$$

Optimal timing weight:
$$
w^*_t \propto E[f_t | \text{Signal}_t]
$$

### 9.9.4 Implementation & Application

```{r timing-backtest, fig.cap="Factor timing backtest. Conditional exposure based on macro indicators."}
# Backtest simple timing strategy
ff5[, vol_regime := frollapply(Mkt_RF, 63, sd) * sqrt(252)]  # 3-month realised vol
ff5[, high_vol := vol_regime > median(vol_regime, na.rm = TRUE)]

# Some factors work better in low-vol regimes (simulated relationship)
ff5 <- ff5[!is.na(high_vol)]

# Strategy: overweight quality (RMW) in high vol, value (HML) in low vol
ff5[, timed_factor := ifelse(high_vol, RMW * 1.5 + HML * 0.5, RMW * 0.5 + HML * 1.5)]
ff5[, untimed_factor := (RMW + HML) / 2]

ff5[, cum_timed := cumprod(1 + timed_factor) - 1]
ff5[, cum_untimed := cumprod(1 + untimed_factor) - 1]

# Statistics
cat("=== Factor Timing Results ===\n\n")
timing_stats <- data.table(
    Strategy = c("Untimed (50/50)", "Timed"),
    `Ann. Return` = sprintf("%.2f%%", c(
        annualised_return(ff5$untimed_factor) * 100,
        annualised_return(ff5$timed_factor) * 100)),
    `Ann. Vol` = sprintf("%.1f%%", c(
        sd(ff5$untimed_factor, na.rm = TRUE) * sqrt(252) * 100,
        sd(ff5$timed_factor, na.rm = TRUE) * sqrt(252) * 100)),
    Sharpe = sprintf("%.2f", c(
        sharpe_ratio(ff5$untimed_factor),
        sharpe_ratio(ff5$timed_factor)))
)
print(timing_stats)

# Plot
vol_timing <- melt(ff5[, .(date, Untimed = cum_untimed, Timed = cum_timed)],
                   id.vars = "date", variable.name = "Strategy", value.name = "Return")

ggplot(vol_timing, aes(x = date, y = Return, colour = Strategy)) +
    geom_line(linewidth = 0.6) +
    scale_colour_manual(values = c(tc[1], tc[2])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Factor Timing by Volatility Regime",
         subtitle = "Overweight defensive (RMW) in high-vol, value (HML) in low-vol",
         x = NULL, y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "bottom")
```

---

## 9.10 Factor Crowding

### 9.10.1 Prose/Intuition

As factor investing has grown ($1T+ in factor ETFs), crowding has become a real concern:

**What Is Crowding?**

Many investors holding similar positions. When one needs to exit, all try to exit together, causing crashes.

**Signs of Crowding:**

1. **Compressed valuations:** Factor spread narrows (cheap stocks not that cheap anymore)
2. **High correlation:** Factor stocks move together more than fundamentals justify
3. **Short interest concentration:** Many shorting the same losers
4. **Large AUM:** Assets chasing the factor exceed capacity

**Consequences:**

1. **Lower future returns:** Premium erodes as prices adjust
2. **Higher crash risk:** Crowded exits cause sharp reversals
3. **Correlation during stress:** Factors that were uncorrelated become correlated

### 9.10.2 Visual Evidence

```{r factor-crowding-proxy, fig.cap="Factor crowding proxy. Compressed spreads may signal lower future returns."}
# Simulate crowding effect
# As more capital flows into factor, spread compresses and future returns decline

n_years <- 20
n_months <- n_years * 12

# Simulate AUM growth in factor strategies
aum_growth <- cumsum(rnorm(n_months, 0.02, 0.05))
aum_growth <- pmax(0, aum_growth)  # Can't be negative
aum_level <- exp(aum_growth)

# Factor spread (inversely related to AUM due to crowding)
base_spread <- 0.5  # 50% spread normally
spread <- base_spread * exp(-0.3 * log(aum_level))

# Factor returns (related to spread)
factor_return <- 0.005 * spread + rnorm(n_months, 0, 0.03)

crowding_data <- data.table(
    month = 1:n_months,
    aum = aum_level,
    spread = spread,
    returns = factor_return
)

crowding_data[, cum_ret := cumprod(1 + returns) - 1]

# Plot spread vs AUM
p1 <- ggplot(crowding_data, aes(x = aum, y = spread)) +
    geom_point(colour = tc[1], alpha = 0.5) +
    geom_smooth(method = "loess", se = FALSE, colour = tc[2]) +
    labs(title = "Factor Crowding: AUM vs Spread",
         subtitle = "More capital → compressed valuations",
         x = "Factor AUM (normalised)", y = "Valuation Spread") +
    theme_trading()

# Plot spread vs future returns
crowding_data[, fwd_return := shift(frollmean(returns, 12), -12)]

p2 <- ggplot(crowding_data[!is.na(fwd_return)], aes(x = spread, y = fwd_return)) +
    geom_point(colour = tc[3], alpha = 0.5) +
    geom_smooth(method = "lm", se = TRUE, colour = tc[2]) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Spread Predicts Future Returns",
         subtitle = "Wide spread → higher expected returns",
         x = "Valuation Spread", y = "Forward 12M Return") +
    theme_trading()

print(p1)
print(p2)
```

```{r correlation-spike, fig.cap="Factor correlation during crises. Normally uncorrelated factors become correlated during stress."}
# Show factor correlation changes during stress
ff5_stress <- ff5[!is.na(Mkt_RF)]

# Rolling correlation between SMB and HML
# Calculate rolling correlation using a loop approach
n <- nrow(ff5_stress)
smb_hml_cor <- rep(NA_real_, n)
for (i in 63:n) {
    idx <- (i - 62):i
    smb_hml_cor[i] <- cor(ff5_stress$SMB[idx], ff5_stress$HML[idx], use = "complete.obs")
}
ff5_stress[, smb_hml_cor := smb_hml_cor]

# Mark stress periods (large drawdowns)
ff5_stress[, stress_period := mkt_drawdown < -0.15]

# Plot
ggplot(ff5_stress[!is.na(smb_hml_cor)], aes(x = date, y = smb_hml_cor)) +
    geom_rect(data = ff5_stress[stress_period == TRUE & !is.na(smb_hml_cor)],
              aes(xmin = date, xmax = date + 1, ymin = -Inf, ymax = Inf),
              fill = tc[4], alpha = 0.3, inherit.aes = FALSE) +
    geom_line(colour = tc[1], linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title = "Rolling 63-Day Correlation: SMB vs HML",
         subtitle = "Shaded areas = market stress; correlations spike during crises",
         x = NULL, y = "Correlation") +
    theme_trading()
```

### 9.10.3 Mathematical Derivation

**Crowding Measure 1: Valuation Spread Compression**

$$
\text{Crowding}_t = \frac{\text{Spread}_t - \text{Spread}_{\text{median}}}{\text{Spread}_{\text{median}}}
$$

Negative values indicate crowding (spread below normal).

**Crowding Measure 2: Short Interest Concentration**

$$
\text{SI Concentration} = \frac{\sum_{i \in \text{Short}} \text{SI}_i}{\sum_{i} \text{SI}_i}
$$

High concentration in factor shorts indicates crowding.

**Crowding Measure 3: Pairwise Correlation**

$$
\rho_{\text{factor}} = \frac{2}{N(N-1)} \sum_{i < j} \rho_{ij}
$$

for stocks in the factor portfolio. Rising correlation suggests crowding.

**Impact on Expected Returns:**

$$
E[r_{\text{factor}}] = \lambda_{\text{base}} \times (1 - \alpha \times \text{Crowding})
$$

Higher crowding reduces expected premium.

### 9.10.4 Implementation & Application

```{r crowding-monitor, fig.cap="Crowding monitoring dashboard. Track multiple indicators to assess crowding risk."}
# Create crowding indicators
ff5_crowd <- ff5[!is.na(HML)]

# 1. Rolling spread percentile (proxy)
ff5_crowd[, hml_60d := frollsum(HML, 60)]
ff5_crowd[, spread_pct := frank(hml_60d) / .N]

# 2. Rolling correlation (as proxy for synchronisation)
# Calculate average pairwise correlation of factors
n_crowd <- nrow(ff5_crowd)
avg_corr <- rep(NA_real_, n_crowd)
for (i in 63:n_crowd) {
    idx <- (i - 62):i
    mat <- cbind(ff5_crowd$SMB[idx], ff5_crowd$HML[idx],
                 ff5_crowd$RMW[idx], ff5_crowd$CMA[idx])
    cormat <- cor(mat, use = "complete.obs")
    avg_corr[i] <- mean(cormat[lower.tri(cormat)])
}
ff5_crowd[, avg_corr := avg_corr]

# 3. Volatility clustering (high vol = potential unwind)
ff5_crowd[, factor_vol := frollapply(HML, 63, sd) * sqrt(252)]

# Composite crowding score
ff5_crowd <- ff5_crowd[!is.na(avg_corr) & !is.na(spread_pct) & !is.na(factor_vol)]
ff5_crowd[, crowding_score := scale(1 - spread_pct) + scale(avg_corr) + scale(factor_vol)]
ff5_crowd[, crowding_score := (crowding_score - min(crowding_score, na.rm = TRUE)) /
                               (max(crowding_score, na.rm = TRUE) - min(crowding_score, na.rm = TRUE))]

# Plot crowding over time
ggplot(ff5_crowd, aes(x = date, y = crowding_score)) +
    geom_line(colour = tc[1], linewidth = 0.5) +
    geom_hline(yintercept = 0.7, linetype = "dashed", colour = tc[4]) +
    annotate("text", x = min(ff5_crowd$date) + 1000, y = 0.75,
             label = "High Crowding Zone", colour = tc[4], size = 3) +
    labs(title = "Factor Crowding Score Over Time",
         subtitle = "Composite indicator based on spread, correlation, and volatility",
         x = NULL, y = "Crowding Score (0-1)") +
    theme_trading()

# Returns conditional on crowding
ff5_crowd[, crowding_quintile := cut(crowding_score, breaks = 5, labels = FALSE)]
ff5_crowd[, fwd_hml := shift(frollsum(HML, 63), -63)]

cond_returns <- ff5_crowd[!is.na(fwd_hml), .(
    avg_fwd_return = mean(fwd_hml),
    n = .N
), by = crowding_quintile]

ggplot(cond_returns, aes(x = crowding_quintile, y = avg_fwd_return)) +
    geom_col(fill = tc[1], alpha = 0.7) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Forward HML Returns by Crowding Quintile",
         subtitle = "Higher crowding may predict lower future returns",
         x = "Crowding Quintile (1=Low, 5=High)", y = "Average Forward 63-Day Return") +
    theme_trading()
```

---

## Quick Reference

### Rebalancing Guidelines

| Frequency | Pros | Cons | Best For |
|-----------|------|------|----------|
| Daily | Maximum exposure | High costs | HFT strategies |
| Weekly | Fresh signals | Moderate costs | Fast-moving factors |
| Monthly | Industry standard | Balanced | Most factors |
| Quarterly | Low costs | Signal decay | Slow factors |

### Turnover Metrics

| Metric | Formula | Target |
|--------|---------|--------|
| One-way turnover | $\frac{1}{2}\sum\|w_t - w_{t-1}^*\|$ | <100% annual |
| Cost impact | $c \times \text{TO}$ | <50 bps/year |
| Break-even | Premium / TO | >1 (ratio) |

### Factor Timing Signals

| Factor | Timing Signal | Direction |
|--------|---------------|-----------|
| Value | Valuation spread | Wide → overweight |
| Momentum | Market drawdown | Deep → underweight |
| All | Volatility regime | High vol → defensive |

### Crowding Indicators

| Indicator | Calculation | Warning Level |
|-----------|-------------|---------------|
| Spread compression | Current / Historical | <50th percentile |
| Correlation spike | Pairwise factor corr | >0.3 |
| AUM growth | YoY change | >20% |
| Short concentration | Top 10% SI share | >50% |

### Key Guidelines

1. **Rebalance monthly:** Balances exposure maintenance vs costs
2. **Use buffer zones:** Reduce unnecessary turnover
3. **Monitor crowding:** Compressed spreads signal lower future returns
4. **Be skeptical of timing:** Most timing doesn't beat buy-and-hold
5. **Diversify across factors:** Reduces reliance on any single premium
