---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 3: Random Variables"
part: "Part 2: Expectation and Variance"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, probability, random-variables, expectation, variance, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

## 3.5 Expectation (Expected Value)

### The Idea: The Long-Run Average

In plain English, the **expected value** of a random variable is the long-run average you would observe if you could repeat the random experiment infinitely many times. Roll a fair die forever and average the results: you get $3.5$. That is the expected value of a single die roll — a value the die can never actually show.

The expected value is the "centre of gravity" of the probability distribution. If you were to cut out the shape of the PMF or PDF from cardboard, the expected value is where you would balance it on a pin.

### Formal Definition

**Discrete case.** If $X$ is a discrete random variable with PMF $p(x)$, the expected value of $X$ is

$$
E[X] = \sum_{x} x \cdot p(x)
$$

where the sum runs over all possible values of $X$.

Each value $x$ is weighted by its probability $p(x)$. Values that are more likely contribute more to the average.

**Continuous case.** If $X$ is a continuous random variable with PDF $f(x)$, the expected value is

$$
E[X] = \int_{-\infty}^{\infty} x \cdot f(x)\,dx
$$

The logic is identical — we replace the sum with an integral and the PMF with the PDF.

In both cases, we require the sum or integral to converge absolutely. If $\sum_x |x| p(x) = \infty$ (or the analogous integral diverges), the expected value does not exist. The Cauchy distribution is the most famous example of a distribution with no mean.

### Implementing Expected Value from Scratch

```{r expected-value-scratch}
#' Compute the expected value of a discrete random variable
#'
#' @param values Numeric vector of possible values
#' @param probs Numeric vector of corresponding probabilities
#' @return The expected value E[X]
my_expected_value <- function(values, probs) {
  stopifnot(length(values) == length(probs))
  stopifnot(abs(sum(probs) - 1) < 1e-10)
  sum(values * probs)
}

# Expected value of a fair die
my_expected_value(1:6, rep(1/6, 6))

# Expected value of number of heads in 3 coin flips
my_expected_value(0:3, c(1/8, 3/8, 3/8, 1/8))

# E[X] for Uniform(0, 1): should be 0.5
integrate(function(x) x * dunif(x), lower = 0, upper = 1)

# E[X] for Exponential(rate = 2): should be 1/2 = 0.5
integrate(function(x) x * dexp(x, rate = 2), lower = 0, upper = Inf)
```

### Simulation: Convergence of the Running Mean

The **Law of Large Numbers** (which we shall prove rigorously in Chapter 7) guarantees that the sample mean converges to the expected value as the sample size grows. Let us visualise this.

```{r running-mean-convergence, fig.cap="The running mean of simulated die rolls converges to E[X] = 3.5 (dashed line) as the number of rolls increases. The convergence is not monotone — it oscillates around the target but the oscillations shrink."}
set.seed(42)
box::use(ggplot2)

n <- 5000
rolls <- sample(1:6, size = n, replace = TRUE)
running_mean <- cumsum(rolls) / seq_len(n)

df <- data.frame(n = seq_len(n), running_mean = running_mean)

ggplot2$ggplot(df, ggplot2$aes(x = n, y = running_mean)) +
  ggplot2$geom_line(colour = "#0072B2", linewidth = 0.4) +
  ggplot2$geom_hline(yintercept = 3.5, linetype = "dashed", colour = "red",
    linewidth = 0.8) +
  ggplot2$scale_x_log10() +
  ggplot2$labs(
    x = "Number of rolls (log scale)",
    y = "Running mean",
    title = "Running mean of die rolls converging to E[X] = 3.5"
  ) +
  ggplot2$theme_minimal(base_size = 14)
```

### Proof: Linearity of Expectation

**Theorem.** For any random variable $X$ and constants $a, b \in \mathbb{R}$:

$$
E[aX + b] = aE[X] + b
$$

*Proof (discrete case).* Let $X$ have PMF $p(x)$. Then

$$
\begin{aligned}
E[aX + b] &= \sum_{x} (ax + b) \cdot p(x) \\
&= \sum_{x} ax \cdot p(x) + \sum_{x} b \cdot p(x) \\
&= a \sum_{x} x \cdot p(x) + b \sum_{x} p(x) \\
&= a \cdot E[X] + b \cdot 1 \\
&= aE[X] + b \qquad \square
\end{aligned}
$$

In the third line, we used the fact that $a$ and $b$ are constants that factor out of the sum. In the fourth line, we used the definition of $E[X]$ and the normalisation property $\sum_x p(x) = 1$.

The continuous case follows identically, replacing sums with integrals.

**Special cases:**

- $E[aX] = aE[X]$ (constants factor out)
- $E[X + b] = E[X] + b$ (adding a constant shifts the mean)
- $E[c] = c$ for any constant $c$ (set $a = 0$, $b = c$)

### Proof: Additivity of Expectation

**Theorem.** For any two random variables $X$ and $Y$ defined on the same probability space:

$$
E[X + Y] = E[X] + E[Y]
$$

This result is remarkable because it **does not require $X$ and $Y$ to be independent**. It holds for any two random variables whatsoever — they can be correlated, dependent, or related in any way.

*Proof (discrete case).* Let $(X, Y)$ have joint PMF $p(x, y)$. Then

$$
\begin{aligned}
E[X + Y] &= \sum_{x} \sum_{y} (x + y) \cdot p(x, y) \\
&= \sum_{x} \sum_{y} x \cdot p(x, y) + \sum_{x} \sum_{y} y \cdot p(x, y) \\
&= \sum_{x} x \sum_{y} p(x, y) + \sum_{y} y \sum_{x} p(x, y) \\
&= \sum_{x} x \cdot p_X(x) + \sum_{y} y \cdot p_Y(y) \\
&= E[X] + E[Y] \qquad \square
\end{aligned}
$$

In the fourth line, we used the marginal PMFs: $p_X(x) = \sum_y p(x, y)$ and $p_Y(y) = \sum_x p(x, y)$.

Emphasise this: the expectation of a sum is **always** the sum of expectations. This innocent-looking identity is one of the most powerful tools in probability.

```{r additivity-demo}
set.seed(42)

# Demonstrate with correlated random variables
# X and Y are NOT independent, yet E[X+Y] = E[X] + E[Y]
n <- 100000
x <- rnorm(n, mean = 3, sd = 1)
y <- x + rnorm(n, mean = 0, sd = 0.5)  # Y depends on X!

mean(x)              # E[X]
mean(y)              # E[Y]
mean(x) + mean(y)    # E[X] + E[Y]
mean(x + y)          # E[X + Y] -- should match the line above
cor(x, y)            # strongly correlated!
```

Despite the strong correlation, $E[X + Y]$ equals $E[X] + E[Y]$ to within simulation noise. Independence is irrelevant for this result.

### LOTUS: The Law of the Unconscious Statistician

Suppose we want $E[g(X)]$ for some function $g$. The naive approach is: find the distribution of $g(X)$, then compute its expectation. LOTUS says we can skip the first step.

**Theorem (LOTUS).** If $X$ is a discrete random variable with PMF $p(x)$ and $g$ is any function, then

$$
E[g(X)] = \sum_{x} g(x) \cdot p(x)
$$

If $X$ is continuous with PDF $f(x)$, then

$$
E[g(X)] = \int_{-\infty}^{\infty} g(x) \cdot f(x)\,dx
$$

In plain English: to find the expected value of a function of $X$, you do not need to find the distribution of $g(X)$ first. You simply weight $g(x)$ by the original distribution of $X$.

*Proof (discrete case).* Let $Y = g(X)$. For each possible value $y$ of $Y$, the PMF of $Y$ is

$$
p_Y(y) = P(Y = y) = P(g(X) = y) = \sum_{x:\, g(x) = y} p(x)
$$

Therefore

$$
\begin{aligned}
E[Y] &= \sum_{y} y \cdot p_Y(y) \\
&= \sum_{y} y \sum_{x:\, g(x) = y} p(x) \\
&= \sum_{y} \sum_{x:\, g(x) = y} g(x) \cdot p(x) \\
&= \sum_{x} g(x) \cdot p(x) \qquad \square
\end{aligned}
$$

In the last step, we used the fact that as $y$ ranges over all values of $g$ and $x$ ranges over all pre-images of $y$, we cover every $x$ in the support of $X$ exactly once.

```{r lotus-demo}
# LOTUS example: E[X^2] for a fair die
die_vals <- 1:6
die_probs <- rep(1/6, 6)

# Using LOTUS: E[X^2] = sum(x^2 * p(x))
sum(die_vals^2 * die_probs)

# Verify by simulation
set.seed(42)
sim_vals <- sample(1:6, size = 100000, replace = TRUE)
mean(sim_vals^2)

# LOTUS for a continuous RV: E[X^2] for Standard Normal
# Theoretical: E[X^2] = Var(X) + (E[X])^2 = 1 + 0 = 1
integrate(function(x) x^2 * dnorm(x), lower = -Inf, upper = Inf)
```

---

## 3.6 Variance

### The Idea: Measuring Spread

While the expected value tells us where the distribution is centred, it says nothing about how spread out the distribution is. A random variable that always equals 5 and a random variable that is equally likely to be 0 or 10 both have $E[X] = 5$, but they are very different distributions.

The **variance** quantifies this spread by measuring the average squared distance from the mean.

### Formal Definition

The **variance** of a random variable $X$ with mean $\mu = E[X]$ is

$$
\text{Var}(X) = E\left[(X - \mu)^2\right]
$$

In words: variance is the expected value of the squared deviation from the mean.

For a discrete random variable:

$$
\text{Var}(X) = \sum_{x} (x - \mu)^2 \cdot p(x)
$$

For a continuous random variable:

$$
\text{Var}(X) = \int_{-\infty}^{\infty} (x - \mu)^2 \cdot f(x)\,dx
$$

### Derivation of the Computational Formula

In practice, the definition $E[(X - \mu)^2]$ is often inconvenient. We can derive a simpler formula.

**Theorem.** $\text{Var}(X) = E[X^2] - (E[X])^2$.

*Proof.* Expand the square in the definition:

$$
\begin{aligned}
\text{Var}(X) &= E\left[(X - \mu)^2\right] \\
&= E\left[X^2 - 2\mu X + \mu^2\right] \\
&= E[X^2] - 2\mu E[X] + \mu^2 \quad \text{(by linearity of expectation)} \\
&= E[X^2] - 2\mu \cdot \mu + \mu^2 \quad \text{(since } E[X] = \mu\text{)} \\
&= E[X^2] - 2\mu^2 + \mu^2 \\
&= E[X^2] - \mu^2 \\
&= E[X^2] - (E[X])^2 \qquad \square
\end{aligned}
$$

This is the **computational formula** for variance: the mean of the squares minus the square of the mean. It is almost always easier to use than the definition.

### Implementing Variance from Scratch

```{r variance-scratch}
#' Compute the variance of a discrete random variable
#'
#' @param values Numeric vector of possible values
#' @param probs Numeric vector of corresponding probabilities
#' @return The variance Var(X)
my_variance_rv <- function(values, probs) {
  stopifnot(length(values) == length(probs))
  stopifnot(abs(sum(probs) - 1) < 1e-10)

  mu <- sum(values * probs)               # E[X]
  e_x2 <- sum(values^2 * probs)           # E[X^2]
  e_x2 - mu^2                              # Var(X) = E[X^2] - (E[X])^2
}

# Variance of a fair die (exact: (6^2 - 1)/12 = 2.9167)
die_var <- my_variance_rv(1:6, rep(1/6, 6))
die_var

# Variance of number of heads in 3 fair coin flips
# Binomial(3, 0.5): Var = n*p*(1-p) = 0.75
heads_var <- my_variance_rv(0:3, c(1/8, 3/8, 3/8, 1/8))
heads_var
```

### Standard Deviation

The **standard deviation** is simply the square root of the variance:

$$
\text{SD}(X) = \sigma = \sqrt{\text{Var}(X)}
$$

The standard deviation has the same units as $X$, which makes it more interpretable. If $X$ is measured in centimetres, $\text{Var}(X)$ is in cm$^2$ but $\text{SD}(X)$ is in cm.

```{r sd-demo}
sqrt(die_var)     # SD of a fair die
sqrt(heads_var)   # SD of heads in 3 flips
```

### Properties of Variance: Proofs

**Property 1.** $\text{Var}(aX + b) = a^2 \text{Var}(X)$

*Proof.* Let $\mu = E[X]$. Then $E[aX + b] = a\mu + b$ by linearity. So

$$
\begin{aligned}
\text{Var}(aX + b) &= E\left[(aX + b - (a\mu + b))^2\right] \\
&= E\left[(aX - a\mu)^2\right] \\
&= E\left[a^2(X - \mu)^2\right] \\
&= a^2 E\left[(X - \mu)^2\right] \\
&= a^2 \text{Var}(X) \qquad \square
\end{aligned}
$$

Note that the constant $b$ vanishes entirely. Shifting a distribution does not change its spread — only scaling does, and the variance scales by the square of the constant.

**Property 2.** $\text{Var}(X) \geq 0$ always.

*Proof.* $\text{Var}(X) = E[(X - \mu)^2]$. Since $(X - \mu)^2 \geq 0$ for every possible value of $X$, and expectation preserves non-negativity (a weighted average of non-negative quantities is non-negative), we have $E[(X - \mu)^2] \geq 0$. $\square$

**Property 3.** $\text{Var}(X) = 0$ if and only if $X$ is constant with probability 1.

*Proof.*

$(\Leftarrow)$ If $X = c$ with probability 1, then $E[X] = c$ and $(X - c)^2 = 0$ with probability 1, so $\text{Var}(X) = E[0] = 0$.

$(\Rightarrow)$ Suppose $\text{Var}(X) = 0$, i.e., $E[(X - \mu)^2] = 0$. Now $(X - \mu)^2 \geq 0$ always. For the expectation of a non-negative random variable to be zero, the random variable must be zero with probability 1. (If $P((X-\mu)^2 > 0) > 0$, then $E[(X-\mu)^2] > 0$, a contradiction.) Therefore $(X - \mu)^2 = 0$ with probability 1, which means $X = \mu$ with probability 1. $\square$

### Simulation: Comparing Variance to Theory

```{r variance-simulation, fig.cap="Distribution of sample variances from repeated experiments. Each experiment draws 50 observations from a fair die. The theoretical variance (35/12) is shown as a dashed red line."}
set.seed(42)
box::use(ggplot2)

n_experiments <- 10000
sample_size <- 50
theoretical_var <- 35/12

# Simulate many experiments, compute sample variance for each
sample_vars <- replicate(n_experiments, {
  draws <- sample(1:6, size = sample_size, replace = TRUE)
  var(draws)  # R's var() uses Bessel's correction (divides by n-1)
})

theoretical_var
mean(sample_vars)

ggplot2$ggplot(data.frame(v = sample_vars), ggplot2$aes(x = v)) +
  ggplot2$geom_histogram(ggplot2$aes(y = ggplot2::after_stat(density)),
    bins = 50, fill = "#0072B2", alpha = 0.7) +
  ggplot2$geom_vline(xintercept = theoretical_var,
    linetype = "dashed", colour = "red", linewidth = 1) +
  ggplot2$labs(
    x = "Sample variance",
    y = "Density",
    title = sprintf("Distribution of sample variances (n = %d per experiment)", sample_size)
  ) +
  ggplot2$theme_minimal(base_size = 14)
```

The mean of the sample variances (using Bessel's correction) matches the theoretical variance $35/12 \approx 2.917$. This confirms that $s^2 = \frac{1}{n-1}\sum(x_i - \bar{x})^2$ is an unbiased estimator of $\text{Var}(X)$ — precisely the connection to Chapter 1's discussion of why we divide by $n - 1$.

---

## 3.7 Covariance and Correlation (Preview)

When we have two random variables, we often want to know whether they tend to move together. Covariance and correlation measure this co-movement.

### Covariance

The **covariance** of two random variables $X$ and $Y$ is

$$
\text{Cov}(X, Y) = E\left[(X - \mu_X)(Y - \mu_Y)\right]
$$

where $\mu_X = E[X]$ and $\mu_Y = E[Y]$.

By expanding the product (using the same algebraic technique as for variance), we obtain the computational formula:

$$
\text{Cov}(X, Y) = E[XY] - E[X]E[Y]
$$

*Derivation.*

$$
\begin{aligned}
\text{Cov}(X, Y) &= E\left[(X - \mu_X)(Y - \mu_Y)\right] \\
&= E\left[XY - X\mu_Y - \mu_X Y + \mu_X \mu_Y\right] \\
&= E[XY] - \mu_Y E[X] - \mu_X E[Y] + \mu_X \mu_Y \\
&= E[XY] - \mu_X \mu_Y - \mu_X \mu_Y + \mu_X \mu_Y \\
&= E[XY] - \mu_X \mu_Y \\
&= E[XY] - E[X]E[Y] \qquad \square
\end{aligned}
$$

**Interpretation:**

- $\text{Cov}(X, Y) > 0$: $X$ and $Y$ tend to move in the same direction.
- $\text{Cov}(X, Y) < 0$: $X$ and $Y$ tend to move in opposite directions.
- $\text{Cov}(X, Y) = 0$: no *linear* relationship (but there could still be a nonlinear one).

Note that $\text{Cov}(X, X) = \text{Var}(X)$.

### If Independent, Then Uncorrelated

**Theorem.** If $X$ and $Y$ are independent, then $\text{Cov}(X, Y) = 0$.

*Proof.* If $X$ and $Y$ are independent, then $E[XY] = E[X]E[Y]$ (this follows from the factorisation of the joint distribution; we prove this properly in Chapter 6). Therefore:

$$
\text{Cov}(X, Y) = E[XY] - E[X]E[Y] = E[X]E[Y] - E[X]E[Y] = 0 \qquad \square
$$

### The Converse Is False

Zero covariance does **not** imply independence. Here is a classic counterexample.

```{r uncorrelated-not-independent}
set.seed(42)

# Let X ~ Uniform(-1, 1) and Y = X^2
# Then Cov(X, Y) = 0 but X and Y are completely dependent!
n <- 100000
x <- runif(n, -1, 1)
y <- x^2

mean(x)        # ~0
mean(y)        # ~1/3
mean(x * y)    # ~0
cov(x, y)      # ~0
cor(x, y)      # ~0
```

$X$ and $Y = X^2$ are perfectly dependent — knowing $X$ tells you $Y$ exactly — yet their covariance is zero. The covariance only detects *linear* relationships.

### Variance of a Sum

**Theorem.** For any two random variables $X$ and $Y$:

$$
\text{Var}(X + Y) = \text{Var}(X) + \text{Var}(Y) + 2\,\text{Cov}(X, Y)
$$

*Proof.* Let $\mu_X = E[X]$ and $\mu_Y = E[Y]$. Then $E[X + Y] = \mu_X + \mu_Y$.

$$
\begin{aligned}
\text{Var}(X + Y) &= E\left[((X+Y) - (\mu_X + \mu_Y))^2\right] \\
&= E\left[((X - \mu_X) + (Y - \mu_Y))^2\right] \\
&= E\left[(X - \mu_X)^2 + 2(X - \mu_X)(Y - \mu_Y) + (Y - \mu_Y)^2\right] \\
&= E[(X - \mu_X)^2] + 2E[(X - \mu_X)(Y - \mu_Y)] + E[(Y - \mu_Y)^2] \\
&= \text{Var}(X) + 2\,\text{Cov}(X, Y) + \text{Var}(Y) \qquad \square
\end{aligned}
$$

**Corollary.** If $X$ and $Y$ are **independent** (or merely uncorrelated), then $\text{Cov}(X, Y) = 0$ and therefore

$$
\text{Var}(X + Y) = \text{Var}(X) + \text{Var}(Y)
$$

Variances of independent random variables add. This is the probabilistic foundation for the $\sqrt{n}$ that appears everywhere in statistics: if $X_1, \ldots, X_n$ are independent with common variance $\sigma^2$, then $\text{Var}(\bar{X}) = \sigma^2/n$.

```{r var-sum-demo}
set.seed(42)
n <- 100000

# Case 1: Independent X, Y
x1 <- rnorm(n, mean = 5, sd = 2)
y1 <- rnorm(n, mean = 3, sd = 1)

# Independent case: Var(X+Y) should equal Var(X) + Var(Y)
c(Var_X = var(x1), Var_Y = var(y1),
  sum = var(x1) + var(y1), Var_XpY = var(x1 + y1),
  two_cov = 2 * cov(x1, y1))

# Case 2: Positively correlated X, Y
x2 <- rnorm(n, mean = 5, sd = 2)
y2 <- 0.8 * x2 + rnorm(n, mean = 0, sd = 1)

# Correlated case: Var(X+Y) = Var(X) + Var(Y) + 2*Cov(X,Y)
c(Var_X = var(x2), Var_Y = var(y2),
  sum_plus_2cov = var(x2) + var(y2) + 2 * cov(x2, y2),
  Var_XpY = var(x2 + y2))
```

### Correlation

The **Pearson correlation coefficient** standardises covariance to the range $[-1, 1]$:

$$
\rho(X, Y) = \frac{\text{Cov}(X, Y)}{\text{SD}(X) \cdot \text{SD}(Y)}
$$

We shall study correlation in depth in Chapter 6 (Joint Distributions). For now, note that $|\rho| = 1$ if and only if $Y$ is a perfect linear function of $X$.

---

## Communicating to Stakeholders

When presenting expected values to a non-technical audience, use the phrase "average outcome" or "what you would expect on average." The formal definition is less important than the interpretation.

For variance, say "how much the outcomes typically vary" or "the spread of possible results." Standard deviation is often preferable in communication because it shares the units of the original measurement.

When discussing covariance, use directional language: "these two quantities tend to move together" (positive covariance) or "when one goes up, the other tends to go down" (negative covariance). Correlation is easier to interpret because it is bounded between $-1$ and $+1$.

Always warn stakeholders that correlation does not imply causation, and that zero correlation does not mean no relationship — only no *linear* relationship.

---

## Quick Reference

| Concept | Formula | Notes |
|:---|:---|:---|
| **Expected value (discrete)** | $E[X] = \sum_x x\,p(x)$ | Long-run average |
| **Expected value (continuous)** | $E[X] = \int x\,f(x)\,dx$ | |
| **Linearity** | $E[aX + b] = aE[X] + b$ | Always holds |
| **Additivity** | $E[X + Y] = E[X] + E[Y]$ | No independence needed |
| **LOTUS** | $E[g(X)] = \sum g(x)p(x)$ | No need to find dist. of $g(X)$ |
| **Variance (definition)** | $\text{Var}(X) = E[(X-\mu)^2]$ | Average squared deviation |
| **Variance (computational)** | $\text{Var}(X) = E[X^2] - (E[X])^2$ | Easier to compute |
| **Scaling** | $\text{Var}(aX + b) = a^2\text{Var}(X)$ | Constants shift, don't spread |
| **Standard deviation** | $\text{SD}(X) = \sqrt{\text{Var}(X)}$ | Same units as $X$ |
| **Covariance** | $\text{Cov}(X,Y) = E[XY] - E[X]E[Y]$ | Measures linear co-movement |
| **Var of sum** | $\text{Var}(X+Y) = \text{Var}(X) + \text{Var}(Y) + 2\text{Cov}(X,Y)$ | |
| **If independent** | $\text{Cov}(X,Y) = 0$, $\text{Var}(X+Y) = \text{Var}(X) + \text{Var}(Y)$ | Converse false |

---

## Exercises

**Exercise 3.6.** Let $X$ have the PMF $P(X = -1) = 0.3$, $P(X = 0) = 0.4$, $P(X = 2) = 0.3$. Compute $E[X]$, $E[X^2]$, and $\text{Var}(X)$ by hand and verify with R.

**Exercise 3.7.** Prove that $E[X - E[X]] = 0$ for any random variable $X$. Explain in words why this must be true.

**Exercise 3.8.** Let $X \sim \text{Uniform}(a, b)$. (a) Derive $E[X] = (a + b)/2$ by integration. (b) Derive $\text{Var}(X) = (b - a)^2/12$ by integration. (c) Verify both results with simulation in R.

**Exercise 3.9.** Suppose $X$ and $Y$ are random variables with $E[X] = 2$, $E[Y] = 5$, $\text{Var}(X) = 3$, $\text{Var}(Y) = 7$, and $\text{Cov}(X, Y) = -1$. Compute $E[3X - 2Y + 4]$ and $\text{Var}(3X - 2Y + 4)$.

**Exercise 3.10.** Write an R function `my_covariance(x_vals, y_vals, joint_probs)` that computes the covariance of two discrete random variables from their joint probability table. Test it on a $2 \times 2$ example where $X \in \{0, 1\}$, $Y \in \{0, 1\}$, with joint probabilities $P(0,0) = 0.4$, $P(0,1) = 0.1$, $P(1,0) = 0.1$, $P(1,1) = 0.4$. Verify your result using the computational formula $\text{Cov}(X,Y) = E[XY] - E[X]E[Y]$.

**Exercise 3.11.** Construct your own example (different from the one in the text) of two random variables that are uncorrelated ($\text{Cov}(X,Y) = 0$) but not independent. Verify both properties with simulation.
