---
title: "One-Way Analysis of Variance"
subtitle: "Part 1 of Chapter 13: ANOVA"
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 10,
    fig.height = 6
)

box::use(
    data.table[...],
    ggplot2
)
```

## 13.1 Introduction to ANOVA

### 13.1.1 The Problem of Comparing Multiple Groups

In Chapter 10, we learned how to compare two groups using t-tests. But what if we have three or more groups? For example:
- Comparing blood pressure across four age groups
- Comparing test scores across three teaching methods
- Comparing crop yields across five fertiliser treatments

We could run multiple t-tests (Group 1 vs 2, Group 1 vs 3, Group 2 vs 3, etc.), but this approach has a serious problem: **inflated Type I error rate**.

```{r packages}
box::use(
    data.table[...],
    ggplot2
)
```

```{r load_data}
# Load NHANES data
nhanes <- fread("../data/primary/nhanes.csv")
```

### 13.1.2 The Multiple Comparisons Problem

With $k$ groups, we need $\binom{k}{2} = \frac{k(k-1)}{2}$ pairwise comparisons.

```{r multiple_comparisons_problem}
# Calculate family-wise error rate
calculate_fwer <- function(k, alpha = 0.05) {
    n_tests <- choose(k, 2)
    fwer <- 1 - (1 - alpha)^n_tests
    return(c(groups = k, tests = n_tests, fwer = fwer))
}

fwer_table <- data.table(t(sapply(2:10, calculate_fwer)))
setnames(fwer_table, c("Groups", "Pairwise Tests", "FWER"))

cat("Family-Wise Error Rate with Multiple t-tests\n")
cat("=============================================\n\n")
cat("(Assuming α = 0.05 for each test)\n\n")
print(fwer_table, digits = 3)

cat("\nWith 5 groups and 10 pairwise comparisons, there's a 40% chance\n")
cat("of at least one false positive, even when all null hypotheses are true!\n")
```

### 13.1.3 ANOVA: A Single Omnibus Test

**Analysis of Variance (ANOVA)** solves this problem by testing all groups simultaneously with a single test:

$$H_0: \mu_1 = \mu_2 = \cdots = \mu_k$$
$$H_1: \text{At least one } \mu_i \text{ differs}$$

This maintains the Type I error rate at $\alpha$ regardless of the number of groups.

---

## 13.2 The One-Way ANOVA Model

### 13.2.1 Model Specification

The one-way ANOVA model is:

$$Y_{ij} = \mu + \alpha_i + \varepsilon_{ij}$$

Where:
- $Y_{ij}$ is the $j$th observation in group $i$
- $\mu$ is the overall (grand) mean
- $\alpha_i$ is the effect of group $i$ (deviation from grand mean)
- $\varepsilon_{ij} \sim N(0, \sigma^2)$ is the random error

With the constraint $\sum_{i=1}^{k} n_i \alpha_i = 0$ (effects sum to zero).

### 13.2.2 Model Assumptions

ANOVA requires:

1. **Independence**: Observations are independent
2. **Normality**: Residuals are normally distributed within each group
3. **Homoscedasticity**: Equal variances across groups (homogeneity of variance)

```{r anova_example_data}
# Create example: BMI across age groups
bmi_data <- nhanes[!is.na(BMI) & !is.na(AgeDecade) & AgeDecade != ""]
bmi_data <- bmi_data[AgeDecade %in% c("20-29", "30-39", "40-49", "50-59", "60-69")]

# Sample for demonstration
set.seed(123)
bmi_sample <- bmi_data[, .SD[sample(.N, min(80, .N))], by = AgeDecade]

cat("Sample sizes by age group:\n")
print(bmi_sample[, .N, by = AgeDecade][order(AgeDecade)])
```

---

## 13.3 Visualising Group Differences

### 13.3.1 Box Plots

```{r boxplot_anova}
ggplot2$ggplot(bmi_sample, ggplot2$aes(x = AgeDecade, y = BMI, fill = AgeDecade)) +
    ggplot2$geom_boxplot(alpha = 0.7) +
    ggplot2$geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
    ggplot2$scale_fill_brewer(palette = "Set2") +
    ggplot2$labs(
        title = "BMI Distribution Across Age Groups",
        subtitle = "Does mean BMI differ significantly across age decades?",
        x = "Age Decade",
        y = "BMI (kg/m²)"
    ) +
    ggplot2$theme_minimal(base_size = 12) +
    ggplot2$theme(legend.position = "none")
```

### 13.3.2 Group Means with Error Bars

```{r means_plot}
# Calculate means and SE
group_stats <- bmi_sample[, .(
    mean = mean(BMI),
    sd = sd(BMI),
    n = .N,
    se = sd(BMI) / sqrt(.N)
), by = AgeDecade]

# Grand mean
grand_mean <- mean(bmi_sample$BMI)

ggplot2$ggplot(group_stats, ggplot2$aes(x = AgeDecade, y = mean)) +
    ggplot2$geom_hline(yintercept = grand_mean, linetype = "dashed", colour = "#D55E00") +
    ggplot2$geom_errorbar(ggplot2$aes(ymin = mean - 1.96*se, ymax = mean + 1.96*se),
                          width = 0.2, colour = "#0072B2") +
    ggplot2$geom_point(size = 4, colour = "#0072B2") +
    ggplot2$labs(
        title = "Group Means with 95% Confidence Intervals",
        subtitle = "Dashed line shows grand mean",
        x = "Age Decade",
        y = "Mean BMI (kg/m²)"
    ) +
    ggplot2$annotate("text", x = 5.3, y = grand_mean + 0.3,
                     label = sprintf("Grand mean = %.1f", grand_mean),
                     colour = "#D55E00", hjust = 0) +
    ggplot2$theme_minimal(base_size = 12)
```

---

## 13.4 Partitioning Variance

### 13.4.1 The Key Insight

ANOVA works by partitioning the total variability into two components:

$$\underbrace{\sum_{i=1}^{k}\sum_{j=1}^{n_i}(Y_{ij} - \bar{Y}_{..})^2}_{\text{SST: Total}} = \underbrace{\sum_{i=1}^{k} n_i(\bar{Y}_{i.} - \bar{Y}_{..})^2}_{\text{SSB: Between Groups}} + \underbrace{\sum_{i=1}^{k}\sum_{j=1}^{n_i}(Y_{ij} - \bar{Y}_{i.})^2}_{\text{SSW: Within Groups}}$$

Where:
- $\bar{Y}_{..}$ is the grand mean
- $\bar{Y}_{i.}$ is the mean of group $i$

### 13.4.2 Intuition

- **SSB (Between)**: Measures how much group means differ from the grand mean
- **SSW (Within)**: Measures variability within groups (random noise)

If group means are truly different, SSB will be large relative to SSW.

```{r partition_variance}
# Calculate sums of squares manually
grand_mean <- mean(bmi_sample$BMI)
bmi_sample[, group_mean := mean(BMI), by = AgeDecade]

SST <- sum((bmi_sample$BMI - grand_mean)^2)
SSB <- sum(bmi_sample[, .(n = .N, group_mean = first(group_mean)), by = AgeDecade][,
            n * (group_mean - grand_mean)^2])
SSW <- sum((bmi_sample$BMI - bmi_sample$group_mean)^2)

cat("Partitioning Total Variance\n")
cat("===========================\n\n")
cat(sprintf("Grand mean = %.2f\n\n", grand_mean))
cat(sprintf("Total Sum of Squares (SST) = %.2f\n", SST))
cat(sprintf("Between Groups (SSB) = %.2f\n", SSB))
cat(sprintf("Within Groups (SSW) = %.2f\n\n", SSW))
cat(sprintf("Check: SSB + SSW = %.2f + %.2f = %.2f ≈ SST = %.2f ✓\n",
            SSB, SSW, SSB + SSW, SST))
```

### 13.4.3 Visualising Variance Partition

```{r variance_visual}
# Create visual showing partition
variance_data <- data.table(
    Component = c("Between Groups\n(Explained)", "Within Groups\n(Unexplained)"),
    SS = c(SSB, SSW),
    Proportion = c(SSB/SST, SSW/SST)
)

ggplot2$ggplot(variance_data, ggplot2$aes(x = "", y = SS, fill = Component)) +
    ggplot2$geom_col(width = 0.5) +
    ggplot2$geom_text(ggplot2$aes(label = sprintf("%.1f%%", Proportion * 100)),
                      position = ggplot2$position_stack(vjust = 0.5),
                      colour = "white", fontface = "bold", size = 5) +
    ggplot2$scale_fill_manual(values = c("#0072B2", "#D55E00")) +
    ggplot2$labs(
        title = "Variance Partition in ANOVA",
        subtitle = sprintf("η² = SSB/SST = %.3f (proportion explained by group membership)", SSB/SST),
        x = "",
        y = "Sum of Squares",
        fill = "Variance Source"
    ) +
    ggplot2$coord_flip() +
    ggplot2$theme_minimal(base_size = 12)
```

---

## 13.5 The F-Test

### 13.5.1 Mean Squares and the F-Ratio

We convert sums of squares to **mean squares** by dividing by degrees of freedom:

$$\text{MSB} = \frac{\text{SSB}}{k-1}, \quad \text{MSW} = \frac{\text{SSW}}{N-k}$$

Where:
- $k$ is the number of groups
- $N$ is the total sample size

The **F-ratio** is:

$$F = \frac{\text{MSB}}{\text{MSW}} = \frac{\text{Between-group variance}}{\text{Within-group variance}}$$

### 13.5.2 The F Distribution

Under $H_0$ (all means equal), $F \sim F_{k-1, N-k}$.

If group means truly differ, MSB will be inflated, leading to a large F.

```{r f_calculation}
k <- length(unique(bmi_sample$AgeDecade))  # Number of groups
N <- nrow(bmi_sample)                       # Total sample size

df_between <- k - 1
df_within <- N - k

MSB <- SSB / df_between
MSW <- SSW / df_within

F_stat <- MSB / MSW
p_value <- pf(F_stat, df1 = df_between, df2 = df_within, lower.tail = FALSE)

cat("ANOVA F-Test\n")
cat("============\n\n")
cat(sprintf("Number of groups (k) = %d\n", k))
cat(sprintf("Total sample size (N) = %d\n\n", N))
cat(sprintf("Degrees of freedom:\n"))
cat(sprintf("  Between: k - 1 = %d\n", df_between))
cat(sprintf("  Within: N - k = %d\n\n", df_within))
cat(sprintf("Mean Squares:\n"))
cat(sprintf("  MSB = SSB / df = %.2f / %d = %.2f\n", SSB, df_between, MSB))
cat(sprintf("  MSW = SSW / df = %.2f / %d = %.2f\n\n", SSW, df_within, MSW))
cat(sprintf("F = MSB / MSW = %.2f / %.2f = %.3f\n", MSB, MSW, F_stat))
cat(sprintf("p-value = P(F > %.3f) = %.4f\n", F_stat, p_value))
```

### 13.5.3 Visualising the F Distribution

```{r f_distribution_visual}
# Generate F distribution
f_data <- data.table(x = seq(0, 6, length.out = 500))
f_data[, density := df(x, df1 = df_between, df2 = df_within)]

# Critical value
f_crit <- qf(0.95, df1 = df_between, df2 = df_within)

ggplot2$ggplot(f_data, ggplot2$aes(x = x, y = density)) +
    ggplot2$geom_line(colour = "#0072B2", linewidth = 1) +
    ggplot2$geom_area(data = f_data[x >= f_crit], fill = "#D55E00", alpha = 0.3) +
    ggplot2$geom_vline(xintercept = F_stat, colour = "#009E73", linewidth = 1.2, linetype = "dashed") +
    ggplot2$geom_vline(xintercept = f_crit, colour = "#D55E00", linewidth = 1, linetype = "dotted") +
    ggplot2$annotate("text", x = F_stat + 0.3, y = max(f_data$density) * 0.8,
                     label = sprintf("F = %.2f", F_stat), colour = "#009E73") +
    ggplot2$annotate("text", x = f_crit + 0.3, y = max(f_data$density) * 0.6,
                     label = sprintf("F_crit = %.2f", f_crit), colour = "#D55E00") +
    ggplot2$labs(
        title = sprintf("F Distribution (df₁ = %d, df₂ = %d)", df_between, df_within),
        subtitle = "Green dashed line: observed F; Orange shaded: rejection region (α = 0.05)",
        x = "F",
        y = "Density"
    ) +
    ggplot2$theme_minimal(base_size = 12)
```

---

## 13.6 The ANOVA Table

### 13.6.1 Traditional Format

```{r anova_table}
cat("ANOVA Table\n")
cat("===========\n\n")
cat("Source         df        SS          MS         F      p-value\n")
cat("---------------------------------------------------------------\n")
cat(sprintf("Between      %4d  %10.2f  %10.2f  %8.3f  %.4f\n",
            df_between, SSB, MSB, F_stat, p_value))
cat(sprintf("Within       %4d  %10.2f  %10.2f\n", df_within, SSW, MSW))
cat(sprintf("Total        %4d  %10.2f\n\n", N - 1, SST))

if (p_value < 0.05) {
    cat("Conclusion: Reject H₀. There is significant evidence that\n")
    cat("           at least one group mean differs from the others.\n")
} else {
    cat("Conclusion: Fail to reject H₀. No significant differences\n")
    cat("           detected among group means.\n")
}
```

### 13.6.2 Using R's aov() Function

```{r aov_function}
# Fit ANOVA using aov()
anova_model <- aov(BMI ~ AgeDecade, data = bmi_sample)

cat("\nR Output: aov() and summary()\n")
cat("=============================\n\n")
print(summary(anova_model))
```

---

## 13.7 Effect Sizes

### 13.7.1 Eta-Squared (η²)

The proportion of variance explained by group membership:

$$\eta^2 = \frac{\text{SSB}}{\text{SST}}$$

This is analogous to R² in regression.

```{r eta_squared}
eta_sq <- SSB / SST

cat("Eta-Squared (η²)\n")
cat("================\n\n")
cat(sprintf("η² = SSB / SST = %.2f / %.2f = %.4f\n\n", SSB, SST, eta_sq))
cat(sprintf("Interpretation: %.1f%% of the variance in BMI is\n", eta_sq * 100))
cat("               explained by age group membership.\n\n")

cat("Effect size guidelines (Cohen):\n")
cat("  η² = 0.01: Small effect\n")
cat("  η² = 0.06: Medium effect\n")
cat("  η² = 0.14: Large effect\n")
```

### 13.7.2 Omega-Squared (ω²)

A less biased estimator of population effect size:

$$\omega^2 = \frac{\text{SSB} - (k-1) \cdot \text{MSW}}{\text{SST} + \text{MSW}}$$

```{r omega_squared}
omega_sq <- (SSB - (k - 1) * MSW) / (SST + MSW)

cat("Omega-Squared (ω²)\n")
cat("==================\n\n")
cat(sprintf("ω² = (SSB - (k-1)×MSW) / (SST + MSW)\n"))
cat(sprintf("   = (%.2f - %d × %.2f) / (%.2f + %.2f)\n",
            SSB, k-1, MSW, SST, MSW))
cat(sprintf("   = %.4f\n\n", omega_sq))
cat("Note: ω² is a more conservative (less biased) estimate than η².\n")
```

---

## 13.8 Checking ANOVA Assumptions

### 13.8.1 Normality Check

```{r normality_check}
# Q-Q plots by group
ggplot2$ggplot(bmi_sample, ggplot2$aes(sample = BMI)) +
    ggplot2$stat_qq(colour = "#0072B2", alpha = 0.5) +
    ggplot2$stat_qq_line(colour = "#D55E00") +
    ggplot2$facet_wrap(~AgeDecade) +
    ggplot2$labs(
        title = "Normality Check: Q-Q Plots by Group",
        subtitle = "Points should follow the reference line",
        x = "Theoretical Quantiles",
        y = "Sample Quantiles"
    ) +
    ggplot2$theme_minimal(base_size = 11)
```

### 13.8.2 Homogeneity of Variance (Levene's Test)

```{r levene_test}
# Manual Levene's test (using median for robustness)
levene_data <- copy(bmi_sample)
levene_data[, median_group := median(BMI), by = AgeDecade]
levene_data[, abs_deviation := abs(BMI - median_group)]

levene_model <- aov(abs_deviation ~ AgeDecade, data = levene_data)
levene_result <- summary(levene_model)[[1]]

cat("Levene's Test for Homogeneity of Variance\n")
cat("==========================================\n\n")
cat("H₀: All group variances are equal\n")
cat("H₁: At least one group variance differs\n\n")
cat(sprintf("F(%d, %d) = %.3f\n",
            levene_result$Df[1], levene_result$Df[2], levene_result$`F value`[1]))
cat(sprintf("p-value = %.4f\n\n", levene_result$`Pr(>F)`[1]))

if (levene_result$`Pr(>F)`[1] < 0.05) {
    cat("Warning: Evidence of unequal variances. Consider Welch's ANOVA.\n")
} else {
    cat("No significant evidence against homogeneity of variance.\n")
}
```

### 13.8.3 Visual Check for Equal Variances

```{r variance_visual_check}
# Compare SDs across groups
var_check <- bmi_sample[, .(
    n = .N,
    mean = mean(BMI),
    sd = sd(BMI),
    var = var(BMI)
), by = AgeDecade][order(AgeDecade)]

cat("Variance by Group\n")
cat("=================\n\n")
print(var_check)

cat(sprintf("\nRatio of largest to smallest variance: %.2f\n",
            max(var_check$var) / min(var_check$var)))
cat("Rule of thumb: If ratio > 3, consider Welch's ANOVA.\n")
```

---

## 13.9 Welch's ANOVA

When variances are unequal, **Welch's ANOVA** provides a robust alternative that does not assume homogeneity:

```{r welch_anova}
# Welch's ANOVA
welch_result <- oneway.test(BMI ~ AgeDecade, data = bmi_sample, var.equal = FALSE)

cat("Welch's ANOVA (Does not assume equal variances)\n")
cat("================================================\n\n")
cat(sprintf("F(%.1f, %.1f) = %.3f\n",
            welch_result$parameter[1], welch_result$parameter[2], welch_result$statistic))
cat(sprintf("p-value = %.4f\n\n", welch_result$p.value))

# Compare with regular ANOVA
regular_result <- oneway.test(BMI ~ AgeDecade, data = bmi_sample, var.equal = TRUE)
cat("Comparison:\n")
cat(sprintf("  Regular ANOVA: F = %.3f, p = %.4f\n",
            regular_result$statistic, regular_result$p.value))
cat(sprintf("  Welch's ANOVA: F = %.3f, p = %.4f\n",
            welch_result$statistic, welch_result$p.value))
```

---

## 13.10 ANOVA as Linear Regression

### 13.10.1 The Connection

ANOVA is a special case of linear regression with categorical predictors. Using dummy coding:

```{r anova_as_regression}
# Fit as regression with dummy variables
reg_model <- lm(BMI ~ AgeDecade, data = bmi_sample)

cat("ANOVA as Linear Regression\n")
cat("==========================\n\n")
cat("Regression coefficients:\n")
print(round(coef(reg_model), 3))

cat("\nInterpretation:\n")
cat("  (Intercept): Mean BMI for reference group (20-29)\n")
cat("  AgeDecade30-39: Difference from reference (30-39 vs 20-29)\n")
cat("  etc.\n")
```

```{r anova_from_lm}
# Compare ANOVA tables
cat("\nANOVA table from lm():\n")
print(anova(reg_model))

cat("\nANOVA table from aov():\n")
print(summary(anova_model))

cat("\nNote: The F-statistics and p-values are identical!\n")
```

---

## 13.11 Communicating to Stakeholders

### 13.11.1 Reporting ANOVA Results

When reporting ANOVA:

1. State the research question
2. Report descriptive statistics by group
3. Report the F-statistic, degrees of freedom, and p-value
4. Report effect size (η² or ω²)
5. If significant, report post-hoc comparisons (next section)

```{r anova_report}
cat("Example Write-Up\n")
cat("================\n\n")

cat("Research Question: Does BMI differ across adult age groups?\n\n")

cat("Descriptive Statistics:\n")
descriptives <- bmi_sample[, .(
    M = round(mean(BMI), 2),
    SD = round(sd(BMI), 2),
    n = .N
), by = AgeDecade][order(AgeDecade)]
print(descriptives)

cat("\n\nResults:\n")
cat("A one-way ANOVA was conducted to compare BMI across five age\n")
cat(sprintf("decades. There was a statistically significant difference,\n"))
cat(sprintf("F(%d, %d) = %.2f, p = %.3f, η² = %.3f.\n\n",
            df_between, df_within, F_stat, p_value, eta_sq))

if (p_value < 0.05) {
    cat("This represents a ")
    if (eta_sq < 0.06) {
        cat("small")
    } else if (eta_sq < 0.14) {
        cat("medium")
    } else {
        cat("large")
    }
    cat(" effect. Post-hoc comparisons are needed to\n")
    cat("determine which specific groups differ.\n")
}
```

---

## Quick Reference

### Key Formulas

| Quantity | Formula |
|----------|---------|
| SST (Total) | $\sum\sum (Y_{ij} - \bar{Y}_{..})^2$ |
| SSB (Between) | $\sum n_i (\bar{Y}_{i.} - \bar{Y}_{..})^2$ |
| SSW (Within) | $\sum\sum (Y_{ij} - \bar{Y}_{i.})^2$ |
| MSB | $\text{SSB} / (k-1)$ |
| MSW | $\text{SSW} / (N-k)$ |
| F | $\text{MSB} / \text{MSW}$ |
| η² | $\text{SSB} / \text{SST}$ |
| ω² | $(\text{SSB} - (k-1)\text{MSW}) / (\text{SST} + \text{MSW})$ |

### Assumptions

| Assumption | Check | If Violated |
|------------|-------|-------------|
| Independence | Study design | Serious—use mixed models |
| Normality | Q-Q plots, Shapiro-Wilk | Robust with large n; use Kruskal-Wallis |
| Homoscedasticity | Levene's test, variance ratio | Use Welch's ANOVA |

### R Functions

| Function | Purpose |
|----------|---------|
| `aov(y ~ group, data)` | Fit one-way ANOVA |
| `summary(aov_model)` | ANOVA table |
| `oneway.test(y ~ group, var.equal = FALSE)` | Welch's ANOVA |
| `TukeyHSD(aov_model)` | Post-hoc comparisons |

