---
title: "Two-Way and Factorial ANOVA"
subtitle: "Part 2 of Chapter 13: ANOVA"
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 10,
    fig.height = 6,
    collapse = FALSE, results = 'hold'
)

box::use(
    data.table[...],
    ggplot2
)
```

## Table of Contents

## 13.12 Introduction to Two-Way ANOVA

### 13.12.1 Why Two Factors?

One-way ANOVA examines the effect of a single factor. But often we want to study:
- The effects of **two factors** simultaneously
- Whether the factors **interact** (does the effect of one depend on the other?)

Examples:
- Effect of drug type AND dosage on blood pressure
- Effect of teaching method AND class size on test scores
- Effect of gender AND age group on BMI

```{r packages}
box::use(
    data.table[...],
    ggplot2
)
```

```{r load_data}
# Load NHANES data
nhanes <- fread("../../../data/primary/nhanes.csv")
```

### 13.12.2 The Two-Way ANOVA Model

$$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk}$$

Where:
- $\mu$ is the grand mean
- $\alpha_i$ is the main effect of Factor A (level $i$)
- $\beta_j$ is the main effect of Factor B (level $j$)
- $(\alpha\beta)_{ij}$ is the **interaction** effect
- $\varepsilon_{ijk} \sim N(0, \sigma^2)$

---

## 13.13 Example: BMI by Gender and Age

```{r two_way_data}
# Create two-way data: BMI by Gender and Age group
two_way_data <- nhanes[!is.na(BMI) & !is.na(Gender) & !is.na(AgeDecade) &
                        AgeDecade != "" & Gender != ""]
two_way_data <- two_way_data[AgeDecade %in% c("20-29", "40-49", "60-69")]
two_way_data[, AgeGroup := factor(AgeDecade)]
two_way_data[, Gender := factor(Gender)]

# Sample for demonstration
set.seed(456)
sample_2way <- two_way_data[, .SD[sample(.N, min(50, .N))],
                             by = .(Gender, AgeGroup)]

cat("Sample sizes by cell:\n")
print(dcast(sample_2way[, .N, by = .(Gender, AgeGroup)],
            Gender ~ AgeGroup, value.var = "N"))
```

---

## 13.14 Visualising Two-Way Effects

### 13.14.1 Cell Means Plot

```{r cell_means_plot}
# Calculate cell means
cell_means <- sample_2way[, .(
    mean = mean(BMI),
    se = sd(BMI) / sqrt(.N),
    n = .N
), by = .(Gender, AgeGroup)]

ggplot2$ggplot(cell_means, ggplot2$aes(x = AgeGroup, y = mean, colour = Gender, group = Gender)) +
    ggplot2$geom_errorbar(ggplot2$aes(ymin = mean - se, ymax = mean + se),
                          width = 0.1, position = ggplot2$position_dodge(0.2)) +
    ggplot2$geom_line(position = ggplot2$position_dodge(0.2), linewidth = 1) +
    ggplot2$geom_point(position = ggplot2$position_dodge(0.2), size = 3) +
    ggplot2$scale_colour_manual(values = c("#0072B2", "#D55E00")) +
    ggplot2$labs(
        title = "Cell Means: BMI by Gender and Age Group",
        subtitle = "Non-parallel lines suggest an interaction",
        x = "Age Group",
        y = "Mean BMI (kg/m²)"
    ) +
    ggplot2$theme_minimal(base_size = 12)
```

### 13.14.2 Interpreting Interaction Plots

- **Parallel lines**: No interaction—the effect of age is the same for both genders
- **Non-parallel lines**: Interaction—the effect of age differs by gender
- **Crossing lines**: Strong interaction—the direction of the effect changes

```{r interaction_examples}
# Create examples of different patterns
set.seed(789)
n <- 30

# No interaction (parallel lines)
no_int <- data.table(
    Factor_A = rep(c("Low", "High"), each = n * 2),
    Factor_B = rep(c("Group 1", "Group 2"), each = n, times = 2),
    Y = c(rnorm(n, 10, 2), rnorm(n, 15, 2),  # Low: 10, 15
          rnorm(n, 14, 2), rnorm(n, 19, 2))  # High: 14, 19 (same difference of 5)
)

# With interaction (non-parallel)
with_int <- data.table(
    Factor_A = rep(c("Low", "High"), each = n * 2),
    Factor_B = rep(c("Group 1", "Group 2"), each = n, times = 2),
    Y = c(rnorm(n, 10, 2), rnorm(n, 12, 2),  # Low: 10, 12 (diff = 2)
          rnorm(n, 14, 2), rnorm(n, 22, 2))  # High: 14, 22 (diff = 8)
)

# Calculate means
no_int_means <- no_int[, .(Y = mean(Y)), by = .(Factor_A, Factor_B)]
with_int_means <- with_int[, .(Y = mean(Y)), by = .(Factor_A, Factor_B)]

no_int_means[, Pattern := "No Interaction\n(Parallel Lines)"]
with_int_means[, Pattern := "Interaction Present\n(Non-Parallel Lines)"]

pattern_data <- rbind(no_int_means, with_int_means)

ggplot2$ggplot(pattern_data, ggplot2$aes(x = Factor_A, y = Y, colour = Factor_B, group = Factor_B)) +
    ggplot2$geom_line(linewidth = 1.2) +
    ggplot2$geom_point(size = 4) +
    ggplot2$facet_wrap(~Pattern) +
    ggplot2$scale_colour_manual(values = c("#0072B2", "#D55E00")) +
    ggplot2$labs(
        title = "Recognising Interactions in Cell Mean Plots",
        x = "Factor A",
        y = "Response (Y)",
        colour = "Factor B"
    ) +
    ggplot2$theme_minimal(base_size = 12)
```

---

## 13.15 Partitioning Variance in Two-Way ANOVA

### 13.15.1 The Decomposition

$$\text{SST} = \text{SS}_A + \text{SS}_B + \text{SS}_{AB} + \text{SSE}$$

Where:
- $\text{SS}_A$: Main effect of Factor A
- $\text{SS}_B$: Main effect of Factor B
- $\text{SS}_{AB}$: Interaction effect
- $\text{SSE}$: Error (within cells)

### 13.15.2 Degrees of Freedom

| Source | df |
|--------|-----|
| Factor A | $a - 1$ |
| Factor B | $b - 1$ |
| A × B Interaction | $(a-1)(b-1)$ |
| Error | $N - ab$ |
| Total | $N - 1$ |

Where $a$ and $b$ are the number of levels for each factor.

---

## 13.16 Running Two-Way ANOVA in R

```{r two_way_anova}
# Fit two-way ANOVA
model_2way <- aov(BMI ~ Gender * AgeGroup, data = sample_2way)

cat("Two-Way ANOVA Results\n")
cat("=====================\n\n")
print(summary(model_2way))
```

### 13.16.1 Interpreting the Output

```{r interpret_2way}
# Extract results
anova_results <- summary(model_2way)[[1]]

cat("Interpretation\n")
cat("==============\n\n")

# Main effect of Gender
cat(sprintf("1. Main Effect of Gender:\n"))
cat(sprintf("   F(1, %d) = %.2f, p = %.4f\n",
            anova_results$Df[4], anova_results$`F value`[1], anova_results$`Pr(>F)`[1]))
if (anova_results$`Pr(>F)`[1] < 0.05) {
    cat("   → Significant main effect: BMI differs between genders.\n\n")
} else {
    cat("   → No significant main effect of gender.\n\n")
}

# Main effect of Age
cat(sprintf("2. Main Effect of Age Group:\n"))
cat(sprintf("   F(2, %d) = %.2f, p = %.4f\n",
            anova_results$Df[4], anova_results$`F value`[2], anova_results$`Pr(>F)`[2]))
if (anova_results$`Pr(>F)`[2] < 0.05) {
    cat("   → Significant main effect: BMI differs across age groups.\n\n")
} else {
    cat("   → No significant main effect of age group.\n\n")
}

# Interaction
cat(sprintf("3. Interaction (Gender × Age):\n"))
cat(sprintf("   F(2, %d) = %.2f, p = %.4f\n",
            anova_results$Df[4], anova_results$`F value`[3], anova_results$`Pr(>F)`[3]))
if (anova_results$`Pr(>F)`[3] < 0.05) {
    cat("   → Significant interaction: The effect of age on BMI\n")
    cat("     depends on gender (or vice versa).\n")
} else {
    cat("   → No significant interaction: The effect of age on BMI\n")
    cat("     is similar for both genders.\n")
}
```

---

## 13.17 Effect Sizes for Two-Way ANOVA

### 13.17.1 Partial Eta-Squared

Partial η² measures the proportion of variance explained by each factor, **controlling for other factors**:

$$\eta^2_p = \frac{\text{SS}_{\text{effect}}}{\text{SS}_{\text{effect}} + \text{SS}_{\text{error}}}$$

```{r partial_eta_squared}
# Calculate partial eta-squared for each effect
SS_gender <- anova_results$`Sum Sq`[1]
SS_age <- anova_results$`Sum Sq`[2]
SS_interaction <- anova_results$`Sum Sq`[3]
SS_error <- anova_results$`Sum Sq`[4]
SS_total <- sum(anova_results$`Sum Sq`)

eta_p_gender <- SS_gender / (SS_gender + SS_error)
eta_p_age <- SS_age / (SS_age + SS_error)
eta_p_int <- SS_interaction / (SS_interaction + SS_error)

cat("Partial Eta-Squared (η²p)\n")
cat("=========================\n\n")
cat(sprintf("Gender:       η²p = %.4f (%.1f%% of variance)\n", eta_p_gender, eta_p_gender * 100))
cat(sprintf("Age Group:    η²p = %.4f (%.1f%% of variance)\n", eta_p_age, eta_p_age * 100))
cat(sprintf("Interaction:  η²p = %.4f (%.1f%% of variance)\n\n", eta_p_int, eta_p_int * 100))

cat("Effect size guidelines:\n")
cat("  η²p ≈ 0.01: Small\n")
cat("  η²p ≈ 0.06: Medium\n")
cat("  η²p ≈ 0.14: Large\n")
```

---

## 13.18 When Interaction Is Significant

### 13.18.1 Focus on Simple Effects

When there's a significant interaction, the main effects become difficult to interpret. Instead, examine **simple effects**—the effect of one factor at each level of the other.

```{r simple_effects}
cat("Simple Effects Analysis\n")
cat("=======================\n\n")

# Effect of Age within each Gender
cat("Effect of Age Group within each Gender:\n\n")
for (g in c("female", "male")) {
    subset_data <- sample_2way[Gender == g]
    simple_model <- aov(BMI ~ AgeGroup, data = subset_data)
    result <- summary(simple_model)[[1]]
    cat(sprintf("  %s: F(2, %d) = %.2f, p = %.4f\n",
                g, result$Df[2], result$`F value`[1], result$`Pr(>F)`[1]))
}

cat("\n\nEffect of Gender within each Age Group:\n\n")
for (age in c("20-29", "40-49", "60-69")) {
    subset_data <- sample_2way[AgeGroup == age]
    simple_model <- aov(BMI ~ Gender, data = subset_data)
    result <- summary(simple_model)[[1]]
    cat(sprintf("  %s: F(1, %d) = %.2f, p = %.4f\n",
                age, result$Df[2], result$`F value`[1], result$`Pr(>F)`[1]))
}
```

---

## 13.19 Post-Hoc Comparisons

### 13.19.1 Tukey's HSD

After finding a significant main effect (in the absence of interaction), use post-hoc tests:

```{r tukey_hsd}
# One-way ANOVA for age (ignoring gender for simplicity)
one_way_model <- aov(BMI ~ AgeGroup, data = sample_2way)

cat("Tukey's Honest Significant Difference\n")
cat("=====================================\n\n")

tukey_result <- TukeyHSD(one_way_model, "AgeGroup")
print(tukey_result)
```

### 13.19.2 Visualising Post-Hoc Results

```{r tukey_visual}
# Extract Tukey results
tukey_df <- as.data.table(tukey_result$AgeGroup, keep.rownames = "comparison")

ggplot2$ggplot(tukey_df, ggplot2$aes(x = comparison, y = diff)) +
    ggplot2$geom_errorbar(ggplot2$aes(ymin = lwr, ymax = upr), width = 0.2, colour = "#0072B2") +
    ggplot2$geom_point(size = 3, colour = "#D55E00") +
    ggplot2$geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    ggplot2$coord_flip() +
    ggplot2$labs(
        title = "Tukey's HSD: Pairwise Comparisons",
        subtitle = "Error bars show 95% CI; comparisons not crossing zero are significant",
        x = "Comparison",
        y = "Mean Difference in BMI"
    ) +
    ggplot2$theme_minimal(base_size = 12)
```

---

## 13.20 Factorial Designs

### 13.20.1 Extending to More Factors

The principles extend to three or more factors:

$$Y_{ijkl} = \mu + \alpha_i + \beta_j + \gamma_k + (\alpha\beta)_{ij} + (\alpha\gamma)_{ik} + (\beta\gamma)_{jk} + (\alpha\beta\gamma)_{ijk} + \varepsilon_{ijkl}$$

With three factors, we have:
- 3 main effects
- 3 two-way interactions
- 1 three-way interaction

```{r three_way_example}
# Example with three factors (simulated)
set.seed(101)
n_per_cell <- 10

# Create 2 × 2 × 2 factorial design
factorial_data <- CJ(
    Factor_A = c("A1", "A2"),
    Factor_B = c("B1", "B2"),
    Factor_C = c("C1", "C2")
)
factorial_data <- factorial_data[rep(1:.N, each = n_per_cell)]

# Generate response with main effects
factorial_data[, Y := 50 +
    5 * (Factor_A == "A2") +
    3 * (Factor_B == "B2") +
    -2 * (Factor_C == "C2") +
    4 * (Factor_A == "A2" & Factor_B == "B2") +  # A × B interaction
    rnorm(.N, 0, 5)]

# Fit three-way ANOVA
model_3way <- aov(Y ~ Factor_A * Factor_B * Factor_C, data = factorial_data)

cat("Three-Way Factorial ANOVA\n")
cat("=========================\n\n")
print(summary(model_3way))
```

### 13.20.2 Interpreting Higher-Order Interactions

- **Two-way interaction**: Effect of A depends on level of B
- **Three-way interaction**: The A × B interaction differs across levels of C

Higher-order interactions are difficult to interpret and often require graphical analysis.

```{r three_way_visual}
# Visualise three-way interaction
means_3way <- factorial_data[, .(Y = mean(Y)), by = .(Factor_A, Factor_B, Factor_C)]

ggplot2$ggplot(means_3way, ggplot2$aes(x = Factor_A, y = Y, colour = Factor_B, group = Factor_B)) +
    ggplot2$geom_line(linewidth = 1) +
    ggplot2$geom_point(size = 3) +
    ggplot2$facet_wrap(~Factor_C, labeller = ggplot2$labeller(Factor_C = c(C1 = "Factor C = C1", C2 = "Factor C = C2"))) +
    ggplot2$scale_colour_manual(values = c("#0072B2", "#D55E00")) +
    ggplot2$labs(
        title = "Visualising a Three-Way Interaction",
        subtitle = "If the pattern differs across panels, a three-way interaction exists",
        x = "Factor A",
        y = "Mean Response",
        colour = "Factor B"
    ) +
    ggplot2$theme_minimal(base_size = 12)
```

---

## 13.21 Type I, II, and III Sums of Squares

### 13.21.1 The Issue with Unbalanced Designs

When cell sizes are unequal, the sums of squares can be calculated in different ways:

- **Type I (Sequential)**: Order of entry matters
- **Type II (Hierarchical)**: Tests each effect adjusted for all others except higher-order interactions
- **Type III (Marginal)**: Tests each effect adjusted for all other effects

```{r ss_types}
# Create unbalanced data
set.seed(202)
unbalanced <- sample_2way[sample(.N, 150)]

cat("Cell Sizes (Unbalanced Design):\n")
print(dcast(unbalanced[, .N, by = .(Gender, AgeGroup)],
            Gender ~ AgeGroup, value.var = "N"))

cat("\n\nType I SS (order: Gender, AgeGroup):\n")
print(anova(lm(BMI ~ Gender + AgeGroup, data = unbalanced)))

cat("\nType I SS (order: AgeGroup, Gender):\n")
print(anova(lm(BMI ~ AgeGroup + Gender, data = unbalanced)))

cat("\nNote: The SS values differ depending on order!\n")
cat("For unbalanced designs, Type III SS is often preferred.\n")
```

### 13.21.2 When to Use Each Type

| Type | Use When |
|------|----------|
| Type I | Sequential hypotheses, order matters |
| Type II | No interaction expected, main effects only |
| Type III | Unbalanced designs, interaction present |

---

## 13.22 Checking Assumptions

### 13.22.1 Residual Analysis

```{r assumptions_2way}
# Extract residuals
sample_2way[, residual := residuals(model_2way)]
sample_2way[, fitted := fitted(model_2way)]

# Residuals vs Fitted
p1 <- ggplot2$ggplot(sample_2way, ggplot2$aes(x = fitted, y = residual)) +
    ggplot2$geom_point(alpha = 0.5, colour = "#0072B2") +
    ggplot2$geom_hline(yintercept = 0, linetype = "dashed", colour = "#D55E00") +
    ggplot2$geom_smooth(method = "loess", se = FALSE, colour = "#009E73") +
    ggplot2$labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
    ggplot2$theme_minimal()

# Q-Q Plot
p2 <- ggplot2$ggplot(sample_2way, ggplot2$aes(sample = residual)) +
    ggplot2$stat_qq(colour = "#0072B2", alpha = 0.5) +
    ggplot2$stat_qq_line(colour = "#D55E00") +
    ggplot2$labs(title = "Normal Q-Q Plot", x = "Theoretical Quantiles", y = "Sample Quantiles") +
    ggplot2$theme_minimal()

print(p1)
print(p2)
```

### 13.22.2 Homogeneity Across Cells

```{r homogeneity_cells}
# Check variance across cells
cell_var <- sample_2way[, .(
    n = .N,
    mean = mean(BMI),
    var = var(BMI)
), by = .(Gender, AgeGroup)]

cat("Variance by Cell\n")
cat("================\n\n")
print(cell_var)

cat(sprintf("\nVariance ratio (max/min): %.2f\n",
            max(cell_var$var) / min(cell_var$var)))
```

---

## 13.23 Communicating to Stakeholders

### 13.23.1 Reporting Two-Way ANOVA

```{r two_way_report}
cat("Example Write-Up: Two-Way ANOVA\n")
cat("================================\n\n")

cat("Research Question: Does BMI differ by gender and age group, and\n")
cat("is there an interaction between these factors?\n\n")

cat("Method:\n")
cat("A two-way ANOVA was conducted with Gender (male, female) and\n")
cat("Age Group (20-29, 40-49, 60-69) as between-subjects factors.\n\n")

cat("Results:\n")

# Main effect of Gender
if (anova_results$`Pr(>F)`[1] < 0.05) {
    cat(sprintf("There was a significant main effect of Gender, F(1, %d) = %.2f,\n",
                anova_results$Df[4], anova_results$`F value`[1]))
    cat(sprintf("p = %.3f, η²p = %.3f. ", anova_results$`Pr(>F)`[1], eta_p_gender))
} else {
    cat(sprintf("The main effect of Gender was not significant, F(1, %d) = %.2f,\n",
                anova_results$Df[4], anova_results$`F value`[1]))
    cat(sprintf("p = %.3f. ", anova_results$`Pr(>F)`[1]))
}

# Main effect of Age
if (anova_results$`Pr(>F)`[2] < 0.05) {
    cat(sprintf("\n\nThere was a significant main effect of Age Group, F(2, %d) = %.2f,\n",
                anova_results$Df[4], anova_results$`F value`[2]))
    cat(sprintf("p = %.3f, η²p = %.3f. ", anova_results$`Pr(>F)`[2], eta_p_age))
} else {
    cat(sprintf("\n\nThe main effect of Age Group was not significant, F(2, %d) = %.2f,\n",
                anova_results$Df[4], anova_results$`F value`[2]))
    cat(sprintf("p = %.3f. ", anova_results$`Pr(>F)`[2]))
}

# Interaction
if (anova_results$`Pr(>F)`[3] < 0.05) {
    cat(sprintf("\n\nThe Gender × Age Group interaction was significant, F(2, %d) = %.2f,\n",
                anova_results$Df[4], anova_results$`F value`[3]))
    cat(sprintf("p = %.3f, η²p = %.3f, indicating that the effect of age on BMI\n",
                anova_results$`Pr(>F)`[3], eta_p_int))
    cat("differs between genders.\n")
} else {
    cat(sprintf("\n\nThe Gender × Age Group interaction was not significant, F(2, %d) = %.2f,\n",
                anova_results$Df[4], anova_results$`F value`[3]))
    cat(sprintf("p = %.3f, suggesting that the effect of age on BMI is similar\n",
                anova_results$`Pr(>F)`[3]))
    cat("for both genders.\n")
}
```

---

## Quick Reference

### Two-Way ANOVA Formulas

| Source | SS Formula | df |
|--------|------------|-----|
| Factor A | $\sum n_{i.}(\bar{Y}_{i.} - \bar{Y}_{..})^2$ | $a - 1$ |
| Factor B | $\sum n_{.j}(\bar{Y}_{.j} - \bar{Y}_{..})^2$ | $b - 1$ |
| A × B | $\sum n_{ij}(\bar{Y}_{ij} - \bar{Y}_{i.} - \bar{Y}_{.j} + \bar{Y}_{..})^2$ | $(a-1)(b-1)$ |
| Error | $\sum (Y_{ijk} - \bar{Y}_{ij})^2$ | $N - ab$ |

### Interpretation Guide

| Scenario | What to Report |
|----------|----------------|
| Significant interaction | Focus on simple effects, graph the interaction |
| No interaction | Report and interpret main effects separately |
| Non-significant main effect | Effect may still exist at some levels (check simple effects) |

### R Functions

| Function | Purpose |
|----------|---------|
| `aov(y ~ A * B, data)` | Two-way ANOVA with interaction |
| `aov(y ~ A + B, data)` | Two-way ANOVA without interaction |
| `interaction.plot(A, B, y)` | Base R interaction plot |
| `TukeyHSD(model)` | Post-hoc comparisons |
| `summary(model)` | ANOVA table |

### Effect Size Guidelines

| Effect Size | η²p Value | Interpretation |
|-------------|-----------|----------------|
| Small | 0.01 | Minimal practical importance |
| Medium | 0.06 | Moderate practical importance |
| Large | 0.14 | Substantial practical importance |

