---
title: "Machine Learning Foundations for Finance"
subtitle: "Adapting ML methods for financial prediction"
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
box::use(
    data.table[...],
    ggplot2[...]
)

box::use(
    ../modules/data[load_market, load_factors, filter_dates],
    ../modules/stats[sharpe_ratio, annualised_return, annualised_vol, max_drawdown],
    ../modules/viz[theme_trading, trading_colors]
)

knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 10,
    fig.height = 6,
    fig.path = ""
)

tc <- unlist(trading_colors)
```

## 11.1 Why Finance is Different

### 11.1.1 Prose & Intuition

Machine learning in finance faces unique challenges that don't exist in other domains. Unlike image classification where a cat remains a cat, financial relationships are non-stationary—patterns that worked yesterday may not work tomorrow. Markets adapt; when an edge becomes known, it erodes.

The **signal-to-noise ratio** in financial data is extremely low. Most price movements are noise, not signal. A classifier that's 51% accurate sounds terrible for ImageNet but could be enormously valuable in trading. The challenge is distinguishing genuine predictive signals from spurious correlations.

**Overfitting** is the central enemy. With limited data and unlimited features, it's trivially easy to fit past data perfectly while having zero predictive power. Financial ML requires paranoid cross-validation and healthy skepticism of any apparent pattern.

The reward function in finance isn't classification accuracy—it's risk-adjusted returns. A model that's correct 90% of the time but wrong during crashes is worthless. Financial ML must focus on when the model is wrong, not just how often.

### 11.1.2 Visual Evidence

```{r load-data}
# Load market data
spy <- load_market("SPY")
ff5 <- load_factors("ff5", freq = "daily")

start_date <- as.Date("2005-01-01")
end_date <- as.Date("2023-12-31")

spy <- filter_dates(spy, start_date, end_date)
ff5 <- filter_dates(ff5, start_date, end_date)

# Merge
data <- merge(spy[, .(date, price = adjusted, ret = returns)], ff5, by = "date", all = FALSE)
data <- data[!is.na(ret)]
```

```{r signal-vs-noise, fig.cap="Financial returns are dominated by noise. The distribution shows fat tails and near-zero mean."}
# Demonstrate signal-to-noise ratio
ret_mean <- mean(data$ret, na.rm = TRUE)
ret_sd <- sd(data$ret, na.rm = TRUE)
snr <- abs(ret_mean) / ret_sd

cat(sprintf("Daily mean return: %.4f%%\n", ret_mean * 100))
cat(sprintf("Daily std deviation: %.2f%%\n", ret_sd * 100))
cat(sprintf("Signal-to-noise ratio: %.4f\n", snr))
cat(sprintf("Annualised SNR: %.4f\n", snr * sqrt(252)))

# Plot return distribution
ggplot(data, aes(x = ret)) +
    geom_histogram(aes(y = after_stat(density)), bins = 100,
                   fill = tc[1], alpha = 0.7, colour = "white") +
    stat_function(fun = dnorm, args = list(mean = ret_mean, sd = ret_sd),
                  colour = tc[2], linewidth = 1) +
    geom_vline(xintercept = ret_mean, colour = tc[3], linewidth = 1, linetype = "dashed") +
    scale_x_continuous(labels = scales::percent_format(), limits = c(-0.1, 0.1)) +
    labs(title = "SPY Daily Return Distribution",
         subtitle = sprintf("Mean: %.4f%% (barely visible); Noise dominates signal", ret_mean * 100),
         x = "Daily Return", y = "Density") +
    theme_trading()
```

```{r regime-changes, fig.cap="Financial regimes change over time. Relationships that hold in one period may break in another."}
# Demonstrate regime changes: rolling correlation between SPY and HML (value)
data[, rolling_cor_hml := {
    n <- .N
    result <- rep(NA_real_, n)
    for (i in 252:n) {
        idx <- (i - 251):i
        result[i] <- cor(ret[idx], HML[idx], use = "complete.obs")
    }
    result
}]

ggplot(data[!is.na(rolling_cor_hml)], aes(x = date, y = rolling_cor_hml)) +
    geom_line(colour = tc[1], linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title = "Rolling 252-Day Correlation: SPY vs Value Factor (HML)",
         subtitle = "Relationship changes dramatically over time—non-stationarity",
         x = NULL, y = "Correlation") +
    theme_trading()
```

```{r overfitting-demo, fig.cap="Overfitting demonstration. In-sample R² improves with model complexity but out-of-sample degrades."}
# Simulate overfitting with polynomial features
set.seed(42)

# Create feature: lagged returns
data[, lag1 := shift(ret, 1)]
data[, lag2 := shift(ret, 2)]
data[, lag5 := shift(ret, 5)]
data[, lag10 := shift(ret, 10)]
data[, lag20 := shift(ret, 20)]

# Filter valid data
train_data <- data[!is.na(lag20)]

# Split: first 70% train, last 30% test
n <- nrow(train_data)
train_idx <- 1:floor(n * 0.7)
test_idx <- (floor(n * 0.7) + 1):n

train_set <- train_data[train_idx]
test_set <- train_data[test_idx]

# Fit models with increasing complexity
models <- list()
complexities <- c(1, 2, 5, 10, 20)
results <- data.table(
    complexity = integer(),
    r2_train = numeric(),
    r2_test = numeric()
)

for (p in complexities) {
    # Create polynomial features
    if (p == 1) {
        formula <- ret ~ lag1
    } else if (p == 2) {
        formula <- ret ~ lag1 + lag2
    } else if (p == 5) {
        formula <- ret ~ lag1 + lag2 + lag5 + I(lag1^2) + I(lag1*lag2)
    } else if (p == 10) {
        formula <- ret ~ lag1 + lag2 + lag5 + lag10 + I(lag1^2) + I(lag2^2) +
                        I(lag1*lag2) + I(lag1*lag5) + I(lag2*lag5) + I(lag1^3)
    } else {
        formula <- ret ~ lag1 + lag2 + lag5 + lag10 + lag20 + I(lag1^2) + I(lag2^2) +
                        I(lag5^2) + I(lag1*lag2) + I(lag1*lag5) + I(lag2*lag5) +
                        I(lag1*lag10) + I(lag1^3) + I(lag2^3) + I(lag1^2*lag2) +
                        I(lag1*lag2^2) + I(lag1*lag2*lag5) + I(lag1^4) + I(lag1^2*lag5) + I(lag1*lag5^2)
    }

    model <- lm(formula, data = train_set)

    # Calculate R²
    pred_train <- predict(model, train_set)
    pred_test <- predict(model, test_set)

    ss_tot_train <- sum((train_set$ret - mean(train_set$ret))^2)
    ss_res_train <- sum((train_set$ret - pred_train)^2)
    r2_train <- 1 - ss_res_train / ss_tot_train

    ss_tot_test <- sum((test_set$ret - mean(test_set$ret))^2)
    ss_res_test <- sum((test_set$ret - pred_test)^2)
    r2_test <- 1 - ss_res_test / ss_tot_test

    results <- rbind(results, data.table(complexity = p, r2_train = r2_train, r2_test = r2_test))
}

# Plot
plot_results <- melt(results, id.vars = "complexity", variable.name = "Set", value.name = "R2")
plot_results[, Set := fifelse(Set == "r2_train", "In-Sample (Train)", "Out-of-Sample (Test)")]

ggplot(plot_results, aes(x = factor(complexity), y = R2 * 100, fill = Set)) +
    geom_col(position = "dodge", alpha = 0.8) +
    scale_fill_manual(values = c(tc[1], tc[4])) +
    geom_hline(yintercept = 0, colour = "grey50") +
    labs(title = "Overfitting Demonstration: In-Sample vs Out-of-Sample R²",
         subtitle = "More features improve training fit but degrade test performance",
         x = "Number of Features", y = "R² (%)", fill = "") +
    theme_trading() +
    theme(legend.position = "top")
```

### 11.1.3 Mathematical Derivation

**Signal-to-Noise Ratio**

For daily returns with mean $\mu$ and volatility $\sigma$:

$$
SNR_{daily} = \frac{|\mu|}{\sigma}
$$

Typical values: $\mu \approx 0.04\%$, $\sigma \approx 1\%$, so $SNR \approx 0.04$.

Annualised SNR: $SNR_{ann} = SNR_{daily} \times \sqrt{252} \approx 0.6$

**Bias-Variance Tradeoff**

Expected prediction error:

$$
E[(y - \hat{f}(x))^2] = \underbrace{\text{Var}(\hat{f}(x))}_{\text{Variance}} + \underbrace{[\text{Bias}(\hat{f}(x))]^2}_{\text{Bias}^2} + \underbrace{\sigma^2_\epsilon}_{\text{Irreducible}}
$$

In finance, the irreducible error ($\sigma^2_\epsilon$) dominates. Reducing it below a certain level is impossible.

**False Discovery Rate**

When testing $m$ hypotheses with significance level $\alpha$:

$$
E[\text{False Discoveries}] = m_0 \cdot \alpha
$$

where $m_0$ is the number of true null hypotheses. With 1000 features tested, even at $\alpha = 0.01$, expect 10 false discoveries.

### 11.1.4 Implementation & Application

```{r proper-cv, fig.cap="Time series cross-validation. Walk-forward validation respects temporal structure."}
# Demonstrate proper time series cross-validation
# Walk-forward: train on past, test on future

# Create simple features
cv_data <- copy(train_data)
cv_data[, pred_feature := shift(ret, 1) + shift(ret, 2)]

# Walk-forward CV parameters
window_train <- 252 * 2  # 2 years training
window_test <- 63  # 3 months testing
step_size <- 63

# Perform walk-forward
cv_results <- list()
start_idx <- window_train + 1

while (start_idx + window_test <= nrow(cv_data)) {
    train_idx <- (start_idx - window_train):(start_idx - 1)
    test_idx <- start_idx:(min(start_idx + window_test - 1, nrow(cv_data)))

    train_cv <- cv_data[train_idx]
    test_cv <- cv_data[test_idx]

    # Simple model: predict sign based on feature
    threshold <- median(train_cv$pred_feature, na.rm = TRUE)
    train_cv[, signal := fifelse(pred_feature > threshold, 1, -1)]
    test_cv[, signal := fifelse(pred_feature > threshold, 1, -1)]

    # Calculate returns
    test_cv[, strat_ret := shift(signal, 1) * ret]

    cv_results[[length(cv_results) + 1]] <- data.table(
        fold = length(cv_results) + 1,
        start_date = cv_data[test_idx[1], date],
        end_date = cv_data[test_idx[length(test_idx)], date],
        n_days = length(test_idx),
        mean_ret = mean(test_cv$strat_ret, na.rm = TRUE),
        sharpe = mean(test_cv$strat_ret, na.rm = TRUE) / sd(test_cv$strat_ret, na.rm = TRUE) * sqrt(252)
    )

    start_idx <- start_idx + step_size
}

cv_results <- rbindlist(cv_results)

# Plot CV results over time
ggplot(cv_results, aes(x = start_date, y = sharpe)) +
    geom_col(aes(fill = sharpe > 0), alpha = 0.8) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_hline(yintercept = mean(cv_results$sharpe), colour = tc[2], linetype = "dashed") +
    scale_fill_manual(values = c(tc[4], tc[1]), guide = "none") +
    labs(title = "Walk-Forward Cross-Validation Results",
         subtitle = sprintf("Mean OOS Sharpe: %.2f; Consistency matters more than magnitude",
                           mean(cv_results$sharpe)),
         x = "Fold Start Date", y = "Annualised Sharpe") +
    theme_trading()
```

```{r feature-importance, fig.cap="Feature importance decay. Even significant features have limited predictive power."}
# Demonstrate feature importance with proper testing
# Using Fama-French factors as features

ff_features <- copy(data)
ff_features <- ff_features[!is.na(lag20)]

# Create lagged factor features
ff_features[, mkt_lag1 := shift(Mkt_RF, 1)]
ff_features[, smb_lag1 := shift(SMB, 1)]
ff_features[, hml_lag1 := shift(HML, 1)]
ff_features[, rmw_lag1 := shift(RMW, 1)]
ff_features[, cma_lag1 := shift(CMA, 1)]

# Correlations with future returns
feature_cors <- ff_features[!is.na(mkt_lag1), .(
    mkt = cor(ret, mkt_lag1, use = "complete.obs"),
    smb = cor(ret, smb_lag1, use = "complete.obs"),
    hml = cor(ret, hml_lag1, use = "complete.obs"),
    rmw = cor(ret, rmw_lag1, use = "complete.obs"),
    cma = cor(ret, cma_lag1, use = "complete.obs"),
    lag1 = cor(ret, lag1, use = "complete.obs"),
    lag5 = cor(ret, lag5, use = "complete.obs")
)]

feature_cors_long <- melt(feature_cors, variable.name = "Feature", value.name = "Correlation")

ggplot(feature_cors_long, aes(x = reorder(Feature, abs(Correlation)), y = Correlation)) +
    geom_col(fill = tc[1], alpha = 0.8) +
    geom_hline(yintercept = 0, colour = "grey50") +
    coord_flip() +
    labs(title = "Feature Correlations with Future Returns",
         subtitle = "All correlations are extremely small—signal is weak",
         x = "Feature", y = "Correlation") +
    theme_trading()
```

---

## 11.2 Feature Engineering for Finance

### 11.2.1 Prose & Intuition

Good features are the foundation of any ML model. In finance, raw price data is typically not useful—we need to transform it into informative signals. The best features capture meaningful economic relationships, not arbitrary patterns.

**Technical features** include returns, volatility, momentum, mean reversion signals, and volume-based indicators. These capture market microstructure and investor behaviour.

**Fundamental features** include valuation ratios, earnings surprises, and quality metrics. These capture economic value.

**Macro features** include interest rates, credit spreads, economic indicators, and cross-asset correlations. These capture regime and environment.

Feature engineering in finance requires domain knowledge. A feature that makes economic sense is more likely to persist than one discovered purely through data mining.

### 11.2.2 Visual Evidence

```{r momentum-feature, fig.cap="Momentum feature engineering. Different lookback periods capture different signals."}
# Create momentum features at various horizons
data[, mom_5d := price / shift(price, 5) - 1]
data[, mom_20d := price / shift(price, 20) - 1]
data[, mom_60d := price / shift(price, 60) - 1]
data[, mom_252d := price / shift(price, 252) - 1]

# Correlation between momentum and forward returns
mom_features <- data[!is.na(mom_252d)]
mom_features[, fwd_ret_20d := shift(frollsum(ret, 20), -20)]

mom_cors <- mom_features[!is.na(fwd_ret_20d), .(
    `5-day` = cor(mom_5d, fwd_ret_20d),
    `20-day` = cor(mom_20d, fwd_ret_20d),
    `60-day` = cor(mom_60d, fwd_ret_20d),
    `252-day` = cor(mom_252d, fwd_ret_20d)
)]

mom_cors_long <- melt(mom_cors, variable.name = "Horizon", value.name = "Correlation")

ggplot(mom_cors_long, aes(x = Horizon, y = Correlation)) +
    geom_col(fill = tc[1], alpha = 0.8) +
    geom_hline(yintercept = 0, colour = "grey50") +
    labs(title = "Momentum Feature Correlation with Forward 20-Day Returns",
         subtitle = "Different horizons capture different effects",
         x = "Momentum Lookback", y = "Correlation") +
    theme_trading()
```

```{r volatility-features, fig.cap="Volatility features capture risk regime. High vol predicts different returns than low vol."}
# Create volatility features
data[, vol_20d := frollapply(ret, 20, sd) * sqrt(252)]
data[, vol_60d := frollapply(ret, 60, sd) * sqrt(252)]
data[, vol_ratio := vol_20d / vol_60d]

# Vol regime and forward returns
vol_data <- data[!is.na(vol_ratio)]
vol_data[, vol_quintile := cut(vol_20d, breaks = quantile(vol_20d, probs = seq(0, 1, 0.2), na.rm = TRUE),
                               labels = c("Q1 (Low)", "Q2", "Q3", "Q4", "Q5 (High)"),
                               include.lowest = TRUE)]

# Forward returns by volatility quintile
vol_returns <- vol_data[!is.na(vol_quintile), .(
    fwd_ret = mean(shift(ret, -20), na.rm = TRUE) * 20,
    fwd_vol = mean(shift(vol_20d, -20), na.rm = TRUE)
), by = vol_quintile]

ggplot(vol_returns, aes(x = vol_quintile, y = fwd_ret * 100)) +
    geom_col(fill = tc[1], alpha = 0.8) +
    labs(title = "Forward 20-Day Return by Current Volatility Quintile",
         subtitle = "Volatility regime affects expected returns",
         x = "Volatility Quintile", y = "Forward 20-Day Return (%)") +
    theme_trading()
```

```{r feature-interactions, fig.cap="Feature interactions can be informative. Momentum in low vol vs high vol regimes."}
# Momentum effect by volatility regime
vol_data[, vol_regime := fifelse(vol_20d > median(vol_20d, na.rm = TRUE), "High Vol", "Low Vol")]
vol_data[, mom_quintile := cut(mom_60d, breaks = quantile(mom_60d, probs = seq(0, 1, 0.2), na.rm = TRUE),
                                labels = c("Q1 (Low)", "Q2", "Q3", "Q4", "Q5 (High)"),
                                include.lowest = TRUE)]

# Forward returns by momentum quintile and vol regime
interaction_returns <- vol_data[!is.na(mom_quintile) & !is.na(vol_regime), .(
    fwd_ret = mean(shift(ret, -20), na.rm = TRUE) * 20
), by = .(vol_regime, mom_quintile)]

ggplot(interaction_returns, aes(x = mom_quintile, y = fwd_ret * 100, fill = vol_regime)) +
    geom_col(position = "dodge", alpha = 0.8) +
    scale_fill_manual(values = c(tc[1], tc[4])) +
    labs(title = "Momentum Effect by Volatility Regime",
         subtitle = "Momentum may work differently in high vs low volatility",
         x = "Momentum Quintile", y = "Forward 20-Day Return (%)", fill = "Vol Regime") +
    theme_trading() +
    theme(legend.position = "top")
```

### 11.2.3 Mathematical Derivation

**Return Features**

Simple return: $r_t = \frac{P_t - P_{t-1}}{P_{t-1}}$

Log return: $r_t = \ln\left(\frac{P_t}{P_{t-1}}\right)$

Cumulative return: $R_{t,t+n} = \sum_{i=1}^{n} r_{t+i}$ (for log returns)

**Volatility Features**

Realised volatility: $\sigma_t = \sqrt{\sum_{i=t-n+1}^{t} r_i^2}$

Exponential moving volatility: $\sigma_t^{EMA} = \sqrt{\lambda \sigma_{t-1}^2 + (1-\lambda) r_t^2}$

**Momentum Features**

Price momentum: $\text{MOM}_t^{(n)} = \frac{P_t}{P_{t-n}} - 1$

Return momentum: $\text{MOM}_t^{(n)} = \sum_{i=t-n+1}^{t} r_i$

**Feature Normalisation**

Z-score: $z_t = \frac{x_t - \mu_{rolling}}{\sigma_{rolling}}$

Cross-sectional rank: $\text{rank}_t = \frac{\text{rank}(x_t)}{\text{N}}$

### 11.2.4 Implementation & Application

```{r feature-pipeline, fig.cap="Complete feature engineering pipeline. Systematic feature construction and normalisation."}
# Complete feature engineering pipeline
feature_data <- copy(data)

# 1. Price-based features
feature_data[, ret_1d := ret]
feature_data[, ret_5d := frollsum(ret, 5)]
feature_data[, ret_20d := frollsum(ret, 20)]

# 2. Volatility features
feature_data[, vol_10d := frollapply(ret, 10, sd) * sqrt(252)]
feature_data[, vol_20d := frollapply(ret, 20, sd) * sqrt(252)]
feature_data[, vol_ratio := vol_10d / vol_20d]

# 3. Mean reversion features (z-score of price vs moving average)
feature_data[, ma_20 := frollmean(price, 20)]
feature_data[, ma_60 := frollmean(price, 60)]
feature_data[, price_ma_ratio := price / ma_20 - 1]
feature_data[, ma_cross := ma_20 / ma_60 - 1]

# 4. Factor features (lagged)
feature_data[, mkt_lag := shift(Mkt_RF, 1)]
feature_data[, smb_lag := shift(SMB, 1)]
feature_data[, hml_lag := shift(HML, 1)]

# 5. Normalise features (rolling z-score)
normalise_features <- c("ret_5d", "ret_20d", "vol_ratio", "price_ma_ratio", "ma_cross")
for (feat in normalise_features) {
    feature_data[, paste0(feat, "_z") := {
        mu <- frollmean(get(feat), 252)
        sig <- frollapply(get(feat), 252, sd)
        (get(feat) - mu) / sig
    }]
}

# Filter valid data
feature_data <- feature_data[!is.na(ma_cross_z)]

# Show feature correlations
feature_cols <- c("ret_5d_z", "ret_20d_z", "vol_ratio_z", "price_ma_ratio_z", "ma_cross_z")
feature_cors <- cor(feature_data[, ..feature_cols], use = "complete.obs")

# Melt for heatmap
feature_cors_dt <- as.data.table(feature_cors, keep.rownames = "Feature1")
feature_cors_long <- melt(feature_cors_dt, id.vars = "Feature1",
                          variable.name = "Feature2", value.name = "Correlation")

ggplot(feature_cors_long, aes(x = Feature1, y = Feature2, fill = Correlation)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", Correlation)), colour = "white", size = 3) +
    scale_fill_gradient2(low = tc[4], mid = "white", high = tc[1], midpoint = 0) +
    labs(title = "Feature Correlation Matrix",
         subtitle = "Low correlation between features = good diversification",
         x = NULL, y = NULL) +
    theme_trading() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r target-engineering, fig.cap="Target engineering matters. Different targets capture different signals."}
# Different target definitions
feature_data[, target_1d := shift(ret, -1)]
feature_data[, target_5d := shift(frollsum(ret, 5), -5)]
feature_data[, target_sign := fifelse(shift(ret, -1) > 0, 1, 0)]
feature_data[, target_quintile := cut(shift(ret, -1),
                                       breaks = quantile(shift(ret, -1), probs = seq(0, 1, 0.2), na.rm = TRUE),
                                       labels = 1:5, include.lowest = TRUE)]

# Base rate for sign prediction
base_rate <- mean(feature_data$target_sign, na.rm = TRUE)
cat(sprintf("Base rate (positive days): %.1f%%\n", base_rate * 100))

# Distribution of targets
target_dist <- data.table(
    Target = c("1-Day Return", "5-Day Return"),
    Mean = c(mean(feature_data$target_1d, na.rm = TRUE),
             mean(feature_data$target_5d, na.rm = TRUE)),
    SD = c(sd(feature_data$target_1d, na.rm = TRUE),
           sd(feature_data$target_5d, na.rm = TRUE))
)

print(target_dist)
```

---

## 11.3 Cross-Validation for Finance

### 11.3.1 Prose & Intuition

Standard k-fold cross-validation is invalid for financial time series. It violates the fundamental principle that we can only train on the past to predict the future. Using future data in training—even indirectly—leads to severe overfitting.

**Walk-forward validation** is the gold standard. Train on a fixed historical window, test on the next period, then roll forward. This mimics actual trading: you only know the past when making decisions.

**Purging** removes observations near the train/test boundary to prevent information leakage. If predicting 5-day returns, observations within 5 days of the boundary must be excluded.

**Embargo** extends purging to account for serial correlation. Even with purging, adjacent observations may be correlated enough to leak information.

The key insight: if your backtesting methodology wouldn't work for actual trading, it's invalid for evaluation.

### 11.3.2 Visual Evidence

```{r cv-schemes, fig.cap="Cross-validation schemes compared. Walk-forward respects temporal order; k-fold does not."}
# Visualise different CV schemes
n_obs <- 100
cv_viz <- data.table(
    obs = 1:n_obs,
    k_fold = rep(1:5, each = 20),
    walk_forward = c(rep("Train", 60), rep("Test", 40))
)

# K-fold visualisation
kfold_plot <- ggplot(cv_viz, aes(x = obs, y = 1, fill = factor(k_fold))) +
    geom_tile(height = 0.8) +
    scale_fill_manual(values = c(tc[1], tc[2], tc[3], tc[4], "grey50")) +
    labs(title = "K-Fold CV: Folds are Mixed in Time",
         subtitle = "Problem: Test data comes before training data",
         x = "Observation", y = NULL, fill = "Fold") +
    theme_trading() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          legend.position = "none")

# Walk-forward visualisation
wf_plot <- ggplot(cv_viz, aes(x = obs, y = 1, fill = walk_forward)) +
    geom_tile(height = 0.8) +
    scale_fill_manual(values = c(tc[1], tc[4])) +
    labs(title = "Walk-Forward: Train Then Test",
         subtitle = "Correct: Only train on past, test on future",
         x = "Observation", y = NULL, fill = "") +
    theme_trading() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          legend.position = "top")

print(kfold_plot)
print(wf_plot)
```

```{r purging-demo, fig.cap="Purging prevents information leakage. Observations near the boundary are excluded."}
# Demonstrate purging
# Without purging: correlation between train tail and test head
# With purging: gap eliminates leakage

# Autocorrelation of features
acf_data <- acf(feature_data$ret_5d_z[!is.na(feature_data$ret_5d_z)], lag.max = 20, plot = FALSE)
acf_df <- data.table(lag = 0:20, acf = acf_data$acf[1:21])

ggplot(acf_df, aes(x = lag, y = acf)) +
    geom_col(fill = tc[1], alpha = 0.7) +
    geom_hline(yintercept = c(-1, 1) * 1.96 / sqrt(nrow(feature_data)), linetype = "dashed", colour = tc[2]) +
    labs(title = "Feature Autocorrelation (5-Day Return Z-Score)",
         subtitle = "Significant autocorrelation requires purging to prevent leakage",
         x = "Lag (Days)", y = "Autocorrelation") +
    theme_trading()
```

### 11.3.3 Mathematical Derivation

**Walk-Forward Validation**

For time series $\{(x_t, y_t)\}_{t=1}^{T}$:

Training window at fold $k$: $t \in [t_k - w_{train}, t_k)$

Test window at fold $k$: $t \in [t_k, t_k + w_{test})$

where $w_{train}$ and $w_{test}$ are window sizes.

**Purging**

If target $y_t$ depends on returns from $t$ to $t+h$:

Purge training observations where $t_{train} \in [t_{test} - h, t_{test})$

**Combinatorial Purged Cross-Validation**

For $N$ paths and $k$ test groups:

$$
\text{Number of combinations} = \binom{N}{k}
$$

Each combination provides one test fold with proper purging.

**Information Coefficient Estimation**

$$
IC = \text{Corr}(\hat{y}, y)
$$

Standard error of IC under null:

$$
SE(IC) = \frac{1}{\sqrt{n_{eff}}}
$$

where $n_{eff}$ accounts for serial correlation.

### 11.3.4 Implementation & Application

```{r walk-forward-implementation, fig.cap="Walk-forward implementation with purging. Proper CV for financial prediction."}
# Implement proper walk-forward CV with purging
walk_forward_cv <- function(data, feature_cols, target_col,
                            train_window = 504, test_window = 63,
                            purge_window = 5, step_size = 63) {

    results <- list()
    n <- nrow(data)
    start_idx <- train_window + purge_window + 1

    while (start_idx + test_window <= n) {
        # Training indices (with purge gap)
        train_end <- start_idx - purge_window - 1
        train_start <- max(1, train_end - train_window + 1)
        train_idx <- train_start:train_end

        # Test indices
        test_idx <- start_idx:(start_idx + test_window - 1)

        # Extract data
        train <- data[train_idx]
        test <- data[test_idx]

        # Simple linear model
        formula <- as.formula(paste(target_col, "~", paste(feature_cols, collapse = " + ")))
        model <- tryCatch({
            lm(formula, data = train)
        }, error = function(e) NULL)

        if (!is.null(model)) {
            # Predictions
            pred <- predict(model, test)
            actual <- test[[target_col]]

            # Metrics
            ic <- cor(pred, actual, use = "complete.obs")
            rmse <- sqrt(mean((pred - actual)^2, na.rm = TRUE))

            # Trading metric: if prediction > 0, go long
            test_copy <- copy(test)
            test_copy[, signal := fifelse(pred > 0, 1, -1)]
            test_copy[, strat_ret := shift(signal, 1) * ret]
            sharpe <- mean(test_copy$strat_ret, na.rm = TRUE) / sd(test_copy$strat_ret, na.rm = TRUE) * sqrt(252)

            results[[length(results) + 1]] <- data.table(
                fold = length(results) + 1,
                train_start = data[train_start, date],
                test_start = data[start_idx, date],
                ic = ic,
                rmse = rmse,
                sharpe = sharpe
            )
        }

        start_idx <- start_idx + step_size
    }

    return(rbindlist(results))
}

# Run CV
cv_results <- walk_forward_cv(
    data = feature_data,
    feature_cols = c("ret_5d_z", "vol_ratio_z", "price_ma_ratio_z"),
    target_col = "target_1d",
    train_window = 504,  # 2 years
    test_window = 63,    # 3 months
    purge_window = 5,
    step_size = 63
)

# Summary
cat(sprintf("\nWalk-Forward CV Results (%d folds):\n", nrow(cv_results)))
cat(sprintf("  Mean IC: %.4f (t-stat: %.2f)\n",
           mean(cv_results$ic),
           mean(cv_results$ic) / (sd(cv_results$ic) / sqrt(nrow(cv_results)))))
cat(sprintf("  Mean Sharpe: %.2f\n", mean(cv_results$sharpe)))

# Plot IC over time
ggplot(cv_results, aes(x = test_start, y = ic)) +
    geom_col(aes(fill = ic > 0), alpha = 0.8) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_hline(yintercept = mean(cv_results$ic), colour = tc[2], linetype = "dashed") +
    scale_fill_manual(values = c(tc[4], tc[1]), guide = "none") +
    labs(title = "Information Coefficient Over Time (Walk-Forward CV)",
         subtitle = sprintf("Mean IC: %.4f; Consistency indicates robustness",
                           mean(cv_results$ic)),
         x = "Test Period Start", y = "IC") +
    theme_trading()
```

```{r cv-stability, fig.cap="CV stability analysis. A good model shows consistent performance across folds."}
# Rolling Sharpe stability
cv_results[, rolling_sharpe := frollmean(sharpe, 4)]

ggplot(cv_results, aes(x = test_start)) +
    geom_col(aes(y = sharpe), fill = tc[1], alpha = 0.5) +
    geom_line(aes(y = rolling_sharpe), colour = tc[2], linewidth = 1, na.rm = TRUE) +
    geom_hline(yintercept = 0, colour = "grey50") +
    labs(title = "Sharpe Ratio Stability Across CV Folds",
         subtitle = "Rolling 4-fold average shows regime dependency",
         x = "Test Period Start", y = "Annualised Sharpe") +
    theme_trading()

# Distribution of Sharpes
ggplot(cv_results, aes(x = sharpe)) +
    geom_histogram(bins = 15, fill = tc[1], alpha = 0.7, colour = "white") +
    geom_vline(xintercept = 0, colour = "grey50") +
    geom_vline(xintercept = mean(cv_results$sharpe), colour = tc[2], linetype = "dashed") +
    labs(title = "Distribution of Out-of-Sample Sharpe Ratios",
         subtitle = sprintf("Mean: %.2f; Std: %.2f", mean(cv_results$sharpe), sd(cv_results$sharpe)),
         x = "Sharpe Ratio", y = "Count") +
    theme_trading()
```
