---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 20: Logistic Regression"
part: "Part 2: Assessment and Extensions"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, logistic-regression, classification, roc, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

```{r packages, include=FALSE}
box::use(ggplot2)
box::use(data.table)
```

```{r dt-symbols, include=FALSE}
.N <- data.table::.N
.SD <- data.table::.SD
```

# Chapter 20 -- Logistic Regression

## Part 2: Assessment and Extensions

In Part 1 we fitted logistic regression models and interpreted coefficients as odds ratios. But how well does the model *predict*? In this part we develop a crash course on classification metrics -- sensitivity, specificity, precision, recall, ROC curves, and AUC -- and then extend logistic regression to multinomial and ordinal outcomes.

---

## 20.6 Classification Metrics

### 20.6.1 The Confusion Matrix

When we use logistic regression for prediction, we convert the estimated probability $\hat{p}_i$ into a class label using a **threshold** $c$ (typically 0.5):

$$\hat{Y}_i = \begin{cases} 1 & \text{if } \hat{p}_i \ge c \\ 0 & \text{if } \hat{p}_i < c \end{cases}$$

The confusion matrix tabulates predictions against actual values:

|  | Predicted Positive | Predicted Negative |
|:---:|:---:|:---:|
| **Actual Positive** | True Positive (TP) | False Negative (FN) |
| **Actual Negative** | False Positive (FP) | True Negative (TN) |

```{r confusion-matrix}
data_dir <- "../../../data"
heart <- data.table$fread(file.path(data_dir, "classic/heart_disease.csv"),
  header = FALSE,
  col.names = c("age", "sex", "cp", "trestbps", "chol", "fbs", "restecg",
                "thalach", "exang", "oldpeak", "slope", "ca", "thal", "target"))
heart[, target := as.integer(target > 0)]

set.seed(42)
n <- nrow(heart)
train_idx <- sample(seq_len(n), floor(0.7 * n))
train <- heart[train_idx]
test <- heart[-train_idx]

fit_heart <- glm(target ~ age + sex + cp + trestbps + chol + thalach + oldpeak,
                 data = train, family = binomial)

test[, p_hat := predict(fit_heart, newdata = test, type = "response")]
test[, y_hat := as.integer(p_hat >= 0.5)]

conf_mat <- table(Predicted = test$y_hat, Actual = test$target)
conf_mat
```

### 20.6.2 Metrics Derived from the Confusion Matrix

```{r classification-metrics}
tp <- conf_mat["1", "1"]
tn <- conf_mat["0", "0"]
fp <- conf_mat["1", "0"]
fn <- conf_mat["0", "1"]

metrics <- data.frame(
  Metric = c("Sensitivity (Recall, TPR)", "Specificity (TNR)",
             "Precision (PPV)", "Negative Predictive Value (NPV)",
             "Accuracy", "F1 Score"),
  Formula = c("TP / (TP + FN)", "TN / (TN + FP)",
              "TP / (TP + FP)", "TN / (TN + FN)",
              "(TP + TN) / n", "2 * Prec * Rec / (Prec + Rec)"),
  Value = round(c(
    tp / (tp + fn),  # sensitivity
    tn / (tn + fp),  # specificity
    tp / (tp + fp),  # precision
    tn / (tn + fn),  # NPV
    (tp + tn) / nrow(test),  # accuracy
    2 * (tp / (tp + fp)) * (tp / (tp + fn)) / ((tp / (tp + fp)) + (tp / (tp + fn)))  # F1
  ), 3)
)
knitr::kable(metrics, caption = "Classification metrics for the heart disease model at threshold 0.5.")
```

### 20.6.3 The Sensitivity-Specificity Trade-Off

Lowering the threshold increases sensitivity (we catch more true positives) but decreases specificity (we generate more false positives). The optimal threshold depends on the cost of each type of error.

```{r threshold-tradeoff, fig.cap="Sensitivity and specificity as a function of the classification threshold. Lowering the threshold increases sensitivity at the expense of specificity."}
thresholds <- seq(0.01, 0.99, by = 0.01)
trade_dt <- data.table$rbindlist(lapply(thresholds, function(c) {
  y_hat <- as.integer(test$p_hat >= c)
  tp_c <- sum(y_hat == 1 & test$target == 1)
  tn_c <- sum(y_hat == 0 & test$target == 0)
  fp_c <- sum(y_hat == 1 & test$target == 0)
  fn_c <- sum(y_hat == 0 & test$target == 1)
  data.table$data.table(
    threshold = c,
    Sensitivity = tp_c / max(tp_c + fn_c, 1),
    Specificity = tn_c / max(tn_c + fp_c, 1)
  )
}))

trade_long <- data.table$melt(trade_dt, id.vars = "threshold",
                               variable.name = "metric", value.name = "value")

ggplot2$ggplot(trade_long, ggplot2$aes(x = threshold, y = value, colour = metric)) +
  ggplot2$geom_line(linewidth = 1) +
  ggplot2$labs(
    x = "Classification threshold",
    y = "Value",
    colour = "Metric",
    title = "Sensitivity vs specificity trade-off"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "bottom")
```

---

## 20.7 ROC Curves

### 20.7.1 What Is an ROC Curve?

The **Receiver Operating Characteristic (ROC) curve** plots sensitivity (TPR) against $1 - \text{specificity}$ (FPR) as the threshold varies. It summarises the model's discriminative ability across *all* thresholds simultaneously.

```{r roc-curve-scratch, fig.cap="ROC curve for the heart disease logistic regression, built from scratch. The diagonal represents random guessing; the curve's distance above the diagonal measures discriminative power."}
roc_dt <- data.table$rbindlist(lapply(thresholds, function(c) {
  y_hat <- as.integer(test$p_hat >= c)
  tp_c <- sum(y_hat == 1 & test$target == 1)
  tn_c <- sum(y_hat == 0 & test$target == 0)
  fp_c <- sum(y_hat == 1 & test$target == 0)
  fn_c <- sum(y_hat == 0 & test$target == 1)
  data.table$data.table(
    threshold = c,
    TPR = tp_c / max(tp_c + fn_c, 1),
    FPR = fp_c / max(fp_c + tn_c, 1)
  )
}))

# Sort by FPR for proper curve
roc_dt <- roc_dt[order(FPR, TPR)]

ggplot2$ggplot(roc_dt, ggplot2$aes(x = FPR, y = TPR)) +
  ggplot2$geom_line(colour = "#0072B2", linewidth = 1.2) +
  ggplot2$geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  ggplot2$labs(
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)",
    title = "ROC curve for heart disease prediction"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

### 20.7.2 Area Under the Curve (AUC)

The **AUC** (area under the ROC curve) is a single-number summary of discriminative performance:

- $\text{AUC} = 0.5$: no better than random guessing.
- $\text{AUC} = 1.0$: perfect discrimination.
- $0.7 \le \text{AUC} < 0.8$: acceptable.
- $0.8 \le \text{AUC} < 0.9$: excellent.
- $\text{AUC} \ge 0.9$: outstanding.

We compute it using the trapezoidal rule:

```{r auc-scratch}
# Sort by FPR ascending, then TPR ascending
roc_sorted <- roc_dt[order(FPR, TPR)]
auc_val <- 0
for (i in 2:nrow(roc_sorted)) {
  dx <- roc_sorted$FPR[i] - roc_sorted$FPR[i - 1]
  avg_y <- (roc_sorted$TPR[i] + roc_sorted$TPR[i - 1]) / 2
  auc_val <- auc_val + dx * avg_y
}

data.frame(AUC = round(auc_val, 4))
```

**Concordance interpretation**: the AUC equals the probability that a randomly chosen positive case has a higher predicted probability than a randomly chosen negative case.

```{r concordance-check}
# Verify via concordance
pos_probs <- test$p_hat[test$target == 1]
neg_probs <- test$p_hat[test$target == 0]
concordant <- 0
total <- 0
for (p in pos_probs) {
  concordant <- concordant + sum(p > neg_probs) + 0.5 * sum(p == neg_probs)
  total <- total + length(neg_probs)
}
c(AUC_trapezoidal = round(auc_val, 4),
  AUC_concordance = round(concordant / total, 4))
```

---

## 20.8 Calibration

A model is **well-calibrated** if, among all observations with predicted probability $\hat{p} \approx 0.3$, roughly 30% are actually positive. We assess calibration with a **calibration plot**.

```{r calibration-plot, fig.cap="Calibration plot for the heart disease model. Perfect calibration lies on the diagonal. Points above the line indicate the model underestimates the probability."}
test[, prob_bin := cut(p_hat, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)]
cal_dt <- test[, .(
  mean_predicted = mean(p_hat),
  mean_observed = mean(target),
  n = .N
), by = prob_bin]

ggplot2$ggplot(cal_dt, ggplot2$aes(x = mean_predicted, y = mean_observed)) +
  ggplot2$geom_point(ggplot2$aes(size = n), colour = "#0072B2") +
  ggplot2$geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  ggplot2$labs(
    x = "Mean predicted probability",
    y = "Observed proportion",
    size = "n in bin",
    title = "Calibration plot for heart disease model"
  ) +
  ggplot2$scale_size_continuous(range = c(2, 8)) +
  ggplot2$theme_minimal(base_size = 13)
```

Good discrimination (high AUC) does not guarantee good calibration. A model can rank patients correctly but assign the wrong absolute probabilities. In medical applications, calibration is often more important than discrimination.

---

## 20.9 Comparing Models

```{r model-comparison}
fit_simple <- glm(target ~ age + sex + cp, data = train, family = binomial)
fit_full <- glm(target ~ age + sex + cp + trestbps + chol + thalach + oldpeak,
                data = train, family = binomial)

test[, p_simple := predict(fit_simple, newdata = test, type = "response")]
test[, p_full := predict(fit_full, newdata = test, type = "response")]

# AUC for both models (using concordance)
compute_auc <- function(p_hat, y) {
  pos <- p_hat[y == 1]
  neg <- p_hat[y == 0]
  conc <- 0
  for (p in pos) conc <- conc + sum(p > neg) + 0.5 * sum(p == neg)
  conc / (length(pos) * length(neg))
}

data.frame(
  Model = c("Simple (age + sex + cp)", "Full (7 predictors)"),
  AUC = round(c(
    compute_auc(test$p_simple, test$target),
    compute_auc(test$p_full, test$target)
  ), 4),
  AIC = round(c(AIC(fit_simple), AIC(fit_full)), 1)
)
```

```{r roc-comparison, fig.cap="ROC curves for simple and full models. The full model has slightly better discrimination."}
roc_compare <- function(p_hat, y, label) {
  thresholds <- seq(0.01, 0.99, by = 0.01)
  data.table$rbindlist(lapply(thresholds, function(c) {
    y_hat <- as.integer(p_hat >= c)
    tp_c <- sum(y_hat == 1 & y == 1)
    fp_c <- sum(y_hat == 1 & y == 0)
    fn_c <- sum(y_hat == 0 & y == 1)
    tn_c <- sum(y_hat == 0 & y == 0)
    data.table$data.table(
      FPR = fp_c / max(fp_c + tn_c, 1),
      TPR = tp_c / max(tp_c + fn_c, 1),
      model = label
    )
  }))
}

roc_both <- data.table$rbindlist(list(
  roc_compare(test$p_simple, test$target, "Simple"),
  roc_compare(test$p_full, test$target, "Full")
))

ggplot2$ggplot(roc_both, ggplot2$aes(x = FPR, y = TPR, colour = model)) +
  ggplot2$geom_line(linewidth = 1) +
  ggplot2$geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  ggplot2$labs(
    x = "False Positive Rate",
    y = "True Positive Rate",
    colour = "Model",
    title = "ROC comparison: simple vs full model"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "bottom")
```

---

## 20.10 Predicted Probability Distribution

```{r pred-dist, fig.cap="Distribution of predicted probabilities by actual class. Good models separate the two distributions."}
pred_dist <- data.table$data.table(
  p_hat = test$p_full,
  actual = factor(test$target, levels = c(0, 1), labels = c("No disease", "Disease"))
)

ggplot2$ggplot(pred_dist, ggplot2$aes(x = p_hat, fill = actual)) +
  ggplot2$geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  ggplot2$labs(
    x = "Predicted probability",
    y = "Count",
    fill = "Actual status",
    title = "Distribution of predicted probabilities by class"
  ) +
  ggplot2$scale_fill_manual(values = c("#0072B2", "#D55E00")) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "bottom")
```

---

## 20.11 Multinomial Logistic Regression

When the response has $K > 2$ unordered categories, we extend to **multinomial logistic regression**. We pick a reference category (say category 1) and model $K - 1$ log-odds simultaneously:

$$\log\frac{P(Y = k)}{P(Y = 1)} = \mathbf{x}^\top \boldsymbol{\beta}_k, \quad k = 2, \ldots, K$$

The probabilities are:

$$P(Y = k) = \frac{\exp(\mathbf{x}^\top \boldsymbol{\beta}_k)}{1 + \sum_{j=2}^{K} \exp(\mathbf{x}^\top \boldsymbol{\beta}_j)}$$

This is sometimes called the **softmax** function in machine learning.

```{r multinomial-demo}
# Demonstrate with iris data (3 species)
iris_dt <- data.table$fread(file.path(data_dir, "classic/iris.csv"))

# Use nnet::multinom for multinomial logistic
if (requireNamespace("nnet", quietly = TRUE)) {
  fit_multi_iris <- nnet::multinom(
    Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
    data = iris_dt, trace = FALSE
  )
  summary(fit_multi_iris)
}
```

```{r multinomial-pred, fig.cap="Predicted class probabilities for the iris data, plotted against petal length. The three sigmoid-like curves partition the probability space."}
pred_grid_iris <- data.table$data.table(
  Sepal.Length = median(iris_dt$Sepal.Length),
  Sepal.Width = median(iris_dt$Sepal.Width),
  Petal.Length = seq(1, 7, length.out = 200),
  Petal.Width = median(iris_dt$Petal.Width)
)

if (requireNamespace("nnet", quietly = TRUE)) {
  probs <- predict(fit_multi_iris, newdata = pred_grid_iris, type = "probs")
  pred_long <- data.table$data.table(
    Petal.Length = rep(pred_grid_iris$Petal.Length, 3),
    probability = c(probs[, 1], probs[, 2], probs[, 3]),
    species = rep(colnames(probs), each = 200)
  )

  ggplot2$ggplot(pred_long, ggplot2$aes(x = Petal.Length, y = probability, colour = species)) +
    ggplot2$geom_line(linewidth = 1.1) +
    ggplot2$labs(
      x = "Petal length (cm)",
      y = "Predicted probability",
      colour = "Species",
      title = "Multinomial logistic regression: iris species probabilities"
    ) +
    ggplot2$theme_minimal(base_size = 13) +
    ggplot2$theme(legend.position = "bottom")
}
```

---

## 20.12 Ordinal Logistic Regression

When the response categories have a natural order (e.g., "mild", "moderate", "severe"), we use the **proportional odds model**:

$$\log\frac{P(Y \le k)}{P(Y > k)} = \alpha_k - \mathbf{x}^\top \boldsymbol{\beta}, \quad k = 1, \ldots, K-1$$

The key assumption is that the $\boldsymbol{\beta}$ coefficients are the same across all cutpoints -- only the intercepts $\alpha_k$ differ. This is the **proportional odds assumption**.

```{r ordinal-demo}
# Create an ordinal outcome from heart disease data
heart[, severity := cut(target, breaks = c(-0.5, 0.5, 1.5),
                         labels = c("healthy", "disease"))]

# For a proper ordinal demo, create severity from continuous variables
heart[, risk_group := cut(oldpeak, breaks = c(-Inf, 0.5, 2.0, Inf),
                           labels = c("low", "moderate", "high"), ordered_result = TRUE)]

if (requireNamespace("MASS", quietly = TRUE)) {
  fit_ord <- MASS::polr(risk_group ~ age + sex + thalach, data = heart, Hess = TRUE)
  summary(fit_ord)
}
```

The proportional odds assumption can be tested using the Brant test. If it fails, a multinomial model may be more appropriate.

---

## 20.13 Practical Workflow Summary

A complete logistic regression analysis follows these steps:

1. **Explore** the data: check class balance, missing values, correlations.
2. **Fit** the model using `glm(..., family = binomial)`.
3. **Test** coefficients with Wald and LRT.
4. **Interpret** via odds ratios and predicted probability curves.
5. **Assess** discrimination (ROC/AUC) and calibration.
6. **Compare** models using AIC, LRT, or cross-validated AUC.
7. **Report** odds ratios with CIs and a predicted probability plot.

---

## Communicating to Stakeholders

When presenting classification model performance:

- **Do not report accuracy alone.** If 95% of patients are healthy, a model that always says "healthy" achieves 95% accuracy but is useless. Report sensitivity and specificity separately.
- **Explain the ROC curve visually.** "The curve shows how well the model separates patients with and without disease. The closer the curve hugs the top-left corner, the better."
- **Choose the right metric for the context.** In screening for a serious disease, sensitivity (catching all true cases) matters more than specificity. In a spam filter, precision (not annoying users with false alarms) may matter more.
- **Calibration matters.** If you tell a patient "your risk is 40%", it had better be roughly 40%. A well-calibrated model is an honest model.

---

## Quick Reference

| Metric | Formula | Interpretation |
|:-------|:--------|:---------------|
| **Sensitivity** | TP / (TP + FN) | P(predict + | actually +) |
| **Specificity** | TN / (TN + FP) | P(predict - | actually -) |
| **Precision** | TP / (TP + FP) | P(actually + | predict +) |
| **NPV** | TN / (TN + FN) | P(actually - | predict -) |
| **Accuracy** | (TP + TN) / n | Overall correct rate |
| **F1 Score** | Harmonic mean of Prec and Rec | Balanced measure |
| **AUC** | Area under ROC | P(score+ > score-) |
| **Multinomial** | `nnet::multinom()` | Unordered K categories |
| **Ordinal** | `MASS::polr()` | Ordered categories |

---

## Exercises

1. **Threshold selection.** For the heart disease model, find the threshold that maximises Youden's index $J = \text{sensitivity} + \text{specificity} - 1$. How does this threshold differ from 0.5?

2. **Cross-validated AUC.** Implement 5-fold cross-validation for the heart disease logistic regression. Compute the AUC on each held-out fold and report the mean and standard deviation.

3. **Cost-sensitive threshold.** Suppose a false negative (missed disease) costs 10 times as much as a false positive. Find the optimal threshold that minimises the expected cost.

4. **Calibration improvement.** Fit a logistic regression, then recalibrate it using Platt scaling (fit a logistic regression of the outcome on the predicted probabilities). Compare calibration plots before and after.

5. **Multinomial extension.** Using the iris data, fit a multinomial logistic regression with all four predictors. Compute the overall accuracy and per-class sensitivity. Which species is hardest to classify?

6. **Brier score.** The Brier score is $\frac{1}{n}\sum(y_i - \hat{p}_i)^2$. Compute it for both the simple and full heart disease models. Which model has a lower (better) Brier score? How does it relate to AUC?
