---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 16: Multiple Linear Regression"
part: "Part 2: Gauss-Markov and Inference"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, regression, gauss-markov, inference, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

```{r packages, include=FALSE}
box::use(ggplot2)
box::use(data.table)
```

```{r dt-symbols, include=FALSE}
.N <- data.table::.N
.SD <- data.table::.SD
```

# Chapter 16 -- Multiple Linear Regression

## Part 2: Gauss-Markov and Inference

In Part 1 we derived $\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$ and showed that OLS is orthogonal projection. Now we ask: is this the *best* estimator? The **Gauss-Markov theorem** gives a precise and affirmative answer. We then derive the full sampling distribution under normality, develop $t$-tests for individual coefficients and $F$-tests for groups of coefficients, and interpret R's `summary()` and `anova()` output.

---

## 16.8 Properties of the OLS Estimator

### 16.8.1 Unbiasedness

**Theorem.** Under the assumptions $E[\boldsymbol{\varepsilon}] = \mathbf{0}$ and $\text{rank}(\mathbf{X}) = p$, the OLS estimator is unbiased: $E[\hat{\boldsymbol{\beta}}] = \boldsymbol{\beta}$.

*Proof.*

$$\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'(\mathbf{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon}) = \boldsymbol{\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\boldsymbol{\varepsilon}$$

Taking expectations (with $\mathbf{X}$ treated as fixed):

$$E[\hat{\boldsymbol{\beta}}] = \boldsymbol{\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'E[\boldsymbol{\varepsilon}] = \boldsymbol{\beta} + \mathbf{0} = \boldsymbol{\beta} \quad \square$$

### 16.8.2 Covariance Matrix

**Theorem.** $\text{Cov}(\hat{\boldsymbol{\beta}}) = \sigma^2(\mathbf{X}'\mathbf{X})^{-1}$.

*Proof.* From $\hat{\boldsymbol{\beta}} - \boldsymbol{\beta} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\boldsymbol{\varepsilon}$:

$$\text{Cov}(\hat{\boldsymbol{\beta}}) = E[(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta})(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta})']$$

$$= (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}' \, E[\boldsymbol{\varepsilon}\boldsymbol{\varepsilon}'] \, \mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}$$

$$= (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}' \, \sigma^2\mathbf{I} \, \mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}$$

$$= \sigma^2 (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{X}(\mathbf{X}'\mathbf{X})^{-1} = \sigma^2(\mathbf{X}'\mathbf{X})^{-1} \quad \square$$

The standard error of $\hat{\beta}_j$ is therefore:

$$\text{SE}(\hat{\beta}_j) = s\sqrt{[(\mathbf{X}'\mathbf{X})^{-1}]_{jj}}$$

where $s = \sqrt{\text{SSE}/(n - p)}$ is the residual standard error.

---

## 16.9 The Gauss-Markov Theorem -- Full Proof

### 16.9.1 Statement

**Gauss-Markov Theorem.** Under the conditions:

1. $E[\boldsymbol{\varepsilon}] = \mathbf{0}$
2. $\text{Cov}(\boldsymbol{\varepsilon}) = \sigma^2\mathbf{I}$
3. $\text{rank}(\mathbf{X}) = p$

the OLS estimator $\hat{\boldsymbol{\beta}}$ is the **Best Linear Unbiased Estimator** (BLUE) of $\boldsymbol{\beta}$. That is, for any linear function $\mathbf{a}'\boldsymbol{\beta}$, the OLS estimator $\mathbf{a}'\hat{\boldsymbol{\beta}}$ has the smallest variance among all linear unbiased estimators of $\mathbf{a}'\boldsymbol{\beta}$.

### 16.9.2 Proof

Let $\tilde{\boldsymbol{\beta}} = \mathbf{C}\mathbf{Y}$ be any other linear estimator of $\boldsymbol{\beta}$, where $\mathbf{C}$ is a $p \times n$ matrix. Write $\mathbf{C} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}' + \mathbf{D}$, where $\mathbf{D} = \mathbf{C} - (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'$ is the "deviation" from the OLS solution.

**Step 1: Unbiasedness constraint.**

$$E[\tilde{\boldsymbol{\beta}}] = \mathbf{C}\mathbf{X}\boldsymbol{\beta} = [(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}' + \mathbf{D}]\mathbf{X}\boldsymbol{\beta} = \boldsymbol{\beta} + \mathbf{D}\mathbf{X}\boldsymbol{\beta}$$

For $\tilde{\boldsymbol{\beta}}$ to be unbiased for all $\boldsymbol{\beta}$, we need $\mathbf{D}\mathbf{X} = \mathbf{0}$.

**Step 2: Variance of $\tilde{\boldsymbol{\beta}}$.**

$$\text{Cov}(\tilde{\boldsymbol{\beta}}) = \sigma^2 \mathbf{C}\mathbf{C}' = \sigma^2[(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}' + \mathbf{D}][(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}' + \mathbf{D}]'$$

Expanding:

$$= \sigma^2\left[(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{X}(\mathbf{X}'\mathbf{X})^{-1} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{D}' + \mathbf{D}\mathbf{X}(\mathbf{X}'\mathbf{X})^{-1} + \mathbf{D}\mathbf{D}'\right]$$

Since $\mathbf{D}\mathbf{X} = \mathbf{0}$, the cross-terms vanish:

$$\text{Cov}(\tilde{\boldsymbol{\beta}}) = \sigma^2(\mathbf{X}'\mathbf{X})^{-1} + \sigma^2\mathbf{D}\mathbf{D}'$$

**Step 3: Conclusion.**

The matrix $\mathbf{D}\mathbf{D}'$ is positive semi-definite (since for any vector $\mathbf{v}$, we have $\mathbf{v}'\mathbf{D}\mathbf{D}'\mathbf{v} = \|\mathbf{D}'\mathbf{v}\|^2 \ge 0$). Therefore:

$$\text{Cov}(\tilde{\boldsymbol{\beta}}) - \text{Cov}(\hat{\boldsymbol{\beta}}) = \sigma^2\mathbf{D}\mathbf{D}' \succeq \mathbf{0}$$

In particular, for any $\mathbf{a}$:

$$\text{Var}(\mathbf{a}'\tilde{\boldsymbol{\beta}}) = \mathbf{a}'\text{Cov}(\tilde{\boldsymbol{\beta}})\mathbf{a} \ge \mathbf{a}'\text{Cov}(\hat{\boldsymbol{\beta}})\mathbf{a} = \text{Var}(\mathbf{a}'\hat{\boldsymbol{\beta}})$$

with equality if and only if $\mathbf{D}'\mathbf{a} = \mathbf{0}$. Thus OLS is BLUE. $\square$

### 16.9.3 What BLUE Does and Does Not Mean

BLUE says OLS is optimal **among linear unbiased estimators**. It does not say:

- OLS is optimal among all estimators (biased estimators like ridge regression can have lower MSE).
- OLS is the MLE (it is, but only under normality).
- The assumptions hold in practice (they must be checked via diagnostics).

---

## 16.10 Sampling Distributions Under Normality

If $\boldsymbol{\varepsilon} \sim N(\mathbf{0}, \sigma^2\mathbf{I})$, then:

$$\hat{\boldsymbol{\beta}} \sim N\!\left(\boldsymbol{\beta},\, \sigma^2(\mathbf{X}'\mathbf{X})^{-1}\right)$$

and:

$$\frac{(n - p)s^2}{\sigma^2} = \frac{\mathbf{e}'\mathbf{e}}{\sigma^2} \sim \chi^2_{n-p}$$

Moreover, $\hat{\boldsymbol{\beta}}$ and $s^2$ are independent (because $\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$ depends on $\mathbf{H}\mathbf{Y}$ while $\mathbf{e} = (\mathbf{I} - \mathbf{H})\mathbf{Y}$, and for normal vectors, uncorrelatedness implies independence).

---

## 16.11 $t$-Tests for Individual Coefficients

### 16.11.1 The Test Statistic

For testing $H_0: \beta_j = 0$ vs $H_1: \beta_j \ne 0$:

$$t_j = \frac{\hat{\beta}_j}{\text{SE}(\hat{\beta}_j)} = \frac{\hat{\beta}_j}{s\sqrt{[(\mathbf{X}'\mathbf{X})^{-1}]_{jj}}} \sim t_{n-p} \quad \text{under } H_0$$

### 16.11.2 Implementation and Interpretation

```{r mtcars-t-tests}
data_dir <- "../../../data"
mtcars_dt <- data.table$fread(file.path(data_dir, "classic/mtcars.csv"))
fit <- lm(mpg ~ hp + wt + qsec, data = mtcars_dt)
s <- summary(fit)
s
```

```{r t-tests-scratch}
# Compute t-tests from scratch
X_mat <- cbind(1, as.matrix(mtcars_dt[, .(hp, wt, qsec)]))
n <- nrow(X_mat)
p <- ncol(X_mat)

XtX_inv <- solve(crossprod(X_mat))
sigma_hat <- s$sigma
beta_hat <- coef(fit)

se <- sigma_hat * sqrt(diag(XtX_inv))
t_stat <- beta_hat / se
p_val <- 2 * pt(abs(t_stat), df = n - p, lower.tail = FALSE)

data.frame(
  parameter = names(beta_hat),
  estimate = beta_hat,
  std_error = se,
  t_value = t_stat,
  p_value = p_val
)
```

**Interpretation**: the $t$-test for each coefficient tests whether that predictor has a non-zero **partial** effect (i.e., holding all other predictors constant). A non-significant $t$-test does not mean the variable is unimportant -- it may be collinear with other predictors.

---

## 16.12 The $F$-Test for Overall Model Significance

### 16.12.1 The Hypothesis

$$H_0: \beta_1 = \beta_2 = \cdots = \beta_{p-1} = 0 \quad \text{(all slopes are zero)}$$

$$H_1: \text{at least one } \beta_j \ne 0$$

### 16.12.2 The Test Statistic

$$F = \frac{(\text{SST} - \text{SSE})/(p - 1)}{\text{SSE}/(n - p)} = \frac{\text{SSReg}/(p-1)}{\text{MSE}} \sim F_{p-1,\, n-p} \quad \text{under } H_0$$

```{r overall-f-test}
SST <- sum((mtcars_dt$mpg - mean(mtcars_dt$mpg))^2)
SSE <- sum(residuals(fit)^2)
SSReg <- SST - SSE

F_stat <- (SSReg / (p - 1)) / (SSE / (n - p))
p_val_F <- pf(F_stat, df1 = p - 1, df2 = n - p, lower.tail = FALSE)

data.frame(
  quantity = c("F-statistic", "df1", "df2", "p-value"),
  value = c(F_stat, p - 1, n - p, p_val_F)
)
```

---

## 16.13 The $F$-Test for Nested Models

### 16.13.1 Comparing a Full and Reduced Model

Suppose we want to test whether a subset of predictors can be dropped. Let the **full model** have $p$ parameters and the **reduced model** have $q$ parameters (with $q < p$). The test statistic is:

$$F = \frac{(\text{SSE}_{\text{reduced}} - \text{SSE}_{\text{full}})/(p - q)}{\text{SSE}_{\text{full}}/(n - p)} \sim F_{p-q,\, n-p}$$

This tests $H_0$: the extra $p - q$ parameters in the full model are all zero.

### 16.13.2 Example: Does horsepower add anything?

```{r nested-f-test}
fit_full <- lm(mpg ~ hp + wt + qsec, data = mtcars_dt)
fit_reduced <- lm(mpg ~ wt + qsec, data = mtcars_dt)

anova(fit_reduced, fit_full)
```

```{r nested-f-scratch}
SSE_full <- sum(residuals(fit_full)^2)
SSE_reduced <- sum(residuals(fit_reduced)^2)
p_full <- length(coef(fit_full))
p_reduced <- length(coef(fit_reduced))

F_nested <- ((SSE_reduced - SSE_full) / (p_full - p_reduced)) /
            (SSE_full / (n - p_full))
p_val_nested <- pf(F_nested, df1 = p_full - p_reduced,
                    df2 = n - p_full, lower.tail = FALSE)

data.frame(
  quantity = c("Extra SS", "F-statistic", "p-value"),
  value = c(SSE_reduced - SSE_full, F_nested, p_val_nested)
)
```

### 16.13.3 Sequential (Type I) ANOVA

```{r sequential-anova}
anova(fit_full)
```

The sequential ANOVA table tests each predictor **given those already in the model**. The order matters! `hp` is tested after the intercept, `wt` after the intercept and `hp`, and `qsec` after the intercept, `hp`, and `wt`.

---

## 16.14 Confidence Intervals and Prediction

### 16.14.1 Confidence Intervals for Coefficients

A $(1-\alpha)$ confidence interval for $\beta_j$ is:

$$\hat{\beta}_j \pm t_{n-p, \alpha/2} \cdot \text{SE}(\hat{\beta}_j)$$

```{r ci-mlr}
confint(fit, level = 0.95)
```

```{r ci-scratch}
t_crit <- qt(0.975, df = n - p)
ci_lower <- beta_hat - t_crit * se
ci_upper <- beta_hat + t_crit * se

data.frame(
  parameter = names(beta_hat),
  lower = ci_lower,
  upper = ci_upper
)
```

### 16.14.2 Prediction at New Points

```{r predict-mlr}
new_data <- data.frame(hp = c(100, 150, 200), wt = c(2.5, 3.0, 3.5), qsec = c(18, 17, 16))

# Confidence interval for mean response
predict(fit, newdata = new_data, interval = "confidence", level = 0.95)

# Prediction interval for individual response
predict(fit, newdata = new_data, interval = "prediction", level = 0.95)
```

As in simple regression, the prediction interval is wider because it includes the irreducible error $\sigma^2$.

---

## 16.15 Simulation: Verifying the Gauss-Markov Theorem

Let us empirically verify that OLS has the smallest variance among linear unbiased estimators.

```{r gauss-markov-sim, fig.cap="Distribution of slope estimates for beta_1 from OLS (blue) and a competing linear unbiased estimator (orange). OLS has smaller variance, confirming the Gauss-Markov theorem."}
set.seed(42)
n_sims <- 5000
n <- 100
p_true <- 3  # intercept + 2 predictors

beta_true <- c(1, 2, -1)  # intercept, beta1, beta2
sigma_true <- 3

b1_ols <- numeric(n_sims)
b1_alt <- numeric(n_sims)

# Fixed design matrix
x1 <- runif(n, 0, 10)
x2 <- runif(n, 0, 10)
X_mat <- cbind(1, x1, x2)

# Alternative linear unbiased estimator:
# C = (X'X)^{-1}X' + D where DX = 0
# Choose D as a random matrix satisfying DX = 0
XtX_inv <- solve(crossprod(X_mat))
ols_mat <- XtX_inv %*% t(X_mat)

# Project a random matrix onto null space of X'
set.seed(123)
D_raw <- matrix(rnorm(p_true * n), nrow = p_true)
# D must satisfy DX = 0, so D = D_raw * (I - H)' where H = X(X'X)^{-1}X'
H <- X_mat %*% XtX_inv %*% t(X_mat)
D <- 0.01 * D_raw %*% (diag(n) - H)  # small perturbation
alt_mat <- ols_mat + D

for (k in seq_len(n_sims)) {
  eps <- rnorm(n, 0, sigma_true)
  y <- X_mat %*% beta_true + eps
  b1_ols[k] <- (ols_mat %*% y)[2]
  b1_alt[k] <- (alt_mat %*% y)[2]
}

sim_data <- data.table$data.table(
  estimate = c(b1_ols, b1_alt),
  method = rep(c("OLS", "Alternative LUE"), each = n_sims)
)

ggplot2$ggplot(sim_data, ggplot2$aes(x = estimate, fill = method)) +
  ggplot2$geom_density(alpha = 0.5) +
  ggplot2$geom_vline(xintercept = beta_true[2], linetype = "dashed") +
  ggplot2$scale_fill_manual(values = c("OLS" = "#0072B2", "Alternative LUE" = "#E69F00")) +
  ggplot2$labs(
    x = expression(hat(beta)[1]),
    y = "Density",
    title = "OLS vs alternative linear unbiased estimator"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

```{r gauss-markov-variance}
data.frame(
  method = c("OLS", "Alternative LUE"),
  mean = c(mean(b1_ols), mean(b1_alt)),
  variance = c(var(b1_ols), var(b1_alt))
)
```

Both estimators are unbiased (mean $\approx 2$), but OLS has smaller variance -- exactly as the Gauss-Markov theorem predicts.

---

## 16.16 Putting It Together: mtcars Full Analysis

```{r mtcars-full, fig.cap="Diagnostic plots for the mtcars multiple regression model."}
par(mfrow = c(2, 2))
plot(fit, which = 1:4, col = "#0072B2", pch = 16, cex = 0.8)
par(mfrow = c(1, 1))
```

```{r mtcars-observed-vs-fitted, fig.cap="Observed vs fitted mpg for the mtcars model. Points cluster near the diagonal, indicating a good fit."}
mtcars_dt[, fitted_mpg := fitted(fit)]

ggplot2$ggplot(mtcars_dt, ggplot2$aes(x = fitted_mpg, y = mpg)) +
  ggplot2$geom_point(colour = "#0072B2", size = 2.5) +
  ggplot2$geom_abline(intercept = 0, slope = 1, colour = "#D55E00", linewidth = 0.8) +
  ggplot2$labs(
    x = "Fitted mpg",
    y = "Observed mpg",
    title = sprintf("Observed vs fitted (R^2 = %.3f)", summary(fit)$r.squared)
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

---

## Communicating to Stakeholders

When presenting multiple regression results:

- **Lead with the research question**, not the mathematics. "We investigated which car characteristics most strongly predict fuel efficiency."
- **Present the coefficients table** with confidence intervals, not just point estimates. "Holding horsepower and acceleration constant, each additional 1000 lbs of weight reduces fuel efficiency by about 3.9 mpg (95% CI: 2.2 to 5.6)."
- **Explain partial effects carefully.** The coefficient on `wt` assumes the other predictors are held constant. In reality, heavier cars also tend to have more horsepower -- the partial effect isolates the weight effect.
- **The overall $F$-test** answers: "do these predictors collectively tell us anything about fuel efficiency?" The individual $t$-tests answer: "does this predictor add information beyond what the others already provide?"

---

## Quick Reference

| Quantity | Formula |
|:---------|:--------|
| **$\text{Cov}(\hat{\boldsymbol{\beta}})$** | $\sigma^2(\mathbf{X}'\mathbf{X})^{-1}$ |
| **$\text{SE}(\hat{\beta}_j)$** | $s\sqrt{[(\mathbf{X}'\mathbf{X})^{-1}]_{jj}}$ |
| **$t$-test** | $t_j = \hat{\beta}_j / \text{SE}(\hat{\beta}_j) \sim t_{n-p}$ |
| **Overall $F$** | $F = \text{MSReg}/\text{MSE} \sim F_{p-1, n-p}$ |
| **Nested $F$** | $F = \frac{(\text{SSE}_R - \text{SSE}_F)/(p-q)}{\text{SSE}_F/(n-p)}$ |
| **Gauss-Markov** | OLS is BLUE under $E[\boldsymbol{\varepsilon}]=\mathbf{0}$, $\text{Cov}(\boldsymbol{\varepsilon})=\sigma^2\mathbf{I}$ |

---

## Exercises

1. **Gauss-Markov proof details.** Fill in the algebra showing that the cross-terms in $\text{Cov}(\tilde{\boldsymbol{\beta}})$ vanish when $\mathbf{D}\mathbf{X} = \mathbf{0}$. Specifically, show that $(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{D}' = \mathbf{0}$.

2. **When BLUE is not best.** Simulate data with $n = 20$, $p = 15$, and compare OLS to ridge regression ($\hat{\boldsymbol{\beta}}_{\text{ridge}} = (\mathbf{X}'\mathbf{X} + \lambda\mathbf{I})^{-1}\mathbf{X}'\mathbf{Y}$). Show that ridge can have lower MSE than OLS despite being biased.

3. **Partial $F$-test.** Using the mtcars data, test whether the pair (hp, qsec) can be dropped from the model `mpg ~ hp + wt + qsec` simultaneously. Carry out the test from scratch and verify with `anova()`.

4. **Confidence ellipsoid.** Show that a joint $(1-\alpha)$ confidence region for $(\beta_1, \beta_2)$ in a two-predictor model is an ellipse. Hint: the pivotal quantity $(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta})'(\mathbf{X}'\mathbf{X})(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta})/(ps^2) \sim F_{p, n-p}$.

5. **Type I vs Type III SS.** Show with a concrete example (use the mtcars data) that the sequential (Type I) sums of squares depend on the order of predictors, but the partial (Type III) sums of squares do not.

6. **Simulation study.** Generate 1000 datasets from $Y = 1 + 2X_1 - X_2 + \varepsilon$ with $n = 50$ and $\varepsilon \sim N(0, 9)$. For each, compute all four $t$-statistics (including the intercept). Verify that the distribution of each $t$-statistic is approximately $t_{47}$.
