---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 24: Experimental Design"
part: "Part 1: Principles and Practice"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, experimental-design, randomisation, power-analysis, sample-size, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

```{r packages, include=FALSE}
box::use(ggplot2)
box::use(data.table)
```

```{r dt-symbols, include=FALSE}
.N <- data.table::.N
.SD <- data.table::.SD
```

# Chapter 24 -- Experimental Design

## Part 1: Principles and Practice

All the statistical methods in this book -- from $t$-tests to Bayesian regression -- depend on how the data were collected. A flawed study design cannot be rescued by clever analysis. In this chapter we examine the principles that make an experiment valid: **randomisation**, **blocking**, **replication**, and **blinding**. We revisit sample size and power calculations with a design perspective, and we study the most common pitfalls that invalidate conclusions.

---

## 24.1 Observational Studies vs Experiments

### 24.1.1 The Key Distinction

An **observational study** records data without intervening. A **randomised experiment** assigns treatments to units under the investigator's control.

| Feature | Observational | Experimental |
|:--------|:-------------|:-------------|
| **Treatment assignment** | Nature / the subject | Investigator (random) |
| **Confounding** | Always a threat | Controlled by randomisation |
| **Causal inference** | Association only (in general) | Causation (if well-designed) |
| **Ethics** | Sometimes the only option | Requires equipoise |
| **Examples** | NHANES, cohort studies | Lind's scurvy trial, RCTs |

### 24.1.2 Simpson's Paradox and Confounding

Confounding occurs when a third variable is associated with both the treatment and the outcome. The classic example: a treatment appears harmful overall but is beneficial within every subgroup, because sicker patients receive the treatment more often.

```{r simpsons-paradox, fig.cap="Simpson's paradox: the overall trend (dashed) is positive, but within each group the trend is negative. Ignoring the confounding group variable reverses the apparent effect."}
set.seed(42)
n_per <- 100

group_a <- data.table$data.table(
  x = rnorm(n_per, 2, 1),
  group = "A"
)
group_a[, y := 10 - 0.8 * x + rnorm(n_per, 0, 0.5)]

group_b <- data.table$data.table(
  x = rnorm(n_per, 5, 1),
  group = "B"
)
group_b[, y := 15 - 0.8 * x + rnorm(n_per, 0, 0.5)]

group_c <- data.table$data.table(
  x = rnorm(n_per, 8, 1),
  group = "C"
)
group_c[, y := 20 - 0.8 * x + rnorm(n_per, 0, 0.5)]

simpson_dt <- data.table$rbindlist(list(group_a, group_b, group_c))

ggplot2$ggplot(simpson_dt, ggplot2$aes(x = x, y = y)) +
  ggplot2$geom_point(ggplot2$aes(colour = group), alpha = 0.5) +
  ggplot2$geom_smooth(method = "lm", se = FALSE, linetype = "dashed", colour = "black") +
  ggplot2$geom_smooth(ggplot2$aes(colour = group), method = "lm", se = FALSE) +
  ggplot2$labs(
    x = "Treatment dose", y = "Outcome",
    title = "Simpson's paradox: overall positive trend masks within-group negative trends"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "top")
```

---

## 24.2 Randomisation: Why It Works

### 24.2.1 The Mathematical Foundation

Randomisation balances **all** covariates -- measured and unmeasured -- in expectation. If we randomly assign $n$ subjects to treatment or control, then for any covariate $X$:

$$E[\bar{X}_{\text{treat}}] = E[\bar{X}_{\text{control}}]$$

More formally, randomisation makes the treatment assignment **independent** of potential outcomes (the Rubin causal model):

$$(Y(0), Y(1)) \perp\!\!\!\perp Z$$

where $Y(0)$ and $Y(1)$ are potential outcomes and $Z$ is the treatment indicator. This is the foundation of causal inference.

### 24.2.2 Simple Randomisation

Each subject is independently assigned to treatment with probability $1/2$.

```{r simple-randomisation}
set.seed(42)
n_subjects <- 100
assignments <- sample(c("Treatment", "Control"), n_subjects, replace = TRUE)
table(assignments)
```

Simple randomisation can produce unbalanced groups by chance. With 100 subjects, the probability of a split worse than 60/40 is:

```{r imbalance-prob}
p_imbalance <- 2 * pbinom(39, 100, 0.5)
sprintf("P(|n_treat - 50| > 10) = %.4f", p_imbalance)
```

### 24.2.3 Blocked Randomisation

To ensure equal group sizes, use **permuted blocks**. Within each block of size $2k$, exactly $k$ are assigned to treatment.

```{r blocked-randomisation}
set.seed(42)
block_size <- 4
n_blocks <- 25
n_total <- block_size * n_blocks

blocked <- unlist(lapply(1:n_blocks, function(b) {
  block <- rep(c("Treatment", "Control"), each = block_size / 2)
  sample(block)
}))

table(blocked)
sprintf("Perfect balance: %d treatment, %d control", sum(blocked == "Treatment"), sum(blocked == "Control"))
```

---

## 24.3 Stratified Randomisation

When a known covariate (e.g., disease severity, sex) strongly affects the outcome, we **stratify**: perform separate blocked randomisation within each stratum. This guarantees balance on that covariate.

```{r stratified}
set.seed(42)
strata <- c(rep("Mild", 40), rep("Moderate", 40), rep("Severe", 20))

stratified_assign <- function(stratum_labels) {
  dt <- data.table$data.table(id = seq_along(stratum_labels), stratum = stratum_labels)
  dt[, treatment := {
    n <- .N
    block <- rep(c("Treatment", "Control"), length.out = n)
    sample(block)
  }, by = stratum]
  dt
}

strat_dt <- stratified_assign(strata)
strat_dt[, .N, by = .(stratum, treatment)]
```

---

## 24.4 Blinding

**Single-blind:** the subject does not know which treatment they receive.
**Double-blind:** neither the subject nor the assessor knows.

Blinding prevents:

- **Placebo effect:** subjects improve because they expect to.
- **Observer bias:** assessors unconsciously rate treated subjects more favourably.
- **Differential attrition:** subjects in the control group drop out more often.

---

## 24.5 Sample Size and Power Revisited

### 24.5.1 The Design Perspective

In Chapter 10 we introduced power ($1 - \beta$) and Type II error. At the design stage, we solve for the **sample size** needed to detect a meaningful effect with adequate power.

For a two-sample $t$-test with equal groups:

$$n = \frac{2(z_{\alpha/2} + z_\beta)^2 \sigma^2}{\delta^2}$$

where $\delta = \mu_1 - \mu_2$ is the minimum clinically important difference and $\sigma^2$ is the common variance.

### 24.5.2 Power Calculations with `pwr`

```{r power-ttest}
box::use(pwr)

power_result <- pwr$pwr.t.test(
  d = 0.5,
  sig.level = 0.05,
  power = 0.80,
  type = "two.sample"
)

sprintf("Required n per group: %.0f (Cohen's d = 0.5, power = 0.80)", ceiling(power_result$n))
```

```{r power-curve, fig.cap="Power curves for the two-sample t-test at different effect sizes. Larger effects require fewer subjects; small effects demand large samples."}
effect_sizes <- c(0.2, 0.5, 0.8)
ns <- seq(5, 200, by = 5)

power_list <- list()
for (d in effect_sizes) {
  powers <- sapply(ns, function(n) {
    pwr$pwr.t.test(d = d, n = n, sig.level = 0.05, type = "two.sample")$power
  })
  power_list[[length(power_list) + 1]] <- data.table$data.table(
    n = ns, power = powers, effect = paste0("d = ", d)
  )
}
power_dt <- data.table$rbindlist(power_list)

ggplot2$ggplot(power_dt, ggplot2$aes(x = n, y = power, colour = effect)) +
  ggplot2$geom_line(linewidth = 1.1) +
  ggplot2$geom_hline(yintercept = 0.80, linetype = "dashed", alpha = 0.5) +
  ggplot2$labs(
    x = "Sample size per group", y = "Power",
    title = "Power curves for the two-sample t-test (alpha = 0.05)",
    colour = "Effect size"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "top")
```

### 24.5.3 Power for Proportions

```{r power-proportions}
prop_power <- pwr$pwr.2p.test(
  h = pwr$ES.h(p1 = 0.3, p2 = 0.2),
  sig.level = 0.05,
  power = 0.80
)

sprintf("Required n per group: %.0f (to detect 30%% vs 20%% success rate)", ceiling(prop_power$n))
```

### 24.5.4 Power for Chi-Squared Tests

```{r power-chi}
chi_power <- pwr$pwr.chisq.test(
  w = 0.3,
  df = 2,
  sig.level = 0.05,
  power = 0.80
)

sprintf("Required n: %.0f (chi-squared, w = 0.3, df = 2)", ceiling(chi_power$N))
```

---

## 24.6 Common Pitfalls

### 24.6.1 Selection Bias

Subjects are not representative of the target population. Example: a trial of a new antidepressant recruits only from a single specialist clinic; results may not generalise to primary care.

### 24.6.2 Survivorship Bias

Analysing only the survivors creates a distorted picture. Example: studying successful companies to find common traits ignores the failed companies with the same traits.

```{r survivorship, fig.cap="Survivorship bias: we only observe the survivors (blue), missing the full distribution. The observed mean is biased upwards."}
set.seed(42)
full_pop <- rnorm(1000, mean = 50, sd = 20)
survivors <- full_pop[full_pop > 30]

surv_dt <- data.table$data.table(
  value = c(full_pop, survivors),
  group = c(rep("Full population", length(full_pop)), rep("Survivors only", length(survivors)))
)

ggplot2$ggplot(surv_dt, ggplot2$aes(x = value, fill = group)) +
  ggplot2$geom_histogram(
    ggplot2$aes(y = ggplot2::after_stat(density)),
    bins = 40, alpha = 0.6, colour = "white", position = "identity"
  ) +
  ggplot2$geom_vline(xintercept = mean(full_pop), linetype = "dashed", colour = "#D55E00") +
  ggplot2$geom_vline(xintercept = mean(survivors), linetype = "dashed", colour = "#0072B2") +
  ggplot2$scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  ggplot2$labs(
    x = "Value", y = "Density",
    title = sprintf("Survivorship bias: full mean = %.1f, survivor mean = %.1f",
                    mean(full_pop), mean(survivors)),
    fill = NULL
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "top")
```

### 24.6.3 Confounding

An unmeasured variable drives both the apparent cause and the effect. Randomisation eliminates confounding; observational studies must rely on statistical adjustment (imperfect).

### 24.6.4 Multiple Comparisons

Testing many hypotheses inflates the family-wise error rate (Chapter 12). A study that tests 20 outcomes at $\alpha = 0.05$ expects one false positive even if all null hypotheses are true.

### 24.6.5 P-Hacking and the Garden of Forking Paths

Researchers may (consciously or not) try multiple analyses until one yields $p < 0.05$. Pre-registration -- specifying the analysis plan before seeing the data -- is the antidote.

---

## 24.7 Case Study: The Streptomycin Tuberculosis Trial

The 1948 British Medical Research Council trial of streptomycin for pulmonary tuberculosis is widely considered the first modern randomised controlled trial.

```{r strep-tb}
data_dir <- "../../../data"
strep <- data.table$fread(file.path(data_dir, "medical/strep_tb.csv"))
str(strep)
```

```{r strep-analysis}
outcome_tab <- strep[, .N, by = .(arm, improved)]
knitr::kable(outcome_tab, caption = "Outcomes in the streptomycin TB trial by treatment arm.")
```

```{r strep-test}
cont_tab <- table(strep$arm, strep$improved)
fisher_result <- fisher.test(cont_tab)
sprintf("Fisher's exact test p-value: %.4f", fisher_result$p.value)
sprintf("Odds ratio: %.3f (95%% CI: %.3f to %.3f)",
        fisher_result$estimate, fisher_result$conf.int[1], fisher_result$conf.int[2])
```

**Design features of this trial:**

- **Randomisation:** sealed envelopes opened only at admission.
- **Concealment:** clinicians did not know the allocation before the patient was enrolled.
- **Blinding:** radiographers assessing X-rays did not know treatment assignment.
- **Pre-specified outcome:** improvement at 6 months assessed by independent radiologists.

---

## 24.8 Case Study: Lind's Scurvy Trial Design

James Lind's 1747 trial is often cited as the first clinical trial, though its design has important limitations.

```{r scurvy-design}
scurvy <- data.table$fread(file.path(data_dir, "medical/scurvy.csv"))

scurvy_summary <- scurvy[, .(n = .N), by = treatment]
knitr::kable(scurvy_summary, caption = "Lind's scurvy trial: treatment allocation.")
```

**Design critique:**

- **No randomisation.** Lind selected twelve similar patients but assigned treatments by convenience.
- **Tiny sample.** Two per group -- far too small for any statistical test to be meaningful.
- **No blinding.** Patients and Lind both knew the treatments.
- **No placebo control.** "Elixir vitriol" and "cider" groups are not true placebos.
- **Strength.** All twelve had similar severity, and treatments were administered simultaneously under identical conditions.

Despite these limitations, the dramatic effect of citrus fruit was visible even with $n = 2$ per group. This illustrates that large effects can be detected even in poor designs -- but subtle effects require proper methodology.

---

## 24.9 Case Study: Licorice Gargle RCT

A modern randomised controlled trial evaluating whether gargling with licorice solution reduces sore throat after intubation.

```{r licorice}
licorice <- data.table$fread(file.path(data_dir, "medical/licorice_gargle.csv"))
str(licorice)
```

```{r licorice-analysis, fig.cap="Sore throat severity scores by treatment group in the licorice gargle RCT. The licorice group shows lower post-operative throat pain."}
lic_dt <- data.table$copy(licorice)
lic_dt[, treat_label := ifelse(treat == 1, "Licorice", "Control")]

ggplot2$ggplot(lic_dt[!is.na(throat_pain_score)],
               ggplot2$aes(x = treat_label, y = throat_pain_score, fill = treat_label)) +
  ggplot2$geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  ggplot2$scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  ggplot2$labs(
    x = "Treatment", y = "Sore throat pain score",
    title = "Licorice gargle RCT: post-operative throat pain"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "none")
```

```{r licorice-test}
test_lic <- wilcox.test(
  throat_pain_score ~ treat,
  data = lic_dt[!is.na(throat_pain_score)]
)

sprintf("Wilcoxon rank-sum test p-value: %.4f", test_lic$p.value)
```

```{r licorice-power}
lic_d <- lic_dt[!is.na(throat_pain_score), {
  m <- tapply(throat_pain_score, treat_label, mean)
  s <- tapply(throat_pain_score, treat_label, sd)
  pooled_sd <- sqrt(mean(s^2))
  list(cohens_d = abs(diff(m)) / pooled_sd, pooled_sd = pooled_sd)
}]

post_hoc_power <- pwr$pwr.t.test(
  d = lic_d$cohens_d,
  n = nrow(lic_dt) / 2,
  sig.level = 0.05,
  type = "two.sample"
)

sprintf("Observed Cohen's d: %.3f", lic_d$cohens_d)
sprintf("Post-hoc power: %.3f", post_hoc_power$power)
```

---

## 24.10 The Design Checklist

Before collecting data, answer:

1. **Research question.** What precisely are we testing?
2. **Primary outcome.** One pre-specified primary endpoint.
3. **Sample size.** Power calculation based on the minimum clinically important effect.
4. **Randomisation scheme.** Simple, blocked, or stratified.
5. **Blinding.** Single, double, or impossible? Justify.
6. **Inclusion/exclusion criteria.** Who is eligible? This determines generalisability.
7. **Analysis plan.** Pre-specify the statistical test and any adjustments.
8. **Registration.** Register the protocol before data collection begins.

---

## Communicating to Stakeholders

- **Randomisation is not arbitrary.** "We randomly assigned patients to groups so that the groups are comparable on all factors -- including ones we did not measure."
- **Power calculations prevent waste.** "We calculated that 64 patients per group are needed to detect a meaningful improvement. Fewer patients would risk missing a real effect; more would be unnecessary."
- **Blinding prevents bias.** "Neither the patients nor the doctors evaluating outcomes knew which treatment each patient received, preventing unconscious bias."
- **Observational studies are valuable but limited.** "This study shows an association, but we cannot be certain it is causal because we could not control for all confounders."

---

## Quick Reference

| Concept | Details |
|:--------|:--------|
| **Randomisation** | Ensures $(Y(0), Y(1)) \perp Z$; eliminates confounding |
| **Blocking** | Balance within known strata (e.g., sex, site) |
| **Stratification** | Separate randomisation within strata |
| **Blinding** | Single (subject), double (+ assessor), triple (+ analyst) |
| **Power** | $1 - \beta$; want $\geq 0.80$ conventionally |
| **Sample size (t-test)** | $n = 2(z_{\alpha/2} + z_\beta)^2 \sigma^2 / \delta^2$ |
| **`pwr` package** | `pwr.t.test()`, `pwr.2p.test()`, `pwr.chisq.test()` |
| **Selection bias** | Non-representative sample |
| **Survivorship bias** | Analysing only those who "survived" some filter |
| **Confounding** | Third variable associated with both cause and effect |
| **P-hacking** | Multiple analyses until $p < 0.05$; pre-registration prevents |
| **Pre-registration** | Protocol specified before data collection |

---

## Exercises

1. **Randomisation balance.** Simulate 10,000 simple randomisations of 50 subjects. For each, compute the difference in group sizes. What fraction have an imbalance worse than 60/40? Now repeat with blocked randomisation (block size 4). Compare.

2. **Power calculation.** A clinical trial aims to detect a 5-point improvement in pain score (SD = 12) with 80% power at $\alpha = 0.05$. How many patients per group are needed? What happens if the true SD is 15?

3. **Confounding simulation.** Generate data where $X$ (treatment) and $Z$ (confounder) are correlated ($r = 0.6$), and the outcome $Y = 0.5 Z + \epsilon$ depends only on $Z$, not on $X$. Fit a simple regression of $Y$ on $X$ and observe the spurious effect. Then fit $Y$ on $X + Z$ and verify the effect of $X$ vanishes.

4. **Survivorship bias.** Generate 1000 observations from $N(0, 1)$ and keep only those with $|x| < 2$ (96% survival). Compute the variance of the survivors and compare with 1. Explain the bias.

5. **Design critique.** Read the description of Lind's scurvy trial. List three specific design improvements that would strengthen the trial if it were conducted today.

6. **Power for logistic regression.** Using simulation, estimate the sample size needed to detect an odds ratio of 2.0 for a binary predictor (50% prevalence) in a logistic regression with 80% power. (Hint: simulate data, fit the model, and count rejections.)
