---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 2: Probability — The Language"
part: "Part 1: Sets, Axioms, and Counting"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, probability, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

## 2.1 What Is Probability?

Before we can do statistics — before we can estimate parameters, test hypotheses, or fit models — we need a language for talking about uncertainty. That language is **probability theory**.

In plain English: probability is a number between 0 and 1 that tells us how likely something is to happen. A probability of 0 means the event is impossible; a probability of 1 means it is certain; a probability of 0.5 means it is a coin flip. Everything else falls somewhere in between.

But what *is* a probability, really? There are two main schools of thought, and both are useful.

### The Frequentist Interpretation

The frequentist says: the probability of an event is the **long-run relative frequency** with which it occurs. If we flip a fair coin infinitely many times, the fraction of heads converges to 0.5. In notation:

$$
P(A) = \lim_{n \to \infty} \frac{n_A}{n}
$$

where $n_A$ is the number of times event $A$ occurred in $n$ trials.

Let us watch this convergence happen in R:

```{r freq-convergence, fig.cap="Convergence of the relative frequency of heads to 0.5 as the number of coin flips increases. The dashed line marks the true probability."}
set.seed(42)

n_flips <- 10000L
flips <- sample(c(0L, 1L), size = n_flips, replace = TRUE)
cumulative_prop <- cumsum(flips) / seq_len(n_flips)

box::use(ggplot2)

df <- data.frame(
  flip = seq_len(n_flips),
  proportion = cumulative_prop
)

ggplot2$ggplot(df, ggplot2$aes(x = flip, y = proportion)) +
  ggplot2$geom_line(colour = "#0072B2", linewidth = 0.4) +
  ggplot2$geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  ggplot2$scale_x_log10(labels = scales::comma) +
  ggplot2$labs(
    x = "Number of flips (log scale)",
    y = "Proportion of heads",
    title = "Frequentist convergence: coin flips"
  ) +
  ggplot2$theme_minimal(base_size = 14)
```

Notice how the line is volatile for small $n$ but settles down as $n$ grows. This is the **law of large numbers** at work — we will prove it rigorously in Chapter 7.

### The Subjective (Bayesian) Interpretation

The Bayesian says: probability is a **degree of belief**. When a meteorologist says "there is a 30% chance of rain tomorrow," she is not imagining infinitely many tomorrows. She is expressing her uncertainty given the evidence she has.

Under this interpretation, two rational people with different background information can assign different probabilities to the same event — and both can be "correct" relative to their information.

We will not take sides in this textbook. Both interpretations lead to the same mathematical axioms, which is what matters. We will use frequentist reasoning when it is natural (e.g., repeated experiments) and Bayesian reasoning when *it* is natural (e.g., updating beliefs). The axioms are the same either way.

---

## 2.2 Sample Spaces and Events

### Crash Course: Set Theory Notation

Probability is built on set theory, so we need to be comfortable with the notation. If you have seen this before, treat this as a quick refresher.

**Sets and elements.** A *set* is a collection of distinct objects. We write sets with curly braces. The set of outcomes when rolling a die is $\{1, 2, 3, 4, 5, 6\}$. If $x$ is an element of set $A$, we write $x \in A$. If $x$ is not in $A$, we write $x \notin A$.

**Subsets.** If every element of $A$ is also in $B$, we write $A \subseteq B$ ("$A$ is a subset of $B$"). If $A \subseteq B$ and $A \neq B$, we write $A \subset B$ (proper subset).

**Union.** The union $A \cup B$ is the set of elements in $A$ **or** $B$ (or both). In plain English: union means **OR**.

$$
A \cup B = \{x : x \in A \text{ or } x \in B\}
$$

**Intersection.** The intersection $A \cap B$ is the set of elements in $A$ **and** $B$. In plain English: intersection means **AND**.

$$
A \cap B = \{x : x \in A \text{ and } x \in B\}
$$

**Complement.** The complement $A^c$ (also written $\bar{A}$ or $A'$) is the set of elements **not** in $A$. In plain English: complement means **NOT**.

$$
A^c = \{x : x \notin A\}
$$

**Empty set.** The empty set $\emptyset$ contains no elements. It is a subset of every set.

**Universal set.** The universal set $\Omega$ contains all elements under consideration. We always have $A \subseteq \Omega$ and $A^c = \Omega \setminus A$.

**Disjoint (mutually exclusive) sets.** Two sets $A$ and $B$ are disjoint if $A \cap B = \emptyset$ — they share no elements.

### De Morgan's Laws

These are two identities that let us move between unions and intersections via complements. They are fundamental and appear everywhere in probability.

**De Morgan's First Law:**

$$
(A \cup B)^c = A^c \cap B^c
$$

In plain English: "not (A or B)" is the same as "(not A) and (not B)." If you are *not* in the union, then you must be outside *both* sets.

**Proof.** We show set equality by showing each side is a subset of the other.

($\subseteq$): Let $x \in (A \cup B)^c$. Then $x \notin A \cup B$, which means $x \notin A$ and $x \notin B$. Therefore $x \in A^c$ and $x \in B^c$, so $x \in A^c \cap B^c$.

($\supseteq$): Let $x \in A^c \cap B^c$. Then $x \in A^c$ and $x \in B^c$, which means $x \notin A$ and $x \notin B$. Therefore $x \notin A \cup B$, so $x \in (A \cup B)^c$.

Since each side is a subset of the other, $(A \cup B)^c = A^c \cap B^c$. $\blacksquare$

**De Morgan's Second Law:**

$$
(A \cap B)^c = A^c \cup B^c
$$

In plain English: "not (A and B)" is the same as "(not A) or (not B)." If you are *not* in both sets simultaneously, then you must be outside at least one of them.

**Proof.** Again, we show set equality by double inclusion.

($\subseteq$): Let $x \in (A \cap B)^c$. Then $x \notin A \cap B$, which means it is *not* the case that ($x \in A$ and $x \in B$). By logical negation, $x \notin A$ or $x \notin B$. Therefore $x \in A^c$ or $x \in B^c$, so $x \in A^c \cup B^c$.

($\supseteq$): Let $x \in A^c \cup B^c$. Then $x \in A^c$ or $x \in B^c$, meaning $x \notin A$ or $x \notin B$. In either case, it is not true that ($x \in A$ and $x \in B$), so $x \notin A \cap B$, and thus $x \in (A \cap B)^c$.

Therefore $(A \cap B)^c = A^c \cup B^c$. $\blacksquare$

### Sample Spaces and Events

Now we can define the key objects of probability.

**Definition (Sample space).** The *sample space* $\Omega$ is the set of all possible outcomes of a random experiment.

**Definition (Event).** An *event* is a subset of $\Omega$. An event $A$ *occurs* if the outcome of the experiment is an element of $A$.

Let us enumerate some sample spaces in R.

```{r sample-spaces}
# Single fair die
die <- 1:6
die

# Two dice: sample space has 36 outcomes
two_dice <- expand.grid(die1 = 1:6, die2 = 1:6)
nrow(two_dice)

# Event: sum equals 7
event_sum7 <- two_dice[two_dice$die1 + two_dice$die2 == 7, ]
event_sum7
nrow(event_sum7) / nrow(two_dice)
```

```{r card-deck}
# Standard deck of 52 cards
suits <- c("Hearts", "Diamonds", "Clubs", "Spades")
ranks <- c("A", 2:10, "J", "Q", "K")
deck <- expand.grid(rank = ranks, suit = suits, stringsAsFactors = FALSE)
nrow(deck)

# Event: drawing a heart
hearts <- deck[deck$suit == "Hearts", ]
nrow(hearts) / nrow(deck)

# Event: drawing a face card (J, Q, K)
face_cards <- deck[deck$rank %in% c("J", "Q", "K"), ]
nrow(face_cards) / nrow(deck)
```

---

## 2.3 Axioms of Probability

We now state the formal foundation. All of probability theory is built on just three axioms, laid down by Andrey Kolmogorov in 1933.

**Kolmogorov's Axioms.** Let $\Omega$ be a sample space and let $P$ be a function that assigns a real number $P(A)$ to each event $A \subseteq \Omega$. Then $P$ is a *probability measure* if it satisfies:

1. **Non-negativity.** For every event $A$, $P(A) \geq 0$.
2. **Normalisation.** $P(\Omega) = 1$.
3. **Countable additivity.** If $A_1, A_2, \ldots$ are mutually exclusive events (i.e., $A_i \cap A_j = \emptyset$ for all $i \neq j$), then

$$
P\left(\bigcup_{i=1}^{\infty} A_i\right) = \sum_{i=1}^{\infty} P(A_i)
$$

In plain English:
1. Probabilities cannot be negative.
2. Something must happen — the probability of the entire sample space is 1.
3. If events cannot occur simultaneously, the probability that at least one of them occurs is the sum of their individual probabilities.

These three axioms are deceptively simple, but a vast amount follows from them. Let us derive the most important consequences.

### Consequence 1: Complement Rule

**Theorem.** For any event $A$, $P(A^c) = 1 - P(A)$.

**Proof.** Since $A$ and $A^c$ are mutually exclusive (they share no outcomes) and $A \cup A^c = \Omega$, Axiom 3 gives us:

$$
P(\Omega) = P(A \cup A^c) = P(A) + P(A^c)
$$

By Axiom 2, $P(\Omega) = 1$. Rearranging:

$$
P(A^c) = 1 - P(A) \qquad \blacksquare
$$

In plain English: the probability that something *does not* happen equals one minus the probability that it *does* happen.

### Consequence 2: Probability of the Empty Set

**Theorem.** $P(\emptyset) = 0$.

**Proof.** The empty set is the complement of the sample space: $\emptyset = \Omega^c$. By the complement rule:

$$
P(\emptyset) = P(\Omega^c) = 1 - P(\Omega) = 1 - 1 = 0 \qquad \blacksquare
$$

In plain English: the probability that *nothing* happens is zero — something always happens.

### Consequence 3: Inclusion-Exclusion (Two Events)

**Theorem.** For any two events $A$ and $B$:

$$
P(A \cup B) = P(A) + P(B) - P(A \cap B)
$$

**Proof.** We decompose $A \cup B$ into three mutually exclusive pieces:

$$
A \cup B = (A \cap B^c) \cup (A \cap B) \cup (A^c \cap B)
$$

These three sets are pairwise disjoint, so by Axiom 3:

$$
P(A \cup B) = P(A \cap B^c) + P(A \cap B) + P(A^c \cap B)
$$

Now observe that $A = (A \cap B^c) \cup (A \cap B)$, where the two pieces are disjoint. So:

$$
P(A) = P(A \cap B^c) + P(A \cap B) \implies P(A \cap B^c) = P(A) - P(A \cap B)
$$

Similarly, $B = (A^c \cap B) \cup (A \cap B)$, so:

$$
P(B) = P(A^c \cap B) + P(A \cap B) \implies P(A^c \cap B) = P(B) - P(A \cap B)
$$

Substituting back:

$$
P(A \cup B) = [P(A) - P(A \cap B)] + P(A \cap B) + [P(B) - P(A \cap B)]
$$

$$
= P(A) + P(B) - P(A \cap B) \qquad \blacksquare
$$

In plain English: when computing "A or B," adding $P(A) + P(B)$ double-counts the overlap, so we subtract $P(A \cap B)$ once.

### General Inclusion-Exclusion

The principle generalises. For events $A_1, A_2, \ldots, A_n$:

$$
P\left(\bigcup_{i=1}^{n} A_i\right) = \sum_{i} P(A_i) - \sum_{i<j} P(A_i \cap A_j) + \sum_{i<j<k} P(A_i \cap A_j \cap A_k) - \cdots + (-1)^{n+1} P(A_1 \cap A_2 \cap \cdots \cap A_n)
$$

We state this without proof; the argument is an induction extending the two-event case above.

### Verifying the Axioms with Simulation

Let us verify these consequences with brute-force computation in R.

```{r axiom-verification}
set.seed(42)

# Simulate 100,000 rolls of a fair die
n_sim <- 100000L
rolls <- sample(1:6, size = n_sim, replace = TRUE)

# Event A: roll is even; Event B: roll >= 4
A <- rolls %% 2 == 0          # {2, 4, 6}
B <- rolls >= 4                # {4, 5, 6}

# Complement rule: P(A^c) should equal 1 - P(A)
p_A <- mean(A)
p_Ac <- mean(!A)
round(c(P_A = p_A, P_Ac = p_Ac, sum = p_A + p_Ac), 4)

# Inclusion-exclusion: P(A ∪ B) = P(A) + P(B) - P(A ∩ B)
p_B <- mean(B)
p_AandB <- mean(A & B)
p_AorB <- mean(A | B)
p_AorB_formula <- p_A + p_B - p_AandB

round(c(simulated = p_AorB, formula = p_AorB_formula, theoretical = 5/6), 4)
```

The simulated values closely match the theoretical values, as expected.

---

## 2.4 Counting

When outcomes are equally likely, probability reduces to counting:

$$
P(A) = \frac{|A|}{|\Omega|} = \frac{\text{number of favourable outcomes}}{\text{total number of outcomes}}
$$

This means we need systematic ways to count. Welcome to **combinatorics**.

### Crash Course: Combinatorics

#### The Multiplication Principle

If there are $n_1$ ways to do task 1, $n_2$ ways to do task 2, ..., and $n_k$ ways to do task $k$, and the tasks are performed in sequence, then there are $n_1 \times n_2 \times \cdots \times n_k$ total ways to complete all tasks.

In plain English: if you have 3 shirts and 4 pairs of trousers, you have $3 \times 4 = 12$ outfits.

```{r multiplication-principle}
# Number of licence plates: 3 letters then 4 digits
n_letters <- 26
n_digits <- 10
n_plates <- n_letters^3 * n_digits^4
format(n_plates, big.mark = ",")
```

#### Permutations

A *permutation* is an **ordered** arrangement. The number of ways to arrange $k$ items chosen from $n$ distinct items is:

$$
P(n, k) = \frac{n!}{(n-k)!}
$$

**Derivation.** For the first position, we have $n$ choices. For the second, $n-1$ choices (one item is used). For the third, $n-2$. Continuing, for the $k$-th position, we have $n - k + 1$ choices. By the multiplication principle:

$$
P(n, k) = n \times (n-1) \times (n-2) \times \cdots \times (n-k+1) = \frac{n!}{(n-k)!}
$$

The special case $P(n, n) = n!$ is the number of ways to arrange all $n$ items.

#### Combinations

A *combination* is an **unordered** selection. The number of ways to choose $k$ items from $n$ distinct items (where order does not matter) is:

$$
\binom{n}{k} = \frac{n!}{k!(n-k)!}
$$

**Derivation from permutations.** Each combination of $k$ items can be arranged in $k!$ ways. So permutations over-count combinations by a factor of $k!$:

$$
P(n, k) = \binom{n}{k} \times k! \implies \binom{n}{k} = \frac{P(n, k)}{k!} = \frac{n!}{k!(n-k)!}
$$

We read $\binom{n}{k}$ as "$n$ choose $k$."

### Implementation from Scratch

Let us build these functions ourselves before using the built-in versions.

```{r counting-from-scratch}
my_factorial <- function(n) {
  stopifnot(n >= 0, n == floor(n))
  if (n <= 1) return(1L)
  result <- 1
  for (i in 2:n) {
    result <- result * i
  }
  result
}

my_permutations <- function(n, k) {
  stopifnot(k <= n, k >= 0)
  my_factorial(n) / my_factorial(n - k)
}

my_choose <- function(n, k) {
  stopifnot(k <= n, k >= 0)
  my_factorial(n) / (my_factorial(k) * my_factorial(n - k))
}

# Verify against built-in functions
c(ours = my_factorial(10), builtin = factorial(10))
c(ours = my_choose(52, 5), builtin = choose(52, 5))
my_permutations(10, 3)
```

### Worked Example: Lottery Odds

In a typical "6/49" lottery, a player chooses 6 numbers from 1 to 49. The order does not matter. The total number of possible combinations is:

$$
\binom{49}{6} = \frac{49!}{6! \cdot 43!} = 13{,}983{,}816
$$

```{r lottery}
total_combinations <- choose(49, 6)
format(total_combinations, big.mark = ",")
signif(1 / total_combinations, 4)
```

### Worked Example: Poker Hands

A poker hand is 5 cards from a standard 52-card deck. Let us compute the probability of being dealt certain hands.

```{r poker}
total_hands <- choose(52, 5)
format(total_hands, big.mark = ",")

# Four of a kind: choose the rank (13 ways), choose which 4 suits (1 way),
# choose the 5th card from remaining 48
four_of_a_kind <- 13 * choose(4, 4) * 48
signif(four_of_a_kind / total_hands, 4)

# Flush: all 5 cards same suit, minus straight flushes
# Choose suit (4), choose 5 ranks from 13
flush_all <- 4 * choose(13, 5)
straight_flush <- 40  # 10 per suit × 4 suits
flush_only <- flush_all - straight_flush
signif(flush_only / total_hands, 4)

# Full house: three of one rank, two of another
full_house <- 13 * choose(4, 3) * 12 * choose(4, 2)
signif(full_house / total_hands, 4)
```

### The Birthday Problem

This is one of the most famous results in combinatorics, and it surprises almost everyone.

**Question.** In a room of $n$ people, what is the probability that at least two share a birthday? (Ignore leap years; assume 365 equally likely birthdays.)

**Derivation.** It is easier to compute the complement: the probability that *all* birthdays are distinct.

For the first person, any birthday is fine: $365/365$ choices. For the second person, to avoid a match, they must have one of the remaining 364 days: $364/365$. For the third: $363/365$. And so on.

$$
P(\text{all distinct}) = \frac{365}{365} \times \frac{364}{365} \times \frac{363}{365} \times \cdots \times \frac{365 - n + 1}{365} = \prod_{i=0}^{n-1} \frac{365 - i}{365}
$$

Therefore:

$$
P(\text{at least one match}) = 1 - \prod_{i=0}^{n-1} \frac{365 - i}{365}
$$

The surprise: with just $n = 23$ people, this probability exceeds 50%.

```{r birthday-formula}
birthday_prob <- function(n) {
  if (n > 365) return(1)
  p_all_distinct <- prod((365:(365 - n + 1)) / 365)
  1 - p_all_distinct
}

# Key values
for (n in c(10, 20, 23, 30, 50, 70)) {
  cat(sprintf("n = %2d: P(match) = %.4f\n", n, birthday_prob(n)))
}
```

Let us verify this with simulation.

```{r birthday-simulation}
set.seed(42)

simulate_birthday <- function(n_people, n_sim = 50000L) {
  matches <- 0L
  for (i in seq_len(n_sim)) {
    birthdays <- sample(1:365, size = n_people, replace = TRUE)
    if (any(duplicated(birthdays))) {
      matches <- matches + 1L
    }
  }
  matches / n_sim
}

# Compare formula and simulation
group_sizes <- c(10, 20, 23, 30, 40, 50, 60, 70, 80)
formula_probs <- vapply(group_sizes, birthday_prob, numeric(1))
sim_probs <- vapply(group_sizes, simulate_birthday, numeric(1))

comparison <- data.frame(
  n = group_sizes,
  formula = round(formula_probs, 4),
  simulation = round(sim_probs, 4)
)
print(comparison)
```

```{r birthday-plot, fig.cap="The birthday problem: probability that at least two people in a group of n share a birthday. The 50% threshold is crossed at n = 23."}
box::use(ggplot2)

n_vals <- 2:80
probs <- vapply(n_vals, birthday_prob, numeric(1))

df_birthday <- data.frame(n = n_vals, prob = probs)

ggplot2$ggplot(df_birthday, ggplot2$aes(x = n, y = prob)) +
  ggplot2$geom_line(colour = "#0072B2", linewidth = 1) +
  ggplot2$geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  ggplot2$geom_vline(xintercept = 23, linetype = "dotted", colour = "#D55E00") +
  ggplot2$annotate("text", x = 26, y = 0.45, label = "n = 23", colour = "#D55E00",
           hjust = 0, size = 5) +
  ggplot2$scale_y_continuous(labels = scales::percent_format()) +
  ggplot2$labs(
    x = "Number of people in the room",
    y = "P(at least one shared birthday)",
    title = "The Birthday Problem"
  ) +
  ggplot2$theme_minimal(base_size = 14)
```

Why is the answer so low? The key insight is that we are not asking whether someone shares *your* birthday — we are asking whether *any* pair matches. With 23 people there are $\binom{23}{2} = 253$ distinct pairs, each a potential match.

---

## Communicating to Stakeholders

When presenting probability concepts to non-technical audiences, keep these guidelines in mind:

1. **Avoid jargon.** Say "chance" or "likelihood" rather than "probability measure." Say "at least one" rather than "union."
2. **Use concrete examples.** "If we tested 1,000 people..." is far more impactful than "$P(A) = 0.03$."
3. **The complement trick is your friend.** It is often easier (and more intuitive) to say "there is a 5% chance this will *not* work" than to say "there is a 95% chance it will."
4. **Beware of large numbers.** People have poor intuition for very small probabilities. Frame lottery odds as "you would have to buy a ticket every week for 270,000 years" rather than "1 in 14 million."
5. **The birthday problem is a great icebreaker.** It demonstrates that human intuition about probability is systematically flawed — which is *why* we need the formal tools.

---

## Quick Reference

| Concept | Formula / Notation | Plain English |
|---|---|---|
| Sample space | $\Omega$ | Set of all possible outcomes |
| Event | $A \subseteq \Omega$ | A subset of outcomes we care about |
| Union | $A \cup B$ | A **or** B (or both) |
| Intersection | $A \cap B$ | A **and** B |
| Complement | $A^c$ | **Not** A |
| De Morgan I | $(A \cup B)^c = A^c \cap B^c$ | Not(A or B) = (not A) and (not B) |
| De Morgan II | $(A \cap B)^c = A^c \cup B^c$ | Not(A and B) = (not A) or (not B) |
| Axiom 1 | $P(A) \geq 0$ | Probabilities are non-negative |
| Axiom 2 | $P(\Omega) = 1$ | Something must happen |
| Axiom 3 | Countable additivity | Disjoint probabilities add |
| Complement rule | $P(A^c) = 1 - P(A)$ | Prob of not-A = 1 minus prob of A |
| Inclusion-exclusion | $P(A \cup B) = P(A) + P(B) - P(A \cap B)$ | Subtract the double-counted overlap |
| Permutations | $P(n,k) = \frac{n!}{(n-k)!}$ | Ordered arrangements |
| Combinations | $\binom{n}{k} = \frac{n!}{k!(n-k)!}$ | Unordered selections |
| Birthday threshold | $n = 23 \Rightarrow P > 0.5$ | 23 people, 50% chance of shared birthday |

---

## Exercises

**Exercise 2.1.** Let $\Omega = \{1, 2, 3, 4, 5, 6, 7, 8, 9, 10\}$, $A = \{1, 2, 3, 4, 5\}$, $B = \{3, 4, 5, 6, 7\}$. Compute (by hand, then verify in R): (a) $A \cup B$, (b) $A \cap B$, (c) $A^c$, (d) $(A \cup B)^c$, (e) $A^c \cap B^c$ (verify De Morgan's law).

**Exercise 2.2.** Prove from the axioms that if $A \subseteq B$, then $P(A) \leq P(B)$. (*Hint:* write $B = A \cup (B \cap A^c)$ and note the union is disjoint.)

**Exercise 2.3.** A committee of 5 must be formed from 8 men and 6 women. (a) How many committees are possible? (b) How many have exactly 3 women? (c) What is the probability of at least one woman? Compute by formula and verify by enumeration in R.

**Exercise 2.4.** In a group of 50 people, what is the probability that at least two share a birthday? What about at least *three* sharing a birthday? (For the second part, simulation is acceptable; the exact formula is harder.)

**Exercise 2.5.** You roll two fair dice. Let $A$ = "sum is 8" and $B$ = "at least one die shows a 5." Compute $P(A)$, $P(B)$, $P(A \cup B)$, and $P(A \cap B)$ exactly, and verify inclusion-exclusion holds.

**Exercise 2.6.** Prove Boole's inequality from the axioms: for any events $A_1, \ldots, A_n$,

$$
P\left(\bigcup_{i=1}^{n} A_i\right) \leq \sum_{i=1}^{n} P(A_i)
$$

(*Hint:* induction on $n$, using inclusion-exclusion and the fact that $P(A \cap B) \geq 0$.)

**Exercise 2.7 (Computational).** Write an R function that computes the probability of getting *exactly* $k$ matches in the birthday problem (exactly $k$ people sharing the same birthday, where a "match" means a pair). Use simulation with at least 100,000 trials.
