---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 2: Probability — The Language"
part: "Part 3: Bayes' Theorem"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, probability, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

## 2.7 Bayes' Theorem

Bayes' theorem is the single most important result in this chapter. It tells us how to **update our beliefs** when we observe new evidence. Everything in Bayesian statistics (Chapters 22–23) grows from this one formula — but even frequentists use it constantly.

### Derivation

We derive Bayes' theorem from the definition of conditional probability. No magic is involved; it is pure algebra.

**Step 1.** Recall the definition of conditional probability:

$$
P(A \mid B) = \frac{P(A \cap B)}{P(B)}
$$

**Step 2.** By the multiplication rule (rearranging the definition):

$$
P(A \cap B) = P(B \mid A) \, P(A)
$$

**Step 3.** Substitute Step 2 into Step 1:

$$
P(A \mid B) = \frac{P(B \mid A) \, P(A)}{P(B)}
$$

That is Bayes' theorem. Let us state it formally.

**Theorem (Bayes' Theorem).** For events $A$ and $B$ with $P(B) > 0$:

$$
\boxed{P(A \mid B) = \frac{P(B \mid A) \, P(A)}{P(B)}}
$$

Each term has a name:

| Term | Name | Plain English |
|---|---|---|
| $P(A \mid B)$ | **Posterior** | Updated belief about $A$ after seeing $B$ |
| $P(B \mid A)$ | **Likelihood** | How probable is the evidence $B$ if $A$ is true? |
| $P(A)$ | **Prior** | Initial belief about $A$ before seeing evidence |
| $P(B)$ | **Evidence** (or marginal likelihood) | Overall probability of observing $B$ |

In plain English: **posterior is proportional to likelihood times prior**. The evidence $P(B)$ acts as a normalising constant to ensure the posterior is a valid probability.

### The Expanded Form

When $A$ can be decomposed into a partition $A_1, A_2, \ldots, A_n$ of the sample space, we can expand $P(B)$ using the law of total probability:

$$
P(B) = \sum_{i=1}^{n} P(B \mid A_i) \, P(A_i)
$$

Substituting:

$$
P(A_j \mid B) = \frac{P(B \mid A_j) \, P(A_j)}{\sum_{i=1}^{n} P(B \mid A_i) \, P(A_i)}
$$

This is the form we use most often in practice, because we rarely know $P(B)$ directly — we compute it by summing over all possible causes.

---

### Worked Example 1: Medical Diagnostic Test

This is the classic example, and the numbers are worth committing to memory because the result is so counterintuitive.

**Setup:**
- A disease has a prevalence of 1%: $P(D) = 0.01$
- The test has a sensitivity of 99%: $P(+ \mid D) = 0.99$
- The test has a specificity of 95%: $P(- \mid D^c) = 0.95$, so $P(+ \mid D^c) = 0.05$

**Question:** If a randomly selected person tests positive, what is $P(D \mid +)$?

**Solution.** By Bayes' theorem with the expanded denominator:

$$
P(D \mid +) = \frac{P(+ \mid D) \, P(D)}{P(+ \mid D) \, P(D) + P(+ \mid D^c) \, P(D^c)}
$$

Substituting:

$$
P(D \mid +) = \frac{0.99 \times 0.01}{0.99 \times 0.01 + 0.05 \times 0.99}
$$

$$
= \frac{0.0099}{0.0099 + 0.0495}
$$

$$
= \frac{0.0099}{0.0594}
$$

$$
\approx 0.167
$$

Only about **16.7%** — not 99%, not 95%, but less than one in five.

```{r bayes-medical}
# Bayes' theorem: medical test
p_disease <- 0.01
sensitivity <- 0.99  # P(+ | D)
specificity <- 0.95  # P(- | D^c)
p_false_pos <- 1 - specificity  # P(+ | D^c)

# Numerator: P(+ | D) * P(D)
numerator <- sensitivity * p_disease

# Denominator: P(+) via law of total probability
p_positive <- sensitivity * p_disease + p_false_pos * (1 - p_disease)

# Posterior
p_disease_given_pos <- numerator / p_positive
round(p_disease_given_pos, 4)
```

**Why is this so counterintuitive?** The problem is that 1% prevalence means for every 1 truly sick person, there are 99 healthy people. Even with only a 5% false positive rate, that 5% of 99 healthy people ($\approx 5$ false positives) swamps the 1 true positive.

Let us see how the posterior changes as we vary the prevalence:

```{r bayes-prevalence-sweep, fig.cap="Positive predictive value (P(Disease | Positive test)) as a function of disease prevalence. Even an excellent test has poor predictive value when the condition is rare."}
box::use(ggplot2)

prevalences <- seq(0.001, 0.5, length.out = 500)

ppv <- function(prev, sens = 0.99, spec = 0.95) {
  fp_rate <- 1 - spec
  num <- sens * prev
  denom <- sens * prev + fp_rate * (1 - prev)
  num / denom
}

df_ppv <- data.frame(
  prevalence = prevalences,
  ppv = vapply(prevalences, ppv, numeric(1))
)

ggplot2$ggplot(df_ppv, ggplot2$aes(x = prevalence, y = ppv)) +
  ggplot2$geom_line(colour = "#0072B2", linewidth = 1.2) +
  ggplot2$geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  ggplot2$geom_vline(xintercept = 0.01, linetype = "dotted", colour = "#D55E00") +
  ggplot2$annotate("text", x = 0.03, y = 0.15,
           label = "Prevalence = 1%\nPPV \u2248 17%",
           hjust = 0, colour = "#D55E00", size = 4) +
  ggplot2$scale_x_continuous(labels = scales::percent_format()) +
  ggplot2$scale_y_continuous(labels = scales::percent_format()) +
  ggplot2$labs(
    x = "Disease prevalence",
    y = "P(Disease | Positive test)",
    title = "How prevalence affects positive predictive value",
    subtitle = "Sensitivity = 99%, Specificity = 95%"
  ) +
  ggplot2$theme_minimal(base_size = 14)
```

---

### Worked Example 2: Quality Control in Manufacturing

**Setup.** A factory receives components from three suppliers:
- Supplier A provides 60% of components, with a 2% defect rate
- Supplier B provides 25% of components, with a 5% defect rate
- Supplier C provides 15% of components, with a 10% defect rate

A randomly selected component is found to be defective. Which supplier most probably produced it?

**Solution.** We want $P(A \mid D)$, $P(B \mid D)$, $P(C \mid D)$ where $D$ = defective.

First, compute $P(D)$ by the law of total probability:

$$
P(D) = P(D \mid A)P(A) + P(D \mid B)P(B) + P(D \mid C)P(C)
$$

$$
= 0.02 \times 0.60 + 0.05 \times 0.25 + 0.10 \times 0.15
$$

$$
= 0.012 + 0.0125 + 0.015 = 0.0395
$$

Now apply Bayes' theorem to each supplier:

$$
P(A \mid D) = \frac{P(D \mid A)P(A)}{P(D)} = \frac{0.012}{0.0395} \approx 0.304
$$

$$
P(B \mid D) = \frac{P(D \mid B)P(B)}{P(D)} = \frac{0.0125}{0.0395} \approx 0.316
$$

$$
P(C \mid D) = \frac{P(D \mid C)P(C)}{P(D)} = \frac{0.015}{0.0395} \approx 0.380
$$

```{r bayes-quality}
# Supplier data
suppliers <- c("A", "B", "C")
shares <- c(0.60, 0.25, 0.15)
defect_rates <- c(0.02, 0.05, 0.10)

# P(D) -- law of total probability
p_defect <- sum(shares * defect_rates)
p_defect

# Bayes' theorem for each supplier
posteriors <- (defect_rates * shares) / p_defect
names(posteriors) <- suppliers
round(posteriors, 3)

suppliers[which.max(posteriors)]
```

Notice that **Supplier C is the most probable source**, even though Supplier A delivers three times as many components. Supplier C's higher defect rate more than compensates for its smaller share.

```{r bayes-quality-plot, fig.cap="Prior (market share) vs. posterior (probability of being the source given a defect) for each supplier. Bayes' theorem shifts weight toward higher-defect-rate suppliers."}
box::use(ggplot2)

df_suppliers <- data.frame(
  supplier = rep(suppliers, 2),
  probability = c(shares, posteriors),
  type = rep(c("Prior (market share)", "Posterior (given defective)"), each = 3)
)

ggplot2$ggplot(df_suppliers, ggplot2$aes(x = supplier, y = probability, fill = type)) +
  ggplot2$geom_col(position = "dodge", width = 0.6) +
  ggplot2$scale_fill_manual(values = c("Prior (market share)" = "#56B4E9",
                                "Posterior (given defective)" = "#D55E00")) +
  ggplot2$scale_y_continuous(labels = scales::percent_format()) +
  ggplot2$labs(
    x = "Supplier", y = "Probability",
    title = "Bayes' theorem in quality control",
    fill = NULL
  ) +
  ggplot2$theme_minimal(base_size = 14) +
  ggplot2$theme(legend.position = "top")
```

---

### Worked Example 3: Email Spam Filtering (Naive Bayes Intuition)

Modern spam filters use a simplified version of Bayes' theorem called **Naive Bayes**. The idea is straightforward.

Suppose we want to classify an email as spam ($S$) or ham ($H$, not spam). We observe certain words $w_1, w_2, \ldots, w_k$ in the email. By Bayes' theorem:

$$
P(S \mid w_1, \ldots, w_k) = \frac{P(w_1, \ldots, w_k \mid S) \, P(S)}{P(w_1, \ldots, w_k)}
$$

The "naive" assumption is that words are conditionally independent given the class:

$$
P(w_1, \ldots, w_k \mid S) \approx \prod_{i=1}^{k} P(w_i \mid S)
$$

This assumption is almost certainly *wrong* (words are correlated), but it works remarkably well in practice.

```{r naive-bayes-demo}
set.seed(42)

# Toy spam filter
# Prior probabilities
p_spam <- 0.3
p_ham <- 0.7

# Word likelihoods: P(word | spam) and P(word | ham)
words <- c("free", "offer", "meeting", "report", "click", "unsubscribe")
p_word_given_spam <- c(0.80, 0.70, 0.10, 0.05, 0.60, 0.50)
p_word_given_ham  <- c(0.05, 0.10, 0.60, 0.50, 0.05, 0.02)

names(p_word_given_spam) <- words
names(p_word_given_ham) <- words

# Classify an email containing "free" and "click"
email_words <- c("free", "click")

# Naive Bayes: multiply likelihoods
likelihood_spam <- prod(p_word_given_spam[email_words])
likelihood_ham  <- prod(p_word_given_ham[email_words])

# Unnormalised posteriors
unnorm_spam <- likelihood_spam * p_spam
unnorm_ham  <- likelihood_ham * p_ham

# Normalise
evidence <- unnorm_spam + unnorm_ham
p_spam_given_email <- unnorm_spam / evidence
p_ham_given_email  <- unnorm_ham / evidence

round(c(
  "P(spam | 'free','click')" = p_spam_given_email,
  "P(ham | 'free','click')" = p_ham_given_email
), 4)

# Now try a different email: "meeting" and "report"
email_words2 <- c("meeting", "report")
likelihood_spam2 <- prod(p_word_given_spam[email_words2])
likelihood_ham2  <- prod(p_word_given_ham[email_words2])
unnorm_spam2 <- likelihood_spam2 * p_spam
unnorm_ham2  <- likelihood_ham2 * p_ham
evidence2 <- unnorm_spam2 + unnorm_ham2

round(c(
  "P(spam | 'meeting','report')" = unnorm_spam2 / evidence2,
  "P(ham | 'meeting','report')" = unnorm_ham2 / evidence2
), 4)
```

### Implementing Bayes' Theorem from Scratch

Let us write a general-purpose function for Bayes' theorem.

```{r bayes-function}
bayes_theorem <- function(prior, likelihood, evidence = NULL) {
  # prior: named vector of prior probabilities P(H_i)
  # likelihood: named vector of likelihoods P(E | H_i)
  # evidence: P(E), computed automatically if NULL
  #
  # Returns: named vector of posterior probabilities P(H_i | E)

  stopifnot(
    length(prior) == length(likelihood),
    all(prior >= 0),
    abs(sum(prior) - 1) < 1e-10
  )

  if (is.null(evidence)) {
    evidence <- sum(likelihood * prior)
  }

  posterior <- (likelihood * prior) / evidence
  names(posterior) <- names(prior)
  posterior
}

# Test with medical example
result <- bayes_theorem(
  prior = c(Disease = 0.01, Healthy = 0.99),
  likelihood = c(Disease = 0.99, Healthy = 0.05)
)
round(result, 4)

# Test with supplier example
result2 <- bayes_theorem(
  prior = c(A = 0.60, B = 0.25, C = 0.15),
  likelihood = c(A = 0.02, B = 0.05, C = 0.10)
)
round(result2, 4)
```

---

### Sequential Bayesian Updating

One of the most powerful aspects of Bayes' theorem is that it can be applied **sequentially**. The posterior from one observation becomes the prior for the next.

Suppose we are trying to determine whether a coin is fair. We start with some prior belief, then update each time we observe a flip.

Mathematically, if we observe data $D_1, D_2, \ldots, D_n$ sequentially:

$$
P(\theta \mid D_1) \propto P(D_1 \mid \theta) \, P(\theta)
$$

$$
P(\theta \mid D_1, D_2) \propto P(D_2 \mid \theta) \, P(\theta \mid D_1)
$$

The posterior after seeing $D_1$ becomes the prior for processing $D_2$. This "ratchet" property means **the order of observations does not matter** — the final posterior is the same regardless of the order in which data arrive.

```{r sequential-bayes}
set.seed(42)

# Sequential Bayesian updating for a coin
# Hypothesis: theta = P(heads)
# We use a grid approximation (discretise theta)

theta_grid <- seq(0, 1, length.out = 1001)
# Prior: uniform (we know nothing)
prior <- rep(1, length(theta_grid))
prior <- prior / sum(prior)  # normalise

# True coin bias
true_theta <- 0.7

# Generate data
n_flips <- 100L
flips <- rbinom(n_flips, size = 1, prob = true_theta)

# Store posteriors at key points
snapshots <- list()
snapshot_at <- c(0, 1, 5, 10, 25, 50, 100)

# Record prior
snapshots[["n = 0"]] <- prior

# Update sequentially
current <- prior
for (i in seq_len(n_flips)) {
  # Likelihood for this single observation
  if (flips[i] == 1) {
    lik <- theta_grid  # P(H | theta) = theta
  } else {
    lik <- 1 - theta_grid  # P(T | theta) = 1 - theta
  }

  # Bayes update
  current <- current * lik
  current <- current / sum(current)  # normalise

  if (i %in% snapshot_at) {
    snapshots[[paste0("n = ", i)]] <- current
  }
}

# Build data frame for plotting
df_list <- lapply(names(snapshots), function(nm) {
  data.frame(
    theta = theta_grid,
    density = snapshots[[nm]] * length(theta_grid),  # scale for visualisation
    stage = nm,
    stringsAsFactors = FALSE
  )
})
df_bayes <- do.call(rbind, df_list)
df_bayes$stage <- factor(df_bayes$stage, levels = paste0("n = ", snapshot_at))
```

```{r sequential-bayes-plot, fig.cap="Sequential Bayesian updating: starting from a uniform prior, the posterior concentrates around the true parameter (theta = 0.7, dashed line) as more coin flips are observed."}
box::use(ggplot2)

ggplot2$ggplot(df_bayes, ggplot2$aes(x = theta, y = density)) +
  ggplot2$geom_line(colour = "#0072B2", linewidth = 0.8) +
  ggplot2$geom_vline(xintercept = true_theta, linetype = "dashed",
             colour = "#D55E00", linewidth = 0.6) +
  ggplot2$facet_wrap(~ stage, scales = "free_y", ncol = 2) +
  ggplot2$labs(
    x = expression(theta ~ "(probability of heads)"),
    y = "Posterior density",
    title = "Sequential Bayesian updating: coin bias estimation",
    subtitle = "Dashed line = true \u03b8 = 0.7. Prior is uniform."
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(strip.text = ggplot2$element_text(face = "bold"))
```

Observe how the posterior evolves:
- At $n = 0$ (the prior), it is flat — we know nothing.
- After just 1 flip, we see a slight tilt.
- By $n = 10$, the posterior is starting to peak near 0.7.
- By $n = 50$ and $n = 100$, the posterior is tightly concentrated around the true value.

This is the Bayesian learning machine at work: **data overwhelms the prior**.

The key fact underlying Bayesian updating is that if our prior for $\theta$ (the probability of heads) is $\text{Beta}(\alpha, \beta)$, and we observe $k$ heads in $n$ flips, the posterior is $\text{Beta}(\alpha + k, \beta + n - k)$. This is called a **conjugate prior** -- the posterior has the same functional form as the prior. We will study this rigorously in Chapters 22--23.

### Prior-to-Posterior Evolution

Let us visualise the full evolution from prior to posterior across increasing sample sizes, showing how data gradually sharpens our inference.

```{r prior-posterior-evolution, fig.cap="Beta-Binomial Bayesian updating. The prior (alpha = 2, beta = 2) is mildly informative. As the sample size grows, the posterior concentrates around the true theta = 0.7."}
set.seed(42)
box::use(ggplot2)

# Beta-Binomial conjugate updating
true_theta <- 0.7
alpha0 <- 2  # prior parameter
beta0 <- 2   # prior parameter

# Generate coin flips
n_total <- 200L
flips <- rbinom(n_total, size = 1, prob = true_theta)
cumulative_heads <- cumsum(flips)
cumulative_tails <- seq_len(n_total) - cumulative_heads

# Create density curves at selected sample sizes
sample_sizes <- c(0, 5, 20, 50, 100, 200)
theta_seq <- seq(0.001, 0.999, length.out = 500)

df_list <- lapply(sample_sizes, function(n) {
  if (n == 0) {
    a <- alpha0
    b <- beta0
  } else {
    a <- alpha0 + cumulative_heads[n]
    b <- beta0 + cumulative_tails[n]
  }
  data.frame(
    theta = theta_seq,
    density = dbeta(theta_seq, a, b),
    label = paste0("n = ", n, "  (α=", a, ", β=", b, ")"),
    n = n,
    stringsAsFactors = FALSE
  )
})

df_evo <- do.call(rbind, df_list)
df_evo$label <- factor(df_evo$label,
  levels = unique(df_evo$label[order(df_evo$n)]))

ggplot2$ggplot(df_evo, ggplot2$aes(x = theta, y = density, colour = label)) +
  ggplot2$geom_line(linewidth = 1) +
  ggplot2$geom_vline(xintercept = true_theta, linetype = "dashed",
             colour = "grey30", linewidth = 0.6) +
  ggplot2$scale_colour_viridis_d(option = "D", end = 0.9) +
  ggplot2$labs(
    x = expression(theta),
    y = "Density",
    colour = "Sample size",
    title = "Bayesian learning: Beta-Binomial conjugate model",
    subtitle = expression(paste("Prior: Beta(2, 2). True ", theta, " = 0.7 (dashed line)"))
  ) +
  ggplot2$theme_minimal(base_size = 14) +
  ggplot2$theme(legend.position = "right")
```

The figure shows a key property: **the posterior width shrinks as $\sqrt{n}$**. More data means more certainty, regardless of where we started. With enough data, different priors lead to essentially the same posterior — the data dominate.

---

### Connection to Part IV of This Book

Bayes' theorem is not just a formula — it is a **philosophy of inference**. In Chapters 22–23, we will develop this idea fully:

- We will replace the discrete events of this chapter with continuous **prior distributions** and **posterior distributions**.
- We will use **Markov chain Monte Carlo (MCMC)** to compute posteriors when analytic solutions do not exist.
- We will compare Bayesian and frequentist approaches to the same problem and see when each has advantages.

For now, the key takeaway is: **probability is the language, and Bayes' theorem is the grammar for updating beliefs with evidence.**

---

## 2.8 Summary and the Road Ahead

This chapter established the language of probability. Let us review what we built:

1. **Sets and sample spaces** — the stage on which probability plays out.
2. **Kolmogorov's axioms** — three simple rules from which everything follows.
3. **Counting** — permutations, combinations, and the birthday problem.
4. **Conditional probability** — how learning changes probabilities.
5. **Independence** — when learning changes nothing.
6. **Bayes' theorem** — the master equation for updating beliefs.

Every subsequent chapter depends on this material:
- **Chapter 3** introduces random variables, which are functions on sample spaces.
- **Chapters 4–5** develop specific probability distributions.
- **Chapters 8–11** use probability to build and test statistical estimators.
- **Chapters 22–23** take Bayes' theorem to its full Bayesian conclusion.

Probability is not a topic to master once and move on from. It is the foundation that supports the entire structure of statistics.

---

## Communicating to Stakeholders

When using Bayesian reasoning in applied settings:

1. **Lead with the conclusion, then explain.** "Given the test result, there is a 17% chance of disease" is more useful than walking through the derivation first. Show the derivation only when asked.

2. **Prior, evidence, posterior.** This three-part structure is a natural narrative: "Before the data, we believed X. The data showed Y. After updating, we now believe Z." Decision-makers understand stories.

3. **Sensitivity analysis.** Show how the conclusion changes under different assumptions about the prior. If the conclusion is robust to the prior, stakeholders gain confidence. If it is sensitive, that is valuable information too.

4. **Do not hide the prior.** In Bayesian analysis, the prior is an explicit assumption. This is a *strength*, not a weakness. Frequentist methods have implicit assumptions too — at least Bayesian assumptions are on the table.

5. **Use the natural frequency framing.** "Out of 10,000 people screened, we expect 100 to have the condition and 495 healthy people to test positive falsely. So among the roughly 594 positive tests, only 99 are true positives." This is concrete and convincing.

---

## Quick Reference

| Concept | Formula | Plain English |
|---|---|---|
| Bayes' theorem | $P(A \mid B) = \frac{P(B \mid A) P(A)}{P(B)}$ | Posterior ∝ likelihood × prior |
| Expanded form | $P(A_j \mid B) = \frac{P(B \mid A_j) P(A_j)}{\sum_i P(B \mid A_i) P(A_i)}$ | Denominator by total probability |
| Prior | $P(A)$ | Belief before evidence |
| Likelihood | $P(B \mid A)$ | Probability of evidence given hypothesis |
| Posterior | $P(A \mid B)$ | Updated belief after evidence |
| Evidence | $P(B)$ | Normalising constant |
| Sequential updating | Today's posterior = tomorrow's prior | Data can arrive one at a time |
| Conjugate prior | Prior and posterior share same family | Makes computation easy |
| Beta-Binomial | $\text{Beta}(\alpha, \beta) + k \text{ heads} \to \text{Beta}(\alpha+k, \beta+n-k)$ | Coin flipping model |
| PPV | $\frac{\text{sensitivity} \times \text{prevalence}}{\text{sensitivity} \times \text{prevalence} + (1-\text{specificity}) \times (1-\text{prevalence})}$ | Positive predictive value |

---

## Exercises

**Exercise 2.15.** Derive Bayes' theorem from scratch, starting only from the definition of conditional probability. (Reproduce the derivation given in this chapter, but from memory.)

**Exercise 2.16.** A disease has prevalence 5%. A test has sensitivity 90% and specificity 90%. (a) Compute $P(D \mid +)$. (b) The person tests positive and takes the test a *second* time (independently), which is also positive. Now what is $P(D \mid + , +)$? (*Hint:* use the posterior from (a) as the prior for (b).)

**Exercise 2.17.** In the spam filtering example, an email contains the words "free," "offer," and "unsubscribe." Using the word likelihoods defined in the text, compute $P(\text{spam} \mid \text{email})$ using Naive Bayes. Is the email spam or ham?

**Exercise 2.18 (Computational).** Implement the Beta-Binomial Bayesian updating model in R. Start with a $\text{Beta}(1, 1)$ prior (uniform). Generate 500 coin flips with $\theta = 0.6$. Plot the posterior after every 50 observations using ggplot2 (10 panels in a facet grid).

```{r ex-beta-binom-starter, eval=FALSE}
# Starter code for Exercise 2.18
set.seed(42)
box::use(ggplot2)

true_theta <- 0.6
n_total <- 500L
flips <- rbinom(n_total, size = 1, prob = true_theta)

alpha0 <- 1
beta0 <- 1

# Build the plot data at n = 50, 100, 150, ..., 500
snapshot_ns <- seq(50, 500, by = 50)
theta_grid <- seq(0.001, 0.999, length.out = 500)

# Your code here: compute posterior Beta parameters at each snapshot
# and create a faceted ggplot2 figure.
```

**Exercise 2.19.** Prove that Bayes' theorem with a uniform prior over $n$ equally likely hypotheses reduces to the classical counting formula:

$$
P(H_j \mid E) = \frac{P(E \mid H_j)}{\sum_{i=1}^n P(E \mid H_i)}
$$

**Exercise 2.20 (Proof).** Show that Bayesian updating is *commutative*: if we observe data $D_1$ then $D_2$, or $D_2$ then $D_1$, the final posterior is the same. (*Hint:* write out the sequential updates and show the result depends only on $P(D_1, D_2 \mid \theta)$, which is the same regardless of ordering.)

**Exercise 2.21 (Computational).** The "German tank problem." During WWII, the Allies estimated the number of German tanks by looking at serial numbers on captured tanks. Suppose tanks are numbered $1, 2, \ldots, N$ and you capture tanks with serial numbers $\{43, 102, 77, 215, 156\}$.

(a) The frequentist estimator (minimum variance unbiased estimator) for $N$ is:

$$
\hat{N} = m \left(1 + \frac{1}{k}\right) - 1
$$

where $m = \max(\text{serial numbers})$ and $k$ is the number of captured tanks. Compute $\hat{N}$.

(b) Implement a Bayesian solution using a uniform prior on $N$ from 1 to 1000. Compute and plot the posterior distribution of $N$ given the observed serial numbers. Compare the posterior mean and mode to the frequentist estimator.

```{r german-tank-starter, eval=FALSE}
# Starter code for Exercise 2.21
observed <- c(43, 102, 77, 215, 156)
k <- length(observed)
m <- max(observed)

# (a) Frequentist estimate
n_hat_freq <- m * (1 + 1/k) - 1
cat("Frequentist estimate:", n_hat_freq, "\n")

# (b) Bayesian approach
# Prior: uniform on N = 215, 216, ..., 1000
# (N must be at least max(observed) = 215)
N_grid <- 215:1000

# Likelihood: P(observed | N) = 1 / choose(N, k) for N >= m
# (each set of k serial numbers from 1..N equally likely)
log_likelihood <- -lchoose(N_grid, k)

# Your code: compute and normalise the posterior, plot it
```

**Exercise 2.22.** (Conceptual) Explain in your own words why the medical testing example is so surprising. What makes the base rate so important? Under what circumstances would a positive test result actually indicate a high probability of disease?
