---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 22: Bayesian Thinking"
part: "Part 1: Framework and Conjugate Priors"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, bayesian, conjugate-priors, posterior, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

```{r packages, include=FALSE}
box::use(ggplot2)
box::use(data.table)
```

```{r dt-symbols, include=FALSE}
.N <- data.table::.N
.SD <- data.table::.SD
```

# Chapter 22 -- Bayesian Thinking

## Part 1: Framework and Conjugate Priors

Throughout this book we have treated parameters as fixed but unknown quantities and used data to make inferences about them. This is the **frequentist** perspective. In the **Bayesian** framework we take a fundamentally different view: parameters are random variables with probability distributions that encode our uncertainty. Data update that uncertainty via Bayes' theorem, yielding a **posterior distribution** that combines prior beliefs with observed evidence.

This chapter derives the Bayesian machinery from first principles, develops the most important conjugate families, and shows how posterior updating works in practice.

---

## 22.1 From Bayes' Theorem to Bayesian Inference

### 22.1.1 Review of Bayes' Theorem

In Chapter 2 we proved Bayes' theorem for events:

$$P(A \mid B) = \frac{P(B \mid A) \, P(A)}{P(B)}$$

Now let $\theta$ be a continuous parameter and $\mathbf{y} = (y_1, \dots, y_n)$ the observed data. Replace events with densities:

$$\pi(\theta \mid \mathbf{y}) = \frac{f(\mathbf{y} \mid \theta) \, \pi(\theta)}{m(\mathbf{y})}$$

where:

- $\pi(\theta)$ is the **prior** density -- what we believe about $\theta$ before seeing data,
- $f(\mathbf{y} \mid \theta) = L(\theta; \mathbf{y})$ is the **likelihood** (Chapter 8),
- $m(\mathbf{y}) = \int f(\mathbf{y} \mid \theta) \, \pi(\theta) \, d\theta$ is the **marginal likelihood** (a normalising constant),
- $\pi(\theta \mid \mathbf{y})$ is the **posterior** density -- updated belief after observing data.

### 22.1.2 The Proportionality Shortcut

Since $m(\mathbf{y})$ does not depend on $\theta$, the entire Bayesian programme can be summarised:

$$\boxed{\text{posterior} \propto \text{likelihood} \times \text{prior}}$$

$$\pi(\theta \mid \mathbf{y}) \propto L(\theta; \mathbf{y}) \, \pi(\theta)$$

This proportionality is the engine of Bayesian statistics. We identify the posterior's kernel (the $\theta$-dependent part) and recognise it as a known distribution.

---

## 22.2 A Simple Example: Estimating a Proportion

Suppose we observe $y$ successes in $n$ independent Bernoulli trials. The likelihood is:

$$L(p; y, n) = \binom{n}{y} p^y (1-p)^{n-y} \propto p^y (1-p)^{n-y}$$

If we adopt a $\text{Beta}(\alpha, \beta)$ prior on $p$:

$$\pi(p) \propto p^{\alpha - 1}(1-p)^{\beta - 1}$$

then the posterior is:

$$\pi(p \mid y) \propto p^{y + \alpha - 1}(1-p)^{n - y + \beta - 1}$$

This is the kernel of a $\text{Beta}(y + \alpha, n - y + \beta)$ distribution. The prior "pseudo-counts" $\alpha$ and $\beta$ combine additively with the data counts $y$ and $n - y$.

```{r beta-binomial-example, fig.cap="Beta-Binomial conjugate updating. Prior (dashed), likelihood (dotted, rescaled), and posterior (solid). Observe how the posterior is a compromise between prior and likelihood."}
set.seed(42)

alpha_prior <- 2
beta_prior <- 5
n_obs <- 20
y_obs <- 12

alpha_post <- alpha_prior + y_obs
beta_post <- beta_prior + (n_obs - y_obs)

p_grid <- seq(0, 1, length.out = 500)
prior_dens <- dbeta(p_grid, alpha_prior, beta_prior)
post_dens <- dbeta(p_grid, alpha_post, beta_post)
lik_vals <- dbinom(y_obs, n_obs, p_grid)
lik_scaled <- lik_vals / max(lik_vals) * max(post_dens)

plot_dt <- data.table$data.table(
  p = rep(p_grid, 3),
  density = c(prior_dens, lik_scaled, post_dens),
  Component = rep(c("Prior Beta(2,5)", "Likelihood (scaled)", "Posterior Beta(14,13)"), each = 500)
)

ggplot2$ggplot(plot_dt, ggplot2$aes(x = p, y = density, colour = Component, linetype = Component)) +
  ggplot2$geom_line(linewidth = 1.1) +
  ggplot2$scale_linetype_manual(values = c("dashed", "dotted", "solid")) +
  ggplot2$scale_colour_manual(values = c("#0072B2", "#D55E00", "#009E73")) +
  ggplot2$labs(
    x = "p", y = "Density",
    title = "Beta-Binomial conjugate updating: 12 successes in 20 trials"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "top")
```

**Key observation.** The posterior mean is:

$$E[p \mid y] = \frac{\alpha + y}{\alpha + \beta + n} = \frac{2 + 12}{2 + 5 + 20} = \frac{14}{27} \approx 0.519$$

This lies between the prior mean $\alpha / (\alpha + \beta) = 2/7 \approx 0.286$ and the MLE $y/n = 0.6$. The posterior is a **precision-weighted compromise** between prior and data.

---

## 22.3 Conjugate Priors

### 22.3.1 Definition

A family of priors $\mathcal{F}$ is **conjugate** to a likelihood if, whenever $\pi(\theta) \in \mathcal{F}$, the posterior $\pi(\theta \mid \mathbf{y}) \in \mathcal{F}$. Conjugacy keeps the posterior in the same distributional family, making updates analytically tractable.

### 22.3.2 Why Conjugacy Matters

1. **Closed-form posteriors** -- no numerical integration needed.
2. **Interpretable updates** -- prior hyperparameters have clear meaning (pseudo-observations).
3. **Sequential updating** -- today's posterior becomes tomorrow's prior.

### 22.3.3 The Beta-Binomial Conjugate Family (Detailed)

We have already seen the result. Let us state it formally.

**Theorem (Beta-Binomial conjugacy).** If $Y \mid p \sim \text{Binomial}(n, p)$ and $p \sim \text{Beta}(\alpha, \beta)$, then:

$$p \mid y \sim \text{Beta}(\alpha + y, \; \beta + n - y)$$

*Proof.* The posterior kernel is:

$$\pi(p \mid y) \propto \underbrace{p^y(1-p)^{n-y}}_{\text{likelihood}} \cdot \underbrace{p^{\alpha-1}(1-p)^{\beta-1}}_{\text{prior}} = p^{\alpha + y - 1}(1-p)^{\beta + n - y - 1}$$

which is the kernel of $\text{Beta}(\alpha + y, \beta + n - y)$. $\square$

---

## 22.4 Normal-Normal Conjugacy

### 22.4.1 Known Variance

Suppose $Y_1, \dots, Y_n \mid \mu \overset{\text{iid}}{\sim} N(\mu, \sigma^2)$ with $\sigma^2$ known, and $\mu \sim N(\mu_0, \tau_0^2)$. The posterior is:

$$\mu \mid \mathbf{y} \sim N(\mu_n, \tau_n^2)$$

where:

$$\tau_n^2 = \left(\frac{1}{\tau_0^2} + \frac{n}{\sigma^2}\right)^{-1}, \qquad \mu_n = \tau_n^2 \left(\frac{\mu_0}{\tau_0^2} + \frac{n \bar{y}}{\sigma^2}\right)$$

*Derivation.* The log-posterior kernel is:

$$\log \pi(\mu \mid \mathbf{y}) = \text{const} - \frac{1}{2\tau_0^2}(\mu - \mu_0)^2 - \frac{n}{2\sigma^2}(\mu - \bar{y})^2$$

Expanding and completing the square in $\mu$ yields the result. The posterior precision (inverse variance) is the sum of the prior precision and the data precision:

$$\frac{1}{\tau_n^2} = \underbrace{\frac{1}{\tau_0^2}}_{\text{prior precision}} + \underbrace{\frac{n}{\sigma^2}}_{\text{data precision}}$$

The posterior mean $\mu_n$ is the **precision-weighted average** of the prior mean and the sample mean.

```{r normal-normal, fig.cap="Normal-Normal conjugate updating. As sample size increases, the posterior concentrates around the sample mean."}
set.seed(42)
true_mu <- 5
sigma <- 2
mu_0 <- 3
tau_0 <- 2

sample_sizes <- c(1, 5, 20, 100)
y_all <- rnorm(100, true_mu, sigma)
mu_grid <- seq(-1, 9, length.out = 500)

plot_list <- list()
for (n in sample_sizes) {
  y_bar <- mean(y_all[1:n])
  tau_n2 <- 1 / (1/tau_0^2 + n/sigma^2)
  mu_n <- tau_n2 * (mu_0/tau_0^2 + n * y_bar / sigma^2)
  plot_list[[length(plot_list) + 1]] <- data.table$data.table(
    mu = mu_grid,
    density = dnorm(mu_grid, mu_n, sqrt(tau_n2)),
    n = paste0("n = ", n)
  )
}

prior_dt <- data.table$data.table(
  mu = mu_grid,
  density = dnorm(mu_grid, mu_0, tau_0),
  n = "Prior"
)

all_dt <- data.table$rbindlist(c(list(prior_dt), plot_list))
all_dt[, n := factor(n, levels = c("Prior", paste0("n = ", sample_sizes)))]

ggplot2$ggplot(all_dt, ggplot2$aes(x = mu, y = density, colour = n)) +
  ggplot2$geom_line(linewidth = 1) +
  ggplot2$geom_vline(xintercept = true_mu, linetype = "dashed", alpha = 0.5) +
  ggplot2$labs(
    x = expression(mu), y = "Density",
    title = "Normal-Normal updating: posterior concentrates with more data",
    colour = NULL
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "top")
```

---

## 22.5 Gamma-Poisson Conjugacy

### 22.5.1 Derivation

If $Y_1, \dots, Y_n \mid \lambda \overset{\text{iid}}{\sim} \text{Poisson}(\lambda)$ and $\lambda \sim \text{Gamma}(\alpha, \beta)$, then:

$$\lambda \mid \mathbf{y} \sim \text{Gamma}\!\left(\alpha + \sum y_i, \; \beta + n\right)$$

*Proof.* The likelihood is:

$$L(\lambda) \propto \lambda^{\sum y_i} e^{-n\lambda}$$

The prior kernel is $\lambda^{\alpha - 1} e^{-\beta \lambda}$. The posterior kernel is:

$$\lambda^{\alpha + \sum y_i - 1} \, e^{-(\beta + n)\lambda}$$

which is $\text{Gamma}(\alpha + \sum y_i, \beta + n)$. $\square$

**Interpretation.** The prior hyperparameters encode $\alpha - 1$ pseudo-events observed over $\beta$ pseudo-time-units. The data contribute $\sum y_i$ observed events over $n$ units.

```{r gamma-poisson, fig.cap="Gamma-Poisson conjugate updating for count data."}
set.seed(42)
alpha_gp <- 3
beta_gp <- 1
true_lambda <- 4

y_pois <- rpois(30, true_lambda)
sum_y <- sum(y_pois)
n_pois <- length(y_pois)

lambda_grid <- seq(0, 10, length.out = 500)

gp_dt <- data.table$data.table(
  lambda = rep(lambda_grid, 3),
  density = c(
    dgamma(lambda_grid, alpha_gp, beta_gp),
    dgamma(lambda_grid, alpha_gp + sum_y, beta_gp + n_pois),
    dgamma(lambda_grid, sum_y + 1, n_pois)
  ),
  Component = rep(c("Prior Gamma(3,1)", "Posterior", "Jeffreys posterior"), each = 500)
)

ggplot2$ggplot(gp_dt, ggplot2$aes(x = lambda, y = density, colour = Component, linetype = Component)) +
  ggplot2$geom_line(linewidth = 1.1) +
  ggplot2$geom_vline(xintercept = true_lambda, linetype = "dashed", alpha = 0.4) +
  ggplot2$scale_colour_manual(values = c("#D55E00", "#009E73", "#0072B2")) +
  ggplot2$labs(
    x = expression(lambda), y = "Density",
    title = "Gamma-Poisson updating: 30 Poisson observations"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "top")
```

---

## 22.6 Summary of Conjugate Families

| Likelihood | Prior | Posterior | Updated hyperparameters |
|:-----------|:------|:----------|:------------------------|
| $\text{Binomial}(n, p)$ | $\text{Beta}(\alpha, \beta)$ | $\text{Beta}(\alpha + y, \beta + n - y)$ | Add successes and failures |
| $N(\mu, \sigma^2)$ (known $\sigma^2$) | $N(\mu_0, \tau_0^2)$ | $N(\mu_n, \tau_n^2)$ | Precision-weighted mean |
| $\text{Poisson}(\lambda)$ | $\text{Gamma}(\alpha, \beta)$ | $\text{Gamma}(\alpha + \sum y_i, \beta + n)$ | Add events and units |
| $\text{Exp}(\lambda)$ | $\text{Gamma}(\alpha, \beta)$ | $\text{Gamma}(\alpha + n, \beta + \sum y_i)$ | Add observations and total time |
| $N(\mu, \sigma^2)$ (known $\mu$) | $\text{Inv-Gamma}(\alpha, \beta)$ | $\text{Inv-Gamma}(\alpha + n/2, \beta + S/2)$ | Add half-counts and half-SS |

---

## 22.7 Sequential Updating

A powerful consequence of Bayes' theorem is that updating is **sequential**. Suppose we observe data in two batches, $\mathbf{y}_1$ then $\mathbf{y}_2$. By Bayes:

$$\pi(\theta \mid \mathbf{y}_1, \mathbf{y}_2) \propto f(\mathbf{y}_2 \mid \theta) \cdot \underbrace{f(\mathbf{y}_1 \mid \theta) \, \pi(\theta)}_{\propto \, \pi(\theta \mid \mathbf{y}_1)}$$

The posterior after $\mathbf{y}_1$ serves as the prior for $\mathbf{y}_2$. The final posterior is the same whether we update sequentially or all at once. This is particularly natural for streaming data.

```{r sequential-updating, fig.cap="Sequential Bayesian updating: data arrive one observation at a time and the posterior continuously sharpens."}
set.seed(42)
n_seq <- 50
y_seq <- rbinom(n_seq, 1, 0.6)

alpha_seq <- 1
beta_seq <- 1

snapshots <- c(0, 1, 5, 10, 25, 50)
p_grid <- seq(0, 1, length.out = 300)
seq_list <- list()

for (k in snapshots) {
  a_k <- alpha_seq + sum(y_seq[seq_len(k)])
  b_k <- beta_seq + k - sum(y_seq[seq_len(k)])
  seq_list[[length(seq_list) + 1]] <- data.table$data.table(
    p = p_grid,
    density = dbeta(p_grid, a_k, b_k),
    stage = paste0("After ", k, " obs")
  )
}

seq_dt <- data.table$rbindlist(seq_list)
seq_dt[, stage := factor(stage, levels = paste0("After ", snapshots, " obs"))]

ggplot2$ggplot(seq_dt, ggplot2$aes(x = p, y = density, colour = stage)) +
  ggplot2$geom_line(linewidth = 1) +
  ggplot2$geom_vline(xintercept = 0.6, linetype = "dashed", alpha = 0.5) +
  ggplot2$labs(
    x = "p", y = "Density",
    title = "Sequential Bayesian updating: Bernoulli data arriving one at a time",
    colour = NULL
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "top")
```

---

## 22.8 Interactive: Watch the Posterior Update

<InteractivePlot>

```{r bayesian-updating-interactive, eval=FALSE}
box::use(shiny)
box::use(ggplot2)

shiny$shinyApp(
  ui = shiny$fluidPage(
    shiny$titlePanel("Bayesian Updating: Data Arrive One at a Time"),
    shiny$sidebarLayout(
      shiny$sidebarPanel(
        shiny$sliderInput("true_p", "True probability:",
          min = 0.05, max = 0.95, value = 0.6, step = 0.05
        ),
        shiny$numericInput("alpha0", "Prior alpha:", value = 1, min = 0.1, step = 0.5),
        shiny$numericInput("beta0", "Prior beta:", value = 1, min = 0.1, step = 0.5),
        shiny$sliderInput("n_obs", "Number of observations:",
          min = 0, max = 200, value = 0, step = 1, animate = TRUE
        ),
        shiny$actionButton("new_data", "Generate new data")
      ),
      shiny$mainPanel(
        shiny$plotOutput("posterior_plot", Height = "450px"),
        shiny$verbatimTextOutput("summary_text")
      )
    )
  ),
  server = function(input, output, session) {
    data_store <- shiny$reactiveVal(NULL)

    shiny$observeEvent(input$new_data, {
      data_store(rbinom(200, 1, input$true_p))
    }, ignoreNULL = FALSE)

    output$posterior_plot <- shiny$renderPlot({
      d <- data_store()
      shiny$req(d)
      n <- min(input$n_obs, length(d))
      a_post <- input$alpha0 + sum(d[seq_len(n)])
      b_post <- input$beta0 + n - sum(d[seq_len(n)])

      p_grid <- seq(0, 1, length.out = 500)
      df <- data.frame(
        p = rep(p_grid, 2),
        density = c(dbeta(p_grid, input$alpha0, input$beta0),
                    dbeta(p_grid, a_post, b_post)),
        Component = rep(c("Prior", "Posterior"), each = 500)
      )

      ggplot2$ggplot(df, ggplot2$aes(x = p, y = density, colour = Component, linetype = Component)) +
        ggplot2$geom_line(linewidth = 1.2) +
        ggplot2$geom_vline(xintercept = input$true_p, linetype = "dashed") +
        ggplot2$scale_colour_manual(values = c("#0072B2", "#D55E00")) +
        ggplot2$labs(
          x = "p", y = "Density",
          title = sprintf("After %d observations (successes: %d)", n, sum(d[seq_len(n)]))
        ) +
        ggplot2$theme_minimal(base_size = 14) +
        ggplot2$theme(legend.position = "top")
    })

    output$summary_text <- shiny$renderText({
      d <- data_store()
      shiny$req(d)
      n <- min(input$n_obs, length(d))
      a_post <- input$alpha0 + sum(d[seq_len(n)])
      b_post <- input$beta0 + n - sum(d[seq_len(n)])
      sprintf(
        "Posterior: Beta(%.1f, %.1f)\nPosterior mean: %.4f\nPosterior 95%% CI: [%.4f, %.4f]\nMLE (y/n): %s",
        a_post, b_post,
        a_post / (a_post + b_post),
        qbeta(0.025, a_post, b_post),
        qbeta(0.975, a_post, b_post),
        if (n > 0) sprintf("%.4f", sum(d[seq_len(n)]) / n) else "NA"
      )
    })
  }
)
```

</InteractivePlot>

---

## 22.9 Posterior Summaries

Once we have the posterior $\pi(\theta \mid \mathbf{y})$, we summarise it:

- **Posterior mean:** $E[\theta \mid \mathbf{y}]$ -- the Bayes estimator under squared-error loss.
- **Posterior median:** the 50th percentile -- the Bayes estimator under absolute-error loss.
- **MAP (maximum a posteriori):** $\arg\max_\theta \pi(\theta \mid \mathbf{y})$ -- the mode, analogous to the MLE.
- **Credible intervals:** discussed in detail in Part 2.

For the Beta-Binomial example above:

```{r posterior-summaries}
post_mean <- alpha_post / (alpha_post + beta_post)
post_median <- qbeta(0.5, alpha_post, beta_post)
post_mode <- (alpha_post - 1) / (alpha_post + beta_post - 2)
mle <- y_obs / n_obs

summary_df <- data.frame(
  Estimator = c("Posterior mean", "Posterior median", "MAP", "MLE"),
  Value = round(c(post_mean, post_median, post_mode, mle), 4)
)
knitr::kable(summary_df, caption = "Point estimates for p from the Beta-Binomial model.")
```

---

## 22.10 The Prior as Information

The strength of the prior is controlled by the hyperparameters. For the Beta-Binomial:

- $\alpha + \beta$ controls the **concentration** -- larger values produce a tighter prior, expressing stronger beliefs.
- An $\alpha + \beta$ of $k$ is equivalent to having already seen $k$ pseudo-observations.

```{r prior-strength, fig.cap="Effect of prior strength on the posterior. Strong priors resist updating; weak priors yield quickly to the data."}
set.seed(42)
priors <- list(
  list(a = 1, b = 1, label = "Flat: Beta(1,1)"),
  list(a = 2, b = 5, label = "Weak: Beta(2,5)"),
  list(a = 20, b = 50, label = "Strong: Beta(20,50)")
)

p_grid <- seq(0, 1, length.out = 500)
strength_list <- list()

for (pr in priors) {
  a_post <- pr$a + y_obs
  b_post <- pr$b + (n_obs - y_obs)
  strength_list[[length(strength_list) + 1]] <- data.table$data.table(
    p = rep(p_grid, 2),
    density = c(dbeta(p_grid, pr$a, pr$b), dbeta(p_grid, a_post, b_post)),
    type = rep(c("Prior", "Posterior"), each = 500),
    prior_label = pr$label
  )
}

strength_dt <- data.table$rbindlist(strength_list)

ggplot2$ggplot(strength_dt, ggplot2$aes(x = p, y = density, colour = type, linetype = type)) +
  ggplot2$geom_line(linewidth = 1) +
  ggplot2$facet_wrap(~ prior_label, scales = "free_y") +
  ggplot2$geom_vline(xintercept = y_obs / n_obs, linetype = "dashed", alpha = 0.4) +
  ggplot2$scale_colour_manual(values = c("#D55E00", "#0072B2")) +
  ggplot2$labs(
    x = "p", y = "Density",
    title = "Prior strength determines how much the data shift the posterior"
  ) +
  ggplot2$theme_minimal(base_size = 12) +
  ggplot2$theme(legend.position = "top")
```

---

## 22.11 Real Data: Scurvy Trial

In 1747, James Lind treated twelve scorbutic sailors with six different remedies, two per group. Let us perform a Bayesian analysis of whether citrus fruit cures scurvy.

```{r scurvy-bayesian}
data_dir <- "../../../data"
scurvy <- data.table$fread(file.path(data_dir, "medical/scurvy.csv"))
str(scurvy)
```

```{r scurvy-analysis, fig.cap="Posterior distributions for cure probability by treatment group in Lind's scurvy trial."}
scurvy[, improved := as.integer(
  gum_rot_d6 %in% c("0_no_change", "1_mild") |
  skin_sores_d6 %in% c("0_no_change", "1_mild") |
  weakness_of_the_knees_d6 %in% c("0_no_change", "1_mild")
)]

treat_summary <- scurvy[, .(
  n = .N,
  y = sum(improved, na.rm = TRUE)
), by = treatment]

p_grid <- seq(0, 1, length.out = 500)
scurvy_list <- list()

for (i in seq_len(nrow(treat_summary))) {
  row <- treat_summary[i]
  a_post <- 1 + row$y
  b_post <- 1 + row$n - row$y
  scurvy_list[[i]] <- data.table$data.table(
    p = p_grid,
    density = dbeta(p_grid, a_post, b_post),
    treatment = row$treatment
  )
}

scurvy_post <- data.table$rbindlist(scurvy_list)

ggplot2$ggplot(scurvy_post, ggplot2$aes(x = p, y = density, colour = treatment)) +
  ggplot2$geom_line(linewidth = 1) +
  ggplot2$labs(
    x = "P(improvement)", y = "Density",
    title = "Posterior distributions for treatment efficacy (flat Beta(1,1) prior)",
    colour = "Treatment"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "right")
```

With only two patients per group, the posteriors are wide -- but citrus fruit still emerges as the most promising treatment. The Bayesian approach honestly represents our uncertainty given the tiny sample.

---

## Communicating to Stakeholders

- **Explain "updating beliefs".** "We started with a vague belief about the cure rate, then combined it with the trial data to get a refined estimate. The wider the curve, the more uncertain we are."
- **Posteriors are intuitive.** "There is a 95% probability that the true cure rate lies in this interval" -- this is what people think a confidence interval means, and in the Bayesian framework it actually does.
- **Prior disclosure.** Always report the prior used, and show how results change under different priors (sensitivity analysis, discussed in Part 2).
- **Avoid jargon.** Use "updated estimate" rather than "posterior"; "initial assumption" rather than "prior"; "range of plausible values" rather than "credible interval".

---

## Quick Reference

| Concept | Details |
|:--------|:--------|
| **Bayes' theorem** | $\pi(\theta \mid \mathbf{y}) \propto L(\theta; \mathbf{y}) \, \pi(\theta)$ |
| **Conjugate prior** | Prior and posterior in the same family |
| **Beta-Binomial** | $\text{Beta}(\alpha, \beta) \to \text{Beta}(\alpha + y, \beta + n - y)$ |
| **Normal-Normal** | Precision-weighted mean; $1/\tau_n^2 = 1/\tau_0^2 + n/\sigma^2$ |
| **Gamma-Poisson** | $\text{Gamma}(\alpha, \beta) \to \text{Gamma}(\alpha + \sum y_i, \beta + n)$ |
| **Posterior mean** | Precision-weighted average of prior mean and MLE |
| **Sequential updating** | Posterior from batch 1 is the prior for batch 2 |
| **Prior strength** | $\alpha + \beta$ pseudo-observations (Beta); $1/\tau_0^2$ precision (Normal) |
| **MAP** | Mode of posterior; coincides with penalised MLE |

---

## Exercises

1. **Beta-Binomial derivation.** Starting from a $\text{Beta}(3, 3)$ prior and observing $y = 7$ successes in $n = 10$ trials, derive the posterior distribution. Compute the posterior mean, median, mode, and a 95% equal-tailed credible interval.

2. **Normal-Normal updating.** Suppose $\sigma^2 = 4$ is known, the prior is $\mu \sim N(0, 25)$, and you observe $n = 16$ data points with $\bar{y} = 3.2$. Find the posterior distribution of $\mu$ and compute the posterior probability that $\mu > 2$.

3. **Gamma-Poisson.** A hospital records the number of emergency admissions per day. Over 30 days, the total count is 180. Using a $\text{Gamma}(2, 0.5)$ prior, find the posterior distribution of the daily rate $\lambda$. How does the posterior compare with the MLE?

4. **Sequential equivalence.** Show algebraically that for the Beta-Binomial model, updating sequentially with $y_1$ from $n_1$ trials and then $y_2$ from $n_2$ trials gives the same posterior as updating once with $y_1 + y_2$ from $n_1 + n_2$ trials.

5. **Prior sensitivity.** For the scurvy data, refit the citrus treatment model with priors $\text{Beta}(1, 1)$, $\text{Beta}(0.5, 0.5)$ (Jeffreys), and $\text{Beta}(10, 10)$ (sceptical). Plot all three posteriors on the same graph. How much does the prior affect the conclusion?

6. **Conjugate for the exponential.** Derive the conjugate prior for the exponential likelihood $f(y \mid \lambda) = \lambda e^{-\lambda y}$. Show that it is a Gamma distribution and find the updated hyperparameters after observing $\mathbf{y}$.
