---
title: "Pairs Trading"
---

```{r setup, include=FALSE}
box::use(
    ../modules/data[load_market, filter_dates],
    ../modules/stats[sharpe_ratio, annualised_return, annualised_vol, max_drawdown],
    ../modules/viz[theme_trading, trading_colors]
)

box::use(
    data.table[...],
    ggplot2[...]
)

tc <- unlist(trading_colors)

knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 10,
    fig.height = 6,
    fig.path = "../figures/08-2_"
)

set.seed(42)
```

# Pairs Trading

While single-asset mean reversion struggles because prices can trend indefinitely, *spreads* between related assets often exhibit genuine mean-reversion. If Exxon and Chevron are fundamentally similar, any divergence in their relative prices should eventually correct.

Pairs trading exploits these temporary mispricings: go long the relatively cheap asset, short the relatively expensive one, and wait for convergence. The strategy is market-neutral—profits come from the spread, not market direction.

---

## 8.4 Pairs Selection

### 8.4.1 Prose/Intuition

The first challenge in pairs trading: finding pairs that actually converge.

**Fundamental pairs:** Companies in the same industry, with similar business models, competing for the same customers. Examples:
- Exxon (XOM) and Chevron (CVX)
- Visa (V) and Mastercard (MA)
- JPMorgan (JPM) and Goldman Sachs (GS)

These pairs have economic logic—fundamental arbitrageurs keep them aligned.

**Statistical pairs:** Pairs identified purely by historical correlation or cointegration, without regard to business fundamentals. These are riskier—the relationship may be spurious or may break.

The best pairs combine both: fundamental logic *and* statistical evidence of mean-reversion.

### 8.4.2 Visual Evidence

```{r pairs-price-comparison, fig.cap="Exxon (XOM) and Chevron (CVX) prices over time. While prices diverge in absolute terms, they tend to move together over the long run."}
# Load pair data - using energy sector pair
xom <- load_market("XOM")
cvx <- load_market("CVX")

# Align dates
xom <- filter_dates(xom, as.Date("2015-01-01"), as.Date("2023-12-31"))
cvx <- filter_dates(cvx, as.Date("2015-01-01"), as.Date("2023-12-31"))

# Merge
pair_data <- merge(
    xom[, .(date, XOM = adjusted)],
    cvx[, .(date, CVX = adjusted)],
    by = "date"
)

# Normalise prices for comparison (start at 100)
pair_data[, XOM_norm := XOM / XOM[1] * 100]
pair_data[, CVX_norm := CVX / CVX[1] * 100]

# Plot
pair_long <- melt(pair_data[, .(date, XOM = XOM_norm, CVX = CVX_norm)],
                  id.vars = "date", variable.name = "Stock", value.name = "Price")

ggplot(pair_long, aes(x = date, y = Price, colour = Stock)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[1], tc[2])) +
    labs(title = "XOM vs CVX (Normalised Prices)",
         subtitle = "Prices start at 100; movement shows relative performance",
         x = NULL, y = "Normalised Price") +
    theme_trading() +
    theme(legend.position = "bottom")
```

```{r spread-time-series, fig.cap="The spread between XOM and CVX (price ratio or difference). A mean-reverting spread oscillates around a stable level."}
# Calculate spread (price ratio)
pair_data[, ratio := XOM / CVX]
pair_data[, ratio_mean := mean(ratio)]
pair_data[, ratio_std := sd(ratio)]
pair_data[, ratio_z := (ratio - ratio_mean) / ratio_std]

# Plot ratio
p1 <- ggplot(pair_data, aes(x = date, y = ratio)) +
    geom_line(colour = tc[1], linewidth = 0.6) +
    geom_hline(yintercept = pair_data$ratio_mean[1], linetype = "dashed", colour = tc[3]) +
    labs(title = "XOM/CVX Price Ratio", x = NULL, y = "Ratio") +
    theme_trading()

# Plot z-score of ratio
p2 <- ggplot(pair_data, aes(x = date, y = ratio_z)) +
    geom_rect(aes(xmin = min(date), xmax = max(date), ymin = 2, ymax = 4),
              fill = tc[4], alpha = 0.1) +
    geom_rect(aes(xmin = min(date), xmax = max(date), ymin = -4, ymax = -2),
              fill = tc[3], alpha = 0.1) +
    geom_line(colour = tc[2], linewidth = 0.6) +
    geom_hline(yintercept = c(-2, 0, 2), linetype = c("dashed", "dotted", "dashed"),
               colour = c(tc[3], "grey50", tc[4])) +
    labs(title = "XOM/CVX Ratio Z-Score",
         subtitle = "Entry zones: |z| > 2",
         x = NULL, y = "Z-Score") +
    theme_trading()

print(p1)
print(p2)
```

### 8.4.3 Mathematical Derivation

**Correlation vs Cointegration:**

Two series can be highly correlated but *not* cointegrated:
- **Correlation:** Measures co-movement of *returns*
- **Cointegration:** Requires the *spread* to be stationary (mean-reverting)

Example: Two random walks with correlated increments are correlated but not cointegrated—their spread is still a random walk.

**Distance Method:**

The distance method ranks pairs by the sum of squared deviations (SSD) of normalised prices:

$$
d_{ij} = \sum_{t=1}^{T} \left(\frac{P_{i,t}}{P_{i,0}} - \frac{P_{j,t}}{P_{j,0}}\right)^2
$$

Lower distance → more similar price paths → better pair candidates.

**Minimum Distance Selection:**

Given $n$ assets, we compute $\binom{n}{2}$ pairwise distances and select pairs with the smallest distances. These have tracked each other historically and may continue to do so.

However, distance alone doesn't guarantee mean reversion—the spread could be a random walk that happened to be small historically.

```{r distance-method, fig.cap="Pairs selection using the distance method. We normalise prices and compute sum of squared deviations. Lower distance indicates more similar price paths."}
# Load multiple stocks for screening
symbols <- c("KO", "MCD", "XOM", "CVX", "JPM", "GS", "AAPL", "MSFT")
all_data <- data.table()

for (sym in symbols) {
    dt <- load_market(sym)
    dt <- filter_dates(dt, as.Date("2020-01-01"), as.Date("2022-12-31"))
    dt[, symbol := sym]
    all_data <- rbind(all_data, dt[, .(date, symbol, adjusted)])
}

# Pivot to wide format
wide_data <- dcast(all_data, date ~ symbol, value.var = "adjusted")
wide_data <- wide_data[complete.cases(wide_data)]

# Normalise prices
for (sym in symbols) {
    wide_data[, (paste0(sym, "_norm")) := get(sym) / get(sym)[1]]
}

# Calculate pairwise distances
calc_distance <- function(data, sym1, sym2) {
    col1 <- paste0(sym1, "_norm")
    col2 <- paste0(sym2, "_norm")
    sum((data[[col1]] - data[[col2]])^2)
}

# All pairs
pairs <- combn(symbols, 2, simplify = FALSE)
distances <- data.table()

for (pair in pairs) {
    d <- calc_distance(wide_data, pair[1], pair[2])
    distances <- rbind(distances, data.table(
        Stock1 = pair[1],
        Stock2 = pair[2],
        Distance = d
    ))
}

distances <- distances[order(Distance)]

cat("=== Top Pairs by Distance Method ===\n\n")
print(head(distances, 10))

# Visualise top pair
top_pair <- distances[1]
cat(sprintf("\nTop pair: %s / %s (distance = %.2f)\n", top_pair$Stock1, top_pair$Stock2, top_pair$Distance))

# Plot top pair
pair_cols <- c(paste0(top_pair$Stock1, "_norm"), paste0(top_pair$Stock2, "_norm"))
plot_dt <- wide_data[, c("date", pair_cols), with = FALSE]
setnames(plot_dt, pair_cols, c("Stock1", "Stock2"))
plot_long <- melt(plot_dt, id.vars = "date", variable.name = "Stock", value.name = "Norm_Price")
plot_long[, Stock := fifelse(Stock == "Stock1", top_pair$Stock1, top_pair$Stock2)]

ggplot(plot_long, aes(x = date, y = Norm_Price, colour = Stock)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[1], tc[2])) +
    labs(title = sprintf("Top Pair by Distance: %s vs %s", top_pair$Stock1, top_pair$Stock2),
         subtitle = sprintf("Distance = %.2f", top_pair$Distance),
         x = NULL, y = "Normalised Price") +
    theme_trading() +
    theme(legend.position = "bottom")
```

### 8.4.4 Implementation & Application

```{r sector-pairs-screening, fig.cap="Sector-based pairs screening. We identify potential pairs within each sector based on distance, then filter by fundamental similarity."}
# Sector-based screening example
# In practice, you'd have sector classifications; here we use industry knowledge

sectors <- list(
    ConsumerStaples = c("KO", "MCD"),
    Oil = c("XOM", "CVX"),
    Banks = c("JPM", "GS"),
    Tech = c("AAPL", "MSFT")
)

cat("=== Sector-Based Pairs ===\n\n")

sector_results <- data.table()
for (sector in names(sectors)) {
    pair <- sectors[[sector]]
    d <- calc_distance(wide_data, pair[1], pair[2])

    # Calculate correlation
    corr <- cor(wide_data[[pair[1]]], wide_data[[pair[2]]])

    sector_results <- rbind(sector_results, data.table(
        Sector = sector,
        Pair = paste(pair, collapse = "/"),
        Distance = round(d, 2),
        Correlation = round(corr, 3)
    ))
}

print(sector_results)
```

---

## 8.5 Cointegration

### 8.5.1 Prose/Intuition

Cointegration is the proper statistical foundation for pairs trading. Two series $X$ and $Y$ are cointegrated if:
1. Both are non-stationary (contain unit roots)
2. Some linear combination $Y - \beta X$ is stationary

The stationary combination is the *spread*. Its stationarity guarantees mean reversion—the statistical basis for pairs trading.

**Engle-Granger Framework:**
1. Regress $Y$ on $X$: $Y_t = \alpha + \beta X_t + \varepsilon_t$
2. Test if residuals $\varepsilon_t$ are stationary (ADF test)
3. If stationary → cointegrated → spread mean-reverts

The coefficient $\beta$ is the *hedge ratio*—how many shares of $X$ to short for each share of $Y$ long.

### 8.5.2 Visual Evidence

```{r cointegration-vs-correlation, fig.cap="Cointegration vs correlation. Two random walks can be correlated (blue/red) without being cointegrated. Cointegrated series (green) have a stationary spread."}
# Simulate cointegrated vs merely correlated pairs
n <- 500

# Correlated but not cointegrated (two random walks with correlated increments)
rho <- 0.8
eps1 <- rnorm(n)
eps2 <- rho * eps1 + sqrt(1 - rho^2) * rnorm(n)

x_corr <- cumsum(eps1)
y_corr <- cumsum(eps2)

# Cointegrated (Y follows X with stationary noise)
x_coint <- cumsum(rnorm(n))
y_coint <- x_coint + rnorm(n, 0, 2)  # Y = X + stationary noise

# Create data.table
sim_dt <- data.table(
    t = 1:n,
    X_correlated = x_corr,
    Y_correlated = y_corr,
    X_cointegrated = x_coint,
    Y_cointegrated = y_coint
)

# Spreads
sim_dt[, spread_correlated := Y_correlated - X_correlated]
sim_dt[, spread_cointegrated := Y_cointegrated - X_cointegrated]

# Plot prices
p1 <- ggplot(sim_dt) +
    geom_line(aes(x = t, y = X_correlated, colour = "X"), linewidth = 0.6) +
    geom_line(aes(x = t, y = Y_correlated, colour = "Y"), linewidth = 0.6) +
    scale_colour_manual(values = c(tc[1], tc[2])) +
    labs(title = "Correlated but NOT Cointegrated", x = "Time", y = "Price", colour = "") +
    theme_trading() +
    theme(legend.position = "bottom")

p2 <- ggplot(sim_dt) +
    geom_line(aes(x = t, y = X_cointegrated, colour = "X"), linewidth = 0.6) +
    geom_line(aes(x = t, y = Y_cointegrated, colour = "Y"), linewidth = 0.6) +
    scale_colour_manual(values = c(tc[1], tc[2])) +
    labs(title = "Cointegrated", x = "Time", y = "Price", colour = "") +
    theme_trading() +
    theme(legend.position = "bottom")

print(p1)
print(p2)

# Plot spreads
p3 <- ggplot(sim_dt) +
    geom_line(aes(x = t, y = spread_correlated), colour = tc[4], linewidth = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title = "Spread (Correlated Pair) - Random Walk", x = "Time", y = "Spread") +
    theme_trading()

p4 <- ggplot(sim_dt) +
    geom_line(aes(x = t, y = spread_cointegrated), colour = tc[3], linewidth = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title = "Spread (Cointegrated Pair) - Mean-Reverting", x = "Time", y = "Spread") +
    theme_trading()

print(p3)
print(p4)
```

### 8.5.3 Mathematical Derivation

**Engle-Granger Two-Step:**

Step 1: Estimate the cointegrating regression:
$$
Y_t = \alpha + \beta X_t + \varepsilon_t
$$

Step 2: Test residuals for stationarity using ADF test:
$$
\Delta \varepsilon_t = \gamma \varepsilon_{t-1} + \sum_{i=1}^{p} \delta_i \Delta \varepsilon_{t-i} + u_t
$$

Test $H_0: \gamma = 0$ (unit root, no cointegration) vs $H_1: \gamma < 0$ (stationary, cointegrated).

**ADF Critical Values:**

Engle-Granger critical values are more stringent than standard ADF because we're testing estimated residuals. At 5% level:
- 1 regressor: -3.37
- 2 regressors: -3.77

**Hedge Ratio Estimation:**

The OLS estimate $\hat{\beta}$ gives the hedge ratio. But OLS is biased when both series are non-stationary.

Alternatives:
- **Total Least Squares (TLS):** Minimises perpendicular distance, not vertical
- **Rolling regression:** Allows hedge ratio to change over time
- **Kalman filter:** Optimal dynamic hedge ratio estimation

**Spread Construction:**

Once we have $\hat{\beta}$:
$$
\text{Spread}_t = Y_t - \hat{\beta} X_t
$$

If cointegrated, this spread is stationary and mean-reverting.

```{r cointegration-test, fig.cap="Cointegration test for XOM/CVX. We regress one on the other and test if residuals are stationary. Significant negative ADF statistic indicates cointegration."}
# Cointegration test implementation
adf_test <- function(x, max_lag = 1) {
    n <- length(x)
    dx <- diff(x)
    x_lag <- x[-n]

    # ADF regression: dx = gamma * x_lag + delta * dx_lag + error
    if (max_lag > 0) {
        dx_lags <- embed(dx, max_lag + 1)[, -1, drop = FALSE]
        x_lag <- x_lag[-(1:max_lag)]
        dx <- dx[-(1:max_lag)]

        # Combine
        reg_data <- data.table(dx = dx, x_lag = x_lag)
        for (i in 1:max_lag) {
            reg_data[, paste0("dx_lag", i) := dx_lags[, i]]
        }

        formula <- as.formula(paste("dx ~ x_lag +", paste(paste0("dx_lag", 1:max_lag), collapse = " + ")))
        fit <- lm(formula, data = reg_data)
    } else {
        fit <- lm(dx ~ x_lag)
    }

    gamma <- coef(fit)["x_lag"]
    se_gamma <- summary(fit)$coefficients["x_lag", "Std. Error"]
    t_stat <- gamma / se_gamma

    list(gamma = gamma, t_stat = t_stat, se = se_gamma)
}

# Test XOM/CVX
# Step 1: Cointegrating regression
pair_data[, `:=`(
    xom_log = log(XOM),
    cvx_log = log(CVX)
)]

fit_coint <- lm(xom_log ~ cvx_log, data = pair_data)
beta <- coef(fit_coint)["cvx_log"]
alpha <- coef(fit_coint)["(Intercept)"]

pair_data[, residuals := xom_log - alpha - beta * cvx_log]

# Step 2: ADF test on residuals
adf_result <- adf_test(pair_data$residuals, max_lag = 1)

cat("=== Engle-Granger Cointegration Test: XOM/CVX ===\n\n")
cat("Step 1: Cointegrating Regression\n")
cat(sprintf("  log(XOM) = %.4f + %.4f * log(CVX)\n", alpha, beta))
cat(sprintf("  Hedge ratio (β): %.4f\n\n", beta))

cat("Step 2: ADF Test on Residuals\n")
cat(sprintf("  ADF t-statistic: %.3f\n", adf_result$t_stat))
cat(sprintf("  5%% Critical value: -3.37 (approx)\n"))
cat(sprintf("  Cointegrated: %s\n", ifelse(adf_result$t_stat < -3.37, "YES", "NO")))

# Plot residuals (spread)
ggplot(pair_data, aes(x = date, y = residuals)) +
    geom_line(colour = tc[1], linewidth = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title = "Cointegrating Residuals (Spread)",
         subtitle = sprintf("ADF stat = %.2f (5%% critical: -3.37)", adf_result$t_stat),
         x = NULL, y = "Residual (log spread)") +
    theme_trading()
```

### 8.5.4 Implementation & Application

```{r hedge-ratio-estimation, fig.cap="Hedge ratio estimation and comparison. OLS gives a point estimate; rolling regression shows how the relationship evolves over time."}
# Compare OLS vs rolling hedge ratio
pair_data[, hedge_ratio_ols := beta]

# Rolling regression (60-day window)
window <- 60
rolling_beta <- numeric(nrow(pair_data))
rolling_beta[1:(window-1)] <- NA

for (i in window:nrow(pair_data)) {
    fit <- lm(xom_log ~ cvx_log, data = pair_data[(i - window + 1):i])
    rolling_beta[i] <- coef(fit)["cvx_log"]
}

pair_data[, hedge_ratio_rolling := rolling_beta]

# Plot
ggplot(pair_data[!is.na(hedge_ratio_rolling)], aes(x = date)) +
    geom_line(aes(y = hedge_ratio_rolling), colour = tc[1], linewidth = 0.7) +
    geom_hline(yintercept = beta, colour = tc[2], linetype = "dashed") +
    labs(title = "Hedge Ratio: OLS vs Rolling (60-day)",
         subtitle = sprintf("OLS β = %.3f (dashed line)", beta),
         x = NULL, y = "Hedge Ratio (β)") +
    theme_trading()
```

---

## 8.6 Trading the Spread

### 8.6.1 Prose/Intuition

With a cointegrated pair and hedge ratio, we trade the spread:

**Entry:**
- When spread is too high (z > 2): Short Y, long X (spread will fall)
- When spread is too low (z < -2): Long Y, short X (spread will rise)

**Exit:**
- When spread returns to mean (z ≈ 0): Close both legs

**Position sizing:**
- Long Y shares, short $\beta \times$ X shares
- This creates a market-neutral position

The strategy profits from *convergence*, not market direction. If both stocks rise or fall together, profits/losses on the long leg offset those on the short leg.

### 8.6.2 Visual Evidence

```{r spread-trading-signals, fig.cap="Pairs trading signals on the XOM/CVX spread. Green triangles indicate long spread (long XOM, short CVX); red triangles indicate short spread."}
# Spread z-score and signals
pair_data[, spread := xom_log - beta * cvx_log]
pair_data[, spread_mean := frollmean(spread, 60)]
pair_data[, spread_std := frollapply(spread, 60, sd)]
pair_data[, spread_z := (spread - spread_mean) / spread_std]

pair_data <- pair_data[!is.na(spread_z)]

# Signals
entry_threshold <- 2
exit_threshold <- 0

pair_data[, long_spread := spread_z < -entry_threshold]
pair_data[, short_spread := spread_z > entry_threshold]

# Plot
ggplot(pair_data, aes(x = date, y = spread_z)) +
    geom_rect(aes(xmin = min(date), xmax = max(date), ymin = 2, ymax = 4),
              fill = tc[4], alpha = 0.1) +
    geom_rect(aes(xmin = min(date), xmax = max(date), ymin = -4, ymax = -2),
              fill = tc[3], alpha = 0.1) +
    geom_line(colour = tc[1], linewidth = 0.6) +
    geom_hline(yintercept = c(-2, 0, 2), linetype = c("dashed", "dotted", "dashed"),
               colour = c(tc[3], "grey50", tc[4])) +
    geom_point(data = pair_data[long_spread == TRUE], aes(y = spread_z),
               colour = tc[3], size = 2, shape = 24, fill = tc[3]) +
    geom_point(data = pair_data[short_spread == TRUE], aes(y = spread_z),
               colour = tc[4], size = 2, shape = 25, fill = tc[4]) +
    labs(title = "XOM/CVX Spread Z-Score with Trading Signals",
         subtitle = "Green = long spread (long XOM, short CVX); Red = short spread",
         x = NULL, y = "Z-Score") +
    theme_trading()
```

### 8.6.3 Mathematical Derivation

**Spread Definition:**

$$
\text{Spread}_t = \log(Y_t) - \hat{\beta} \log(X_t)
$$

(Using log prices for return calculations.)

**Z-Score:**

$$
z_t = \frac{\text{Spread}_t - \bar{\mu}_t}{\hat{\sigma}_t}
$$

where $\bar{\mu}_t$ and $\hat{\sigma}_t$ are rolling mean and standard deviation.

**Optimal Entry Threshold:**

From OU model with parameters $(\theta, \mu, \sigma)$, the optimal entry maximises expected profit minus costs:

$$
z^* = \arg\max_z \left[ |z| \cdot \hat{\sigma} - 2c - \frac{|z| \cdot \hat{\sigma}}{\theta \cdot t_{hold}} \right]
$$

In practice, z = 2 is commonly used (corresponds to ~2.5% probability under normality).

**Expected Holding Time:**

$$
E[T] \approx \frac{|z_{\text{entry}}|}{\theta}
$$

With $\theta = 0.05$ and $z = 2$: $E[T] \approx 40$ days.

**Expected Profit:**

Per trade profit (before costs):
$$
E[\text{Profit}] = |z_{\text{entry}}| \cdot \hat{\sigma} \cdot \text{Position Size}
$$

With z = 2 and σ = 2% spread volatility: 4% expected profit per trade.

### 8.6.4 Implementation & Application

```{r pairs-trading-backtest, fig.cap="Full pairs trading backtest on XOM/CVX. The strategy profits from spread convergence while remaining market-neutral."}
# Full pairs trading implementation
pair_data[, xom_ret := c(NA, diff(xom_log))]
pair_data[, cvx_ret := c(NA, diff(cvx_log))]
pair_data <- pair_data[!is.na(xom_ret)]

# Trading logic
position <- 0  # 1 = long spread (long XOM, short CVX), -1 = short spread
positions <- numeric(nrow(pair_data))

for (i in 2:nrow(pair_data)) {
    z <- pair_data$spread_z[i - 1]

    # Entry
    if (position == 0) {
        if (z < -entry_threshold) position <- 1    # Long spread
        else if (z > entry_threshold) position <- -1  # Short spread
    }
    # Exit
    else if (position == 1 && z > exit_threshold) {
        position <- 0
    }
    else if (position == -1 && z < -exit_threshold) {
        position <- 0
    }

    positions[i] <- position
}

pair_data[, position := positions]

# Strategy returns (market-neutral: long XOM, short β*CVX)
# Spread return = XOM return - β * CVX return
pair_data[, spread_ret := xom_ret - beta * cvx_ret]
pair_data[, strat_ret := position * spread_ret]

# Transaction costs (10 bps per leg, so 20 bps total)
cost <- 0.002
pair_data[, trade := position != shift(position, 1)]
pair_data[is.na(trade), trade := FALSE]
pair_data[, strat_ret_cost := strat_ret - trade * cost]

# Cumulative returns
pair_data[, cum_xom := exp(cumsum(xom_ret))]
pair_data[, cum_strat := exp(cumsum(strat_ret))]
pair_data[, cum_strat_cost := exp(cumsum(strat_ret_cost))]

# Statistics
cat("=== XOM/CVX Pairs Trading Backtest ===\n")
cat(sprintf("Entry threshold: |z| > %.1f\n", entry_threshold))
cat(sprintf("Exit threshold: z = %.1f\n", exit_threshold))
cat(sprintf("Hedge ratio (β): %.3f\n\n", beta))

# Count trades
n_trades <- sum(pair_data$trade, na.rm = TRUE)

pairs_stats <- data.table(
    Strategy = c("XOM Buy & Hold", "Pairs (no cost)", "Pairs (20bp cost)"),
    `Ann. Return` = sprintf("%.1f%%", c(
        annualised_return(pair_data$xom_ret, type = "simple") * 100,
        annualised_return(pair_data$strat_ret, type = "simple") * 100,
        annualised_return(pair_data$strat_ret_cost, type = "simple") * 100)),
    `Ann. Vol` = sprintf("%.1f%%", c(
        sd(pair_data$xom_ret, na.rm = TRUE) * sqrt(252) * 100,
        sd(pair_data$strat_ret, na.rm = TRUE) * sqrt(252) * 100,
        sd(pair_data$strat_ret_cost, na.rm = TRUE) * sqrt(252) * 100)),
    Sharpe = sprintf("%.2f", c(
        sharpe_ratio(pair_data$xom_ret),
        sharpe_ratio(pair_data$strat_ret),
        sharpe_ratio(pair_data$strat_ret_cost))),
    `Max Drawdown` = sprintf("%.1f%%", c(
        max_drawdown(pair_data$xom_ret) * 100,
        max_drawdown(pair_data$strat_ret) * 100,
        max_drawdown(pair_data$strat_ret_cost) * 100))
)
print(pairs_stats)
cat(sprintf("\nTotal trades: %d (%.1f per year)\n", n_trades, n_trades / (nrow(pair_data) / 252)))

# Plot
pairs_long <- melt(pair_data[, .(date, `XOM Buy & Hold` = cum_xom,
                                  `Pairs (no cost)` = cum_strat,
                                  `Pairs (20bp cost)` = cum_strat_cost)],
                   id.vars = "date", variable.name = "Strategy", value.name = "Growth")

ggplot(pairs_long, aes(x = date, y = Growth, colour = Strategy)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = c(tc[1], tc[2], tc[4])) +
    labs(title = "XOM/CVX Pairs Trading Strategy",
         subtitle = "Market-neutral strategy profits from spread convergence",
         x = NULL, y = "Growth of $1") +
    theme_trading() +
    theme(legend.position = "bottom")
```

```{r rolling-cointegration, fig.cap="Rolling cointegration analysis. The relationship between pairs can weaken over time; monitoring is essential."}
# Rolling cointegration test
window <- 252  # 1 year

rolling_adf <- numeric(nrow(pair_data))
rolling_adf[1:(window-1)] <- NA

for (i in window:nrow(pair_data)) {
    subset <- pair_data[(i - window + 1):i]

    # Cointegrating regression on subset
    fit <- lm(xom_log ~ cvx_log, data = subset)
    resids <- residuals(fit)

    # ADF test
    adf <- adf_test(resids, max_lag = 1)
    rolling_adf[i] <- adf$t_stat
}

pair_data[, rolling_adf := rolling_adf]

# Plot
ggplot(pair_data[!is.na(rolling_adf)], aes(x = date, y = rolling_adf)) +
    geom_line(colour = tc[1], linewidth = 0.6) +
    geom_hline(yintercept = -3.37, linetype = "dashed", colour = tc[3]) +
    geom_hline(yintercept = -2.5, linetype = "dotted", colour = tc[4]) +
    labs(title = "Rolling Cointegration Test (1-Year Window)",
         subtitle = "Below green line: cointegrated at 5%; above red dotted: relationship weakening",
         x = NULL, y = "ADF Statistic") +
    theme_trading()
```

---

## Quick Reference

### Pairs Selection

| Method | Formula | Notes |
|--------|---------|-------|
| Distance | $\sum(P_i^{norm} - P_j^{norm})^2$ | Lower = more similar |
| Correlation | $\rho(r_i, r_j)$ | Return correlation |
| Cointegration | ADF test on spread | Proper statistical test |

### Cointegration Testing

| Step | Procedure | Criterion |
|------|-----------|-----------|
| 1 | Regress $Y$ on $X$ | Get $\hat{\beta}$ |
| 2 | Compute residuals | $\varepsilon = Y - \hat{\beta}X$ |
| 3 | ADF test on residuals | t-stat < -3.37 (5% level) |

### Trading Rules

| Signal | Condition | Action |
|--------|-----------|--------|
| Long Spread | $z < -2$ | Long Y, Short $\beta$X |
| Short Spread | $z > 2$ | Short Y, Long $\beta$X |
| Exit | $z \approx 0$ | Close both legs |

### Key Insights

1. **Cointegration ≠ correlation:** High correlation doesn't imply mean-reverting spread
2. **Hedge ratio matters:** Wrong $\beta$ creates market exposure
3. **Relationships drift:** Monitor cointegration over time
4. **Transaction costs hurt:** Pairs trading is capital-intensive
5. **Fundamental logic helps:** Statistically-selected pairs without economic rationale are risky
