---
title: "Statistics with R"
chapter: "Preface"
coverImage: 13
author: "Dereck Mezquita"
date: 2023-10-20
tags: [statistics, mathematics, probability, data]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
# https://bookdown.org/yihui/rmarkdown-cookbook/hook-html5.html
if (knitr::is_html_output()) knitr::knit_hooks$set(
    plot = function(x, options) {
        cap  <- options$fig.cap
        # x <- paste0("/courses/", x)
        as.character(htmltools::tag(
            "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        ))
    }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng) # optipng = '-o7'
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

# Preface

Welcome to "Statistics with R: A Comprehensive Guide for the R Practitioner"—a definitive reference designed for those who already possess proficiency in R programming and seek to master statistical methodology through a computational lens. This book stands apart from typical statistics textbooks that focus primarily on theory or introductory R tutorials that merely scratch the surface of statistical applications.

Our mission is clear: to bridge the chasm between theoretical statistics and practical implementation, providing a resource that working statisticians, data scientists, bioinformaticians, financial analysts, and researchers can reference daily. Every statistical concept presented herein is demonstrated through efficient R code, leveraging the power of modern packages like data.table for high-performance data manipulation and ggplot2 for publication-quality visualisations.

## Why This Book Is Different

* **Assumes R Proficiency**: Unlike many statistics texts that teach basic R alongside statistics, we presume you're already comfortable with R programming. This allows us to focus exclusively on statistical methodology and its implementation.

* **Practical, Code-First Approach**: Each statistical concept is explained through working code examples using real-world datasets.

* **Focus on Modern Applications**: The examples draw from contemporary domains such as:
  * Bioinformatics and genomics
  * Financial markets and risk assessment
  * Environmental modelling
  * Social network analysis
  * Machine learning applications

* **Performance-Oriented**: We emphasize computational efficiency with data.table and vectorised operations, essential for working with large datasets common in modern statistical practice.

* **Comprehensive Coverage**: From foundational probability theory to cutting-edge topics like Bayesian computation, causal inference, and deep learning applications in statistics.

## How to Use This Book

This is not merely a book to be read sequentially; it is designed as a comprehensive reference that can be consulted at any stage of statistical analysis:

* Need to refresh your understanding of mixture models? Chapter 9 covers Bayesian methods in detail.
* Implementing a time series approach for financial data? Chapter 8 provides practical examples with ARIMA and GARCH models.
* Selecting the appropriate test for complex experimental designs? Chapter 5 guides you through the decision process with concrete implementations.

In every case, you will find rigorous explanations accompanied by efficient, production-ready R code.

## The Statistical Landscape

The statistical landscape continues to evolve rapidly, and this book embraces both classical foundations and cutting-edge methodologies. Our focus remains steadfastly on statistical thinking—understanding when and why particular methods are appropriate—while providing the computational tools to implement these methods with confidence and precision.

We believe that modern statistical practice is inherently computational. By combining strong theoretical understanding with practical implementation skills, statisticians can extract deeper insights from data and communicate findings more effectively.

Let's begin the journey of mastering statistics through the powerful lens of R.

```{r preface_example, echo=TRUE}
box::use(
    data.table[data.table, .N, .SD],
    ggplot2
)

# Creating a demonstration dataset
set.seed(123)
n <- 1000
dt <- data.table(
    x = rnorm(n),
    group = sample(LETTERS[1:5], n, replace = TRUE)
)

# Calculate summary statistics by group
summary_stats <- dt[, .(
    count = .N,
    mean = mean(x),
    median = median(x),
    sd = sd(x),
    q25 = quantile(x, 0.25),
    q75 = quantile(x, 0.75)
), by = group]

# Print the summary statistics
summary_stats

# Visualise the distribution by group
ggplot2$ggplot(dt, ggplot2$aes(x = x, fill = group)) +
    ggplot2$geom_density(alpha = 0.5) +
    ggplot2$labs(
        title = "Density Plot by Group",
        x = "Value",
        y = "Density"
    ) +
    ggplot2$theme_minimal()
```

This example illustrates our approach: clean, efficient code that combines statistical analysis with clear visualisation, all using modern R practices. Throughout this book, you'll find similar examples that build on these foundations to tackle increasingly complex statistical challenges.