---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 18: Analysis of Variance"
part: "Part 1: One-Way ANOVA and Comparisons"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, anova, regression, f-test, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

```{r packages, include=FALSE}
box::use(ggplot2)
box::use(data.table)
```

```{r dt-symbols, include=FALSE}
.N <- data.table::.N
.SD <- data.table::.SD
```

# Chapter 18 -- Analysis of Variance

## Part 1: One-Way ANOVA and Comparisons

Analysis of Variance (ANOVA) is among the most widely used procedures in experimental science. It is often taught as a standalone technique, but the deepest understanding comes from recognising that **ANOVA is regression with categorical predictors**. The $F$-test in ANOVA is exactly the $F$-test for comparing a full model against a reduced (intercept-only) model. In this part we derive the one-way ANOVA $F$-test as a nested model comparison, show that `aov()` and `anova(lm())` give identical results, and connect multiple comparisons to Chapter 12.

---

## 18.1 The One-Way Layout

### 18.1.1 Setup

We observe $n_{j}$ responses in each of $k$ groups (levels of a factor), for a total of $N = \sum_{j=1}^k n_j$ observations. The model is:

$$Y_{ij} = \mu_j + \varepsilon_{ij}, \quad i = 1, \ldots, n_j, \quad j = 1, \ldots, k$$

where $\varepsilon_{ij} \stackrel{\text{iid}}{\sim} N(0, \sigma^2)$.

Equivalently, using the "effects" parametrisation with overall mean $\mu$ and group effects $\alpha_j$:

$$Y_{ij} = \mu + \alpha_j + \varepsilon_{ij}, \quad \sum_{j=1}^k n_j \alpha_j = 0$$

### 18.1.2 The Hypothesis

$$H_0: \mu_1 = \mu_2 = \cdots = \mu_k \quad \text{(all group means are equal)}$$
$$H_1: \text{at least two means differ}$$

Under the effects parametrisation, $H_0: \alpha_1 = \alpha_2 = \cdots = \alpha_k = 0$.

---

## 18.2 ANOVA as Regression

### 18.2.1 Dummy Coding

A categorical variable with $k$ levels is encoded using $k - 1$ dummy (indicator) variables. If group $j = 1$ is the reference category, define:

$$D_{ij}^{(m)} = \begin{cases} 1 & \text{if observation } i \text{ belongs to group } m \\ 0 & \text{otherwise} \end{cases}, \quad m = 2, \ldots, k$$

The regression model becomes:

$$Y_i = \beta_0 + \beta_2 D_i^{(2)} + \beta_3 D_i^{(3)} + \cdots + \beta_k D_i^{(k)} + \varepsilon_i$$

where $\beta_0 = \mu_1$ (the reference group mean) and $\beta_m = \mu_m - \mu_1$ (the difference between group $m$ and the reference).

### 18.2.2 The Full and Reduced Models

- **Full model** (one mean per group): $Y_i = \beta_0 + \beta_2 D_i^{(2)} + \cdots + \beta_k D_i^{(k)} + \varepsilon_i$ -- this has $p = k$ parameters.
- **Reduced model** (one common mean): $Y_i = \beta_0 + \varepsilon_i$ -- this has $p_0 = 1$ parameter.

Testing $H_0: \beta_2 = \cdots = \beta_k = 0$ is equivalent to comparing these two nested models.

### 18.2.3 Deriving the $F$-test

The general $F$-statistic for comparing nested models is:

$$F = \frac{(\text{SSE}_{\text{reduced}} - \text{SSE}_{\text{full}}) / (p - p_0)}{\text{SSE}_{\text{full}} / (N - p)}$$

For the ANOVA problem:

- $\text{SSE}_{\text{reduced}} = \text{SST} = \sum_{j=1}^k \sum_{i=1}^{n_j}(Y_{ij} - \bar{Y}_{\cdot\cdot})^2$ (total SS).
- $\text{SSE}_{\text{full}} = \text{SSW} = \sum_{j=1}^k \sum_{i=1}^{n_j}(Y_{ij} - \bar{Y}_{\cdot j})^2$ (within-group SS).
- $\text{SST} - \text{SSW} = \text{SSB} = \sum_{j=1}^k n_j(\bar{Y}_{\cdot j} - \bar{Y}_{\cdot\cdot})^2$ (between-group SS).
- $p - p_0 = k - 1$, $N - p = N - k$.

Therefore:

$$\boxed{F = \frac{\text{SSB}/(k-1)}{\text{SSW}/(N-k)} = \frac{\text{MSB}}{\text{MSW}} \sim F_{k-1, N-k} \quad \text{under } H_0}$$

This is the one-way ANOVA $F$-test. Under $H_0$, the ratio of mean squares follows an $F$-distribution because both the numerator and denominator are independent $\chi^2$ random variables (divided by their degrees of freedom), by the properties of normal quadratic forms.

---

## 18.3 Building the ANOVA Table from Scratch

```{r anova-scratch}
data_dir <- "../../../data"
penguins_raw <- data.table$fread(file.path(data_dir, "primary/penguins.csv"))
penguins <- penguins_raw[
  complete.cases(penguins_raw[, .(flipper_length_mm, species)])
]

# One-way ANOVA: flipper length by species
y <- penguins$flipper_length_mm
g <- penguins$species

groups <- unique(g)
k <- length(groups)
N <- length(y)
grand_mean <- mean(y)

# Group means and sizes
group_stats <- penguins[, .(
  mean = mean(flipper_length_mm),
  n = .N
), by = species]

# Sums of squares
SST <- sum((y - grand_mean)^2)
SSB <- sum(group_stats$n * (group_stats$mean - grand_mean)^2)
SSW <- SST - SSB

# Mean squares
df_between <- k - 1
df_within <- N - k
MSB <- SSB / df_between
MSW <- SSW / df_within

# F statistic and p-value
F_stat <- MSB / MSW
p_value <- pf(F_stat, df_between, df_within, lower.tail = FALSE)

anova_table <- data.table$data.table(
  Source = c("Between (species)", "Within (residual)", "Total"),
  df = c(df_between, df_within, N - 1),
  SS = c(SSB, SSW, SST),
  MS = c(MSB, MSW, NA),
  `F` = c(F_stat, NA, NA),
  `p-value` = c(p_value, NA, NA)
)

anova_table
```

---

## 18.4 `aov()` = `anova(lm())` -- The Equivalence

### 18.4.1 Using aov()

```{r aov-fit}
fit_aov <- aov(flipper_length_mm ~ species, data = penguins)
summary(fit_aov)
```

### 18.4.2 Using lm() + anova()

```{r lm-anova}
fit_lm <- lm(flipper_length_mm ~ species, data = penguins)
anova(fit_lm)
```

### 18.4.3 Verifying Equivalence

```{r equivalence-check}
# The F-statistics should be identical
c(
  "F from scratch" = F_stat,
  "F from aov" = summary(fit_aov)[[1]]["species", "F value"],
  "F from anova(lm)" = anova(fit_lm)["species", "F value"]
)
```

All three approaches produce the same $F$-statistic. The `aov()` function is simply a wrapper around `lm()` that formats the output as an ANOVA table. Behind the scenes, it fits a linear model with dummy variables.

### 18.4.4 What the Dummy Coefficients Mean

```{r lm-coefs}
summary(fit_lm)
```

R uses treatment (dummy) coding by default with the first level (`Adelie`) as the reference. The intercept $\hat{\beta}_0$ is the mean flipper length for Adelie penguins; `speciesChinstrap` is the difference between Chinstrap and Adelie means; `speciesGentoo` is the difference between Gentoo and Adelie means.

```{r verify-means}
group_stats[, .(
  species,
  group_mean = mean,
  from_coefs = c(
    coef(fit_lm)[1],
    coef(fit_lm)[1] + coef(fit_lm)[2],
    coef(fit_lm)[1] + coef(fit_lm)[3]
  )
)]
```

---

## 18.5 Visualising the Data

```{r penguins-boxplot, fig.cap="Flipper length by species in the Palmer Penguins dataset. Gentoo penguins have markedly longer flippers than Adelie and Chinstrap penguins. The ANOVA F-test confirms that the species means are not all equal."}
ggplot2$ggplot(penguins, ggplot2$aes(x = species, y = flipper_length_mm, fill = species)) +
  ggplot2$geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.alpha = 0.5) +
  ggplot2$geom_jitter(width = 0.15, alpha = 0.2, size = 1) +
  ggplot2$scale_fill_manual(values = c(
    "Adelie" = "#E69F00", "Chinstrap" = "#0072B2", "Gentoo" = "#009E73"
  )) +
  ggplot2$labs(
    x = "Species",
    y = "Flipper length (mm)",
    title = sprintf("Flipper length by species (F = %.1f, p < 0.001)", F_stat)
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "none")
```

```{r penguins-means, fig.cap="Group means with 95 per cent confidence intervals. The intervals for Adelie and Chinstrap overlap slightly, but both are well separated from Gentoo."}
group_stats[, `:=`(
  se = penguins[, .(se = sd(flipper_length_mm) / sqrt(.N)), by = species]$se,
  lower = mean - qt(0.975, n - 1) * penguins[, .(se = sd(flipper_length_mm) / sqrt(.N)), by = species]$se,
  upper = mean + qt(0.975, n - 1) * penguins[, .(se = sd(flipper_length_mm) / sqrt(.N)), by = species]$se
)]

ggplot2$ggplot(group_stats, ggplot2$aes(x = species, y = mean, colour = species)) +
  ggplot2$geom_pointrange(
    ggplot2$aes(ymin = lower, ymax = upper),
    size = 1, linewidth = 0.8
  ) +
  ggplot2$scale_colour_manual(values = c(
    "Adelie" = "#E69F00", "Chinstrap" = "#0072B2", "Gentoo" = "#009E73"
  )) +
  ggplot2$labs(
    x = "Species",
    y = "Flipper length (mm)",
    title = "Group means with 95% CIs"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "none")
```

---

## 18.6 Checking ANOVA Assumptions

ANOVA inherits the same assumptions as regression: independence, normality within each group, and equal variances (homoscedasticity) across groups.

### 18.6.1 Normality within Groups

```{r anova-qq, fig.cap="Q-Q plots of residuals by species. All three groups show approximately normal residuals."}
penguins[, resid := residuals(fit_lm)]

ggplot2$ggplot(penguins, ggplot2$aes(sample = resid)) +
  ggplot2$geom_qq(colour = "#0072B2", alpha = 0.4, size = 1.2) +
  ggplot2$geom_qq_line(colour = "#D55E00", linewidth = 0.8) +
  ggplot2$facet_wrap(~ species) +
  ggplot2$labs(
    x = "Theoretical quantiles",
    y = "Residuals",
    title = "Q-Q plots by species"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

### 18.6.2 Homogeneity of Variance -- Levene's Test from Scratch

Levene's test applies ANOVA to the **absolute deviations** from group medians:

```{r levene-scratch}
# Levene's test from scratch
penguins[, group_median := median(flipper_length_mm), by = species]
penguins[, abs_dev := abs(flipper_length_mm - group_median)]

fit_levene <- lm(abs_dev ~ species, data = penguins)
levene_anova <- anova(fit_levene)

c(
  "Levene F" = levene_anova["species", "F value"],
  "p-value" = levene_anova["species", "Pr(>F)"]
)
```

A non-significant result suggests the equal variance assumption is reasonable.

---

## 18.7 The PlantGrowth Example

The PlantGrowth dataset contains yields from 30 plants under three conditions: control, treatment 1, and treatment 2.

```{r plantgrowth-data}
# PlantGrowth is a built-in R dataset
pg <- data.table$as.data.table(PlantGrowth)
pg[, .(.N, mean = mean(weight), sd = sd(weight)), by = group]
```

```{r plantgrowth-plot, fig.cap="Dried weight of plants under three conditions. Treatment 2 appears to produce heavier plants than the control and treatment 1."}
ggplot2$ggplot(pg, ggplot2$aes(x = group, y = Weight, fill = group)) +
  ggplot2$geom_boxplot(alpha = 0.7) +
  ggplot2$geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  ggplot2$scale_fill_manual(values = c(
    "ctrl" = "#999999", "trt1" = "#E69F00", "trt2" = "#0072B2"
  )) +
  ggplot2$labs(
    x = "Group",
    y = "Dried weight",
    title = "PlantGrowth: dried weight by treatment group"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "none")
```

```{r plantgrowth-anova}
fit_pg <- lm(weight ~ group, data = pg)
anova(fit_pg)
```

The $F$-test is marginally significant ($p \approx 0.016$). But which pairs of groups differ? The $F$-test tells us only that *at least* two means differ -- it does not say which. This is where multiple comparisons come in.

---

## 18.8 Multiple Comparisons -- Connecting to Chapter 12

### 18.8.1 The Problem

With $k = 3$ groups, there are $\binom{3}{2} = 3$ pairwise comparisons. If we perform each at $\alpha = 0.05$, the probability of at least one false rejection under $H_0$ is:

$$1 - (1 - 0.05)^3 = 0.143$$

With $k = 10$ groups, there are 45 comparisons and the family-wise error rate exceeds 0.9. This is exactly the multiple testing problem from Chapter 12.

### 18.8.2 Bonferroni Correction

The simplest approach: test each pair at level $\alpha / m$ where $m = \binom{k}{2}$. This controls the family-wise error rate at $\alpha$ but is conservative (the adjusted $p$-values are larger than necessary).

```{r bonferroni-pg}
# All pairwise t-tests with Bonferroni correction
pairwise.t.test(pg$Weight, pg$group, p.adjust.method = "bonferroni")
```

### 18.8.3 Tukey's Honest Significant Difference (HSD)

Tukey's HSD uses the **studentised range distribution** to account for all pairwise comparisons simultaneously. For balanced designs ($n_j = n$ for all $j$), the critical value is:

$$q_{\alpha, k, N-k} \cdot \frac{s_W}{\sqrt{n}}$$

where $q_{\alpha, k, N-k}$ is the upper $\alpha$ quantile of the studentised range distribution with $k$ groups and $N - k$ degrees of freedom, and $s_W = \sqrt{\text{MSW}}$.

Two means $\bar{Y}_{\cdot j}$ and $\bar{Y}_{\cdot m}$ are significantly different at level $\alpha$ if:

$$|\bar{Y}_{\cdot j} - \bar{Y}_{\cdot m}| > q_{\alpha, k, N-k} \cdot \sqrt{\frac{\text{MSW}}{n}}$$

### 18.8.4 Tukey HSD from Scratch

```{r tukey-scratch}
# Tukey HSD from scratch for PlantGrowth (balanced)
group_means <- pg[, .(mean = mean(weight), n = .N), by = group]
MSW_pg <- anova(fit_pg)["Residuals", "Mean Sq"]
df_w <- anova(fit_pg)["Residuals", "Df"]
n_per <- group_means$n[1]  # balanced design

q_crit <- qtukey(0.95, nmeans = 3, df = df_w)
hsd <- q_crit * sqrt(MSW_pg / n_per)

# All pairwise differences
pairs <- combn(group_means$group, 2, simplify = FALSE)
tukey_results <- data.table$rbindlist(lapply(pairs, function(p) {
  m1 <- group_means[group == p[1], mean]
  m2 <- group_means[group == p[2], mean]
  diff <- m2 - m1
  data.table$data.table(
    comparison = paste(p[2], "-", p[1]),
    diff = diff,
    lower = diff - hsd,
    upper = diff + hsd,
    significant = abs(diff) > hsd
  )
}))

tukey_results
```

### 18.8.5 Comparing with TukeyHSD()

```{r tukey-builtin}
TukeyHSD(aov(weight ~ group, data = pg))
```

### 18.8.6 Visualising Tukey HSD

```{r tukey-plot, fig.cap="Tukey HSD 95 per cent family-wise confidence intervals. Treatment 2 is significantly different from treatment 1 (the interval does not contain zero), but neither treatment is significantly different from the control."}
tukey_obj <- TukeyHSD(aov(weight ~ group, data = pg))
tukey_dt <- data.table$as.data.table(tukey_obj$group, keep.rownames = "comparison")

ggplot2$ggplot(tukey_dt, ggplot2$aes(x = diff, y = comparison)) +
  ggplot2$geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2$geom_pointrange(
    ggplot2$aes(xmin = lwr, xmax = upr),
    colour = "#0072B2", size = 0.8, linewidth = 0.7
  ) +
  ggplot2$labs(
    x = "Difference in means",
    y = "Comparison",
    title = "Tukey HSD: 95% family-wise confidence intervals"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

Treatment 2 vs treatment 1 is the only significant pairwise difference. The comparison between each treatment and the control is not significant at the family-wise $\alpha = 0.05$ level.

---

## 18.9 The Pooled Variance Assumption

### 18.9.1 Why It Matters

The $F$-test assumes a common within-group variance $\sigma^2$. When variances are very unequal, the $F$-test can have inflated Type I error rates. Welch's ANOVA provides a robust alternative:

```{r welch-anova}
# Standard ANOVA
anova(lm(flipper_length_mm ~ species, data = penguins))

# Welch's ANOVA (does not assume equal variances)
oneway.test(flipper_length_mm ~ species, data = penguins, var.equal = FALSE)
```

For the penguins data, the conclusions are essentially identical because the within-group variances are similar. When they are not, Welch's ANOVA is preferred.

---

## 18.10 Effect Size: Eta-Squared and Omega-Squared

### 18.10.1 Eta-Squared

$$\eta^2 = \frac{\text{SSB}}{\text{SST}}$$

This is the $R^2$ of the ANOVA model -- the proportion of variance explained by the grouping factor. It is upward-biased.

### 18.10.2 Omega-Squared

A less biased alternative:

$$\omega^2 = \frac{\text{SSB} - (k-1)\text{MSW}}{\text{SST} + \text{MSW}}$$

```{r effect-sizes}
SSB_pen <- anova(fit_lm)["species", "Sum Sq"]
SSW_pen <- anova(fit_lm)["Residuals", "Sum Sq"]
SST_pen <- SSB_pen + SSW_pen
MSW_pen <- anova(fit_lm)["Residuals", "Mean Sq"]
k_pen <- 3

eta2 <- SSB_pen / SST_pen
omega2 <- (SSB_pen - (k_pen - 1) * MSW_pen) / (SST_pen + MSW_pen)

c(
  "eta-squared" = eta2,
  "omega-squared" = omega2,
  "R-squared from lm" = summary(fit_lm)$r.squared
)
```

$\eta^2 \approx R^2 \approx 0.78$: species explains about 78% of the variance in flipper length. This is a very large effect.

---

## 18.11 Simulated ANOVA: From Null to Alternative

```{r sim-anova, fig.cap="Simulated ANOVA under the null (left: all group means equal) and the alternative (right: one group mean differs). Under the null, the F-statistic follows an F-distribution; under the alternative, it is shifted to the right."}
set.seed(42)
n_sim <- 1000

# Under H0: all means = 5
f_null <- replicate(n_sim, {
  y_sim <- rnorm(90, mean = 5, sd = 2)
  g_sim <- rep(c("A", "B", "C"), each = 30)
  anova(lm(y_sim ~ g_sim))["g_sim", "F value"]
})

# Under H1: means = 5, 5, 7
f_alt <- replicate(n_sim, {
  y_sim <- rnorm(90, mean = rep(c(5, 5, 7), each = 30), sd = 2)
  g_sim <- rep(c("A", "B", "C"), each = 30)
  anova(lm(y_sim ~ g_sim))["g_sim", "F value"]
})

sim_f_dt <- data.table$data.table(
  F_stat = c(f_null, f_alt),
  scenario = rep(c("H0 true", "H1 true"), each = n_sim)
)

ggplot2$ggplot(sim_f_dt, ggplot2$aes(x = F_stat, fill = scenario)) +
  ggplot2$geom_histogram(bins = 50, alpha = 0.6, colour = "white") +
  ggplot2$facet_wrap(~ scenario, scales = "free_y") +
  ggplot2$scale_fill_manual(values = c("H0 true" = "#0072B2", "H1 true" = "#D55E00")) +
  ggplot2$labs(
    x = "F statistic",
    y = "Count",
    title = "Simulated F-statistics: null vs alternative"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "none")
```

Under the null, the $F$-statistics cluster near 1 (the expected value of $F_{2, 87}$ is $87/85 \approx 1.02$). Under the alternative, the distribution shifts dramatically to the right.

---

## 18.12 Power Analysis for One-Way ANOVA

The power of the $F$-test depends on the **non-centrality parameter**:

$$\lambda = \frac{\sum_{j=1}^k n_j(\mu_j - \bar{\mu})^2}{\sigma^2}$$

where $\bar{\mu} = \sum n_j\mu_j / N$. The $F$-statistic under $H_1$ follows a non-central $F$ distribution: $F \sim F_{k-1, N-k, \lambda}$.

```{r power-curve, fig.cap="Power of the one-way ANOVA F-test as a function of the non-centrality parameter lambda. With k = 3 groups and N - k = 87 residual degrees of freedom, power exceeds 0.8 when lambda is greater than approximately 12."}
lambda_seq <- seq(0, 30, by = 0.5)
power_vals <- 1 - pf(
  qf(0.95, df1 = 2, df2 = 87),
  df1 = 2, df2 = 87,
  ncp = lambda_seq
)

pow_dt <- data.table$data.table(lambda = lambda_seq, power = power_vals)

ggplot2$ggplot(pow_dt, ggplot2$aes(x = lambda, y = power)) +
  ggplot2$geom_line(colour = "#0072B2", linewidth = 0.9) +
  ggplot2$geom_hline(yintercept = 0.8, linetype = "dashed", colour = "#D55E00") +
  ggplot2$labs(
    x = expression(lambda ~ "(non-centrality parameter)"),
    y = "Power",
    title = "Power curve for one-way ANOVA (k = 3, N = 90)"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

---

## Communicating to Stakeholders

When presenting ANOVA results:

- **Start with the plot.** Box plots or group mean plots tell the story before any numbers appear. Stakeholders understand "these boxes look different" better than "$F_{2,339} = 594.8$."
- **The $F$-test is an omnibus test.** It tells you *whether* groups differ, not *which* ones. Always follow up with pairwise comparisons (Tukey HSD) to identify specific differences.
- **Report effect sizes alongside $p$-values.** "$p < 0.001$" tells you the effect exists; $\eta^2 = 0.78$ tells you it is large. Both pieces of information are needed.
- **ANOVA is regression.** If your audience is comfortable with regression, explain ANOVA as "a regression with dummy variables." This unifies the framework and makes extensions (ANCOVA, factorial designs) natural.

---

## Quick Reference

| Quantity | Formula |
|:---------|:--------|
| **SSB** (between) | $\sum_{j=1}^k n_j(\bar{Y}_{\cdot j} - \bar{Y}_{\cdot\cdot})^2$ |
| **SSW** (within) | $\sum_{j=1}^k\sum_{i=1}^{n_j}(Y_{ij} - \bar{Y}_{\cdot j})^2$ |
| **$F$-statistic** | $\frac{\text{SSB}/(k-1)}{\text{SSW}/(N-k)}$ |
| **Tukey HSD** | $q_{\alpha,k,N-k}\sqrt{\text{MSW}/n}$ |
| **$\eta^2$** | $\text{SSB}/\text{SST}$ |
| **$\omega^2$** | $[\text{SSB} - (k-1)\text{MSW}] / [\text{SST} + \text{MSW}]$ |

---

## Exercises

1. **ANOVA as regression.** Using the penguins data, construct the design matrix $\mathbf{X}$ (with dummy columns for Chinstrap and Gentoo) by hand. Compute $\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$ and verify that the coefficients match those from `lm(flipper_length_mm ~ species)`.

2. **Deriving the $F$-test.** Starting from the SSB/SSW decomposition, prove that under $H_0$ the ratio $F = \text{MSB}/\text{MSW}$ follows an $F_{k-1, N-k}$ distribution. (*Hint:* use Cochran's theorem.)

3. **Tukey from scratch.** Implement a function `my_tukey_hsd(fit_lm, alpha)` that takes a one-way ANOVA `lm` object and returns all pairwise comparisons with confidence intervals and adjusted $p$-values. Test it on the PlantGrowth data.

4. **Bonferroni vs Tukey.** Using simulation, compare the family-wise error rate and power of Bonferroni and Tukey HSD for $k = 5$ groups with $n = 20$ per group. Which method has greater power?

5. **Welch vs classical.** Simulate one-way ANOVA with $k = 3$ groups where $\sigma_1 = 1$, $\sigma_2 = 1$, $\sigma_3 = 4$ (all means equal). Show that the classical $F$-test has an inflated Type I error rate, while Welch's ANOVA maintains the nominal level.

6. **Penguin effect sizes.** Compute $\eta^2$ and $\omega^2$ for flipper length, body mass, and bill length by species. Which morphometric variable shows the largest between-species variation relative to within-species variation?
