---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 23: Computational Bayesian Methods"
part: "Part 1: MCMC Algorithms"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, mcmc, metropolis-hastings, gibbs-sampling, bayesian, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

```{r packages, include=FALSE}
box::use(ggplot2)
box::use(data.table)
```

```{r dt-symbols, include=FALSE}
.N <- data.table::.N
.SD <- data.table::.SD
```

# Chapter 23 -- Computational Bayesian Methods

## Part 1: MCMC Algorithms

In Chapter 22 we exploited conjugate priors to obtain closed-form posteriors. But most real-world models -- hierarchical models, mixture models, models with many parameters -- do not admit conjugate solutions. The posterior is known only up to a normalising constant:

$$\pi(\boldsymbol{\theta} \mid \mathbf{y}) = \frac{f(\mathbf{y} \mid \boldsymbol{\theta}) \, \pi(\boldsymbol{\theta})}{m(\mathbf{y})}$$

and the integral $m(\mathbf{y}) = \int f(\mathbf{y} \mid \boldsymbol{\theta}) \, \pi(\boldsymbol{\theta}) \, d\boldsymbol{\theta}$ is intractable in high dimensions.

**Markov chain Monte Carlo (MCMC)** solves this problem by constructing a Markov chain whose stationary distribution is the target posterior. We draw correlated samples from this chain and use them to approximate any posterior quantity.

---

## 23.1 Monte Carlo Integration

### 23.1.1 The Core Idea

Suppose we want to compute $E[g(\theta) \mid \mathbf{y}] = \int g(\theta) \, \pi(\theta \mid \mathbf{y}) \, d\theta$. If we can draw $\theta^{(1)}, \dots, \theta^{(S)} \sim \pi(\theta \mid \mathbf{y})$, then by the **law of large numbers**:

$$\hat{g}_S = \frac{1}{S} \sum_{s=1}^S g(\theta^{(s)}) \xrightarrow{a.s.} E[g(\theta) \mid \mathbf{y}] \qquad \text{as } S \to \infty$$

The Monte Carlo standard error is $\text{SE} = \text{sd}(g(\theta^{(s)})) / \sqrt{S}$. With $S = 10{,}000$ draws, we can approximate most posterior summaries to high accuracy.

### 23.1.2 Demonstration: Monte Carlo Integration

```{r mc-integration, fig.cap="Monte Carlo integration converges to the true posterior mean as the number of draws increases."}
set.seed(42)
a_true <- 14
b_true <- 13
true_mean <- a_true / (a_true + b_true)

draws <- rbeta(10000, a_true, b_true)
running_mean <- cumsum(draws) / seq_along(draws)

mc_dt <- data.table$data.table(
  s = seq_along(running_mean),
  estimate = running_mean
)

ggplot2$ggplot(mc_dt, ggplot2$aes(x = s, y = estimate)) +
  ggplot2$geom_line(colour = "#0072B2", alpha = 0.7) +
  ggplot2$geom_hline(yintercept = true_mean, linetype = "dashed", colour = "#D55E00", linewidth = 1) +
  ggplot2$labs(
    x = "Number of Monte Carlo draws (S)",
    y = "Running mean",
    title = sprintf("Monte Carlo estimate converges to true mean = %.4f", true_mean)
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

---

## 23.2 Why MCMC? The Problem with Direct Sampling

For conjugate models we can sample directly from the posterior. But consider a posterior like:

$$\pi(\mu, \sigma^2 \mid \mathbf{y}) \propto \prod_{i=1}^n \frac{1}{\sigma} \exp\!\left[-\frac{(y_i - \mu)^2}{2\sigma^2}\right] \cdot \pi(\mu) \cdot \pi(\sigma^2)$$

With non-conjugate priors, no closed form exists. We cannot use `rbeta()` or `rnorm()`. We need a method that samples from an **unnormalised** density.

---

## 23.3 The Metropolis-Hastings Algorithm

### 23.3.1 The Algorithm

The **Metropolis-Hastings (MH)** algorithm constructs a Markov chain $\{\theta^{(0)}, \theta^{(1)}, \theta^{(2)}, \dots\}$ whose stationary distribution is the target $\pi(\theta \mid \mathbf{y})$.

**Algorithm.** Given current state $\theta^{(t)}$:

1. **Propose** $\theta^* \sim q(\theta^* \mid \theta^{(t)})$ (a proposal distribution).
2. **Compute** the acceptance ratio:

$$\alpha = \min\!\left(1, \; \frac{\pi(\theta^* \mid \mathbf{y}) \, q(\theta^{(t)} \mid \theta^*)}{\pi(\theta^{(t)} \mid \mathbf{y}) \, q(\theta^* \mid \theta^{(t)})}\right)$$

3. **Accept or reject:**
   - With probability $\alpha$, set $\theta^{(t+1)} = \theta^*$.
   - Otherwise, set $\theta^{(t+1)} = \theta^{(t)}$.

The beauty is that $\pi(\theta^* \mid \mathbf{y})$ appears only as a ratio, so the normalising constant cancels:

$$\frac{\pi(\theta^* \mid \mathbf{y})}{\pi(\theta^{(t)} \mid \mathbf{y})} = \frac{f(\mathbf{y} \mid \theta^*) \pi(\theta^*)}{f(\mathbf{y} \mid \theta^{(t)}) \pi(\theta^{(t)})}$$

### 23.3.2 Detailed Balance (Intuition)

The chain satisfies **detailed balance**: the flow of probability from $\theta_a$ to $\theta_b$ equals the flow from $\theta_b$ to $\theta_a$:

$$\pi(\theta_a) \, T(\theta_a \to \theta_b) = \pi(\theta_b) \, T(\theta_b \to \theta_a)$$

where $T(\theta_a \to \theta_b) = q(\theta_b \mid \theta_a) \, \alpha(\theta_a, \theta_b)$. This ensures $\pi(\theta \mid \mathbf{y})$ is the stationary distribution -- the chain spends time in each region proportional to its posterior probability.

### 23.3.3 The Random Walk Metropolis

The most common variant uses a **symmetric** proposal: $q(\theta^* \mid \theta^{(t)}) = q(\theta^{(t)} \mid \theta^*)$, typically:

$$\theta^* = \theta^{(t)} + \epsilon, \qquad \epsilon \sim N(0, \sigma^2_{\text{prop}})$$

The proposal terms cancel, and the acceptance ratio simplifies to:

$$\alpha = \min\!\left(1, \; \frac{\pi(\theta^* \mid \mathbf{y})}{\pi(\theta^{(t)} \mid \mathbf{y})}\right)$$

If the proposed point has higher posterior density, we always accept. If lower, we accept with probability equal to the density ratio -- allowing the chain to occasionally move downhill and explore the full posterior.

---

## 23.4 Implementing Metropolis-Hastings from Scratch

Let us estimate the mean of a normal distribution with known variance using a non-conjugate (Cauchy) prior.

```{r mh-from-scratch}
set.seed(42)

y_data <- rnorm(30, mean = 5, sd = 2)
n_data <- length(y_data)
y_bar <- mean(y_data)
sigma_known <- 2

log_posterior <- function(mu) {
  log_lik <- -n_data / (2 * sigma_known^2) * (mu - y_bar)^2
  log_prior <- dt(mu, df = 1, log = TRUE)
  log_lik + log_prior
}

metropolis_hastings <- function(n_iter, sigma_prop, start = 0) {
  chain <- numeric(n_iter)
  chain[1] <- start
  accepted <- 0

  for (t in 2:n_iter) {
    proposal <- rnorm(1, chain[t - 1], sigma_prop)
    log_alpha <- log_posterior(proposal) - log_posterior(chain[t - 1])

    if (log(runif(1)) < log_alpha) {
      chain[t] <- proposal
      accepted <- accepted + 1
    } else {
      chain[t] <- chain[t - 1]
    }
  }

  list(chain = chain, acceptance_rate = accepted / (n_iter - 1))
}

result <- metropolis_hastings(n_iter = 20000, sigma_prop = 1.5)
sprintf("Acceptance rate: %.3f", result$acceptance_rate)
sprintf("Posterior mean (MCMC): %.4f", mean(result$chain[5001:20000]))
sprintf("Sample mean (MLE): %.4f", y_bar)
```

```{r mh-trace-and-hist, fig.cap="Metropolis-Hastings output: trace plot (left) and posterior histogram (right). The chain mixes well and concentrates around the data mean."}
chain_dt <- data.table$data.table(
  iteration = seq_along(result$chain),
  mu = result$chain
)

p1 <- ggplot2$ggplot(chain_dt[iteration > 500], ggplot2$aes(x = iteration, y = mu)) +
  ggplot2$geom_line(alpha = 0.4, colour = "#0072B2") +
  ggplot2$geom_hline(yintercept = y_bar, linetype = "dashed", colour = "#D55E00") +
  ggplot2$labs(x = "Iteration", y = expression(mu), title = "Trace plot") +
  ggplot2$theme_minimal(base_size = 12)

p2 <- ggplot2$ggplot(chain_dt[iteration > 5000], ggplot2$aes(x = mu)) +
  ggplot2$geom_histogram(
    ggplot2$aes(y = ggplot2::after_stat(density)),
    bins = 50, fill = "#0072B2", alpha = 0.7, colour = "white"
  ) +
  ggplot2$geom_vline(xintercept = y_bar, linetype = "dashed", colour = "#D55E00") +
  ggplot2$labs(x = expression(mu), y = "Density", title = "Posterior histogram") +
  ggplot2$theme_minimal(base_size = 12)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

---

## 23.5 Tuning the Proposal

The proposal standard deviation $\sigma_{\text{prop}}$ controls the trade-off between exploration and acceptance:

- **Too small:** high acceptance rate (~95%), but the chain takes tiny steps and explores slowly.
- **Too large:** the chain proposes extreme values that are usually rejected (~5%), and the chain gets stuck.
- **Optimal:** for a single parameter, an acceptance rate of ~44% is optimal; for many parameters, ~23%.

```{r tuning-demo, fig.cap="Effect of proposal scale on chain behaviour. Too narrow: slow exploration. Too wide: high rejection. A well-tuned proposal explores efficiently."}
set.seed(42)
sigmas <- c(0.1, 1.5, 20)
tuning_list <- list()

for (sig in sigmas) {
  res <- metropolis_hastings(10000, sig)
  tuning_list[[length(tuning_list) + 1]] <- data.table$data.table(
    iteration = 1:10000,
    mu = res$chain,
    sigma_prop = sprintf("sigma = %.1f (accept = %.0f%%)", sig, res$acceptance_rate * 100)
  )
}

tuning_dt <- data.table$rbindlist(tuning_list)

ggplot2$ggplot(tuning_dt[iteration <= 2000], ggplot2$aes(x = iteration, y = mu)) +
  ggplot2$geom_line(alpha = 0.6, colour = "#0072B2") +
  ggplot2$facet_wrap(~ sigma_prop, ncol = 1, scales = "free_y") +
  ggplot2$geom_hline(yintercept = y_bar, linetype = "dashed", colour = "#D55E00") +
  ggplot2$labs(
    x = "Iteration", y = expression(mu),
    title = "Proposal tuning: too narrow, just right, and too wide"
  ) +
  ggplot2$theme_minimal(base_size = 12)
```

---

## 23.6 Gibbs Sampling

### 23.6.1 The Idea

**Gibbs sampling** is a special case of Metropolis-Hastings for multivariate problems. Instead of proposing all parameters at once, we update each parameter in turn from its **full conditional** distribution -- the distribution of that parameter given all others and the data.

For parameters $\theta_1, \theta_2$:

1. Draw $\theta_1^{(t+1)} \sim \pi(\theta_1 \mid \theta_2^{(t)}, \mathbf{y})$
2. Draw $\theta_2^{(t+1)} \sim \pi(\theta_2 \mid \theta_1^{(t+1)}, \mathbf{y})$

Each step is an MH step with acceptance probability 1, because the proposal equals the full conditional.

### 23.6.2 Gibbs as a Special Case of MH

In step 1, the proposal is $q(\theta_1^* \mid \theta_1^{(t)}) = \pi(\theta_1^* \mid \theta_2^{(t)}, \mathbf{y})$. The MH acceptance ratio becomes:

$$\alpha = \frac{\pi(\theta_1^*, \theta_2^{(t)} \mid \mathbf{y}) \, \pi(\theta_1^{(t)} \mid \theta_2^{(t)}, \mathbf{y})}{\pi(\theta_1^{(t)}, \theta_2^{(t)} \mid \mathbf{y}) \, \pi(\theta_1^* \mid \theta_2^{(t)}, \mathbf{y})} = \frac{\cancel{\pi(\theta_2^{(t)} \mid \mathbf{y})} \, \cancel{\pi(\theta_1^{(t)} \mid \theta_2^{(t)}, \mathbf{y})}}{\cancel{\pi(\theta_2^{(t)} \mid \mathbf{y})} \, \cancel{\pi(\theta_1^{(t)} \mid \theta_2^{(t)}, \mathbf{y})}} = 1$$

Every proposal is accepted.

---

## 23.7 Gibbs Sampler: Bivariate Normal

Let us implement a Gibbs sampler for the bivariate normal distribution with known parameters:

$$\begin{pmatrix} X \\ Y \end{pmatrix} \sim N\!\left(\begin{pmatrix} \mu_X \\ \mu_Y \end{pmatrix}, \begin{pmatrix} 1 & \rho \\ \rho & 1 \end{pmatrix}\right)$$

The full conditionals are:

$$X \mid Y = y \sim N(\mu_X + \rho(y - \mu_Y), \; 1 - \rho^2)$$
$$Y \mid X = x \sim N(\mu_Y + \rho(x - \mu_X), \; 1 - \rho^2)$$

```{r gibbs-bivariate}
set.seed(42)

gibbs_bivariate <- function(n_iter, mu_x = 0, mu_y = 0, rho = 0.8) {
  x <- numeric(n_iter)
  y <- numeric(n_iter)
  x[1] <- 0
  y[1] <- 0
  cond_sd <- sqrt(1 - rho^2)

  for (t in 2:n_iter) {
    x[t] <- rnorm(1, mu_x + rho * (y[t - 1] - mu_y), cond_sd)
    y[t] <- rnorm(1, mu_y + rho * (x[t] - mu_x), cond_sd)
  }

  data.table$data.table(iteration = seq_len(n_iter), x = x, y = y)
}

gibbs_samples <- gibbs_bivariate(5000, rho = 0.8)
```

```{r gibbs-scatter, fig.cap="Gibbs sampler for a bivariate normal with rho = 0.8. The samples (after burn-in) reproduce the target correlation structure."}
burn_in <- 500
post_burn <- gibbs_samples[iteration > burn_in]

ggplot2$ggplot(post_burn, ggplot2$aes(x = x, y = y)) +
  ggplot2$geom_point(alpha = 0.15, colour = "#0072B2", size = 0.8) +
  ggplot2$stat_density_2d(colour = "#D55E00", linewidth = 0.5) +
  ggplot2$labs(
    x = "X", y = "Y",
    title = sprintf(
      "Gibbs samples: cor = %.3f (target = 0.800)",
      cor(post_burn$x, post_burn$y)
    )
  ) +
  ggplot2$coord_equal() +
  ggplot2$theme_minimal(base_size = 13)
```

```{r gibbs-path, fig.cap="First 50 iterations of the Gibbs sampler showing the characteristic zig-zag pattern as it alternates between updating X and Y."}
early <- gibbs_samples[iteration <= 50]

ggplot2$ggplot(early, ggplot2$aes(x = x, y = y)) +
  ggplot2$geom_path(alpha = 0.5, colour = "#0072B2") +
  ggplot2$geom_point(size = 1.5, colour = "#D55E00") +
  ggplot2$labs(
    x = "X", y = "Y",
    title = "Gibbs sampler path: first 50 iterations (zig-zag pattern)"
  ) +
  ggplot2$coord_equal() +
  ggplot2$theme_minimal(base_size = 13)
```

---

## 23.8 Interactive: Watch the MCMC Chain Explore

<InteractivePlot>

```{r mcmc-interactive, eval=FALSE}
box::use(shiny)
box::use(ggplot2)

shiny$shinyApp(
  ui = shiny$fluidPage(
    shiny$titlePanel("MCMC: Watch the Chain Explore the Posterior"),
    shiny$sidebarLayout(
      shiny$sidebarPanel(
        shiny$selectInput("algorithm", "Algorithm:",
          choices = c("Random Walk Metropolis" = "rw", "Gibbs Sampler" = "gibbs")
        ),
        shiny$sliderInput("n_show", "Iterations to show:",
          min = 10, max = 5000, value = 100, step = 10, animate = TRUE
        ),
        shiny$sliderInput("sigma_prop", "Proposal SD (Metropolis):",
          min = 0.1, max = 5, value = 1, step = 0.1
        ),
        shiny$sliderInput("rho", "Correlation (Gibbs):",
          min = -0.95, max = 0.95, value = 0.7, step = 0.05
        ),
        shiny$actionButton("rerun", "New chain")
      ),
      shiny$mainPanel(
        shiny$plotOutput("chain_plot", Height = "500px"),
        shiny$verbatimTextOutput("stats_text")
      )
    )
  ),
  server = function(input, output, session) {
    chain_data <- shiny$reactiveVal(NULL)

    shiny$observeEvent(input$rerun, {
      chain_data(NULL)
    }, ignoreNULL = FALSE)

    get_chain <- shiny$reactive({
      input$rerun
      if (input$algorithm == "gibbs") {
        n <- 5000
        rho <- input$rho
        x <- y <- numeric(n)
        csd <- sqrt(1 - rho^2)
        for (t in 2:n) {
          x[t] <- rnorm(1, rho * y[t-1], csd)
          y[t] <- rnorm(1, rho * x[t], csd)
        }
        data.frame(iter = 1:n, x = x, y = y)
      } else {
        n <- 5000
        sp <- input$sigma_prop
        x <- y <- numeric(n)
        for (t in 2:n) {
          px <- rnorm(1, x[t-1], sp)
          py <- rnorm(1, y[t-1], sp)
          log_a <- -(px^2 + py^2)/2 + (x[t-1]^2 + y[t-1]^2)/2
          if (log(runif(1)) < log_a) { x[t] <- px; y[t] <- py
          } else { x[t] <- x[t-1]; y[t] <- y[t-1] }
        }
        data.frame(iter = 1:n, x = x, y = y)
      }
    })

    output$chain_plot <- shiny$renderPlot({
      d <- get_chain()
      k <- min(input$n_show, nrow(d))
      sub <- d[1:k, ]
      ggplot2$ggplot(sub, ggplot2$aes(x = x, y = y)) +
        ggplot2$geom_path(alpha = 0.3, colour = "#0072B2") +
        ggplot2$geom_point(size = 0.5, colour = "#D55E00", alpha = 0.5) +
        ggplot2$labs(title = sprintf("MCMC chain: %d iterations", k)) +
        ggplot2$coord_equal(xlim = c(-4, 4), ylim = c(-4, 4)) +
        ggplot2$theme_minimal(base_size = 14)
    })

    output$stats_text <- shiny$renderText({
      d <- get_chain()
      k <- min(input$n_show, nrow(d))
      sub <- d[1:k, ]
      sprintf("Mean X: %.3f  Mean Y: %.3f  Cor(X,Y): %.3f",
              mean(sub$x), mean(sub$y),
              if (k > 2) cor(sub$x, sub$y) else NA)
    })
  }
)
```

</InteractivePlot>

---

## 23.9 Burn-In and Thinning

### 23.9.1 Burn-In

The chain starts from an arbitrary point and may take many iterations to reach the high-probability region. We discard the first $B$ iterations as **burn-in**. A common choice is to discard the first 10--50% of the chain.

### 23.9.2 Thinning

Successive MCMC samples are correlated. **Thinning** -- keeping every $k$-th sample -- reduces correlation but wastes samples. Modern practice favours running longer chains rather than thinning aggressively.

```{r burnin-demo, fig.cap="Effect of burn-in: the chain starts far from the target and takes several hundred iterations to settle. Discarding these early iterations removes the initialisation bias."}
set.seed(42)
bad_start <- metropolis_hastings(10000, sigma_prop = 1.5, start = -20)

burn_dt <- data.table$data.table(
  iteration = seq_along(bad_start$chain),
  mu = bad_start$chain
)

ggplot2$ggplot(burn_dt, ggplot2$aes(x = iteration, y = mu)) +
  ggplot2$geom_line(alpha = 0.5, colour = "#0072B2") +
  ggplot2$geom_hline(yintercept = y_bar, linetype = "dashed", colour = "#D55E00") +
  ggplot2$geom_vline(xintercept = 1000, linetype = "dotted", colour = "#009E73", linewidth = 1) +
  ggplot2$annotate("text", x = 1200, y = -15, label = "Burn-in ends", colour = "#009E73", hjust = 0) +
  ggplot2$labs(
    x = "Iteration", y = expression(mu),
    title = "Chain started at mu = -20: burn-in removes initialisation transient"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

---

## 23.10 Autocorrelation in MCMC

Unlike independent Monte Carlo draws, MCMC samples are correlated. The **effective sample size (ESS)** quantifies the information content:

$$\text{ESS} = \frac{S}{1 + 2\sum_{k=1}^{\infty} \rho_k}$$

where $\rho_k$ is the autocorrelation at lag $k$.

```{r autocorrelation, fig.cap="Autocorrelation function for MCMC chains with different proposal scales. A well-tuned chain decorrelates quickly."}
set.seed(42)
chain_good <- metropolis_hastings(10000, 1.5)$chain[5001:10000]
chain_narrow <- metropolis_hastings(10000, 0.1)$chain[5001:10000]

acf_good <- acf(chain_good, lag.max = 50, plot = FALSE)
acf_narrow <- acf(chain_narrow, lag.max = 50, plot = FALSE)

acf_dt <- data.table$data.table(
  lag = rep(acf_good$lag[-1], 2),
  acf = c(acf_good$acf[-1], acf_narrow$acf[-1]),
  chain = rep(c("Well-tuned (sigma=1.5)", "Too narrow (sigma=0.1)"), each = 50)
)

ggplot2$ggplot(acf_dt, ggplot2$aes(x = lag, y = acf, fill = chain)) +
  ggplot2$geom_col(position = "dodge", alpha = 0.8) +
  ggplot2$scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  ggplot2$labs(
    x = "Lag", y = "Autocorrelation",
    title = "Autocorrelation: well-tuned chain decorrelates much faster",
    fill = NULL
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "top")
```

```{r ess-computation}
ess_good <- coda::effectiveSize(chain_good)
ess_narrow <- coda::effectiveSize(chain_narrow)
sprintf("ESS (well-tuned): %.0f out of 5000", ess_good)
sprintf("ESS (too narrow): %.0f out of 5000", ess_narrow)
```

---

## Communicating to Stakeholders

- **Demystify MCMC.** "Instead of solving the equation analytically, we use a computer to draw random samples from the answer. More samples means more precision."
- **Explain convergence.** "We ran the algorithm long enough to be confident it has settled into a stable pattern. The trace plot confirms this."
- **Report ESS, not raw iterations.** "We ran 10,000 iterations, but because of autocorrelation the effective sample size is 3,200 -- still more than adequate."
- **Burn-in is quality control.** "We discarded the first 2,000 iterations to remove the effect of our arbitrary starting point."

---

## Quick Reference

| Concept | Details |
|:--------|:--------|
| **Monte Carlo** | $\hat{g} = \frac{1}{S}\sum g(\theta^{(s)}) \to E[g(\theta) \mid \mathbf{y}]$ |
| **MH acceptance** | $\alpha = \min(1, \frac{\pi^* q(\theta^{(t)} \mid \theta^*)}{\pi^{(t)} q(\theta^* \mid \theta^{(t)})})$ |
| **Random walk MH** | Symmetric proposal; $\alpha = \min(1, \pi^*/\pi^{(t)})$ |
| **Gibbs sampling** | Update each parameter from full conditional; acceptance = 1 |
| **Detailed balance** | $\pi(a) T(a \to b) = \pi(b) T(b \to a)$ |
| **Burn-in** | Discard early iterations (initialisation transient) |
| **ESS** | $S / (1 + 2\sum \rho_k)$; effective independent samples |
| **Optimal acceptance** | ~44% for 1D; ~23% for many parameters |
| **Thinning** | Keep every $k$-th sample; modern: prefer longer chains |

---

## Exercises

1. **MH from scratch.** Implement the Metropolis-Hastings algorithm to sample from a $\text{Gamma}(3, 1)$ distribution using a log-normal proposal. Verify your samples match the true distribution by comparing quantiles.

2. **Proposal tuning.** For the normal posterior example in Section 23.4, run the MH sampler with $\sigma_{\text{prop}} \in \{0.05, 0.5, 1.5, 5, 50\}$. Plot the acceptance rate vs $\sigma_{\text{prop}}$ and identify the optimal range.

3. **Gibbs for normal.** Implement a Gibbs sampler for $\mu$ and $\sigma^2$ of a normal distribution using conjugate priors ($\mu \mid \sigma^2 \sim N(\mu_0, \sigma^2/\kappa_0)$ and $\sigma^2 \sim \text{Inv-Gamma}(\nu_0/2, \nu_0 \sigma_0^2/2)$). Apply to the `airquality` temperature data.

4. **ESS comparison.** Generate 10,000 MCMC samples from a standard normal using (a) independent sampling, (b) MH with $\sigma_{\text{prop}} = 2.4$, and (c) MH with $\sigma_{\text{prop}} = 0.1$. Compute the ESS for each. How many MH samples would you need to match the ESS of 1,000 independent samples?

5. **Multimodal target.** Implement MH for the mixture $0.3 \cdot N(-3, 1) + 0.7 \cdot N(3, 1)$. Does the random walk Metropolis explore both modes? Experiment with different proposal scales.

6. **Gibbs for bivariate.** Modify the bivariate normal Gibbs sampler to handle $\rho = 0.99$. How does the high correlation affect mixing? Compute the ESS and compare with $\rho = 0.5$.
