---
letter: 'b'
word: 'black-scholes'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['monte-carlo-simulation','distribution']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x,options){
    cap<-options$fig.cap
    as.character(
      htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t",cap,"\n",sep="")))
    )
  })
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="black-scholes">Black-Scholes Formula</a> - In **finance**, the **Black-Scholes** (or **Black-Scholes-Merton**) formula provides a closed-form solution for pricing a European call/put option under certain assumptions (continuous compounding, constant volatility, no dividends, etc.).

**Call option price**:
'68859'
C_0 = S_0\Phi(d_1) - K e^{-rT}\Phi(d_2),
'68859'
where
'68859'
d_1 = \frac{\ln\bigl(\frac{S_0}{K}\bigr) + \bigl(r + \tfrac{1}{2}\sigma^2\bigr)T}{\sigma\sqrt{T}}, \quad
d_2 = d_1 - \sigma\sqrt{T}.
'68859'

\(\Phi\) is the cumulative distribution function of the standard normal distribution.

**R demonstration** (computing call price):

```{r}
library(data.table)
library(ggplot2)

# We'll define a function to compute d1, d2, then call price
phi <- function(x) pnorm(x)  # standard normal cdf

black_scholes_call <- function(S0, K, r, sigma, T) {
  d1 <- (log(S0/K) + (r + 0.5*sigma^2)*T)/(sigma*sqrt(T))
  d2 <- d1 - sigma*sqrt(T)
  C0 <- S0*phi(d1) - K*exp(-r*T)*phi(d2)
  C0
}

S0 <- 100
K  <- 105
r  <- 0.02
sigma <- 0.2
T <- 1
C0 <- black_scholes_call(S0, K, r, sigma, T)
C0
```

