---
letter: 'n'
word: 'normal-distribution'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['distribution','mean','variance']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="normal-distribution">Normal Distribution</a> - In **statistics**, the **normal distribution** (or Gaussian) is a continuous probability [distribution](#distribution) with probability density function:

'56956'
f(x) = \frac{1}{\sqrt{2\pi}\,\sigma} \exp\!\Bigl(-\frac{(x - \mu)^2}{2\sigma^2}\Bigr)
'56956'

where '56956'\mu'56956' is the [mean](#mean) and '56956'\sigma^2'56956' is the [variance](#variance).

**Key points**:
- Symmetric, bell-shaped curve.
- Many natural phenomena approximate normality by Central Limit Theorem arguments.

**R demonstration**:

```{r}
library(data.table)
library(ggplot2)

set.seed(123)
dt_norm <- data.table(x = rnorm(1000, mean=5, sd=2))

ggplot(dt_norm, aes(x=x)) +
  geom_histogram(bins=30, fill="lightblue", color="black", aes(y=..density..)) +
  geom_density(color="red", size=1) +
  labs(
    title="Normal Distribution Example",
    x="Value",
    y="Density"
  ) +
  theme_minimal()
```
