---
letter: 's'
word: 'sharpe-ratio'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['standard-deviation','mean','value-at-risk']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x, options){
    cap<-options$fig.cap
    as.character(
      htmltools::tag("Figure",
        list(src=x, alt=cap, paste("\n\t",cap,"\n",sep=""))
      )
    )
  })
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="sharpe-ratio">Sharpe Ratio</a> - In **finance**, the **Sharpe ratio** measures the performance of an investment (or strategy) compared to a risk-free asset, accounting for volatility. 

**Formula**:
$$
\text{Sharpe Ratio} = \frac{\mathbb{E}[R - R_f]}{\sigma_{(R-R_f)}},
$$
where
- \(R\) is the asset/portfolio return,
- \(R_f\) is the risk-free rate,
- \(\sigma\) is the standard deviation of \(R - R_f\).

**Key points**:
- Higher is better, but it can be gamed by different return distributions.
- Assumes returns are normally distributed for best interpretability.

**R demonstration** (compute Sharpe for a random returns series):

```{r}
library(data.table)
set.seed(123)
Rf <- 0.0005  # risk-free daily, e.g. ~0.12% annual
n <- 200
ret <- rnorm(n, mean=0.001, sd=0.01)

dt_sharpe <- data.table(ret=ret)
excess <- dt_sharpe$ret - Rf
sr <- mean(excess)/sd(excess)
sr
```

