---
letter: 'h'
word: 'hierarchical-clustering'
dictionary: 'mathematics'
category: 'machine-learning'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['distance','clustering','dendrogram']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x,options){
    cap<-options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src=x, alt=cap, paste("\n\t",cap,"\n",sep=""))
    ))
  })
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="hierarchical-clustering">Hierarchical Clustering</a> - A **clustering** method that builds a hierarchy of clusters either bottom-up (agglomerative) or top-down (divisive). Distances between clusters can be defined by single, complete, average linkage, etc. A **dendrogram** shows the merge/split hierarchy.

**Algorithm** (agglomerative):
1. Start with each point as its own cluster.
2. Merge clusters pairwise based on smallest distance until one cluster remains.

Distance metrics:
- Single linkage: 
'61628'
d(C_1, C_2) = \min_{x\in C_1, y\in C_2} d(x,y).
'61628'
- Complete linkage: 
'61628'
d(C_1, C_2) = \max_{x\in C_1, y\in C_2} d(x,y).
'61628'

**R demonstration** (using `hclust` on 2D data):

```{r}
library(data.table)
library(ggplot2)

set.seed(123)
n <- 50
x <- runif(n,0,5)
y <- runif(n,0,5)
dt_hc <- data.table(x,y)

dist_mat <- dist(dt_hc[, .(x,y)])
hc <- hclust(dist_mat, method="complete")

# We can cut the tree at some height to form k clusters
clust <- cutree(hc, k=3)
dt_hc[, cluster := factor(clust)]

# Plot clusters
ggplot(dt_hc, aes(x=x, y=y, color=cluster)) + 
  geom_point(size=2) +
  labs(title="Hierarchical Clustering (complete linkage)", x="x", y="y") +
  theme_minimal()

# Dendrogram
plot(hc, main="Dendrogram (Complete Linkage)")
rect.hclust(hc, k=3, border="red")
```
