---
letter: 'm'
word: 'monoid'
dictionary: 'mathematics'
category: 'algebra'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['semigroup','group']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(
    plot=function(x,options){
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src=x,alt=cap,paste("\n\t",cap,"\n",sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300,fig.width=10,fig.height=7)
```

<a id="monoid">Monoid</a> - In **abstract algebra**, a **monoid** is a [semigroup](#semigroup) with an identity element. Specifically, a set \(M\) with an associative binary operation \(*\) and an identity element \(e\) so:

1. Associativity: $(a*b)*c = a*(b*c)$ for all $a,b,c\in M$.
2. Identity: $(a*e = e*a = a)$ for all $a\in M$.

**Key points**:
- A [group](#group) is a monoid where every element also has an inverse.
- Examples: Natural numbers under addition with identity 0, strings under concatenation with identity "" (empty string).

No direct R demonstration typical, but we can show a small "string monoid":

```{r}
library(data.table)

str_monoid_op <- function(a,b) paste0(a,b)  # concatenation
e <- ""  # identity

# Check associativity on a small example
a<-"cat"; b<-"fish"; c<-"food"
assoc_left  <- str_monoid_op(str_monoid_op(a,b), c)
assoc_right <- str_monoid_op(a, str_monoid_op(b,c))
data.table(assoc_left, assoc_right)
```
