---
letter: 'h'
word: hypothesis-testing
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

# We reference "distribution" in the definition, so let's include it in linksTo.
linksTo: ['distribution']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", 
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="hypothesis-testing">Hypothesis Testing</a> - In **statistics**, **hypothesis testing** is a method to decide whether sample data support or refute a particular hypothesis about a population parameter or [distribution](#distribution).

**Common steps**:
1. State the null hypothesis ($H_0$) and an alternative hypothesis ($H_1$).
2. Choose a significance level ($\alpha$) and test statistic.
3. Compute the p-value from sample data.
4. Reject or fail to reject $H_0$ based on whether the p-value is below $\alpha$.

**R demonstration** (example t-test):

```{r}
library(data.table)

set.seed(123)
dt_ht <- data.table(
  groupA = rnorm(20, mean=5, sd=1),
  groupB = rnorm(20, mean=5.5, sd=1)
)

# Let's do a two-sample t-test
res <- t.test(dt_ht$groupA, dt_ht$groupB, var.equal=TRUE)
res
```
