---
letter: 'z'
word: 'z-axis'
dictionary: 'mathematics'
category: 'geometry'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['x-axis','y-axis','coordinate-system']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="z-axis">Z-axis</a> - In a **3D coordinate system**, the **z-axis** is the line orthogonal to both the [x-axis](#x-axis) and [y-axis](#y-axis). Positive values extend “up” or “out” from the origin depending on orientation, while negative values extend in the opposite direction.

**Key points**:
- Commonly used in 3D geometry or 3D coordinate-systems.
- Forms a right-handed system with x and y axes if oriented properly.

Basic R demonstration (no direct 3D in base R, but let's show a conceptual data set):

```{r}
library(data.table)

dt_3d <- data.table(x=0, y=0, z=-3:3)
dt_3d
```
