---
letter: 'z'
word: 'zeckendorf-representation'
dictionary: 'mathematics'
category: 'number-theory'
dataSource: 'assistant'
published: true
comments: true

linksTo: []
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="zeckendorf-representation">Zeckendorf Representation</a> - In **number theory**, **Zeckendorf’s theorem** states every positive integer can be uniquely written as a sum of nonconsecutive Fibonacci numbers. This sum is called the **Zeckendorf representation**. For example, 17 = 13 + 3 + 1 uses Fibonacci numbers (1,2,3,5,8,13, ...).

**Key points**:
- No two consecutive Fibonacci numbers are used in the representation.
- The representation is unique for each positive integer.

**R demonstration** (a naive function to find Zeckendorf representation):

```{r}
library(data.table)

zeckendorf <- function(n) {
  # build fibonacci up to n
  fib <- c(1,2)
  while(tail(fib,1) < n) {
    fib <- c(fib, fib[length(fib)] + fib[length(fib)-1])
  }
  representation <- c()
  remaining <- n
  for (f in rev(fib)) {
    if (f <= remaining) {
      representation <- c(representation, f)
      remaining <- remaining - f
    }
  }
  representation
}

zeckendorf(17)  # expect 13,3,1
```
