---
letter: 'f'
word: fractal
dictionary: 'mathematics'
category: 'geometry'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['fourier-transform','frequency']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.height= 7
)
```

<a id="fractal">Fractal</a> - A **fractal** is a geometric object that often exhibits self-similarity at various scales. Examples include the Mandelbrot set, Julia sets, and natural phenomena (coastlines, etc.).

**Key traits**:
- *Self-similarity*: Zoomed-in portions look similar to the original.
- *Fractional dimension*: Dimension can be non-integer.
- *Often defined recursively or via iterative processes*.

**R demonstration** (a simple iteration for the Koch snowflake boundary length, numerical only):

```{r}
library(data.table)

koch_iterations <- 5
dt_koch <- data.table(step = 0:koch_iterations)

# Start with length 1 for the side of an equilateral triangle
# Each iteration multiplies the total line length by 4/3
dt_koch[, length := (4/3)^step]

dt_koch
```

