---
letter: 'm'
word: 'monte-carlo-simulation'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['distribution','probability','markov-chain']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(
    plot=function(x, options){
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src=x, alt=cap, paste("\n\t",cap,"\n",sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="monte-carlo-simulation">Monte Carlo Simulation</a> - A **Monte Carlo simulation** uses repeated random sampling to estimate numerical results. In **finance**, it’s often used for pricing derivatives, calculating risk metrics, or projecting potential outcomes of uncertain processes.

**Mathematical principle**:
If we want to estimate
$$
\mathbb{E}[f(X)] = \int f(x) p(x) \,dx,
$$
we can approximate by sampling \(X_1,\dots,X_N\) from \(p\) and computing:
$$
\frac{1}{N}\sum_{i=1}^N f(X_i).
$$

**R demonstration** (estimating the price of a European call option via a simple GBM model):

```{r}
library(data.table)
library(ggplot2)

set.seed(123)

# Suppose we have a stock with current price S0, volatility sigma, risk-free rate r,
# time to maturity T, strike K.
S0 <- 100
K  <- 105
r  <- 0.02
sigma <- 0.20
T <- 1
num_sims <- 10000

# Generate terminal stock prices under Geometric Brownian Motion:
# S_T = S0 * exp((r - 0.5*sigma^2)*T + sigma*sqrt(T)*Z)
Z <- rnorm(num_sims)
ST <- S0 * exp((r - 0.5*sigma^2)*T + sigma*sqrt(T)*Z)

# Payoff for European call:
payoffs <- pmax(ST - K, 0)
# Discount payoffs back:
call_value <- exp(-r*T)*mean(payoffs)
call_value

# Visual quick distribution:
dt_call <- data.table(ST=ST, payoffs=payoffs)
ggplot(dt_call, aes(x=ST)) +
  geom_histogram(bins=40, fill="lightblue", color="black") +
  labs(title="Monte Carlo Sim for Terminal Stock Price", x="S(T)", y="Frequency") +
  theme_minimal()
```

