---
letter: 'i'
word: interval
dictionary: 'mathematics'
category: 'analysis'
dataSource: 'assistant'
published: true
comments: true

# Suppose we reference "real-number-line" (not yet created?), so we add it:
linksTo: ['real-number-line']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="interval">Interval</a> - In **analysis**, an **interval** is a connected subset of the [real-number-line](#real-number-line). Common types of intervals include:

- **Open interval**: $(a,b) = \{ x : a < x < b \}$
- **Closed interval**: $[a,b] = \{ x : a \le x \le b \}$
- **Half-open / half-closed**: $(a,b] = \{ x : a < x \le b \}$, etc.

Intervals are the building blocks of basic topology on the real line and are central in defining integrals, continuity, and other concepts of real analysis.

```{r}
# Minimal R demonstration: we can define intervals simply with numeric vectors
my_interval <- 0:5  # representing discrete steps from 0 to 5
my_interval
```
