---
letter: 'l'
word: 'lhopitals-rule'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

# We reference "derivative" and "limit" in the definition, so let's add them.
linksTo: ['derivative','limit']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="lhopitals-rule">L'Hôpital's Rule</a> - In **calculus**, **L'Hôpital's rule** is a result for evaluating certain indeterminate forms of [limit](#limit) expressions. If

56724
\lim_{x \to a} \frac{f(x)}{g(x)}
56724

produces indeterminate forms like $0/0$ or $\infty/\infty$, then (under certain conditions involving differentiability and continuity):

56724
\lim_{x \to a} \frac{f(x)}{g(x)} = \lim_{x \to a} \frac{f'(x)}{g'(x)},
56724

provided the latter limit exists. It relies on the concept of the [derivative](#derivative).

**Simple R demonstration** (symbolic approach would be used in a CAS, but we can do numeric checks):

```{r}
library(data.table)

f <- function(x) x^2 - 1
g <- function(x) x - 1

# Evaluate near x=1 to see 0/0
x_vals <- seq(0.9, 1.1, by=0.01)
dt_lhop <- data.table(
  x = x_vals,
  f_x = f(x_vals),
  g_x = g(x_vals),
  ratio = f(x_vals)/g(x_vals)
)
head(dt_lhop)
```

We can see the ratio near x=1 is close to the ratio of derivatives at that point:
- f'(x) = 2x
- g'(x) = 1
So at x=1, ratio ~ 2(1)/1 = 2.
