---
letter: 'o'
word: 'orthogonal'
dictionary: 'mathematics'
category: 'linear-algebra'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['matrix']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="orthogonal">Orthogonal</a> - In **linear algebra**, vectors (or subspaces) are **orthogonal** if their dot product is zero. A set of vectors is orthogonal if every pair of distinct vectors in the set is orthogonal. A [matrix](#matrix) is an orthogonal matrix if $Q^T Q = I$.

**Key points**:
- Orthogonality generalises the concept of perpendicularity in higher dimensions.
- Orthogonal transformations preserve lengths and angles.

**R demonstration** (check if a matrix is orthogonal):

```{r}
library(data.table)

Q <- matrix(c(0,1, -1,0), nrow=2, byrow=TRUE)
# Q^T Q
test_orth <- t(Q) %*% Q
test_orth
# If Q is orthogonal, test_orth should be the 2x2 identity.
```
