---
letter: 'e'
word: euclidean-algorithm
dictionary: 'mathematics'
category: 'number-theory'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['expectation','eigenvalue']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.height = 7
)
```

<a id="euclidean-algorithm">Euclidean Algorithm</a> - The **Euclidean algorithm** is a process to find the greatest common divisor (GCD) of two integers. It repeatedly applies the rule:

55424
\gcd(a, b) = \gcd(b, a \bmod b),
55424

until the remainder is zero.

**Key points**:
- Efficiently computes the GCD without factorising the numbers fully.
- Fundamental in number theory and applications like rational simplification.

**R demonstration** (using built-in gcd from base R):

```{r}
library(data.table)

# Let's pick two integers
a_val <- 56
b_val <- 98

gcd_val <- gcd(a_val, b_val)
gcd_val

# Or define a custom Euclidean function:
euclid_gcd <- function(a, b) {
  if (b == 0) return(a)
  euclid_gcd(b, a %% b)
}
euclid_gcd(a_val, b_val)
```

