---
letter: 'm'
word: 'markowitz-portfolio'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['variance','mean','covariance']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(
    plot=function(x,options){
      cap<-options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t",cap,"\n",sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="markowitz-portfolio">Markowitz Portfolio Theory</a> - **Modern Portfolio Theory (MPT)** by Harry Markowitz aims to construct an **efficient frontier** of portfolios that maximize [mean](#mean) return for a given level of [variance](#variance), or equivalently minimize variance for a given expected return.

**Mathematical formulation**:
- We have assets with expected returns \(\mu_i\) and a covariance matrix \(\Sigma\).  
- Let \(w\) be the vector of weights. We want to:

$$
\max_{w} \Bigl( w^T \mu - \lambda \, w^T \Sigma \, w \Bigr)
$$
where \(\lambda\) is a risk-aversion parameter. Or we can do a constraint-based minimisation:

$$
\min_{w} \bigl(w^T \Sigma \, w\bigr) \quad \text{subject to } w^T\mu = \mu_{target}, \; \mathbf{1}^T w = 1.
$$

**R demonstration** (a small 3-asset example using a quadratic programming approach):

```{r}
library(data.table)
library(quadprog)
set.seed(123)

# Suppose we have 3 assets
mu <- c(0.08, 0.12, 0.15)  # expected returns
Sigma <- matrix(c(0.01, 0.002, 0.0015,
                  0.002,0.02 , 0.003,
                  0.0015, 0.003, 0.03), nrow=3, byrow=TRUE)
names(mu) <- c("Asset1","Asset2","Asset3")

# We want to minimize w^T Sigma w subject to sum(w)=1
# We can do this with solve.QP:
Dmat <- 2*Sigma
dvec <- rep(0,3)
Amat <- cbind(rep(1,3))  # sum(w)=1
bvec <- 1
res <- solve.QP(Dmat, dvec, Amat, bvec, meq=1)
w <- res$solution
w
# w is the minimum variance portfolio weights

# Let's see the variance and return
var_port <- t(w) %*% Sigma %*% w
ret_port <- sum(w*mu)
var_port
ret_port
```

