---
letter: 'i'
word: induction
dictionary: 'mathematics'
category: 'algebra'
dataSource: 'assistant'
published: true
comments: true

# We reference "arithmetic" in the definition, so we add it to linksTo.
linksTo: ['arithmetic']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="induction">Induction</a> - **Mathematical induction** is a proof technique used to show that a statement holds for all natural numbers. It involves two steps:

1. **Base Case**: Prove the statement for the first natural number (often $n=1$).
2. **Inductive Step**: Assume the statement holds for some $n=k$, and then prove it holds for $n=k+1$.

This relies on the well-ordering principle of the natural numbers.

Consider a simple example with [arithmetic](#arithmetic) progressions:
- We may prove $1 + 2 + \dots + n = \frac{n(n+1)}{2}$ by induction.

No complicated R demonstration is typical here, but we can at least verify sums for a few values:

```{r}
library(data.table)

n_vals <- 1:10
dt_ind <- data.table(
  n     = n_vals,
  sum_n = sapply(n_vals, function(k) sum(1:k)),
  formula = n_vals*(n_vals+1)/2
)

dt_ind
```
