---
letter: 'p'
word: 'poisson-distribution'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['distribution','mean']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="poisson-distribution">Poisson Distribution</a> - In **probability**, the **Poisson distribution** is a discrete [distribution](#distribution) describing the probability of a number of events occurring in a fixed interval, given the events occur with a known average rate (the [mean](#mean)) and independently of the time since the last event.

Its pmf for $k=0,1,2,\dots$ is:

'57323'
P(X=k) = \frac{\lambda^k e^{-\lambda}}{k!},
'57323'

where '57323'\lambda'57323' is both the mean and the variance of the distribution.

**R demonstration**:

```{r}
library(data.table)
library(ggplot2)

lambda_val <- 4
dt_pois <- data.table(k = 0:15)
dt_pois[, prob := dpois(k, lambda_val)]

ggplot(dt_pois, aes(x=factor(k), y=prob)) +
  geom_col(fill="lightblue", color="black") +
  labs(
    title=paste("Poisson distribution with lambda =", lambda_val),
    x="k",
    y="Probability"
  ) +
  theme_minimal()
```
