---
letter: 'b'
word: 'backtesting'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['algorithmic-trading','markov-chain','monte-carlo-simulation']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x,options){
    cap<-options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src=x, alt=cap, paste("\n\t",cap,"\n",sep=""))
    ))
  })
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="backtesting">Backtesting</a> - In **algorithmic trading**, **backtesting** is the process of applying a trading strategy (or model) to historical data to see how it would have performed. This helps gauge performance, risk, and feasibility before live deployment.

**Key steps**:
1. Gather clean historical data.
2. Define entry/exit rules and signal generation.
3. Simulate trades on historical data, accounting for transaction costs, slippage, etc.
4. Collect performance metrics (e.g., returns, drawdowns, Sharpe ratio, max drawdown).

**R demonstration** (toy strategy on synthetic daily data):

```{r}
library(data.table)
library(ggplot2)

set.seed(123)
n <- 200
price <- cumsum(rnorm(n, mean=0.01, sd=0.05)) + 100
dt_bt <- data.table(day=1:n, price=price)

# A naive "moving average crossover" strategy:
fast_ma <- 5
slow_ma <- 20

dt_bt[, fast := frollmean(price, fast_ma)]
dt_bt[, slow := frollmean(price, slow_ma)]

# Signal: go long if fast>slow, else flat
dt_bt[, signal := fifelse(fast>slow, 1, 0)]
# lag signal by 1 to avoid lookahead
dt_bt[, signal := shift(signal,1,0)]
# daily return
dt_bt[, ret_day := shift(price,1,NA)/price - 1]
# strategy daily PnL
dt_bt[, strategy_ret := signal*ret_day]
dt_bt[, cumPNL := cumsum(strategy_ret)]

ggplot(dt_bt, aes(x=day)) +
  geom_line(aes(y=cumPNL, color="Strategy PnL")) +
  geom_line(aes(y=(price-100), color="Price-100"), alpha=0.7) +
  labs(title="Backtesting Simple MA Crossover", x="Day", y="PNL / Price(shifted)") +
  scale_color_manual(values=c("Strategy PnL"="blue","Price-100"="red")) +
  theme_minimal()
```

