---
letter: 'p'
word: 'percentile'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['distribution','median']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="percentile">Percentile</a> - In **statistics**, a **percentile** is a measure used to indicate the value below which a given percentage of observations in a group of observations falls. For example, the 50th percentile is the [median](#median). 

**Key points**:
- Commonly used in test scores, growth charts, and any context where relative standing is measured.
- The [distribution](#distribution) of data helps interpret percentile rank.

**R demonstration** (finding percentiles via `quantile`):

```{r}
library(data.table)

set.seed(123)
dt_p <- data.table(vals = rnorm(100, mean=0, sd=1))
quantile(dt_p$vals, probs=c(0.25, 0.5, 0.75))
```
