---
letter: 'q'
word: 'queueing-theory'
dictionary: 'mathematics'
category: 'probability'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['distribution']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="queueing-theory">Queueing Theory</a> - In **probability** and operations research, **queueing theory** studies the behaviour of waiting lines or queues. Models often involve a Poisson arrival [distribution](#distribution) and exponential service times, e.g., the M/M/1 queue.

**Key points**:
- Performance measures: average waiting time, queue length, server utilisation.
- Widely applied in telecommunications, computer networks, and service systems.

**R demonstration** (simulation of a simple queue, optional approaches exist but we show a conceptual snippet):

```{r}
library(data.table)

set.seed(123)
# We'll simulate interarrival times with rexp(rate=lambda),
# and service times with rexp(rate=mu).

simulate_queue <- function(n_customers, lambda=1, mu=1) {
  # generate interarrival times
  inter_arr <- rexp(n_customers, rate=lambda)
  arrival_times <- cumsum(inter_arr)
  service_times <- rexp(n_customers, rate=mu)
  
  # track when each customer starts service
  start_service <- numeric(n_customers)
  finish_service <- numeric(n_customers)
  
  for(i in seq_len(n_customers)) {
    if(i==1) {
      start_service[i] <- arrival_times[i]
    } else {
      start_service[i] <- max(arrival_times[i], finish_service[i-1])
    }
    finish_service[i] <- start_service[i] + service_times[i]
  }
  
  data.table(
    customer = 1:n_customers,
    arrival = arrival_times,
    start = start_service,
    finish = finish_service,
    wait = start_service - arrival_times
  )
}

dt_queue <- simulate_queue(10, lambda=1, mu=2)
dt_queue
```
