---
letter: 'd'
word: dimension
dictionary: 'mathematics'
category: 'geometry'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['determinant','derivative']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi       = 300,
  fig.width = 10,
  fig.height= 7
)
```

<a id="dimension">Dimension</a> - **Dimension** generally refers to the number of coordinates needed to specify a point in a space:

- In **geometry**, 2D refers to a plane, 3D to space, etc.
- In **linear algebra**, dimension is the cardinality of a basis for a vector space.
- In **data science**, dimension often describes the number of features or columns in a dataset.

**Linear algebra perspective**:
If $V$ is a vector space over a field $\mathbb{F}$ and $\{\mathbf{v}_1, \dots, \mathbf{v}_n\}$ is a basis for $V$, then $\dim(V) = n$.

**R demonstration** (showing dimension of a data.table):

```{r}
library(data.table)

dt_dim <- data.table(
  colA = rnorm(5),
  colB = rnorm(5),
  colC = rnorm(5)
)

# Number of rows
nrow(dt_dim)

# Number of columns (dimension in the sense of data features)
ncol(dt_dim)
```

We have a 5 × 3 data.table, so we can say it has 3 “features” or columns in that sense, but in linear algebra, dimension has a more formal meaning related to basis and span.

