---
letter: 'r'
word: 'rank'
dictionary: 'mathematics'
category: 'linear-algebra'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['matrix','dimension']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="rank">Rank</a> - In **linear algebra**, the **rank** of a [matrix](#matrix) $A$ is the dimension of its column space (or row space). Equivalently, it’s the maximum number of linearly independent columns (or rows). The rank also ties into the concept of [dimension](#dimension) of the image (column space) of a linear transformation.

**Key points**:
- If $\mathrm{rank}(A) < \min(m,n)$ for an $m \times n$ matrix, the matrix is not full rank.
- The rank-nullity theorem: $\mathrm{rank}(A)+ \mathrm{nullity}(A) = n$.

**R demonstration** (finding rank of a matrix):

```{r}
library(MASS)
library(data.table)

A <- matrix(c(1,2,3, 2,4,6, 3,6,9), nrow=3, byrow=TRUE)
A_rank <- qr(A)  # rank via QR decomposition
A_rank
```
