---
letter: 'l'
word: 'limit'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

# We'll reference "derivative" or "integral"? Usually derivative is defined by a limit.
# Let's say we link to "derivative".
linksTo: ['derivative']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="limit">Limit</a> - In **calculus**, a **limit** describes the value that a function (or sequence) “approaches” as the input (or index) moves toward some point. For a function $f(x)$:

56724
\lim_{x \to a} f(x) = L
56724

means that $f(x)$ can be made arbitrarily close to $L$ by taking $x$ sufficiently close to $a$.

**Key role in**:
- Defining the [derivative](#derivative): $f'(x) = \lim_{h \to 0} \dots$.
- Defining continuity and integrals.

**R demonstration** (numeric approximation of a limit at a point):

```{r}
library(data.table)

f <- function(x) (x^2 - 1)/(x - 1)  # Indeterminate at x=1, but simplifies to x+1
x_vals <- seq(0.9, 1.1, by=0.01)
dt_lim <- data.table(
  x = x_vals,
  f_x = f(x_vals)
)
dt_lim
# As x -> 1, f(x)-> 2. 
```
