---
letter: 'a'
word: asymptote
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['algebra','angle']
linkedFrom: ['absolute-value','arithmetic']
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
# Common setup for rendering
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="asymptote">Asymptote</a> - An asymptote of a function is a line (or curve) that the function approaches as the input or output grows large in magnitude.

**Types**:
- **Horizontal**: $y = L$ if $\lim_{x \to \pm\infty} f(x) = L$.
- **Vertical**: $x = c$ if $\lim_{x \to c} \bigl| f(x) \bigr| = \infty$.
- **Oblique (Slant)**: $y = mx + b$ if the function approaches that line as $x \to \pm\infty$.

**Example**: $f(x) = \frac{1}{x}$:
- Horizontal asymptote at $y=0$, since 
  $$\lim_{x\to\pm\infty} \frac{1}{x} = 0.$$
- Vertical asymptote at $x=0$, since 
  $$\lim_{x\to 0} \left|\frac{1}{x}\right| = \infty.$$

To analyse numerically in R:

```{r}
f <- function(x) 1/x

# Large values
large_x <- seq(100, 1000, by=200)
vals_large <- f(large_x)
vals_large

# Near x=0
small_x <- seq(-0.1, 0.1, by=0.05)
vals_small <- f(small_x)
vals_small
```

Observe how $f(x)$ tends to $0$ for large $|x|$ (horizontal asymptote) and diverges as $x$ approaches $0$ (vertical asymptote).

