---
letter: 'w'
word: 'whole-number'
dictionary: 'mathematics'
category: 'arithmetic'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['arithmetic']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="whole-number">Whole Number</a> - A **whole number** typically refers to the non-negative integers:

'58445'
\{0, 1, 2, 3, \dots\}.
'58445'

Depending on convention, “whole numbers” may or may not include zero. They are part of [arithmetic](#arithmetic) operations on integers (add, subtract, multiply, etc.).

**Key Points**:
- Whole numbers are closed under addition and multiplication.
- Subtraction can lead outside the set if the result is negative (unless zero is included, etc.).

No direct R demonstration needed, but we can illustrate basic set membership:

```{r}
whole_nums <- 0:10
whole_nums
```
