---
letter: 'k'
word: 'kurtosis'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['distribution']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="kurtosis">Kurtosis</a> - In **statistics**, **kurtosis** measures the “tailedness” of a [distribution](#distribution). The standard formula for sample kurtosis (excess kurtosis) is often:

56642
\mathrm{excess\ kurtosis} = \frac{\sum (x_i - \bar{x})^4 / n}{(\sum (x_i - \bar{x})^2 / n)^2} - 3.
56642

- **High kurtosis**: heavy tails, outliers are more frequent.
- **Low kurtosis**: light tails, fewer extreme outliers (relative to a normal distribution).

**R demonstration**:

```{r}
library(data.table)
library(e1071)  # for kurtosis function
set.seed(123)
dt_kurt <- data.table(
  normal = rnorm(500, mean=0, sd=1),
  heavy_tail = rt(500, df=3)  # t-dist with df=3, heavier tails
)

k_norm <- e1071::kurtosis(dt_kurt$normal)
k_heavy <- e1071::kurtosis(dt_kurt$heavy_tail)

k_norm
k_heavy
```
