---
letter: 'n'
word: 'null-hypothesis'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['hypothesis-testing']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="null-hypothesis">Null Hypothesis</a> - In **statistics**, the **null hypothesis** (commonly denoted $H_0$) is a baseline assumption or “no change” scenario in [hypothesis-testing](#hypothesis-testing). Typically, $H_0$ states that there is no effect or no difference between groups.

**Key points**:
- We either “reject $H_0$” or “fail to reject $H_0$” based on data evidence.
- The alternative hypothesis $H_1$ or $H_a$ posits the effect or difference.

**R demonstration** (t-test example, focusing on $H_0$ that the population means are equal):

```{r}
library(data.table)

set.seed(123)
dt_null <- data.table(
  groupA = rnorm(20, mean=5, sd=1),
  groupB = rnorm(20, mean=5.2, sd=1)
)

t_res <- t.test(dt_null$groupA, dt_null$groupB, var.equal=TRUE)
t_res
```
