---
letter: '#'
word: 'forall'
dictionary: 'mathematics'
category: 'logic'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['set','universal-quantification']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="forall">∀ (For all)</a> - The **forall** symbol is used in **logic** and set theory to denote “for all elements” in a [set](#set). For example:

$$
\forall x \in A, \quad P(x).
$$

This states that for every element x of set A, the proposition P(x) holds.

**Key points**:
- Central in **universal-quantification** statements.
- Often combined with \(\exists\) (there exists) to form more complex logical formulas.

No special R demonstration needed; it’s a logical/quantifier concept.  
```{r}
cat("∀ is used purely in logical statements: 'For all x in A...'")
```
