---
letter: 'z'
word: 'z-score'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['mean','variance','standard-deviation','distribution']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="z-score">Z-score</a> - In **statistics**, a **z-score** (or standard score) is the number of [standard-deviation](#standard-deviation)s by which the value of an observation differs from the [mean](#mean). For an observation $x$ in a sample:

$$
z = \frac{x - \bar{x}}{s},
$$

where $\bar{x}$ is the sample mean and $s$ is the sample SD (see [variance](#variance) / SD). In a normal [distribution](#distribution), z-scores help locate observations relative to the population mean.

**R demonstration**:

```{r}
library(data.table)

set.seed(123)
vals <- rnorm(10, mean=10, sd=2)
z_scores <- (vals - mean(vals)) / sd(vals)
z_scores
```
