---
letter: 'd'
word: derivative
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['dimension','determinant']
linkedFrom: ['calculus','chain-rule']
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
# Standard setup for rendering
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi       = 300,
  fig.width = 10,
  fig.height= 7
)

# We'll load data.table if needed later
```

<a id="derivative">Derivative</a> - In calculus, the **derivative** of a function $f(x)$ at a point $x = a$ measures the rate at which $f$ changes with respect to $x$.

Formally, the derivative $f'(a)$ is defined by:

$$
f'(a) = \lim_{h \to 0} \frac{f(a + h) - f(a)}{h}.
$$

**Key points**:
- Geometric interpretation: The slope of the tangent line to $y = f(x)$ at $x = a$.
- Practical interpretation: Instantaneous rate of change (e.g. velocity from position).

**Simple R demonstration** (numerical approximation):

```{r}
# We'll approximate the derivative of f(x) = x^2 at x=2 using a small h
f <- function(x) x^2

numeric_derivative <- function(f, a, h = 1e-5) {
  (f(a + h) - f(a)) / h
}

approx_deriv_2 <- numeric_derivative(f, 2)
actual_deriv_2 <- 2 * 2  # derivative of x^2 is 2x, so at x=2 it's 4

approx_deriv_2
actual_deriv_2
```

We see that $f'(2) = 4$ exactly, while our numeric approximation should be close to 4 for a suitably small $h$.

