---
letter: 'w'
word: 'wronskian'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['derivative','determinant']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="wronskian">Wronskian</a> - In differential equations, the **Wronskian** of two (or more) functions is the [determinant](#determinant) of a matrix whose entries are those functions and their [derivative](#derivative)s. For two functions $f$ and $g$:

'58445'
W(f,g)(x) =
\begin{vmatrix}
f(x) & g(x) \\
f'(x) & g'(x)
\end{vmatrix}.
'58445'

If the Wronskian is nonzero at some point, $f$ and $g$ are linearly independent solutions.

**R demonstration** (small numeric check):

```{r}
library(data.table)

f <- function(x) sin(x)
g <- function(x) cos(x)
df <- function(x) cos(x)
dg <- function(x) -sin(x)

wronskian <- function(x) {
  det( rbind( c(f(x), g(x)), c(df(x), dg(x)) ) )
}

x_test <- pi/4
W_val <- wronskian(x_test)
W_val
# We know sin'(x)=cos(x), cos'(x)=-sin(x); check numeric result
```
