---
letter: 's'
word: 'svm'
dictionary: 'mathematics'
category: 'machine-learning'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['kernel','margin']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x,options){
    cap<- options$fig.cap
    as.character(
      htmltools::tag("Figure",list(src=x,alt=cap,paste("\n\t",cap,"\n",sep="")))
    )
  })
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300,fig.width=10,fig.height=7)
```

<a id="svm">Support Vector Machine (SVM)</a> - A **support vector machine** is a powerful method for classification (and sometimes regression) that finds a maximal [margin](#margin) hyperplane separating classes in feature space. If classes are not linearly separable, it uses a [kernel](#kernel) trick to map into higher-dimensional spaces.

**Key points**:
- Minimises hinge loss for classification.
- Creates a decision boundary that maximises margin from support vectors.

**Mathematical form**:
For binary classification, we try to solve:
'60196'
\min_{w,b} \frac{1}{2}\|w\|^2 \quad \text{subject to} \quad y_i(w \cdot x_i + b) \ge 1 \quad \forall i.
'60196'

**R demonstration** (using `e1071` for a small SVM example):

```{r}
library(e1071)
library(data.table)

set.seed(123)
n <- 30
x1 <- runif(n, min=-3, max=3)
x2 <- runif(n, min=-3, max=3)
y_class <- ifelse(x2 > x1 + rnorm(n,0,1), "A","B")

dt_svm <- data.table(x1=x1, x2=x2, y=as.factor(y_class))
fit_svm <- svm(y ~ x1 + x2, data=dt_svm, kernel="linear", scale=TRUE)
summary(fit_svm)

# Grid for plotting
grid_x1 <- seq(-3,3,length.out=50)
grid_x2 <- seq(-3,3,length.out=50)
grid_data <- data.table(expand.grid(x1=grid_x1, x2=grid_x2))
pred_grid <- predict(fit_svm, newdata=grid_data)

plot(x2 ~ x1, data=dt_svm, col=y, pch=19,
     main="SVM Demo (linear kernel)")
contour(grid_x1, grid_x2, matrix(as.numeric(pred_grid),50), add=TRUE, levels=c(1.5))
```
