---
letter: 'k'
word: 'kruskal-algorithm'
dictionary: 'mathematics'
category: 'graph-theory'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['graph']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="kruskal-algorithm">Kruskal's Algorithm</a> - In **graph theory**, **Kruskal's algorithm** finds a minimum spanning tree (MST) of a weighted [graph](#graph) by:

1. Sorting edges in order of increasing weight.
2. Adding edges one by one to the MST, provided they do not form a cycle.
3. Repeating until all vertices are connected or edges are exhausted.

This greedy approach ensures an MST if the graph is connected. 

**R demonstration** (a small example with igraph):

```{r}
library(igraph)

# Create a weighted graph
g <- graph(edges=c("A","B","B","C","A","C","C","D","B","D"),
           directed=FALSE)
E(g)$weight <- c(2, 4, 5, 1, 3)  # just some weights

# Use built-in MST function that uses Kruskal internally
mst_g <- mst(g)
mst_g

# Let's plot
plot(mst_g, vertex.color="lightblue", vertex.size=30,
     edge.label=E(mst_g)$weight)
```
