---
letter: 'p'
word: 'proportion'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: []
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="proportion">Proportion</a> - In **statistics**, a **proportion** represents a fraction of the whole—essentially, how many observations fall into a particular category, divided by the total.  

'57323'
\text{proportion} = \frac{\text{count of interest}}{\text{total count}}.
'57323'

**Key points**:
- Used in categorical data analysis.
- Confidence intervals for a proportion use techniques like the Wald method, Wilson method, etc.

**R demonstration** (simple proportion of “heads” in coin flips):

```{r}
library(data.table)

set.seed(123)
flips <- sample(c("H","T"), 50, replace=TRUE)
prop_heads <- mean(flips=="H")
prop_heads
```
