---
letter: 'i'
word: integral
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

# We reference "derivative" in the definition, so we add it to linksTo.
linksTo: ['derivative']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="integral">Integral</a> - In **calculus**, an **integral** represents the accumulation of quantities or the area under a curve. It is the inverse operation to the [derivative](#derivative) (by the Fundamental Theorem of Calculus).

For a function $f(x)$, the definite integral from $a$ to $b$ is:

56131
\int_a^b f(x)\, dx.
56131

**Key points**:
- **Indefinite integral**: $\int f(x)\,dx = F(x) + C$, where $F'(x)=f(x)$.
- **Riemann sums** approximate integrals by partitioning the interval and summing “area slices.”

**R demonstration** (numeric approximation of an integral via trapezoidal rule):

```{r}
library(data.table)

f <- function(x) x^2
a <- 0
b <- 3
n <- 100
x_vals <- seq(a, b, length.out=n+1)
dx <- (b - a)/n
trapezoid <- sum((f(x_vals[-1]) + f(x_vals[-(n+1)]))/2) * dx
trapezoid  # approximate integral of x^2 from 0 to 3 = 9
```
