---
letter: 'g'
word: generating-function
dictionary: 'mathematics'
category: 'combinatorics'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['gamma-function','graph']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = "")))
    )
  }
)
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="generating-function">Generating Function</a> - A **generating function** is a formal power series whose coefficients encode information about a sequence $(a_n)$. For instance:

$$
G(x) = \sum_{n=0}^{\infty} a_n x^n.
$$

**Key points**:
- Used extensively in combinatorics to derive closed forms or recurrences.
- Different types (ordinary, exponential) depending on the combinatorial interpretation.

**Example**: The ordinary generating function for the sequence $\{1,1,1,\dots\}$ is

$$
\sum_{n=0}^{\infty} x^n = \frac{1}{1-x}, \quad |x|<1.
$$

No direct data.table example is typical here, but generating functions are used in discrete math. One might do symbolic manipulations with external packages. We can, however, illustrate partial sums numerically:

```{r}
library(data.table)

x_val <- 0.2
num_terms <- 10

dt_genfun <- data.table(
  n = 0:(num_terms-1)
)

dt_genfun[, term := x_val^n]
partial_sum <- dt_genfun[, sum(term)]
partial_sum

# Compare to closed-form 1/(1 - x_val)
closed_form <- 1 / (1 - x_val)
closed_form
```

