---
letter: 'k'
word: 'kernel'
dictionary: 'mathematics'
category: 'linear-algebra'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['dimension']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="kernel">Kernel</a> - In **linear algebra**, the **kernel** (or **null space**) of a linear map 
$$T: V \to W$$ is the set of all vectors $\mathbf{v} \in V$ such that $T(\mathbf{v}) = \mathbf{0}$. Symbolically,

56642
\ker(T) = \{ \mathbf{v} : T(\mathbf{v}) = \mathbf{0} \}.
56642

- If $T$ is given by a matrix $A$, then $\ker(T) = \{ \mathbf{x} : A\mathbf{x} = \mathbf{0} \}$.
- The rank-nullity theorem links the dimension of the kernel with the [dimension](#dimension) of the image.

**R demonstration** (finding the kernel of a matrix):

```{r}
library(data.table)

A <- matrix(c(1,2,3, 2,4,6, 1,1,2), nrow=3, byrow=TRUE)

# We can find the null space using MASS::Null
library(MASS)
kerA <- Null(A)  # basis for the kernel
kerA
```
