---
letter: 'k'
word: 'knn'
dictionary: 'mathematics'
category: 'machine-learning'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['distance','classification','regression']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(
    plot=function(x, options){
      cap<-options$fig.cap
      as.character(htmltools::tag("Figure",
        list(src=x, alt=cap, paste("\n\t",cap,"\n",sep=""))
      ))
    }
  )
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300,fig.width=10,fig.height=7)
```

<a id="knn">K-Nearest Neighbors (KNN)</a> - A **KNN** classifier (or regressor) predicts the label (or value) of a new point \(x_{new}\) by looking at the **k** closest points (in some [distance](#distance) metric) in the training set. For classification, it uses a majority vote among neighbors; for regression, it averages the neighbor values.

**Mathematical form** (for classification):
'60830'
\hat{y}(x_{new}) = \text{majority-label}\bigl( y_i : x_i \in \mathcal{N}_k(x_{new})\bigr),
'60830'
where \( \mathcal{N}_k(x_{new}) \) is the set of k nearest neighbors under a chosen distance (often Euclidean).

**R demonstration** (using `class::knn` for classification):

```{r}
library(class)
library(data.table)
library(ggplot2)

set.seed(123)
n <- 100
x1 <- runif(n, 0, 5)
x2 <- runif(n, 0, 5)
y <- ifelse(x1 + x2 + rnorm(n) > 5, "A","B")

dt_knn <- data.table(x1=x1, x2=x2, y=as.factor(y))

# We'll do a train/test split
train_idx <- sample(1:n, size=70)
train <- dt_knn[train_idx]
test  <- dt_knn[-train_idx]

train_x <- as.matrix(train[, .(x1,x2)])
train_y <- train$y
test_x  <- as.matrix(test[, .(x1,x2)])
true_y  <- test$y

pred_knn <- knn(train_x, test_x, cl=train_y, k=3)

accuracy <- mean(pred_knn == true_y)
accuracy

# Plot classification boundary
grid_x1 <- seq(0,5, by=0.1)
grid_x2 <- seq(0,5, by=0.1)
grid_data <- data.table(expand.grid(x1=grid_x1, x2=grid_x2))
grid_mat  <- as.matrix(grid_data[,.(x1,x2)])
grid_data[, pred := knn(train_x, grid_mat, cl=train_y, k=3)]

ggplot() +
  geom_tile(data=grid_data, aes(x=x1, y=x2, fill=pred), alpha=0.4) +
  geom_point(data=dt_knn, aes(x=x1, y=x2, color=y), size=2) +
  scale_fill_manual(values=c("A"="lightblue","B"="salmon")) +
  scale_color_manual(values=c("A"="blue","B"="red")) +
  labs(title="K-Nearest Neighbors (k=3)", x="x1", y="x2") +
  theme_minimal()
```
