---
letter: 'e'
word: eigenvalue
dictionary: 'mathematics'
category: 'linear-algebra'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['entropy','expectation']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.height = 7
)
```

<a id="eigenvalue">Eigenvalue</a> - In linear algebra, an **eigenvalue** of a square matrix $A$ is a scalar $\lambda$ such that there exists a nonzero vector \(\mathbf{v}\) (the **eigenvector**) satisfying:

55424
A\mathbf{v} = \lambda\mathbf{v}.
55424

**Key points**:
- Eigenvalues reveal important properties of linear transformations (e.g., scaling factors in certain directions).
- If $\lambda$ is an eigenvalue, then \(\mathbf{v}\) is an eigenvector corresponding to $\lambda$.
- The polynomial \(\det(A - \lambda I) = 0\) is the characteristic equation that yields eigenvalues.

**R demonstration** (finding eigenvalues of a 2x2 matrix):

```{r}
library(data.table)

# Create a data.table for matrix entries
dtA <- data.table(a=2, b=1, c=1, d=2)
A <- matrix(c(dtA$a, dtA$b, dtA$c, dtA$d), nrow=2, byrow=TRUE)
A

# Compute eigenvalues using base R
eigs <- eigen(A)
eigs$values
eigs$vectors
```

