---
letter: 'x'
word: 'x-coordinate'
dictionary: 'mathematics'
category: 'geometry'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['x-axis','coordinate-system']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure", 
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="x-coordinate">X-coordinate</a> - The **x-coordinate** of a point in a [coordinate-system](#coordinate-system) is its horizontal position, indicating how far left or right it is relative to the y-axis. Typically, an ordered pair is written as \((x, y)\), where $x$ is the x-coordinate.

**Key points**:
- Positive x-coordinates lie to the right of the y-axis.
- Negative x-coordinates lie to the left of the y-axis.
- See also [x-axis](#x-axis) for orientation.

Simple R demonstration (extracting x-coordinates from a data set):

```{r}
library(data.table)

dt_points <- data.table(x=seq(-2,2,1), y=rnorm(5))
dt_points
dt_points[, x]
```
