---
letter: 't'
word: 't-statistic'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['null-hypothesis','mean','variance']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="t-statistic">T-Statistic</a> - In **statistics**, the **t-statistic** is used in a t-test to compare a sample [mean](#mean) to a hypothesised population mean (under the [null-hypothesis](#null-hypothesis)), typically when population [variance](#variance) is unknown. For sample size \(n\):

'57989'
t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}},
'57989'

where \(\bar{x}\) is the sample mean, \(s\) is the sample standard deviation, and \(\mu_0\) is the hypothesised mean.

**R demonstration** (simple t-test):

```{r}
library(data.table)

set.seed(123)
x_vals <- rnorm(20, mean=10, sd=2)
res_ttest <- t.test(x_vals, mu=9)
res_ttest
```
