---
letter: 'y'
word: 'young-tableau'
dictionary: 'mathematics'
category: 'combinatorics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['partition']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="young-tableau">Young Tableau</a> - In **combinatorics** and representation theory, a **Young tableau** is a grid diagram (partition shape) where cells are filled with numbers (or symbols) that obey certain row/column ordering. A **standard Young tableau** uses distinct numbers $1,2,...,n$ that strictly increase left-to-right in rows and top-to-bottom in columns.

**Key points**:
- Related to [partition](#partition) of integers (diagrams shaped by partition lengths).
- Vital in symmetric function theory and representation theory of symmetric groups.

**R demonstration** (no base R function for Young tableaux, but we can show a small layout):

```{r}
library(data.table)

# We'll just define a small 'shape' as row lengths, e.g. partition (3,2)
# Then store a possible filling as a standard Young tableau

young_shape <- list(c(1,2,3), c(4,5))  # conceptual: row1=(1,2,3), row2=(4,5)
young_shape
```
