---
letter: 'o'
word: 'order-statistic'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

# Suppose we reference 'median' since the median is essentially the middle order statistic.
linksTo: ['median']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width=10, fig.height=7)
```

<a id="order-statistic">Order Statistic</a> - In **statistics**, an **order statistic** is one of the values in a sorted sample. Given $n$ data points, the $k$th **order statistic** is the $k$th smallest value. The [median](#median) is a well-known order statistic (middle value for odd $n$).

**Key points**:
- Distribution of order statistics helps in confidence intervals, extreme value theory.
- The minimum is the 1st order statistic, the maximum is the $n$th.

**R demonstration**:

```{r}
library(data.table)

set.seed(123)
x_vals <- sample(1:100, 10) 
dt_ord <- data.table(x = x_vals)
dt_ord_sorted <- dt_ord[order(x)]
dt_ord_sorted[, idx := .I]  # .I is row index in data.table
dt_ord_sorted
```
