---
letter: 'l'
word: 'laplace-transform'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

# We reference "integral" in the definition, so let's add it.
linksTo: ['integral']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="laplace-transform">Laplace Transform</a> - In **calculus**, the **Laplace transform** of a function $f(t)$ (for $t \ge 0$) is defined by the [integral](#integral):

56724
\mathcal{L}\{f(t)\}(s) = \int_0^\infty e^{-st} f(t)\, dt,
56724

assuming the integral converges.

**Key points**:
- Simplifies solving ordinary differential equations by converting them into algebraic equations in the $s$-domain.
- Inverse Laplace transform recovers $f(t)$ from \(\mathcal{L}\{f\}(s)\).

**R demonstration** (no base R function for Laplace transforms, but we can do numeric approximations or use external packages. We show a naive numeric approach for a simple function $e^{-t}$):

```{r}
library(data.table)

f <- function(t) exp(-t)
laplace_numeric <- function(f, s, upper=10, n=1000) {
  # naive numerical approach
  t_vals <- seq(0, upper, length.out=n)
  dt <- (upper - 0)/n
  sum( exp(-s * t_vals) * f(t_vals) ) * dt
}

s_test <- 2
approx_LT <- laplace_numeric(f, s_test, upper=10)
approx_LT
# The exact Laplace transform of e^{-t} is 1/(s+1). For s=2 => 1/3 ~ 0.3333
```
