---
letter: 'a'
word: algebra
dictionary: 'mathematics'
category: 'algebra'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['arithmetic','absolute-value','angle']
linkedFrom: ['asymptote']
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
# Common setup for rendering
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="algebra">Algebra</a> - Algebra is the branch of mathematics that studies symbols and the rules for manipulating them. It extends basic arithmetic by introducing variables to represent unknown or general quantities.

**Scopes of algebra**:
- **Elementary Algebra**:
  - Solving linear and quadratic equations
  - Factorising polynomials
  - Manipulating algebraic expressions
- **Abstract Algebra**:
  - **Groups**: A set with one operation satisfying closure, associativity, identity, and invertibility
  - **Rings**: A set with two operations (addition, multiplication) generalising integer arithmetic
  - **Fields**: A ring in which every nonzero element has a multiplicative inverse

**Example: Solving a linear system**:

$$
\begin{cases}
 x + 2y = 1, \\
 3x - y = 0
\end{cases}
$$

- We can rewrite this system in matrix form and solve it using methods from linear algebra.
- The matrix representation is:

$$
\begin{pmatrix}
 1 & 2 \\
 3 & -1
\end{pmatrix}
\begin{pmatrix}
 x \\
 y
\end{pmatrix}
=
\begin{pmatrix}
 1 \\
 0
\end{pmatrix}.
$$

- Solving $A\mathbf{x} = \mathbf{b}$ typically involves finding the inverse of $A$ (when it exists) or using other factorizations (LU, QR, etc.).

```{r}
A <- matrix(c(1, 2, 3, -1), nrow=2, byrow=TRUE)
b <- c(1, 0)
solve(A, b)
```

Algebra underpins higher mathematics, from geometry (coordinate systems) to analysis (manipulating series expansions) and number theory (factorisation, modular arithmetic).

