---
letter: 'r'
word: 'ring'
dictionary: 'mathematics'
category: 'algebra'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['group','field','algebra']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()) {
  knitr::knit_hooks$set(plot = function(x, options){
    cap <- options$fig.cap
    as.character(
      htmltools::tag("Figure",
        list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
      )
    )
  })
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="ring">Ring</a> - In **abstract algebra**, a **ring** is a set equipped with two binary operations (usually called addition and multiplication), satisfying:

1. (R, +) is an abelian [group](#group).
2. Multiplication is associative.
3. Distributive laws link the two operations:  
   $ a \times (b + c) = a\times b + a\times c$ and $(b + c)\times a = b\times a + c\times a$.

**Key points**:
- Rings may or may not have a multiplicative identity (1). If present, we say it’s a “ring with unity.”
- A [field](#field) is a ring where every nonzero element has a multiplicative inverse.

No direct R demonstration typical for ring structure, but we can mention integer arithmetic as a ring:

```{r}
cat("Integers under + and * form a classic ring.")
```
