---
letter: 'f'
word: 'field'
dictionary: 'mathematics'
category: 'algebra'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['ring','algebra']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()) {
  knitr::knit_hooks$set(plot = function(x, options){
    cap <- options$fig.cap
    as.character(
      htmltools::tag("Figure",
        list(src=x, alt=cap, paste("\n\t", cap, "\n", sep=""))
      )
    )
  })
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="field">Field</a> - In **abstract algebra**, a **field** is a [ring](#ring) in which every nonzero element has a multiplicative inverse. The real numbers $\mathbb{R}$ and rational numbers $\mathbb{Q}$ are classic examples of fields.

**Key points**:
- Both addition and multiplication exist and distribute.
- Every nonzero element is invertible under multiplication.
- Foundation of much of modern mathematics (vector spaces, linear algebra).

No direct R demonstration typical.  
```{r}
cat("Examples: ℚ, ℝ, ℂ all form fields with standard + and *.")
```
