---
letter: 'h'
word: hyperbola
dictionary: 'mathematics'
category: 'geometry'
dataSource: 'assistant'
published: true
comments: true

# We reference "asymptote" in the definition, so include it in linksTo.
linksTo: ['asymptote']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="hyperbola">Hyperbola</a> - A **hyperbola** is a conic section formed by the intersection of a plane and a double cone, but at a steeper angle than that of a parabola. In standard form, a hyperbola centered at the origin can be written as:

55983
\frac{x^2}{a^2} - \frac{y^2}{b^2} = 1
55983

(for the east-west opening case). Hyperbolas have two branches and characteristic [asymptote](#asymptote) lines that the hyperbola approaches as $|x|\to \infty$ or $|y|\to \infty$.

**R demonstration** (plotting a hyperbola segment in ggplot2):

```{r}
library(data.table)
library(ggplot2)

a <- 2
b <- 1
x_vals <- seq(-5, 5, by=0.01)

dt_hyp <- data.table(
  x = x_vals
)

# For x^2/a^2 - y^2/b^2 = 1 => y^2 = (x^2/a^2 - 1)*b^2
# We only plot real solutions (where x^2/a^2 > 1)

dt_hyp[, y_pos := ifelse(abs(x/a) > 1, b*sqrt((x^2/a^2) - 1), NA)]
dt_hyp[, y_neg := ifelse(abs(x/a) > 1, -b*sqrt((x^2/a^2) - 1), NA)]

ggplot() +
  geom_line(aes(x=x, y=y_pos), data=dt_hyp, color="blue") +
  geom_line(aes(x=x, y=y_neg), data=dt_hyp, color="blue") +
  coord_fixed() +
  labs(
    title="Hyperbola for x^2/4 - y^2/1 = 1",
    x="x", 
    y="y"
  ) +
  theme_minimal()
```
