---
letter: 's'
word: 'surjection'
dictionary: 'mathematics'
category: 'functions'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['injection','bijection']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()) {
  knitr::knit_hooks$set(plot = function(x, options){
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
    ))
  })
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="surjection">Surjection</a> - In **functions** (set theory), a **surjection** (or onto function) is a function $f: A \to B$ such that every element in $B$ has at least one preimage in $A$. Formally:

$$
\forall b \in B, \exists a \in A : f(a) = b.
$$

**Key points**:
- Every element of $B$ is “hit” by the function.
- Contrasts with [injection](#injection) (one-to-one). If a function is both surjective and injective, it’s a [bijection](#bijection).

**R demonstration** (not typical, but we can illustrate a partial concept with sets):

```{r}
library(data.table)

A <- 1:5
B <- letters[1:5]

# We'll define f: A->B, f(n) = the nth letter
f_surjective <- function(n) letters[n]

dt_map <- data.table(a = A, f_of_a = sapply(A, f_surjective))
dt_map
# This is surjective over B=letters[1:5], as all letters a through e appear.
```
