---
letter: 'x'
word: 'x-axis'
dictionary: 'mathematics'
category: 'geometry'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['coordinate-system','x-intercept']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="x-axis">X-axis</a> - In a **2D coordinate system**, the **x-axis** is the horizontal line used to measure the x-coordinate of points. Typically, positive values extend to the right from the origin, while negative values extend to the left.

**Key points**:
- Intersects the y-axis at the origin (0,0).
- The slope of the x-axis is 0 (a horizontal line).
- Used to define an [x-intercept](#x-intercept) for curves.

No direct R demonstration needed, but we can show a basic plot:

```{r}
library(ggplot2)
library(data.table)

dt_plot <- data.table(x=-5:5, y=0)
ggplot(dt_plot, aes(x=x, y=y)) +
  geom_line() +
  labs(
    title="X-axis (horizontal)",
    x="x",
    y="y"
  ) +
  theme_minimal()
```
