---
letter: 'n'
word: 'non-euclidean-geometry'
dictionary: 'mathematics'
category: 'geometry'
dataSource: 'assistant'
published: true
comments: true
linksTo: []
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="non-euclidean-geometry">Non-Euclidean Geometry</a> - In **geometry**, **non-Euclidean geometry** refers to either hyperbolic or elliptic geometry (or others) that reject or modify Euclid’s fifth postulate (the parallel postulate). 

**Key points**:
- **Hyperbolic geometry**: infinite lines diverge more rapidly, sums of angles in triangles are < 180°.
- **Elliptic geometry**: lines “curve,” angles in triangles sum to > 180°.

No standard R demonstration, but we might explore transformations or plots for illustrative geometry.  

```{r}
# No direct numeric example, but let's just place a note:
cat("No direct numeric example for non-Euclidean geometry in base R. Consider specialized geometry packages or external tools.")
```
