---
letter: 'g'
word: group
dictionary: 'mathematics'
category: 'algebra'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['gamma-function','graph']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = "")))
    )
  }
)
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="group">Group</a> - In **abstract algebra**, a **group** is a set $G$ together with a binary operation $\cdot$ satisfying:

1. **Closure**: For all $a,b \in G$, $a \cdot b \in G$.
2. **Associativity**: $ (a \cdot b) \cdot c = a \cdot (b \cdot c)$.
3. **Identity**: There exists $e \in G$ such that $e \cdot a = a \cdot e = a$ for all $a \in G$.
4. **Inverse**: For each $a \in G$, there exists $a^{-1} \in G$ with $a \cdot a^{-1} = a^{-1} \cdot a = e$.

**Examples**:
- Integers under addition
- Nonzero real numbers under multiplication
- Symmetry groups in geometry

While direct R demonstration is less obvious, one could illustrate a finite group:

```{r}
library(data.table)

# Let's define a small group table: Z2 x Z2 (Klein group), with elements { (0,0), (0,1), (1,0), (1,1) } under addition mod 2
# We'll store all results in a data.table
dt_group <- data.table(
  elem1 = c("(0,0)","(0,0)","(1,0)","(1,0)","(0,1)","(0,1)","(1,1)","(1,1)"),
  elem2 = c("(0,0)","(1,0)","(0,0)","(1,0)","(0,1)","(1,1)","(0,1)","(1,1)")
)

# We'll define a small function that "adds" these pairs mod 2
add_mod2_pairs <- function(a, b) {
  # parse, e.g. "(1,0)"
  a_vals <- as.integer(unlist(strsplit(gsub("[()]", "", a), ",")))
  b_vals <- as.integer(unlist(strsplit(gsub("[()]", "", b), ",")))
  
  sum_vals <- (a_vals + b_vals) %% 2
  paste0("(", sum_vals[1], ",", sum_vals[2], ")")
}

dt_group[, result := mapply(add_mod2_pairs, elem1, elem2)]
dt_group
```

This small table demonstrates closure in the group. Associativity, identity, and inverses also hold, though not explicitly shown here.

