---
letter: 'l'
word: 'laplacian'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

# We'll reference "dimension" from geometry or linear algebra context.
linksTo: ['dimension']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="laplacian">Laplacian</a> - In **multivariable calculus**, the **Laplacian** of a scalar function $f(x_1, x_2, \dots, x_n)$ is denoted by $\nabla^2 f$ or $\Delta f$, and is defined as:

$$
\nabla^2 f = \sum_{k=1}^n \frac{\partial^2 f}{\partial x_k^2}.
$$

- In 2D: $\nabla^2 f = \frac{\partial^2 f}{\partial x^2} + \frac{\partial^2 f}{\partial y^2}$.
- In 3D: $\nabla^2 f = \frac{\partial^2 f}{\partial x^2} + \frac{\partial^2 f}{\partial y^2} + \frac{\partial^2 f}{\partial z^2}$.
- The concept generalises to higher [dimension](#dimension).  
- The Laplacian is crucial in PDEs like the heat equation and wave equation.

No direct R built-in for second partial derivatives numerically, but we can approximate:

```{r}
library(data.table)

f_xy <- function(x, y) x^2 + y^2

laplacian_approx <- function(f, x, y, h=1e-4) {
  # second partial w.r.t x
  f_xph <- f(x+h, y); f_xmh <- f(x-h, y); f_xyc <- f(x, y)
  d2f_dx2 <- (f_xph - 2*f_xyc + f_xmh)/(h^2)
  
  # second partial w.r.t y
  f_yph <- f(x, y+h); f_ymh <- f(x, y-h)
  d2f_dy2 <- (f_yph - 2*f_xyc + f_ymh)/(h^2)
  
  d2f_dx2 + d2f_dy2
}

laplacian_approx(f_xy, 2, 3)
# For f(x,y)= x^2 + y^2, exact Laplacian = 2 + 2 = 4
```
