---
letter: 'v'
word: 'variance'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['standard-deviation','distribution','mean']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="variance">Variance</a> - In **statistics**, **variance** measures how far a set of numbers (random variable outcomes) spreads out around its [mean](#mean). For a population with values $x_1, x_2, ..., x_n$ and mean $\mu$:

$$
\mathrm{Var}(X) = \frac{1}{n} \sum_{i=1}^n (x_i - \mu)^2.
$$

For a sample-based estimate, the usual unbiased form has denominator $(n-1)$. The square root of variance is the [standard-deviation](#standard-deviation). Variance also helps describe a [distribution](#distribution)’s spread.

**R demonstration** (computing variance):

```{r}
library(data.table)

set.seed(123)
vals <- rnorm(30, mean=5, sd=2)
pop_var   <- mean( (vals - mean(vals))^2 )   # population style
sample_var <- var(vals)  # sample style, uses denominator (n-1)
c(pop_var, sample_var)
```
