---
letter: 'f'
word: frequency
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['factorial','fractal']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.height= 7
)
```

<a id="frequency">Frequency</a> - **Frequency** in mathematics and statistics can refer to:

1. **Statistical frequency**: How often a value appears in a dataset.
2. **Periodic phenomenon**: The number of cycles per unit time (e.g., in sine waves, signals).

**Statistical frequency**:
- Relative frequency = count of event / total observations.
- Frequency table is a basic summary in data analysis.

**Periodic frequency** (in signals):
- If $f(t) = \sin(2\pi \, \nu \, t)$, then $\nu$ is the frequency in cycles per unit time.

**R demonstration** (calculating frequencies in a categorical dataset):

```{r}
library(data.table)

# Suppose a small categorical variable
dt_freq <- data.table(
  category = c("A", "B", "A", "C", "B", "A", "B", "B", "C")
)

# Frequency count
freq_table <- dt_freq[, .N, by=category]
setnames(freq_table, "N", "count")

freq_table[, rel_freq := count / sum(count)]
freq_table
```

