---
letter: 'd'
word: discrete-random-variable
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['distribution','derivative']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi       = 300,
  fig.width = 10,
  fig.height= 7
)
```

<a id="discrete-random-variable">Discrete Random Variable</a> - A **discrete random variable** is one that takes on a countable set of values (often integers). Typical examples include:

- **Number of heads** in $n$ coin tosses
- **Number of customers** arriving at a store in an hour (Poisson process)

**Probability Mass Function (pmf)** for a discrete random variable $X$:

$$
P(X = x) = p(x),
$$

where $\sum_x p(x) = 1$ over all possible $x$.

**R demonstration** (creating a binomial discrete variable):

```{r}
library(data.table)
library(ggplot2)

# Suppose X ~ Binomial(n=10, p=0.3)
n <- 10
p <- 0.3
num_sims <- 1000

# Generate 1000 realisations of X
dt_binom <- data.table(
  X = rbinom(num_sims, size=n, prob=p)
)

# Plot distribution
ggplot(dt_binom, aes(x=factor(X))) +
  geom_bar(fill="lightgreen", colour="black") +
  labs(
    title="Simulation of Binomial(10, 0.3)",
    x="Number of successes",
    y="Frequency"
  ) +
  theme_minimal()
```

