---
letter: 'm'
word: 'mutually-exclusive-events'
dictionary: 'mathematics'
category: 'probability'
dataSource: 'assistant'
published: true
comments: true

# We reference "probability" or "union". Let's say we reference "union".
linksTo: ['union']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="mutually-exclusive-events">Mutually Exclusive Events</a> - In **probability**, two events $A$ and $B$ are **mutually exclusive** (or disjoint) if they cannot happen simultaneously:

56872
A \cap B = \varnothing.
56872

In other words, $P(A \cap B) = 0$. The [union](#union) of mutually exclusive events has a probability that’s just the sum of their individual probabilities:

56872
P(A \cup B) = P(A) + P(B),
56872

since $A$ and $B$ never overlap.

**R demonstration**: no direct R function, but we can illustrate logic:

```{r}
# Suppose events are flipping a coin:
# A = heads, B = tails
# A and B are mutually exclusive.

# We can do a small simulation
set.seed(123)
flips <- sample(c("H","T"), size=100, replace=TRUE)
mean(flips == "H")  # approximate P(A)
mean(flips == "T")  # approximate P(B)
# Overlap: none, because a single flip can't be both H and T
```
