---
letter: 'l'
word: 'likelihood'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

# We'll reference "distribution" in the definition.
linksTo: ['distribution']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="likelihood">Likelihood</a> - In **statistics**, the **likelihood** function measures how well a given model parameter explains observed data. It’s similar to a [distribution](#distribution) but viewed from the parameter’s perspective:

- For data $x$ and parameter $\theta$, the likelihood $L(\theta | x)$ is often expressed as $p(x | \theta)$, the probability of observing $x$ given $\theta$.

**Key points**:
- Maximum likelihood estimation chooses $\hat\theta$ that maximises $L(\theta | x)$.
- Log-likelihood is commonly used for convenience: $\ell(\theta) = \log L(\theta|x)$.

**R demonstration** (fitting a simple normal likelihood):

```{r}
library(data.table)

set.seed(123)
x_data <- rnorm(50, mean=5, sd=2)

lik_fun <- function(mu, sigma, x) {
  # Normal pdf for each x, product as likelihood
  # i.e. prod(dnorm(x, mean=mu, sd=sigma))
  # We'll return negative log-likelihood for convenience
  -sum(dnorm(x, mean=mu, sd=sigma, log=TRUE))
}

# We'll do a quick grid search
mu_seq <- seq(4, 6, by=0.1)
sigma_seq <- seq(1, 3, by=0.1)
res <- data.table()
for(m in mu_seq) {
  for(s in sigma_seq) {
    nll <- lik_fun(m, s, x_data)
    res <- rbind(res, data.table(mu=m, sigma=s, nll=nll))
  }
}
res_min <- res[which.min(nll)]
res_min
```
