---
letter: 'a'
word: 'algorithmic-trading'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['markov-chain','machine-learning','high-frequency-trading']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x,options){
    cap<-options$fig.cap
    as.character(
      htmltools::tag("Figure",
        list(src=x, alt=cap, paste("\n\t",cap,"\n",sep=""))
      )
    )
  })
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="algorithmic-trading">Algorithmic Trading</a> - **Algorithmic trading** uses computer-driven strategies to execute orders based on predefined rules (e.g. technical indicators, [machine-learning](#machine-learning) models, or risk constraints). Key components include:

- Strategy design (signal generation, risk mgmt).
- Implementation (order routing, execution algorithms).
- Monitoring & backtesting.

**Mathematical aspects**:
- Strategies often rely on time-series analysis, probability models (e.g. [Markov-chain](#markov-chain)), or advanced ML.
- Execution can aim to minimize market impact and transaction costs.

**R demonstration** (A minimal example of a “toy” daily returns strategy):

```{r}
library(data.table)
set.seed(123)

# Suppose we have daily returns from a random walk
n <- 200
returns <- rnorm(n, mean=0.0005, sd=0.01)

dt_algo <- data.table(day=1:n, ret=returns)
# A naive “signal”: if ret>0 => “go long next day”, else “stay out”
signal <- sign(dt_algo$ret)  # +1 or -1 or 0
# We'll track PnL by day (with a 1-day lag in signal if we want)
pnls <- shift(signal, fill=0, type="lag")* dt_algo$ret
cumulative_pnl <- cumsum(pnls)

dt_algo[, signal := shift(signal,1,0)]
dt_algo[, pnl := pnls]
dt_algo[, cumPNL := cumulative_pnl]
tail(dt_algo)

# Quick plot
plot(dt_algo$day, dt_algo$cumPNL, type="l",
     main="Toy Algo Trading Strategy (Sign of Return)",
     xlab="Day", ylab="Cumulative PnL")
```

