---
letter: 'm'
word: 'matrix'
dictionary: 'mathematics'
category: 'linear-algebra'
dataSource: 'assistant'
published: true
comments: true

# We reference "determinant" in the definition, so we add it to linksTo.
linksTo: ['determinant']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="matrix">Matrix</a> - A **matrix** is a rectangular array of numbers (or more abstract objects) arranged in rows and columns. Matrices are fundamental in [determinant](#determinant) calculations, linear transformations, and a variety of applications:

$$
\begin{pmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{m1} & a_{m2} & \cdots & a_{mn}
\end{pmatrix}.
$$

**Key operations**:
- Addition and scalar multiplication (element-wise).
- Matrix multiplication.
- Transposition and inversion (if square and invertible).

**R demonstration** (basic matrix creation and operations):

```{r}
library(data.table)

A <- matrix(c(1,2,3,4,5,6), nrow=2, byrow=TRUE)
B <- matrix(c(10,20,30,40,50,60), nrow=2, byrow=TRUE)

A_plus_B <- A + B
A_times_B <- A %*% t(B)  # 2x3 %*% 3x2 => 2x2

A_plus_B
A_times_B
```
