---
letter: 'r'
word: 'real-number'
dictionary: 'mathematics'
category: 'analysis'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['interval','number-line']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="real-number">Real Number</a> - A **real number** is a value on the continuous [number-line](#number-line), including both rational and irrational numbers. The set of real numbers is typically denoted \(\mathbb{R}\). Intervals (see [interval](#interval)) are subsets of real numbers.

**Key points**:
- Complete ordered field: every Cauchy sequence converges in \(\mathbb{R}\).
- Used in nearly all continuous mathematics contexts (calculus, analysis, measurement).

No special R demonstration: standard R numeric types approximate real numbers (double precision floats).

```{r}
library(data.table)
vals <- c(pi, sqrt(2), exp(1))
vals
```
