---
letter: 'i'
word: identity-matrix
dictionary: 'mathematics'
category: 'linear-algebra'
dataSource: 'assistant'
published: true
comments: true

# Suppose we reference "determinant" in the definition with a jump link,
# so we include it in the linksTo array:
linksTo: ['determinant']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="identity-matrix">Identity Matrix</a> - In linear algebra, the **identity matrix** $I_n$ is an $n \times n$ square matrix with ones on the main diagonal and zeros elsewhere:

$$
I_n = 
\begin{pmatrix}
1 & 0 & \cdots & 0 \\
0 & 1 & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & 1
\end{pmatrix}.
$$

**Key points**:
- $I_n$ serves as the multiplicative identity for matrices: $I_n A = A I_n = A$.
- Its [determinant](#determinant) is 1 for all $n$.
- Invertible matrices always have an identity matrix (the “unit” of their multiplicative structure).

**R demonstration** (creating identity matrices):

```{r}
library(data.table)

I2 <- diag(2)
I3 <- diag(3)

I2
I3
```
