---
letter: 'r'
word: 'rational-number'
dictionary: 'mathematics'
category: 'analysis'
dataSource: 'assistant'
published: true
comments: true

linksTo: []
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="rational-number">Rational Number</a> - In **analysis/number theory**, a **rational number** is a number that can be expressed as a fraction $\frac{p}{q}$ with integers $p$ and $q \neq 0$. The set of all rational numbers is typically denoted $\mathbb{Q}$.

**Key points**:
- Dense in the real line but countable.
- Opposite concept: irrational numbers (like $\sqrt{2}$ or $\pi$) cannot be written as a fraction of integers.

Minimal R demonstration: standard numeric types are floating approximations. No direct “rational type” in base R, though external packages exist:

```{r}
library(data.table)
# We'll just store p, q pairs
rat <- function(p,q) p/q

rat(3,2)
rat(22,7)  # approximate 22/7
```
