---
letter: 'j'
word: 'joint-distribution'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

# Suppose we reference "distribution" or "covariance",
# so let's say we add "distribution" in linksTo.
linksTo: ['distribution']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi=300,
  fig.width=10,
  fig.height=7
)
```

<a id="joint-distribution">Joint Distribution</a> - In **statistics**, a **joint distribution** describes the probability distribution of two or more random variables simultaneously. If $X$ and $Y$ are two random variables:

- **Joint pmf** (discrete case): $p_{X,Y}(x,y)$
- **Joint pdf** (continuous case): $f_{X,Y}(x,y)$

It extends the idea of a single-variable [distribution](#distribution) to multiple dimensions.

**R demonstration** (bivariate normal sampling):

```{r}
library(MASS)  # for mvrnorm
library(data.table)
library(ggplot2)

Sigma <- matrix(c(1, 0.5, 0.5, 1), 2, 2)  # Cov matrix
mu <- c(0, 0)
set.seed(123)
dt_joint <- data.table(
  mvrnorm(n=1000, mu=mu, Sigma=Sigma)
)
setnames(dt_joint, c("V1","V2"), c("X","Y"))

# Plot joint distribution via scatter plot
ggplot(dt_joint, aes(x=X, y=Y)) +
  geom_point(alpha=0.5) +
  labs(
    title="Bivariate Normal Scatter",
    x="X",
    y="Y"
  ) +
  theme_minimal()
```
