---
letter: 'z'
word: 'z-transform'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['fourier-transform']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="z-transform">Z-transform</a> - In **discrete-time signal processing**, the **Z-transform** is analogous to the [fourier-transform](#fourier-transform) but for discrete sequences. For a sequence \(x[n]\), the Z-transform is

$$
X(z) = \sum_{n=-\infty}^{\infty} x[n] \, z^{-n}.
$$

**Key points**:
- Used to analyse and design discrete control systems and filters.
- Regions of convergence define when the series converges.

No direct R demonstration in base for Z-transform, but we can show a conceptual example:

```{r}
library(data.table)

# We'll define a small discrete sequence
x <- c(1, 2, 3, 4)
# A naive partial "z-transform" style sum for illustration
Z_transform_partial <- function(x, z) {
  n_range <- seq_along(x)-1  # 0-based indexing
  sum( x * z^(-n_range) )
}

Z_transform_partial(x, 1.1)
```
