---
letter: '#'
word: 'partial-derivative-symbol'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['partial-derivative']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="partial-derivative-symbol">∂ (Partial derivative symbol)</a> - The **∂** symbol indicates a partial derivative, as in $\frac{\partial f}{\partial x}$. It generalises differentiation to functions of multiple variables (see [partial-derivative](#partial-derivative)).

**Key points**:
- ∂ is distinct from d in single-variable calculus.
- Common in PDEs (partial differential equations).

No direct R demonstration needed here, but partial derivatives can be approximated numerically.

```{r}
library(data.table)

f_xy <- function(x,y) x^2 + 3*x*y
partial_x <- function(f, x, y, h=1e-6) (f(x+h,y)-f(x,y))/h
partial_x(f_xy, 2, 3)
```
