---
letter: 'i'
word: 'injection'
dictionary: 'mathematics'
category: 'functions'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['bijection','surjection']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
    ))
  })
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="injection">Injection</a> - In **functions** (set theory), an **injection** (or one-to-one function) is a function $f: A \to B$ such that different elements of $A$ always map to different elements of $B$. Formally:

$$
f(a_1) = f(a_2) \implies a_1 = a_2.
$$

**Key points**:
- No two distinct elements in $A$ share the same image in $B$.
- Contrasts with [surjection](#surjection) (onto) and [bijection](#bijection) (one-to-one and onto).

**R demonstration** (not typical, but we can check uniqueness in a numeric map):

```{r}
library(data.table)

f_injective <- function(x) x^2  # for integers, watch out for collisions at +/-x
x_vals <- c(-2,-1,0,1,2)
f_vals <- f_injective(x_vals)
data.table(x=x_vals, f=f_vals)
# Notice that x^2 is not injective over all integers (f(-2)=f(2)). 
# But restricted to nonnegative x, it can be injective.
```
