---
letter: 'c'
word: covariance
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['calculus']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="covariance">Covariance</a> - In statistics and probability theory, **covariance** measures the joint variability of two random variables $X$ and $Y$:

54861
\mathrm{Cov}(X, Y) = E\bigl[(X - E[X])(Y - E[Y])\bigr].
54861

**Key observations**:
- If $X$ and $Y$ tend to increase together, covariance is positive.
- If one tends to increase when the other decreases, covariance is negative.
- A covariance of zero does not necessarily imply independence (unless under specific conditions, like normality).

**Example in R**:
```{r}
set.seed(123)
X <- rnorm(10, mean=5, sd=2)
Y <- rnorm(10, mean=7, sd=3)
cov(X, Y)
```

Covariance forms the basis of **correlation** (a normalised version of covariance) and is central in statistics (e.g., linear regression, portfolio variance in finance).

