---
letter: 'j'
word: 'jordan-normal-form'
dictionary: 'mathematics'
category: 'linear-algebra'
dataSource: 'assistant'
published: true
comments: true

# Suppose we reference "eigenvalue" in the definition,
# so we add it to linksTo.
linksTo: ['eigenvalue']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi=300,
  fig.width=10,
  fig.height=7
)
```

<a id="jordan-normal-form">Jordan Normal Form</a> - In **linear algebra**, the **Jordan normal form** (or Jordan canonical form) of a matrix is a block diagonal matrix with Jordan blocks, each corresponding to an [eigenvalue](#eigenvalue).

A **Jordan block** for an eigenvalue $\lambda$ looks like:

$$
\begin{pmatrix}
\lambda & 1 & 0 & \cdots & 0 \\
0 & \lambda & 1 & \cdots & 0 \\
\vdots & & \ddots & \ddots & \vdots \\
0 & 0 & \cdots & \lambda & 1 \\
0 & 0 & \cdots & 0 & \lambda 
\end{pmatrix}.
$$

The Jordan form classifies matrices up to similarity transformations and is critical in solving systems of linear differential equations and more.

**R demonstration** (no built-in base R function to compute Jordan form, but we can show a small example):

```{r}
library(data.table)
# Usually, packages like 'jord' or 'expm' might help. 
# We'll just illustrate a 2x2 Jordan block for eigenvalue 3:

J <- matrix(c(3,1,0,3), 2, 2, byrow=TRUE)
J
```
