---
letter: 'd'
word: determinant
dictionary: 'mathematics'
category: 'linear-algebra'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['derivative','dimension']
linkedFrom: ['algebra']
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi       = 300,
  fig.width = 10,
  fig.height= 7
)
```

<a id="determinant">Determinant</a> - For a square matrix $A$, the **determinant** is a scalar that can be computed from the elements of $A$. It provides important information:

- $\det(A) = 0$ indicates $A$ is not invertible (singular).
- $\det(A) \neq 0$ indicates $A$ is invertible (nonsingular).
- Geometrically, for a 2D matrix, the absolute value of the determinant gives the area scaling factor of the linear transformation represented by $A$.

**For a 2×2 matrix**:

$$
A = 
\begin{pmatrix}
a & b \\
c & d
\end{pmatrix},
\quad
\det(A) = ad - bc.
$$

**Example in R**:

```{r}
library(data.table)

# We'll create a small data.table of matrix elements
dt <- data.table(
  a = 2,
  b = 1,
  c = 1,
  d = 3
)

# Convert dt to a matrix
A <- matrix(c(dt$a, dt$b, dt$c, dt$d), nrow=2, byrow=TRUE)

det_A <- det(A)
det_A
```

