---
letter: 'h'
word: histogram
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

# We reference "distribution" in the definition, so add it to linksTo.
linksTo: ['distribution']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", 
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="histogram">Histogram</a> - A **histogram** is a graphical representation of the distribution of numerical data. It groups data into bins (intervals) and displays the count or frequency within each bin, providing a quick visual of how values are spread.

It’s directly related to a [distribution](#distribution) in statistics, visually summarising the frequency or relative frequency of data within specified intervals.

**R demonstration** (constructing a histogram):

```{r}
library(data.table)
library(ggplot2)

set.seed(123)
dt_hist <- data.table(x = rnorm(500, mean=10, sd=2))

ggplot(dt_hist, aes(x=x)) +
  geom_histogram(bins=30, fill="lightblue", colour="black") +
  labs(title="Histogram of Random Normal Data",
       x="Value",
       y="Count") +
  theme_minimal()
```
