---
letter: 'q'
word: 'quaternion'
dictionary: 'mathematics'
category: 'algebra'
dataSource: 'assistant'
published: true
comments: true

linksTo: []
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="quaternion">Quaternion</a> - In **algebra**, a **quaternion** is a hypercomplex number of the form

$$
q = a + b\mathbf{i} + c\mathbf{j} + d\mathbf{k},
$$

where $a,b,c,d \in \mathbb{R}$ and \(\mathbf{i}, \mathbf{j}, \mathbf{k}\) follow certain multiplication rules (\(\mathbf{i}^2=\mathbf{j}^2=\mathbf{k}^2= \mathbf{i}\mathbf{j}\mathbf{k} = -1\)).

**Key points**:
- Noncommutative: \(\mathbf{i}\mathbf{j} = \mathbf{k}\) but \(\mathbf{j}\mathbf{i} = -\mathbf{k}\).
- Used in 3D rotations (e.g., in computer graphics, robotics).

No direct base R demonstration of quaternions, but certain libraries handle them. Let’s just illustrate we can store them as a list:

```{r}
library(data.table)

# We'll store quaternions as lists, no direct operation
q1 <- list(a=1, b=2, c=3, d=-1)
q1
```
