---
letter: 'y'
word: 'yates-correction'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['chi-squared-test','null-hypothesis']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="yates-correction">Yates' Correction</a> - In **statistics**, **Yates' correction** (also called **Yates' continuity correction**) is applied to a [chi-squared-test](#chi-squared-test) for 2×2 contingency tables to reduce bias when sample sizes are small. It adjusts the observed frequencies before computing the chi-squared statistic:

$$
\chi^2 = \sum\frac{(|O - E| - 0.5)^2}{E}.
$$

**Key points**:
- Often recommended if expected frequencies are < 5, though its usage is debated.
- Under the [null-hypothesis](#null-hypothesis), the chi-squared distribution still approximates the test statistic.

No built-in function for Yates-correction in base R, but some `chisq.test` wrappers handle it, or you can manually apply the formula.

```{r}
library(data.table)

# Example of manually applying Yates' correction (conceptual)
observed <- matrix(c(10,4,3,12), nrow=2)
expected <- rowSums(observed) %o% colSums(observed) / sum(observed)

chisq_yates <- sum((abs(observed - expected) - 0.5)^2 / expected)
chisq_yates
```
