---
letter: '#'
word: 'infinity'
dictionary: 'mathematics'
category: 'analysis'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['limit','extended-real-number-line']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="infinity">∞ (Infinity)</a> - In **analysis**, **infinity** (∞) symbolises an unbounded value that grows beyond any finite number. It often appears in contexts of [limit](#limit) and the [extended-real-number-line](#extended-real-number-line).

**Key points**:
- Not a real number in standard arithmetic; it’s a conceptual extension.
- In set theory, there are different *sizes* (cardinalities) of infinity (ℵ₀, 2^ℵ₀, etc.).
- In calculus, writing $x \to ∞$ indicates x grows without bound.

**R demonstration** (conceptual usage of Inf in R):

```{r}
library(data.table)

val_inf <- Inf
val_inf
# Basic check
is.finite(val_inf)  # should be FALSE
```
