---
letter: '#'
word: 'exists'
dictionary: 'mathematics'
category: 'logic'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['forall']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="exists">∃ (There exists)</a> - The **exists** symbol is used in **logic** to express the existence of at least one element satisfying a property. For example:

$$
\exists x \in A : P(x).
$$

means there is at least one x in A for which P(x) holds true.

**Key points**:
- Often used with [forall](#forall) statements to form more complex logical conditions.
- Symbolically pairs with optional uniqueness, e.g. “there exists exactly one.”

No direct R demonstration; it’s a logical concept.  
```{r}
cat("∃ is used in statements like 'There exists x in A such that...'")
```
