---
letter: 'p'
word: 'partial-derivative'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

# We reference 'derivative' or 'gradient' from previous definitions if desired:
linksTo: ['derivative','gradient']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="partial-derivative">Partial Derivative</a> - In **multivariable calculus**, a **partial derivative** of a function
'57323'
f(x_1, x_2, \dots, x_n)
'57323'
with respect to $x_k$ is the derivative treating $x_k$ as the only variable, holding others constant:

'57323'
\frac{\partial f}{\partial x_k}(x_1,\dots,x_n) = 
\lim_{h \to 0} \frac{f(\dots, x_k + h, \dots) - f(\dots, x_k, \dots)}{h}.
'57323'

**Key points**:
- Used in computing the [gradient](#gradient).
- The concept generalises [derivative](#derivative) to higher dimensions.

**R demonstration** (numerical approximation for $f(x,y)= x^2 + 2xy$ wrt $x$):

```{r}
library(data.table)

f_xy <- function(x,y) x^2 + 2*x*y

partial_x <- function(f, x, y, h=1e-6) {
  (f(x+h, y) - f(x, y)) / h
}

val <- partial_x(f_xy, 2, 3)
val
# Compare to analytic partial derivative wrt x: 2x + 2y. 
# At (2,3) => 2*2 + 2*3=4+6=10
```
