---
letter: 'd'
word: distribution
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['discrete-random-variable','covariance']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi       = 300,
  fig.width = 10,
  fig.height= 7
)
```

<a id="distribution">Distribution</a> - In probability and statistics, a **distribution** describes how values of a random variable are spread out. It can be specified by a **probability density function (pdf)** for continuous variables or a **probability mass function (pmf)** for discrete variables.

**Common examples**:
- **Normal distribution**: $X \sim N(\mu, \sigma^2)$
- **Binomial distribution**: Counts successes in $n$ independent Bernoulli trials
- **Poisson distribution**: Counts events in a fixed interval with known average rate

**R demonstration** (sampling from a normal distribution and visualising via ggplot2):

```{r}
library(data.table)
library(ggplot2)

# Create a data.table with 1000 random N(0,1) values
dt_dist <- data.table(
  x = rnorm(1000, mean=0, sd=1)
)

# Plot a histogram
ggplot(dt_dist, aes(x=x)) +
  geom_histogram(bins=30, colour="black", fill="skyblue") +
  geom_density(aes(y=..count..), colour="red", size=1) +
  labs(
    title="Histogram & Density for N(0,1)",
    x="Value",
    y="Count/Density"
  ) +
  theme_minimal()
```

