---
letter: 'c'
word: 'cointegration'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['time-series','stationarity','markov-chain']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x,options){
    cap<-options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src=x, alt=cap, paste("\n\t",cap,"\n",sep="")))
    )
  })
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="cointegration">Cointegration</a> - In **time-series** analysis, two (or more) series \(X_t, Y_t\) are **cointegrated** if a linear combination \(\alpha X_t + \beta Y_t\) is **stationary**, even if \(X_t\) and \(Y_t\) themselves are individually non-stationary (e.g., I(1) processes).

**Key formula**:
A common approach is to test:
'72077'
Z_t = X_t - \beta Y_t
'72077'
for stationarity (via the Engle–Granger two-step method or the Johansen test).

**Finance usage**:
- Pairs trading strategies rely on cointegration between asset prices.
- If they deviate from their equilibrium spread, trade accordingly.

**R demonstration** (synthetic pairs data with partial cointegration):

```{r}
library(data.table)
library(egcm) # Engle-Granger cointegration method if installed
set.seed(123)

n <- 300
x <- cumsum(rnorm(n, mean=0.01, sd=0.05)) + 100  # random walk
y <- 0.5*x + cumsum(rnorm(n, mean=0, sd=0.02))  # partially following x + noise

dt_coin <- data.table(time=1:n, X=x, Y=y)

# We'll use 'egcm' if installed, or just do a quick "lm then test residual" approach
if(requireNamespace("egcm", quietly=TRUE)){
  res_egcm <- egcm::egcm(x, y)
  print(res_egcm)
} else {
  cat("Package 'egcm' not installed, skipping direct cointegration test.\n")
}

# We'll do a quick correlation approach
fit_lm <- lm(Y ~ X, data=dt_coin)
summary(fit_lm)
resid <- fit_lm
# One might do an ADF test on resid for stationarity, skipping code if package not installed

# Plot
library(ggplot2)
dt_coin[, residual := resid]
ggplot(dt_coin, aes(x=time)) +
  geom_line(aes(y=residual)) +
  labs(title="Residual from linear combination X_t - beta*Y_t", x="time", y="residual") +
  theme_minimal()
```

