---
letter: 'y'
word: 'y-intercept'
dictionary: 'mathematics'
category: 'geometry'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['y-axis','polynomial']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="y-intercept">Y-intercept</a> - In **geometry**, the **y-intercept** of a curve is the point(s) where it crosses the [y-axis](#y-axis). Formally, points with x=0:

'58814'
\text{Find } y \text{ such that } f(0)=y.
'58814'

**Key points**:
- Solving for $f(0)$ often involves [polynomial](#polynomial) expressions or other functional forms.
- Graphically, y-intercepts appear where the plot crosses the vertical axis.

**R demonstration** (finding approximate y-intercept for a function):

```{r}
library(data.table)

f <- function(x) x^2 - 4*x + 1
y_intercept <- f(0)
y_intercept
```
