---
letter: 'c'
word: 'capm'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['markowitz-portfolio','variance']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(
    plot=function(x,options){
      cap<-options$fig.cap
      as.character(htmltools::tag("Figure", list(src=x, alt=cap,paste("\n\t",cap,"\n",sep=""))))
    }
  )
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="capm">CAPM (Capital Asset Pricing Model)</a> - The **CAPM** states that the expected return of an asset i is given by:

$$
\mathbb{E}[R_i] = R_f + \beta_i \bigl( \mathbb{E}[R_m] - R_f \bigr),
$$

where:
- \(R_f\) is the risk-free rate,
- \(\mathbb{E}[R_m]\) is the expected return of the market,
- \(\beta_i\) measures the asset’s sensitivity to market movements:
  '68859'
  \beta_i = \frac{\mathrm{Cov}(R_i, R_m)}{\mathrm{Var}(R_m)}.
  '68859'

**Key points**:
- Relies on assumptions from Markowitz portfolio theory.
- Suggests a linear relationship between an asset’s risk (\(\beta\)) and its expected return.

**R demonstration** (estimating beta from synthetic returns data):

```{r}
library(data.table)
library(ggplot2)

set.seed(123)
n <- 200
Rf <- 0.01  # risk-free rate
market_returns <- rnorm(n, mean=0.08, sd=0.04)
asset_returns  <- 0.02 + 1.2*market_returns + rnorm(n, mean=0, sd=0.02) # true beta=1.2

dt_capm <- data.table(market=market_returns, asset=asset_returns)

# Estimate beta = Cov(R_i, R_m)/Var(R_m)
est_beta <- cov(dt_capm$asset, dt_capm$market)/var(dt_capm$market)
est_beta

# We'll also do a quick regression
fit <- lm(asset ~ market, data=dt_capm)
summary(fit)

# Plot
ggplot(dt_capm, aes(x=market, y=asset)) +
  geom_point(alpha=0.4) +
  geom_smooth(method="lm", se=FALSE, color="red") +
  labs(title="CAPM example: Estimating Beta via Regression", x="Market Return", y="Asset Return") +
  theme_minimal()
```
