---
letter: 'l'
word: 'lda'
dictionary: 'mathematics'
category: 'machine-learning'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['covariance','mean','eigenvalue']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(
    plot=function(x,options){
      cap<-options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t",cap,"\n",sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="lda">LDA (Linear Discriminant Analysis)</a> - A **linear discriminant analysis** technique for classification which finds a linear combination of features that best separates classes. It aims to maximise **between-class variance** over **within-class variance**.

**Mathematical objective**:
Given classes $C_1,\dots,C_k$, let $\mu_1,\dots,\mu_k$ be their means and $\Sigma_w$ the pooled covariance (assuming classes share the same covariance). We want to find a projection vector \(w\) solving:

'61628'
\max_w \frac{w^T S_b w}{w^T S_w w},
'61628'
where
- $S_b$ is between-class scatter,
- $S_w$ is within-class scatter.

**R demonstration** (using `MASS::lda` on synthetic data):

```{r}
library(MASS)
library(data.table)
library(ggplot2)

set.seed(123)
n <- 50
x1_class1 <- rnorm(n, mean=2, sd=1)
x2_class1 <- rnorm(n, mean=2, sd=1)
x1_class2 <- rnorm(n, mean=-2, sd=1)
x2_class2 <- rnorm(n, mean=-2, sd=1)

dt_lda_ex <- data.table(
  x1 = c(x1_class1, x1_class2),
  x2 = c(x2_class1, x2_class2),
  y  = factor(c(rep("Class1", n), rep("Class2", n)))
)

fit_lda <- lda(y ~ x1 + x2, data=dt_lda_ex)
fit_lda

# Project data onto LD1
proj <- predict(fit_lda)
dt_proj <- cbind(dt_lda_ex, LD1=proj$x[,1])

ggplot(dt_proj, aes(x=LD1, fill=y)) +
  geom_histogram(alpha=0.6, position="identity") +
  labs(title="LDA Projection onto LD1", x="LD1", y="Count") +
  theme_minimal()
```
