---
letter: 'm'
word: 'manifold'
dictionary: 'mathematics'
category: 'topology'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['topological-space','dimension']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="manifold">Manifold</a> - In **topology** and differential geometry, a **manifold** is a [topological-space](#topological-space) that locally resembles Euclidean space. Formally, an $n$-dimensional manifold is a space where every point has a neighbourhood homeomorphic to \(\mathbb{R}^n\).

**Key points**:
- The concept of [dimension](#dimension) is central: a 2D manifold locally looks like a plane, a 3D manifold like space, etc.
- Smooth manifolds allow calculus-like operations on them.

No direct R demonstration, but we can illustrate how to store a “chart” or local coordinate system conceptually:

```{r}
library(data.table)
cat("Manifolds are an advanced concept. In R, we'd handle geometry libraries for numeric solutions.")
```
