---
letter: '#'
word: 'empty-set'
dictionary: 'mathematics'
category: 'set-theory'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['set','union','intersection']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="empty-set">∅ (Empty set)</a> - In **set theory**, the **empty set** ∅ is the unique set containing no elements. Symbolically:

$$
\varnothing = \{\},
$$

meaning it has cardinality 0. Any operation on ∅ typically yields minimal or neutral results. For instance, union with ∅ returns the other set unchanged.

**Key points**:
- ∅ is a subset of every [set](#set).
- It is distinct from $\{\varnothing\}$, which is a set containing the empty set as an element.

No direct R demonstration needed, but we can show conceptually:

```{r}
library(data.table)

empty_set <- vector(mode="list", length=0)  # a list with length 0
length(empty_set)
```
