---
letter: 'g'
word: gradient
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['gamma-function','group']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = "")))
    )
  }
)
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="gradient">Gradient</a> - In multivariable calculus, the **gradient** of a scalar field $f(x_1, x_2, \dots, x_n)$ is the vector of its partial derivatives:

55763
\nabla f = 
\left(
  \frac{\partial f}{\partial x_1},
  \frac{\partial f}{\partial x_2}, 
  \dots,
  \frac{\partial f}{\partial x_n}
\right).
55763

**Interpretation**:
- Points in the direction of steepest ascent of $f$.
- Magnitude represents the rate of increase in that direction.

**R demonstration** (numerically approximating a gradient for $f(x,y) = x^2 + 2xy$ at $(1,2)$):

```{r}
library(data.table)

f_xy <- function(x, y) x^2 + 2*x*y

numeric_grad <- function(f, x, y, h=1e-6) {
  df_dx <- (f(x+h, y) - f(x, y)) / h
  df_dy <- (f(x, y+h) - f(x, y)) / h
  c(df_dx, df_dy)
}

grad_1_2 <- numeric_grad(f_xy, 1, 2)
grad_1_2  # The exact gradient is (2x + 2y, 2x). At (1,2) => (2*1 + 2*2, 2*1) => (6,2)
```

