---
letter: 'w'
word: 'wavelet'
dictionary: 'mathematics'
category: 'analysis'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['fourier-transform']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="wavelet">Wavelet</a> - In **analysis** and signal processing, a **wavelet** is a function used to decompose and analyse signals at multiple scales. Unlike the [fourier-transform](#fourier-transform) which uses infinite sine and cosine waves, wavelets are localised in both time (space) and frequency.

**Key points**:
- Wavelets are useful for time-frequency or space-scale analysis.
- Common families: Haar, Daubechies, Morlet, etc.

**R demonstration** (no base function for wavelets, but let's show a conceptual signal transform snippet):

```{r}
# This is conceptual, as wavelet packages (e.g. 'wavelets') would be used.
# We'll do a simple mock demonstration of a signal decomposition approach.

library(data.table)
set.seed(123)
signal <- c(rep(0,50), rnorm(50), rep(0,50))

# Suppose we apply a placeholder "wavelet transform"
# (here, just a naive split for illustration)
signal_low  <- filter(signal, rep(1/2,2), sides=2)  # naive "approx" 
signal_high <- signal - signal_low                  # naive "detail"

dt_wav <- data.table(
  idx = seq_along(signal),
  signal = signal,
  approx = signal_low,
  detail = signal_high
)
head(dt_wav, 10)
```
