---
letter: 'd'
word: 'divergence'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['nabla','curl','gradient']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()) {
  knitr::knit_hooks$set(plot = function(x, options){
    cap <- options$fig.cap
    as.character(
      htmltools::tag("Figure",
        list(src=x, alt=cap, paste("\n\t", cap, "\n", sep=""))
      )
    )
  })
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="divergence">Divergence</a> - In **vector calculus**, the **divergence** of a vector field \(\mathbf{F} = (F_x, F_y, F_z)\) is a scalar measure of how much the field “spreads out” (source/sink). Using the [nabla](#nabla) operator ∇:

'59475'
\nabla \cdot \mathbf{F} = \frac{\partial F_x}{\partial x} + \frac{\partial F_y}{\partial y} + \frac{\partial F_z}{\partial z}.
'59475'

**Key points**:
- If divergence is zero everywhere, the field is solenoidal (incompressible).
- Common in fluid dynamics, electromagnetics, etc.

**R demonstration** (approx numeric partials of a simple 3D field):

```{r}
library(data.table)

F <- function(x,y,z) c(x*y, x+z, y*z)  # example vector field
h <- 1e-6

divergence_approx <- function(f, x,y,z, h=1e-6) {
  # f => returns c(Fx, Fy, Fz)
  # partial wrt x
  fx_plus <- f(x+h,y,z); fx <- f(x,y,z)
  dFx_dx <- (fx_plus[1] - fx[1]) / h
  
  # partial wrt y
  fy_plus <- f(x,y+h,z)
  dFy_dy <- (fy_plus[2] - fx[2]) / h
  
  # partial wrt z
  fz_plus <- f(x,y,z+h)
  dFz_dz <- (fz_plus[3] - fx[3]) / h
  
  dFx_dx + dFy_dy + dFz_dz
}

divergence_approx(F, 1,2,3)
```
