---
letter: 'e'
word: expectation
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['entropy','euclidean-algorithm']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.height = 7
)
```

<a id="expectation">Expectation</a> - In probability theory, the **expectation** (or expected value) of a random variable $X$ represents the long-run average outcome of $X$ after many repetitions of an experiment.

For a discrete random variable:

55424
E[X] = \sum_{x} x\, p(x).
55424

For a continuous random variable:

55424
E[X] = \int_{-\infty}^{\infty} x\, f_X(x) \, dx,
55424

where $f_X(x)$ is the probability density function.

**R demonstration** (empirical estimation of expectation):

```{r}
library(data.table)

set.seed(123)
# Suppose X ~ Uniform(0, 10)
X_samples <- runif(10000, min=0, max=10)

dtX <- data.table(X = X_samples)

# Empirical mean
emp_mean <- mean(dtX$X)

# Theoretical expectation for Uniform(0, 10) is 5
theoretical <- 5

emp_mean
theoretical
```

