---
letter: 'y'
word: 'y-axis'
dictionary: 'mathematics'
category: 'geometry'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['coordinate-system','y-intercept']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="y-axis">Y-axis</a> - In a **2D coordinate system**, the **y-axis** is the vertical line used to measure the y-coordinate of points. Typically, positive values extend upward from the origin, while negative values extend downward.

**Key points**:
- Intersects the x-axis at the origin (0,0).
- The slope of the y-axis is undefined (vertical line).
- Used to define a [y-intercept](#y-intercept) for curves.

Basic R demonstration:

```{r}
library(ggplot2)
library(data.table)

dt_plot <- data.table(x=0, y=-5:5)
ggplot(dt_plot, aes(x=x, y=y)) +
  geom_line() +
  labs(
    title="Y-axis (vertical)",
    x="x",
    y="y"
  ) +
  theme_minimal()
```
