---
letter: 'w'
word: 'wedge-product'
dictionary: 'mathematics'
category: 'algebra'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['vector','matrix']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="wedge-product">Wedge Product</a> - In **exterior algebra**, the **wedge product** ($\wedge$) of two [vector](#vector)s $\mathbf{u}, \mathbf{v}$ forms an oriented area element (in 2D) or higher-dimensional analog. For 2D vectors, the wedge product is related to the determinant of a [matrix](#matrix):

$$
\mathbf{u}\wedge\mathbf{v} = 
\begin{vmatrix}
u_1 & v_1 \\
u_2 & v_2
\end{vmatrix}.
$$

**Key points**:
- Anticommutative: $\mathbf{u}\wedge\mathbf{v} = -\mathbf{v}\wedge\mathbf{u}$.
- In higher dimensions, wedge products generalise to forms $\alpha\wedge\beta$.

**R demonstration** (simple 2D wedge product as area determinant):

```{r}
library(data.table)

wedge_2d <- function(u, v) {
  # u, v are c(u1,u2), c(v1,v2)
  u[1]*v[2] - u[2]*v[1]
}

u <- c(2,1)
v <- c(3,4)
wedge_2d(u,v)
```
