---
letter: 'v'
word: 'value-at-risk'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['distribution','monte-carlo-simulation']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x,options){
    cap<-options$fig.cap
    as.character(
      htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t",cap,"\n",sep="")))
    )
  })
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="value-at-risk">Value at Risk (VaR)</a> - In **risk management**, **Value at Risk** at confidence level \(\alpha\) (e.g., 95% or 99%) is the loss threshold that is not exceeded with probability \(\alpha\). More concretely, for a random loss \(L\):

$$
\mathrm{VaR}_{\alpha}(L) = \inf\{\ell : P(L \le \ell) \ge \alpha\}.
$$

**Key points**:
- Often computed via historical simulation or [monte-carlo-simulation](#monte-carlo-simulation).
- Criticised because it doesn’t capture tail risk beyond the threshold.

**R demonstration** (historical VaR from synthetic returns data):

```{r}
library(data.table)
library(ggplot2)

set.seed(123)
n <- 5000
# We'll simulate daily returns from a normal distribution
daily_returns <- rnorm(n, mean=0.0005, sd=0.01)
# Convert to daily PnL or "loss" if negative
# We treat negative returns as positive "loss" for simplicity
losses <- -daily_returns

alpha <- 0.95
var_level <- quantile(losses, alpha)
var_level

# Distribution plot
dt_var <- data.table(loss=losses)
ggplot(dt_var, aes(x=loss)) +
  geom_histogram(bins=40, fill="lightblue", color="black") +
  geom_vline(xintercept=var_level, color="red", lwd=1.2) +
  labs(title=paste0("Histogram of Losses with VaR at ", alpha*100, "%"),
       x="Loss",
       y="Count") +
  theme_minimal()
```

