---
letter: 't'
word: 'topological-space'
dictionary: 'mathematics'
category: 'topology'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['set','boundary']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot=function(x, options){
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src=x, alt=cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="topological-space">Topological Space</a> - In **topology**, a **topological space** is a [set](#set) \(X\) equipped with a collection of open sets that satisfy the axioms:
1. \(\varnothing\) and \(X\) are open.
2. Finite intersections of open sets are open.
3. Arbitrary unions of open sets are open.

**Key points**:
- Generalises notions of continuity, [boundary](#boundary), and “closeness” beyond Euclidean space.
- Foundation for manifold definitions, continuity arguments, convergence, etc.

No direct R demonstration (purely theoretical structure).  
```{r}
cat("A topological space is the core object in topology, with open sets defining 'closeness'.")
```
