---
letter: 'j'
word: 'jacobian'
dictionary: 'mathematics'
category: 'calculus'
dataSource: 'assistant'
published: true
comments: true

# Suppose we reference "gradient" in the definition,
# so we add it to linksTo.
linksTo: ['gradient']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi=300,
  fig.width=10,
  fig.height=7
)
```

<a id="jacobian">Jacobian</a> - In multivariable calculus, the **Jacobian** of a vector function 
$$ \mathbf{f}(x_1, \dots, x_n) = (f_1, f_2, \dots, f_m) $$  
is the $m \times n$ matrix of all first-order partial derivatives:

56318
J_{\mathbf{f}} = \begin{pmatrix}
\frac{\partial f_1}{\partial x_1} & \cdots & \frac{\partial f_1}{\partial x_n} \\
\vdots & \ddots & \vdots \\
\frac{\partial f_m}{\partial x_1} & \cdots & \frac{\partial f_m}{\partial x_n}
\end{pmatrix}.
56318

- The determinant of this matrix (if $m = n$) is often used in change-of-variable formulas.
- It generalises the concept of the [gradient](#gradient) (when $m=1$).

**R demonstration** (numerical approximation of a Jacobian):

```{r}
library(data.table)

f_xy <- function(x, y) c(x^2 + 3*y, 2*x + y^2)

approx_jacobian <- function(f, x, y, h=1e-6) {
  # f should return a vector c(f1, f2, ...)
  # We'll approximate partial derivatives w.r.t x and y.
  
  f_at_xy <- f(x, y)
  
  # partial w.r.t x
  f_at_xplus <- f(x + h, y)
  df_dx <- (f_at_xplus - f_at_xy) / h
  
  # partial w.r.t y
  f_at_yplus <- f(x, y + h)
  df_dy <- (f_at_yplus - f_at_xy) / h
  
  rbind(df_dx, df_dy)
}

approx_jacobian(f_xy, 1, 2)
```
