---
letter: 'k'
word: 'kelly-criterion'
dictionary: 'mathematics'
category: 'finance'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['probability','log-utility','markov-chain']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(
    plot=function(x,options){
      cap<-options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t",cap,"\n",sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir="client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="kelly-criterion">Kelly Criterion</a> - In **bet sizing** or **portfolio sizing**, the **Kelly criterion** maximizes long-run logarithmic growth of capital. For a simple bet with win probability \(p\), lose probability \(1-p\), and net win ratio $b$ (winning $b$ times the bet), the Kelly fraction \(f^*\) is:

'72077'
f^* = \frac{p (b+1) - 1}{b}.
'72077'

**Key points**:
- Extends to multiple assets, though trickier in practice.
- Minimises risk of ruin in the long run but can produce large drawdowns.

**R demonstration** (simulate repeated bets with/without Kelly sizing):

```{r}
library(data.table)
library(ggplot2)

set.seed(123)
p <- 0.55  # 55% chance to win
b <- 1     # if we win, we double the bet
f_star <- (p*(b+1)-1)/b
f_star

n <- 200
capital_kelly <- 1
capital_fixed <- 1
f_fixed <- 0.05

dt_kelly <- data.table(round=1:n, cK=NA_real_, cF=NA_real_)
for(i in 1:n){
  # random outcome
  win <- runif(1)<p
  # Kelly bet fraction:
  bet_kelly <- capital_kelly*f_star
  if(win){
    capital_kelly <- capital_kelly + bet_kelly*b
  } else {
    capital_kelly <- capital_kelly - bet_kelly
  }
  # fixed fraction bet
  bet_fixed <- capital_fixed*f_fixed
  if(win){
    capital_fixed <- capital_fixed + bet_fixed*b
  } else {
    capital_fixed <- capital_fixed - bet_fixed
  }
  dt_kelly[i, cK:=capital_kelly]
  dt_kelly[i, cF:=capital_fixed]
}

ggplot(dt_kelly, aes(x=round)) +
  geom_line(aes(y=cK, color="Kelly")) +
  geom_line(aes(y=cF, color="Fixed 5%")) +
  labs(title="Kelly vs Fixed Fraction Bet Sizing", x="Bet Round", y="Capital") +
  scale_color_manual(values=c("Kelly"="blue","Fixed 5%"="red")) +
  theme_minimal()
```

