---
letter: 'n'
word: 'number-line'
dictionary: 'mathematics'
category: 'analysis'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['interval']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="number-line">Number Line</a> - The **number line** (real line) is a straight line on which every real number corresponds to a unique point. Basic structures like an [interval](#interval) are subsets of the number line:

- Negative numbers extend to the left, positive numbers to the right.
- Zero is typically placed at the origin.

No direct R demonstration is typical, but we can illustrate numeric representations:

```{r}
library(data.table)

vals <- seq(-3, 3, by=1)
vals
```
