---
letter: 'j'
word: 'julia-set'
dictionary: 'mathematics'
category: 'complex-analysis'
dataSource: 'assistant'
published: true
comments: true

# Suppose we reference "fractal" in the definition,
# so we add it to linksTo.
linksTo: ['fractal']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi=300,
  fig.width=10,
  fig.height=7
)
```

<a id="julia-set">Julia Set</a> - In **complex dynamics**, a **Julia set** is the boundary of points in the complex plane describing the behaviour of a complex function, often associated with the iteration of polynomials like $z \mapsto z^2 + c$. Julia sets are typical examples of a [fractal](#fractal).

**Key points**:
- For each complex parameter $c$, there is a distinct Julia set.
- The set often exhibits self-similarity and intricate boundaries.

**R demonstration** (simple iteration to classify points):

```{r}
library(data.table)
library(ggplot2)

# We'll do a basic "escape-time" iteration for z^2 + c, with c = -0.8 + 0.156i
c_val <- complex(real=-0.8, imaginary=0.156)

n <- 400
x_seq <- seq(-1.5, 1.5, length.out=n)
y_seq <- seq(-1.5, 1.5, length.out=n)

max_iter <- 50
threshold <- 2

res <- data.table()
for (ix in seq_along(x_seq)) {
  for (iy in seq_along(y_seq)) {
    z <- complex(real=x_seq[ix], imaginary=y_seq[iy])
    iter <- 0
    while(Mod(z) < threshold && iter < max_iter) {
      z <- z*z + c_val
      iter <- iter + 1
    }
    res <- rbind(
      res,
      data.table(
        x = x_seq[ix],
        y = y_seq[iy],
        iteration = iter
      )
    )
  }
}

ggplot(res, aes(x=x, y=y, color=iteration)) +
  geom_point(shape=15, size=1) +
  scale_color_viridis_c() +
  coord_fixed() +
  labs(
    title="Simple Julia Set (z^2 + c)",
    x="Re(z)",
    y="Im(z)"
  ) +
  theme_minimal()
```
