---
letter: 'u'
word: 'unbiased-estimator'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['expectation','variance']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="unbiased-estimator">Unbiased Estimator</a> - In **statistics**, an **unbiased estimator** is one whose [expectation](#expectation) equals the true parameter being estimated. For example, the sample variance with denominator $(n-1)$ is an unbiased estimator of the population [variance](#variance).

**Key points**:
- Unbiasedness is about the *expected value* of the estimator matching the parameter.
- It does *not* guarantee minimum variance or other optimal properties.

**R demonstration** (comparing biased vs. unbiased sample variance):

```{r}
library(data.table)

set.seed(123)
x_vals <- rnorm(1000, mean=5, sd=2)

biased_var   <- (1/length(x_vals))*sum((x_vals - mean(x_vals))^2)
unbiased_var <- var(x_vals)  # R's 'var' uses denominator (n-1)

c(biased_var, unbiased_var)
```
