---
letter: 'i'
word: intersection
dictionary: 'mathematics'
category: 'set-theory'
dataSource: 'assistant'
published: true
comments: true

# Suppose we reference "union" (which might be found in another set-based definition),
# so we add it to linksTo, even if it's not created yet.
linksTo: ['union']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag(
          "Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="intersection">Intersection</a> - In **set theory**, the **intersection** of two sets $A$ and $B$ is the set of elements that belong to both $A$ and $B$. Symbolically:

56131
A \cap B = \{ x : x \in A \text{ and } x \in B \}.
56131

- Compare this with the [union](#union) $A \cup B$, which combines all elements in either $A$ or $B$.
- The empty set $\varnothing$ results if $A$ and $B$ share no elements.

No special R demonstration is typically needed, but we can illustrate a basic example using sets as vectors:

```{r}
A <- c(1, 2, 3, 4)
B <- c(3, 4, 5, 6)

intersect(A, B)  # yields 3,4
```
