---
letter: 'z'
word: 'zero'
dictionary: 'mathematics'
category: 'arithmetic'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['arithmetic','whole-number','identity-matrix']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="zero">Zero</a> - In **arithmetic**, **zero** (0) is the additive identity. Adding zero to a number leaves it unchanged:

$$
x + 0 = x.
$$

**Key points**:
- It belongs to the [whole-number](#whole-number) system (and integers).
- [Arithmetic](#arithmetic) with zero is straightforward: multiplication by 0 yields 0.
- In linear algebra, the [identity-matrix](#identity-matrix) uses 1 on diagonals but 0 off-diagonal.

Minimal R demonstration:

```{r}
library(data.table)

x <- 5
x_plus_zero <- x + 0
x_plus_zero
```
