---
letter: 'r'
word: 'random-variable'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['distribution','expectation']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="random-variable">Random Variable</a> - In **probability theory**, a **random variable** is a function that assigns a real number to each outcome in a sample space. It links randomness (abstract events) to numerical values for analysis (via a [distribution](#distribution)). 

'57737'
X: \Omega \to \mathbb{R}.
'57737'

**Key points**:
- Can be discrete or continuous.
- The [expectation](#expectation) of a random variable provides a measure of its average outcome.

**R demonstration** (sampling a random variable $X\sim N(0,1)$):

```{r}
library(data.table)
set.seed(123)
dt_rv <- data.table(x = rnorm(10))
dt_rv
```
