---
letter: 'o'
word: 'outlier'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['distribution']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap <- options$fig.cap
      as.character(
        htmltools::tag("Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7)
```

<a id="outlier">Outlier</a> - In **statistics**, an **outlier** is a data point significantly distant from the rest of the [distribution](#distribution). Outliers can arise from measurement errors, heavy-tailed distributions, or genuine extreme events.

**Key points**:
- Outliers can skew means, inflate variances, or distort analyses.
- Detection methods include IQR-based rules, z-scores, or robust statistics.

**R demonstration** (basic detection via boxplot stats):

```{r}
library(data.table)
library(ggplot2)

set.seed(123)
dt_out <- data.table(x = c(rnorm(30, mean=10, sd=1), 25)) # one extreme outlier
ggplot(dt_out, aes(y=x)) + geom_boxplot(fill="lightblue") +
  theme_minimal()

stats <- boxplot.stats(dt_out$x)
stats$out
```
