---
letter: 's'
word: 'sample'
dictionary: 'mathematics'
category: 'statistics'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['distribution']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep=""))
        )
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="sample">Sample</a> - In **statistics**, a **sample** is a subset taken from a larger population, used to infer characteristics (such as a [distribution](#distribution)) of that population. Random sampling is crucial to reduce bias.

**Key points**:
- **Sample size** is the number of observations in the subset.
- Methods of sampling include simple random sampling, stratified sampling, etc.

**R demonstration** (random sample from a vector):

```{r}
library(data.table)

vals <- 1:100
set.seed(123)
sample_vals <- sample(vals, 10)
sample_vals
```
