---
letter: 'u'
word: 'upper-triangular'
dictionary: 'mathematics'
category: 'linear-algebra'
dataSource: 'assistant'
published: true
comments: true

linksTo: ['matrix']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) {
  knitr::knit_hooks$set(
    plot = function(x, options) {
      cap  <- options$fig.cap
      as.character(
        htmltools::tag("Figure",
          list(src = x, alt = cap, paste("\n\t", cap, "\n", sep="")))
      )
    }
  )
}
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="upper-triangular">Upper Triangular</a> - In **linear algebra**, an **upper triangular** [matrix](#matrix) is one where all entries below the main diagonal are zero. Formally, a square matrix $U=(u_{ij})$ is upper triangular if:

$$
u_{ij} = 0 \quad \text{for all } i > j.
$$

**Key points**:
- Common in LU decomposition (where U is the upper triangular factor).
- Determinant is the product of diagonal entries.

**R demonstration** (example of an upper triangular matrix):

```{r}
library(data.table)

U <- matrix(c(1,2,3,0,4,5,0,0,6), nrow=3, byrow=TRUE)
U
# Checking that entries below diagonal are zero:
all(U[lower.tri(U)]==0)
```
