---
letter: 'e'
word: entropy
dictionary: 'mathematics'
category: 'information-theory'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['eigenvalue','expectation']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag("Figure",
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.height = 7
)
```

<a id="entropy">Entropy</a> - In information theory, **entropy** quantifies the average amount of information contained in a random variable’s possible outcomes. For a discrete random variable $X$ with pmf $p(x)$, the Shannon entropy (in bits) is:

55424
H(X) = -\sum_{x} p(x) \log_2\bigl(p(x)\bigr).
55424

**Key points**:
- Entropy is maximised when all outcomes are equally likely.
- Low entropy implies outcomes are more predictable.
- It underpins coding theory, compressions, and measures of uncertainty.

**R demonstration** (computing entropy of a discrete distribution):

```{r}
library(data.table)

entropy_shannon <- function(prob_vec) {
  # Make sure prob_vec sums to 1
  -sum(prob_vec * log2(prob_vec), na.rm=TRUE)
}

dt_prob <- data.table(
  outcome = letters[1:4],
  prob    = c(0.1, 0.4, 0.3, 0.2)  # must sum to 1
)

H <- entropy_shannon(dt_prob$prob)
H
```

