---
letter: 'h'
word: homomorphism
dictionary: 'mathematics'
category: 'algebra'
dataSource: 'assistant'
published: true
comments: true

# We reference "group" in the definition, so add it to linksTo.
linksTo: ['group']
linkedFrom: []

output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", 
      list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)
knitr::opts_knit$set(root.dir = "/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="homomorphism">Homomorphism</a> - In **algebra**, a **homomorphism** is a structure-preserving map between two algebraic structures of the same type. For instance, a **group** homomorphism between groups $G$ and $H$ is a function

55983
\phi: G \to H
55983

such that:

55983
\phi(a \cdot b) = \phi(a) \star \phi(b),
55983

where $\cdot$ is the operation in $G$ and $\star$ is the operation in $H$ (see [group](#group) for more on group operations).

This ensures that the algebraic structure (associativity, identity, inverses) is respected.

No direct R demonstration is typical for homomorphisms, but we can quickly illustrate a trivial map:

```{r}
# As a simple example:
# A map from Z -> Z given by phi(x) = 2*x is a group homomorphism under addition.
phi <- function(x) 2*x

phi(3)  # 6
# phi(a + b) = 2*(a + b) = 2a + 2b = phi(a) + phi(b)
```
