---
letter: 'l'
word: 'llm'
dictionary: 'mathematics'
category: 'machine-learning'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['transformer','neural-network','attention-mechanism']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x,options){
    cap<-options$fig.cap
    as.character(
      htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t",cap,"\n",sep="")))
    )
  })
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300, fig.width=10, fig.height=7)
```

<a id="llm">LLM (Large Language Model)</a> - A **large language model** is typically a **Transformer**-based or similarly advanced architecture with billions (or more) of parameters, trained on massive text corpora to generate coherent text or perform NLP tasks.

**Key points**:
- Uses self-attention to handle long contexts.
- Learns complex linguistic structures, can generate next tokens based on context.

**Mathematical gist**:
At each token step, an LLM computes a probability distribution over the vocabulary:
$$
p(x_{t+1}\mid x_{1:\dots:t}) = \mathrm{softmax}(h_t W_{\text{vocab}}),
$$
where $h_t$ is the hidden representation after attention layers.

**R demonstration** (We can show a mini example of text generation with `keras`, but typically giant LLM training isn't feasible in R. We'll do conceptual snippet):

```{r}
# Conceptual only:
library(data.table)
cat("Training an LLM is typically done in Python with large GPU clusters.\nWe'll do a small toy example with a simple RNN or minimal next-token model.")
```

