---
letter: 'n'
word: 'naive-bayes'
dictionary: 'mathematics'
category: 'machine-learning'
dataSource: 'assistant'
published: true
comments: true
linksTo: ['bayes-theorem','distribution']
linkedFrom: []
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if(knitr::is_html_output()){
  knitr::knit_hooks$set(plot=function(x,options){
    cap<-options$fig.cap
    as.character(
      htmltools::tag("Figure", list(src=x, alt=cap, paste("\n\t",cap,"\n",sep="")))
    )
  })
}
knitr::opts_knit$set(root.dir="/Users/work/Coding/derecksprojects/derecksnotes.com-playground/derecksnotes.com/client/src/app/dictionaries/mathematics/definitions")
knitr::knit_hooks$set(optipng=knitr::hook_optipng)
knitr::opts_chunk$set(dpi=300,fig.width=10,fig.height=7)
```

<a id="naive-bayes">Naive Bayes</a> - In **machine learning**, **Naive Bayes** is a probabilistic classifier applying [Bayes' theorem](#bayes-theorem) with a “naive” (independence) assumption among features given the class. For a class $C_k$ and features $x_1,\dots,x_n$:

$$
P(C_k | x_1,\dots,x_n) \propto P(C_k) \prod_{i=1}^n P(x_i | C_k).
$$

**Key points**:
- Independence assumption simplifies computation of \(P(x_1,\dots,x_n|C_k)\).
- Effective in text classification (bag-of-words assumption).

**R demonstration** (using `e1071::naiveBayes` on synthetic data):

```{r}
library(e1071)
library(data.table)
library(ggplot2)

set.seed(123)
n <- 100
x1 <- rnorm(n, mean=2, sd=1)
x2 <- rnorm(n, mean=-1, sd=1)
cl1 <- data.table(x1, x2, y="Class1")

x1 <- rnorm(n, mean=-2, sd=1)
x2 <- rnorm(n, mean=2,  sd=1)
cl2 <- data.table(x1, x2, y="Class2")

dt_nb <- rbind(cl1, cl2)

fit_nb <- naiveBayes(y ~ x1 + x2, data=dt_nb)
fit_nb

# Predict
grid_x1 <- seq(-5,5, by=0.2)
grid_x2 <- seq(-5,5, by=0.2)
grid_data <- data.table(expand.grid(x1=grid_x1, x2=grid_x2))
grid_data[, pred := predict(fit_nb, newdata=.SD)]

ggplot() +
  geom_tile(data=grid_data, aes(x=x1, y=x2, fill=pred), alpha=0.4) +
  geom_point(data=dt_nb, aes(x=x1, y=x2, color=y), size=2) +
  scale_fill_manual(values=c("Class1"="lightblue","Class2"="salmon")) +
  scale_color_manual(values=c("Class1"="blue","Class2"="red")) +
  labs(title="Naive Bayes Classification", x="x1", y="x2") +
  theme_minimal()
```
