---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 6: Joint Distributions and Multivariate Thinking"
part: "Part 2: Covariance and Correlation"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, probability, distributions, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

# Chapter 6 -- Joint Distributions and Multivariate Thinking

## Part 2: Covariance and Correlation

In Part 1 we learnt to describe how two random variables behave *together* through their joint distribution. Now we develop **numerical summaries** of the relationship between two variables. Just as the mean and variance summarise a single variable, the **covariance** and **correlation** summarise the linear association between a pair of variables.

We shall derive these quantities from first principles, prove the Cauchy-Schwarz inequality (which guarantees that correlation is bounded between $-1$ and $+1$), and carefully distinguish between correlation and dependence -- one of the most important conceptual distinctions in all of statistics.

---

## 6.5 Covariance

### 6.5.1 Definition

The **covariance** of two random variables $X$ and $Y$ is defined as:

$$\text{Cov}(X, Y) = E\bigl[(X - E[X])(Y - E[Y])\bigr].$$

This measures the *average product of deviations from the mean*. If $X$ tends to be above its mean when $Y$ is above its mean (and below when below), the product is positive. If they move in opposite directions, it is negative.

### 6.5.2 The Computational Formula

Expanding the definition:

$$\text{Cov}(X, Y) = E\bigl[XY - X \cdot E[Y] - E[X] \cdot Y + E[X] \cdot E[Y]\bigr].$$

By linearity of expectation:

$$= E[XY] - E[X] \cdot E[Y] - E[X] \cdot E[Y] + E[X] \cdot E[Y]$$

$$\boxed{\text{Cov}(X, Y) = E[XY] - E[X] \cdot E[Y].}$$

This is the **computational formula** -- often easier to use than the definition.

### 6.5.3 Properties of Covariance

We derive the key properties from the definition.

**Property 1: Symmetry.** $\text{Cov}(X, Y) = \text{Cov}(Y, X)$.

*Proof:* $E[(X - \mu_X)(Y - \mu_Y)] = E[(Y - \mu_Y)(X - \mu_X)]$ by commutativity of multiplication. $\square$

**Property 2: Covariance with itself.** $\text{Cov}(X, X) = \text{Var}(X)$.

*Proof:* $E[(X - \mu_X)(X - \mu_X)] = E[(X - \mu_X)^2] = \text{Var}(X)$. $\square$

**Property 3: Linearity in each argument.**

$$\text{Cov}(aX + b, Y) = a \cdot \text{Cov}(X, Y).$$

*Proof:*

$$E\bigl[(aX + b - E[aX + b])(Y - E[Y])\bigr] = E\bigl[(aX + b - a\mu_X - b)(Y - \mu_Y)\bigr]$$
$$= E\bigl[a(X - \mu_X)(Y - \mu_Y)\bigr] = a \cdot \text{Cov}(X, Y). \quad \square$$

The constant $b$ vanishes because it cancels with $E[aX + b] = a\mu_X + b$.

**Property 4: Bilinearity.**

$$\text{Cov}\left(\sum_i a_i X_i, \sum_j b_j Y_j\right) = \sum_i \sum_j a_i b_j \, \text{Cov}(X_i, Y_j).$$

**Property 5: Independence implies zero covariance.**

If $X$ and $Y$ are independent, then $E[XY] = E[X] \cdot E[Y]$, so $\text{Cov}(X, Y) = 0$.

**Warning:** The converse is *false* -- zero covariance does **not** imply independence. We demonstrate this in Section 6.7.

**Property 6: Variance of a sum.**

$$\text{Var}(X + Y) = \text{Var}(X) + \text{Var}(Y) + 2\,\text{Cov}(X, Y).$$

*Proof:*

$$\text{Var}(X + Y) = E\bigl[(X + Y - \mu_X - \mu_Y)^2\bigr]$$
$$= E\bigl[(X - \mu_X)^2 + 2(X - \mu_X)(Y - \mu_Y) + (Y - \mu_Y)^2\bigr]$$
$$= \text{Var}(X) + 2\,\text{Cov}(X, Y) + \text{Var}(Y). \quad \square$$

When $X$ and $Y$ are independent, $\text{Cov}(X,Y) = 0$ and we recover the familiar $\text{Var}(X + Y) = \text{Var}(X) + \text{Var}(Y)$.

### 6.5.4 Implementing Covariance from Scratch

```{r dt-symbols, include=FALSE}
.N <- data.table::.N
.SD <- data.table::.SD
```

```{r my-cov}
box::use(data.table)
box::use(ggplot2)

#' Covariance -- implemented from scratch
#' Uses the unbiased estimator (denominator n-1)
#' @param x numeric vector
#' @param y numeric vector
#' @return scalar covariance
my_cov <- function(x, y) {
  stopifnot(length(x) == length(y))
  n <- length(x)
  sum((x - mean(x)) * (y - mean(y))) / (n - 1)
}

# Verify against R's built-in cov()
set.seed(42)
x <- rnorm(100)
y <- 2 * x + rnorm(100)

c(ours = my_cov(x, y), builtin = cov(x, y))
```

The values match exactly. The denominator $n - 1$ (rather than $n$) gives the unbiased sample covariance, just as for the sample variance.

---

## 6.6 Correlation and the Cauchy-Schwarz Inequality

### 6.6.1 Definition of Correlation

The **Pearson correlation coefficient** between $X$ and $Y$ is the covariance normalised by the standard deviations:

$$\rho_{X,Y} = \frac{\text{Cov}(X, Y)}{\sqrt{\text{Var}(X)} \cdot \sqrt{\text{Var}(Y)}} = \frac{\text{Cov}(X, Y)}{\sigma_X \, \sigma_Y}.$$

The correlation is a **dimensionless** quantity -- it does not depend on the units or scale of $X$ and $Y$. It measures the strength and direction of the *linear* relationship.

### 6.6.2 The Cauchy-Schwarz Inequality

To prove that the correlation coefficient always lies between $-1$ and $+1$, we need the **Cauchy-Schwarz inequality** for expectations:

$$\bigl|E[UV]\bigr|^2 \le E[U^2] \cdot E[V^2]$$

for any random variables $U$ and $V$ with finite second moments.

**Proof.** Consider the non-negative quantity $E[(U - tV)^2] \ge 0$ for any real number $t$. Expanding:

$$E[U^2] - 2t \, E[UV] + t^2 \, E[V^2] \ge 0.$$

This is a quadratic in $t$: $at^2 + bt + c \ge 0$ where $a = E[V^2]$, $b = -2E[UV]$, $c = E[U^2]$.

A quadratic $at^2 + bt + c$ is non-negative for all $t$ if and only if its discriminant is non-positive:

$$b^2 - 4ac \le 0$$

$$4\bigl(E[UV]\bigr)^2 - 4 \, E[U^2] \cdot E[V^2] \le 0$$

$$\boxed{\bigl(E[UV]\bigr)^2 \le E[U^2] \cdot E[V^2].} \quad \square$$

### 6.6.3 Proving $|\rho| \le 1$

Set $U = X - E[X]$ and $V = Y - E[Y]$. Then:

- $E[UV] = \text{Cov}(X, Y)$
- $E[U^2] = \text{Var}(X) = \sigma_X^2$
- $E[V^2] = \text{Var}(Y) = \sigma_Y^2$

Applying Cauchy-Schwarz:

$$\bigl(\text{Cov}(X,Y)\bigr)^2 \le \sigma_X^2 \, \sigma_Y^2$$

$$\left(\frac{\text{Cov}(X,Y)}{\sigma_X \, \sigma_Y}\right)^2 \le 1$$

$$\boxed{|\rho_{X,Y}| \le 1.}$$

Equality holds if and only if $U = tV$ for some constant $t$, i.e., $Y = aX + b$ for constants $a, b$. In other words, $\rho = \pm 1$ if and only if $X$ and $Y$ have a *perfectly linear* relationship.

### 6.6.4 Implementing Correlation from Scratch

```{r my-cor}
#' Pearson correlation -- implemented from scratch
#' @param x numeric vector
#' @param y numeric vector
#' @return scalar correlation
my_cor <- function(x, y) {
  my_cov(x, y) / (sd(x) * sd(y))
}

# Verify
c(ours = my_cor(x, y), builtin = cor(x, y))
```

### 6.6.5 Visualising Different Correlations

```{r correlation-gallery, fig.cap="Scatterplots illustrating different values of the correlation coefficient. Perfect positive correlation (rho = 1) and perfect negative correlation (rho = -1) produce straight lines. As |rho| decreases, the scatter increases. At rho = 0, there is no linear pattern."}
set.seed(42)
n_pts <- 500

make_correlated <- function(rho, n = n_pts) {
  Z1 <- rnorm(n)
  Z2 <- rnorm(n)
  X <- Z1
  Y <- rho * Z1 + sqrt(1 - rho^2) * Z2
  data.table$data.table(X = X, Y = Y)
}

rho_values <- c(-0.95, -0.5, 0, 0.5, 0.95, 1.0)
gallery <- data.table$rbindlist(lapply(rho_values, function(r) {
  dt <- make_correlated(min(r, 0.999))  # avoid sqrt(0) issue at rho=1
  if (r == 1.0) { dt$Y <- dt$X }  # exact linear
  dt[, rho_label := sprintf("rho == %.2f", r)]
  dt
}))

ggplot2$ggplot(gallery, ggplot2$aes(x = X, y = Y)) +
  ggplot2$geom_point(alpha = 0.25, size = 0.8, colour = "#0072B2") +
  ggplot2$facet_wrap(~rho_label, scales = "free", ncol = 3, labeller = ggplot2$label_parsed) +
  ggplot2$labs(x = "X", y = "Y", title = "Gallery of correlation values") +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(strip.text = ggplot2$element_text(face = "bold"))
```

---

## 6.7 Correlation Does Not Imply Dependence

### 6.7.1 The Classic Counterexample

This is one of the most important lessons in statistics: **uncorrelated does not mean independent**. We construct a definitive example.

Let $X \sim N(0, 1)$ and define $Y = X^2$. Then $Y$ is completely determined by $X$ -- they are maximally dependent. But:

$$\text{Cov}(X, Y) = E[XY] - E[X] \cdot E[Y] = E[X^3] - 0 \cdot E[X^2] = E[X^3].$$

For $X \sim N(0,1)$, the third moment is zero (by symmetry of the normal distribution). Therefore:

$$\text{Cov}(X, X^2) = 0 \quad \text{and} \quad \rho(X, X^2) = 0.$$

The variables are **uncorrelated but perfectly dependent**.

### 6.7.2 Simulation

```{r uncorrelated-dependent, fig.cap="X ~ N(0,1) and Y = X^2: a perfect functional relationship (Y is completely determined by X) yet the correlation is zero. The parabolic pattern is invisible to correlation, which only detects linear relationships."}
set.seed(42)
n_sim <- 3000
X_sim <- rnorm(n_sim)
Y_sim <- X_sim^2

dep_dt <- data.table$data.table(X = X_sim, Y = Y_sim)

ggplot2$ggplot(dep_dt, ggplot2$aes(x = X, y = Y)) +
  ggplot2$geom_point(alpha = 0.3, size = 0.8, colour = "#0072B2") +
  ggplot2$labs(
    x = "X", y = expression(Y == X^2),
    title = sprintf("Uncorrelated but dependent (r = %.4f)", cor(X_sim, Y_sim))
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

```{r uncorrelated-verify}
c(cor_XY = cor(X_sim, Y_sim), cov_XY = cov(X_sim, Y_sim))
```

The correlation is essentially zero (up to sampling noise), even though $Y$ is a deterministic function of $X$.

### 6.7.3 The Lesson

Correlation measures **linear** association only. It is blind to nonlinear relationships like $Y = X^2$, $Y = \sin(X)$, or $Y = |X|$. This is why we must always look at scatterplots in addition to computing numerical summaries.

### 6.7.4 Anscombe's Quartet Idea

```{r anscombe, fig.cap="Four datasets with nearly identical means, variances, and correlations, but dramatically different structure. Only Dataset I is well-described by a linear relationship. This demonstrates that summary statistics alone are insufficient."}
# Construct four datasets with similar statistics but different structure
set.seed(42)
n_a <- 50

d1 <- data.table$data.table(
  x = rnorm(n_a, 9, 3),
  type = "I: Linear"
)
d1[, y := 0.5 * x + rnorm(n_a, 0, 1)]

d2_x <- rnorm(n_a, 9, 3)
d2 <- data.table$data.table(
  x = d2_x,
  y = -0.1 * (d2_x - 9)^2 + 8 + rnorm(n_a, 0, 0.5),
  type = "II: Quadratic"
)

d3 <- data.table$data.table(
  x = rnorm(n_a, 9, 3),
  type = "III: Outlier"
)
d3[, y := 0.5 * x + rnorm(n_a, 0, 0.3)]
d3[1, y := 20]  # one big outlier

d4_x <- rep(8, n_a)
d4_x[1] <- 18
d4 <- data.table$data.table(
  x = d4_x + rnorm(n_a, 0, 0.1),
  type = "IV: Leverage point"
)
d4[, y := 0.5 * x + rnorm(n_a, 0, 1)]

anscombe_data <- data.table$rbindlist(list(d1, d2, d3, d4))

ggplot2$ggplot(anscombe_data, ggplot2$aes(x = x, y = y)) +
  ggplot2$geom_point(alpha = 0.6, size = 1.5, colour = "#0072B2") +
  ggplot2$geom_smooth(method = "lm", se = FALSE, colour = "#D55E00", linewidth = 0.8) +
  ggplot2$facet_wrap(~type, scales = "free", ncol = 2) +
  ggplot2$labs(x = "X", y = "Y", title = "Four datasets with similar summary statistics") +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(strip.text = ggplot2$element_text(face = "bold"))
```

```{r anscombe-stats}
anscombe_data[, .(
  mean_x = round(mean(x), 2),
  mean_y = round(mean(y), 2),
  cor = round(cor(x, y), 3)
), by = type]
```

---

## 6.8 Functions of Multiple Random Variables: The Convolution Formula

### 6.8.1 The Problem

Given two independent random variables $X$ and $Y$, what is the distribution of $Z = X + Y$?

### 6.8.2 Discrete Convolution

If $X$ and $Y$ are independent discrete random variables, the PMF of $Z = X + Y$ is:

$$P(Z = z) = \sum_x P(X = x) \cdot P(Y = z - x).$$

This is the **discrete convolution** of the PMFs of $X$ and $Y$.

*Derivation:* By the law of total probability, summing over all ways the sum can equal $z$:

$$P(Z = z) = \sum_x P(X = x, Y = z - x) = \sum_x P(X = x) \cdot P(Y = z - x)$$

where the second equality uses independence.

### 6.8.3 Continuous Convolution

For independent continuous $X$ and $Y$ with PDFs $f_X$ and $f_Y$, the PDF of $Z = X + Y$ is:

$$f_Z(z) = \int_{-\infty}^{\infty} f_X(x) \cdot f_Y(z - x) \, dx = (f_X * f_Y)(z).$$

The symbol $*$ denotes convolution.

### 6.8.4 Example: Sum of Two Uniforms

Let $X, Y \sim \text{Uniform}(0, 1)$ independently. We derive the PDF of $Z = X + Y$.

For $0 < z < 1$:

$$f_Z(z) = \int_0^z 1 \cdot 1 \, dx = z.$$

For $1 < z < 2$:

$$f_Z(z) = \int_{z-1}^1 1 \cdot 1 \, dx = 2 - z.$$

So:

$$f_Z(z) = \begin{cases} z & 0 < z < 1 \\ 2 - z & 1 < z < 2 \\ 0 & \text{otherwise} \end{cases}$$

This is the **triangular distribution** on $[0, 2]$ -- the same shape as the sum-of-two-dice distribution, but continuous.

### 6.8.5 Simulation Verification

```{r convolution-uniform, fig.cap="Sum of two independent Uniform(0,1) random variables. The simulated histogram follows the theoretical triangular density (orange line), confirming the convolution formula."}
set.seed(42)
n_sim <- 50000
Z_sim <- runif(n_sim) + runif(n_sim)

conv_dt <- data.table$data.table(Z = Z_sim)

tri_pdf <- function(z) {
  ifelse(z >= 0 & z < 1, z,
    ifelse(z >= 1 & z <= 2, 2 - z, 0))
}

ggplot2$ggplot(conv_dt, ggplot2$aes(x = Z)) +
  ggplot2$geom_histogram(
    ggplot2$aes(y = ggplot2::after_stat(density)),
    bins = 50, fill = "#0072B2", alpha = 0.6
  ) +
  ggplot2$stat_function(fun = tri_pdf, colour = "#D55E00", linewidth = 1.2) +
  ggplot2$labs(
    x = "Z = X + Y", y = "Density",
    title = "Convolution: sum of two Uniform(0, 1) variables"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

### 6.8.6 Example: Sum of Two Independent Normals

Let $X \sim N(\mu_1, \sigma_1^2)$ and $Y \sim N(\mu_2, \sigma_2^2)$ be independent. Using the convolution integral (or, more elegantly, the MGF technique from Chapter 3), we can show:

$$X + Y \sim N(\mu_1 + \mu_2, \; \sigma_1^2 + \sigma_2^2).$$

This is one of the most important results in probability: **the sum of independent normals is normal**. The means add, and the variances add. We shall return to this property in Part 3 when we study the multivariate normal.

---

## 6.9 Applied Example: NHANES Height and Weight

```{r load-nhanes}
data_dir <- "../../../data"
nhanes <- data.table$fread(file.path(data_dir, "primary/nhanes.csv"))

# Keep adults with complete height and weight data
nhanes_adults <- nhanes[Age >= 18 & !is.na(Height) & !is.na(Weight)]
nhanes_adults[, .N]
```

### 6.9.1 Scatterplot with Correlation

```{r nhanes-scatter, fig.cap="Height vs. weight for adults in NHANES. The correlation is strongly positive, reflecting the biological relationship between the two measurements. The blue line is the least-squares fit."}
r_hw <- cor(nhanes_adults$Height, nhanes_adults$Weight)

ggplot2$ggplot(nhanes_adults, ggplot2$aes(x = Height, y = Weight)) +
  ggplot2$geom_point(alpha = 0.1, size = 0.5, colour = "#0072B2") +
  ggplot2$geom_smooth(method = "lm", se = FALSE, colour = "#D55E00", linewidth = 1) +
  ggplot2$labs(
    x = "Height (cm)",
    y = "Weight (kg)",
    title = sprintf("Height vs. weight (r = %.3f)", r_hw)
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

### 6.9.2 Correlation Matrix

```{r nhanes-cor-matrix}
# Select numeric columns of interest
cols <- c("Height", "Weight", "Age", "BMI")
available <- intersect(cols, names(nhanes_adults))
cor_mat <- cor(nhanes_adults[, ..available], use = "complete.obs")
round(cor_mat, 3)
```

### 6.9.3 Verifying Our Implementation

```{r nhanes-verify}
c(
  my_cov = my_cov(nhanes_adults$Height, nhanes_adults$Weight),
  builtin_cov = cov(nhanes_adults$Height, nhanes_adults$Weight),
  my_cor = my_cor(nhanes_adults$Height, nhanes_adults$Weight),
  builtin_cor = cor(nhanes_adults$Height, nhanes_adults$Weight)
)
```

---

## Communicating to Stakeholders

When presenting covariance and correlation to a non-technical audience:

- Never report covariance alone -- it depends on the units. Always report correlation as well: "The correlation is 0.65, meaning there is a moderately strong positive linear relationship."
- Emphasise that correlation measures **linear** relationships: "A correlation near zero does *not* mean the variables are unrelated -- only that there is no straight-line pattern."
- Always show a scatterplot alongside the correlation number. The Anscombe quartet demonstrates that the same correlation can arise from wildly different data patterns.
- Use language like "as X increases, Y *tends* to increase" rather than "X causes Y to increase." Correlation does not imply causation.
- When comparing groups, show the correlation within each group separately. An overall correlation can be misleading if the groups have different means (Simpson's paradox).

---

## Quick Reference

| Quantity | Formula | Notes |
|:---------|:--------|:------|
| **Covariance** | $\text{Cov}(X,Y) = E[XY] - E[X]E[Y]$ | Units: (units of X) $\times$ (units of Y) |
| **Correlation** | $\rho = \text{Cov}(X,Y) / (\sigma_X \sigma_Y)$ | Dimensionless, $-1 \le \rho \le 1$ |
| **Cauchy-Schwarz** | $(E[UV])^2 \le E[U^2] E[V^2]$ | Guarantees $|\rho| \le 1$ |
| **Var(X+Y)** | $\text{Var}(X) + \text{Var}(Y) + 2\text{Cov}(X,Y)$ | Reduces to sum of variances when independent |
| **Discrete convolution** | $P(Z=z) = \sum_x P(X=x) P(Y=z-x)$ | For $Z = X + Y$, $X \perp Y$ |
| **Continuous convolution** | $f_Z(z) = \int f_X(x) f_Y(z-x) dx$ | For $Z = X + Y$, $X \perp Y$ |
| **R functions** | `cov(x,y)`, `cor(x,y)` | Sample versions with denominator $n-1$ |

**Key conceptual point:** $\text{Cov}(X,Y) = 0 \;\not\!\!\!\Rightarrow\; X \perp Y$. Independence implies zero covariance, but zero covariance does **not** imply independence. Correlation captures only *linear* dependence.

---

## Exercises

### Derivation Exercises

**Exercise 6.10.** Prove the **general variance formula** for a linear combination: if $Y = \sum_{i=1}^n a_i X_i$, then

$$\text{Var}(Y) = \sum_{i=1}^n a_i^2 \text{Var}(X_i) + 2 \sum_{i < j} a_i a_j \text{Cov}(X_i, X_j).$$

*(Hint: expand $\text{Var}(Y) = \text{Cov}(Y, Y)$ and use bilinearity.)*

**Exercise 6.11.** Prove that $\text{Cov}(X + Y, X - Y) = \text{Var}(X) - \text{Var}(Y)$. Under what condition are $X + Y$ and $X - Y$ uncorrelated?

**Exercise 6.12.** Prove that if $\rho(X, Y) = 1$, then there exist constants $a > 0$ and $b$ such that $P(Y = aX + b) = 1$. *(Hint: examine when equality holds in Cauchy-Schwarz.)*

### Computation Exercises

**Exercise 6.13.** Let $(X, Y)$ have joint density $f(x,y) = 2$ on $0 \le y \le x \le 1$. Compute $\text{Cov}(X, Y)$ and $\rho(X, Y)$.

*Steps:*
(a) Find $E[X]$, $E[Y]$, $E[XY]$, $E[X^2]$, $E[Y^2]$.
(b) Compute $\text{Cov}(X,Y)$, $\text{Var}(X)$, $\text{Var}(Y)$.
(c) Compute $\rho$.

```{r ex-6-13, eval=FALSE}
# Numerical verification via simulation
set.seed(42)
U <- runif(100000); V <- runif(100000)
X <- pmax(U, V); Y <- pmin(U, V)
c(EX = mean(X), EY = mean(Y), EXY = mean(X * Y),
  cov = cov(X, Y), cor = cor(X, Y))
```

**Exercise 6.14.** Compute the correlation between $X$ and $Z = X + Y$ where $X$ and $Y$ are independent with $\text{Var}(X) = 4$ and $\text{Var}(Y) = 9$. *(Hint: use properties of covariance.)*

### R Exercises

**Exercise 6.15.** Using the penguins dataset, compute the correlation matrix for bill length, bill depth, flipper length, and body mass. Which pair has the strongest correlation? Does the answer change if you compute the correlation within each species separately?

**Exercise 6.16.** Implement a function `my_cor_matrix(dt, cols)` that takes a data.table and a character vector of column names and returns the sample correlation matrix computed entirely from scratch (using your `my_cov` and `my_cor` functions). Verify that it matches `cor()`.

```{r ex-6-16, eval=FALSE}
my_cor_matrix <- function(dt, cols) {
  p <- length(cols)
  mat <- matrix(NA, p, p, dimnames = list(cols, cols))
  for (i in seq_len(p)) {
    for (j in seq_len(p)) {
      mat[i, j] <- my_cor(dt[[cols[i]]], dt[[cols[j]]])
    }
  }
  mat
}
```

**Exercise 6.17.** Simulate 50,000 draws from the pair $(X, Y)$ where $X \sim N(0,1)$ and $Y = \sin(X)$. Compute $\rho(X, Y)$. Plot the scatterplot. Is the correlation close to zero? Are the variables independent?

### Spot-the-Error Exercises

**Exercise 6.18.** A student computes $\rho(X, Y) = 0.02$ and concludes: "X and Y are independent." What is the error?

**Exercise 6.19.** A colleague reports: "The covariance between height (in cm) and weight (in kg) is 85.3. After converting height to metres, the covariance became 0.853. Clearly the linear relationship got weaker." What is wrong with this reasoning? What quantity should they use instead?
