---
title: "Statistics with R II: Intermediate"
chapter: "Chapter 9: Survival Analysis"
part: "Part 3: Advanced Topics and Extensions"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, survival-analysis, parametric-survival, competing-risks, R, biomedical]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
    plot = function(x, options) {
        cap  <- options$fig.cap
        as.character(htmltools::tag(
            "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
        ))
    }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(dpi = 300, fig.width = 10, fig.height = 7, comment = "#>", warning = FALSE, collapse = FALSE, results = 'hold')
```

# Part 3: Advanced Topics and Extensions

The Kaplan-Meier method and Cox proportional hazards model form the foundation of survival analysis, but many biomedical applications require extensions. This chapter covers parametric survival models, time-dependent covariates, competing risks, and special designs like interval censoring and recurrent events. We conclude with comprehensive guidance on presenting survival analyses in publications and regulatory submissions.

```{r packages, message=FALSE, warning=FALSE}
box::use(
    data.table[...],
    ggplot2
)

library(survival)
```

```{r load_data, message=FALSE}
# Load lung cancer data
data(lung)
lung_dt <- as.data.table(lung)
lung_dt[, event := status - 1]  # 0 = censored, 1 = dead
lung_dt[, sex := factor(sex, levels = c(1, 2), labels = c("Male", "Female"))]

# Load PBC (Primary Biliary Cholangitis) data for competing risks
data(pbc)
pbc_dt <- as.data.table(pbc)
pbc_dt <- pbc_dt[!is.na(trt)]  # Keep only randomised patients

cat("Datasets Loaded:\n")
cat("  Lung cancer:", nrow(lung_dt), "patients\n")
cat("  PBC trial:", nrow(pbc_dt), "patients\n")
```

---

## Table of Contents

## 9.21 Parametric Survival Models

### 9.21.1 Why Parametric Models?

**Prose and Intuition**

The Cox model is semi-parametric—it doesn't specify the baseline hazard. **Parametric models** assume a specific distribution for survival times, offering:

**Advantages**:
- More efficient estimation (use full likelihood)
- Extrapolation beyond observed data
- Explicit survival time predictions
- Clearer interpretation of hazard shape

**Disadvantages**:
- Must specify correct distribution
- Misspecification leads to bias
- Less robust than Cox model

### 9.21.2 Common Parametric Distributions

**Mathematical Derivation**

| Distribution | Survival Function | Hazard Function | Characteristics |
|--------------|-------------------|-----------------|-----------------|
| Exponential | $S(t) = e^{-\lambda t}$ | $h(t) = \lambda$ | Constant hazard |
| Weibull | $S(t) = e^{-(\lambda t)^\gamma}$ | $h(t) = \lambda\gamma(\lambda t)^{\gamma-1}$ | Monotone hazard |
| Log-normal | $S(t) = 1 - \Phi\left(\frac{\log t - \mu}{\sigma}\right)$ | Non-monotone | Hump-shaped hazard |
| Log-logistic | $S(t) = \frac{1}{1 + (\lambda t)^\gamma}$ | Non-monotone | Similar to log-normal |
| Gamma | Complex | Various shapes | Flexible |

**Weibull model** is most common because:
- Includes exponential as special case ($\gamma = 1$)
- $\gamma > 1$: Increasing hazard (aging)
- $\gamma < 1$: Decreasing hazard (early mortality)

```{r parametric_distributions, fig.cap="Different parametric survival distributions have distinct hazard shapes"}
t_vals <- seq(0.01, 3, length.out = 200)

# Different distributions
param_dt <- data.table(
    t = rep(t_vals, 4),
    hazard = c(
        rep(1, 200),                                    # Exponential (λ=1)
        1.5 * (t_vals)^0.5,                            # Weibull (γ=1.5, λ=1)
        0.7 * (t_vals)^(-0.3),                         # Weibull (γ=0.7)
        dlnorm(t_vals, 0, 0.5) / plnorm(t_vals, 0, 0.5, lower.tail = FALSE)  # Log-normal
    ),
    distribution = factor(rep(c("Exponential (constant)",
                                 "Weibull γ>1 (increasing)",
                                 "Weibull γ<1 (decreasing)",
                                 "Log-normal (hump-shaped)"), each = 200))
)

# Cap extreme values
param_dt[hazard > 5, hazard := NA]

ggplot2$ggplot(param_dt[!is.na(hazard)], ggplot2$aes(x = t, y = hazard, colour = distribution)) +
    ggplot2$geom_line(linewidth = 1.2) +
    ggplot2$scale_colour_viridis_d(option = "C") +
    ggplot2$labs(
        title = "Hazard Functions for Different Parametric Distributions",
        subtitle = "Choice of distribution determines assumed hazard shape",
        x = "Time",
        y = "Hazard h(t)",
        colour = ""
    ) +
    ggplot2$theme_minimal() +
    ggplot2$theme(legend.position = "bottom")
```

### 9.21.3 Fitting Parametric Models

```{r parametric_fit}
# Fit various parametric models to lung data
lung_complete <- lung_dt[!is.na(ph.ecog)]

# Exponential
fit_exp <- survreg(Surv(time, event) ~ age + sex + factor(ph.ecog),
                    data = lung_complete, dist = "exponential")

# Weibull
fit_weibull <- survreg(Surv(time, event) ~ age + sex + factor(ph.ecog),
                        data = lung_complete, dist = "weibull")

# Log-normal
fit_lnorm <- survreg(Surv(time, event) ~ age + sex + factor(ph.ecog),
                      data = lung_complete, dist = "lognormal")

# Log-logistic
fit_llogis <- survreg(Surv(time, event) ~ age + sex + factor(ph.ecog),
                       data = lung_complete, dist = "loglogistic")

# Compare AIC
aic_comparison <- data.table(
    Distribution = c("Exponential", "Weibull", "Log-normal", "Log-logistic"),
    AIC = c(AIC(fit_exp), AIC(fit_weibull), AIC(fit_lnorm), AIC(fit_llogis)),
    LogLik = c(logLik(fit_exp), logLik(fit_weibull), logLik(fit_lnorm), logLik(fit_llogis))
)
aic_comparison[, Delta_AIC := AIC - min(AIC)]

cat("Parametric Model Comparison:\n")
cat("============================\n\n")
print(aic_comparison[order(AIC)])

cat("\nBest fitting distribution:", aic_comparison[which.min(AIC), Distribution], "\n")
```

### 9.21.4 Interpreting Parametric Model Output

**Prose and Intuition**

Parametric models (via `survreg`) report coefficients on the **log-time scale**, which is the opposite direction from Cox models:
- Positive coefficient → longer survival (lower hazard)
- Negative coefficient → shorter survival (higher hazard)

To convert to "acceleration factor" interpretation: $\exp(\beta)$ is the multiplicative effect on survival time.

```{r parametric_interpretation}
cat("Weibull Model Summary:\n")
cat("======================\n\n")
print(summary(fit_weibull))

# Convert coefficients to acceleration factors
weibull_coef <- coef(fit_weibull)
acceleration_factors <- exp(weibull_coef)

cat("\nAcceleration Factors (effect on survival time):\n")
cat("===============================================\n\n")
for (i in 1:length(acceleration_factors)) {
    cat(sprintf("  %s: %.3f\n", names(acceleration_factors)[i], acceleration_factors[i]))
}

cat("\nInterpretation:\n")
cat("  Females have", round((acceleration_factors["sexFemale"] - 1) * 100, 1),
    "% longer survival times than males\n")
cat("  ECOG 2 patients have", round((1 - acceleration_factors["factor(ph.ecog)2"]) * 100, 1),
    "% shorter survival than ECOG 0\n")
```

---

## 9.22 Time-Dependent Covariates

### 9.22.1 When Covariates Change Over Time

**Prose and Intuition**

So far, we've assumed covariates are fixed at baseline. But many important predictors change over time:
- Lab values (e.g., CD4 count in HIV)
- Treatment exposure (cumulative dose)
- Disease status (remission/relapse)
- Transplant status (pre/post transplant)

**Time-dependent covariates** require restructuring data so each row represents a time interval, with covariate values for that period.

### 9.22.2 Creating Time-Dependent Data

```{r time_dependent}
# Example: Create a time-varying covariate
# Suppose Karnofsky score decreases over time for some patients

set.seed(42)

# For illustration, create simplified time-varying data
# Patient had Karnofsky score change at day 100
n_patients <- 50
tv_data <- rbindlist(lapply(1:n_patients, function(id) {
    # Baseline
    baseline_kps <- sample(c(60, 70, 80, 90, 100), 1, prob = c(0.1, 0.2, 0.3, 0.3, 0.1))
    event_time <- rweibull(1, shape = 1.5, scale = 300)
    event <- rbinom(1, 1, 0.7)

    # Did KPS change?
    kps_change <- runif(1) < 0.4 & event_time > 100

    if (kps_change) {
        new_kps <- max(baseline_kps - 20, 40)
        rbind(
            data.table(id = id, tstart = 0, tstop = 100, event = 0,
                       kps = baseline_kps, age = 60 + rnorm(1, 0, 10)),
            data.table(id = id, tstart = 100, tstop = event_time, event = event,
                       kps = new_kps, age = 60 + rnorm(1, 0, 10))
        )
    } else {
        data.table(id = id, tstart = 0, tstop = event_time, event = event,
                   kps = baseline_kps, age = 60 + rnorm(1, 0, 10))
    }
}))

cat("Time-Varying Covariate Data Structure:\n")
cat("======================================\n\n")
print(head(tv_data[order(id, tstart)], 10))

# Fit Cox model with time-varying covariate
cox_tv <- coxph(Surv(tstart, tstop, event) ~ kps + age, data = tv_data)

cat("\nCox Model with Time-Varying Karnofsky Score:\n")
print(summary(cox_tv))
```

---

## 9.23 Competing Risks

### 9.23.1 The Problem

**Prose and Intuition**

In many studies, patients can experience multiple types of events:
- Cancer patients may die from cancer OR from other causes
- Transplant patients may experience rejection, infection, or death
- Patients may experience disease recurrence before death

With **competing risks**, experiencing one event precludes others. Standard Kaplan-Meier can overestimate event probabilities because it treats competing events as censoring.

**Key distinction**:
- **Cause-specific hazard**: Risk of a specific event (treating others as censored)
- **Cumulative incidence function (CIF)**: Probability of experiencing the event by time $t$, accounting for competing risks

### 9.23.2 Cumulative Incidence Functions

**Mathematical Derivation**

The **cumulative incidence function** for event type $j$:
$$F_j(t) = P(T \leq t, J = j) = \int_0^t h_j(u) S(u) \, du$$

where:
- $h_j(u)$ = cause-specific hazard for event $j$
- $S(u) = \exp(-\sum_k H_k(u))$ = overall survival (from all causes)

**Key property**: $\sum_j F_j(\infty) = 1$ (everyone eventually has one of the events or is censored)

```{r competing_risks_cif, fig.cap="Cumulative incidence functions for competing risks in PBC data"}
# PBC data has transplant and death as competing events
# status: 0 = censored, 1 = transplant, 2 = dead

# Create competing risk outcome
pbc_dt[, cr_status := factor(status, levels = c(0, 1, 2),
                              labels = c("censored", "transplant", "death"))]

# Fit cumulative incidence
# Using survfit for competing risks
cif_fit <- survfit(Surv(time, factor(status)) ~ 1, data = pbc_dt)

# Extract CIF data
cif_data <- data.table(
    time = cif_fit$time,
    transplant = cif_fit$pstate[, 2],  # State 2 = transplant
    death = cif_fit$pstate[, 3]        # State 3 = death
)

cif_long <- melt(cif_data, id.vars = "time",
                  variable.name = "event", value.name = "probability")

ggplot2$ggplot(cif_long, ggplot2$aes(x = time/365, y = probability, colour = event)) +
    ggplot2$geom_step(linewidth = 1.2) +
    ggplot2$scale_colour_manual(values = c("transplant" = "#0072B2", "death" = "#D55E00"),
                                 labels = c("Transplant", "Death")) +
    ggplot2$scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    ggplot2$labs(
        title = "Cumulative Incidence Functions for Competing Risks",
        subtitle = "Primary Biliary Cholangitis: Death vs Liver Transplant",
        x = "Time (years)",
        y = "Cumulative Incidence",
        colour = "Event"
    ) +
    ggplot2$theme_minimal() +
    ggplot2$theme(legend.position = "bottom")
```

### 9.23.3 Fine-Gray Model for Competing Risks

**Prose and Intuition**

The **Fine-Gray model** directly models the subdistribution hazard—the hazard for a specific event type while keeping patients who experienced other events "at risk":

$$h^{sub}_j(t|X) = h^{sub}_{0j}(t) \exp(\beta' X)$$

This differs from cause-specific Cox models and directly relates to cumulative incidence.

```{r fine_gray}
# Fit Fine-Gray model
if (requireNamespace("cmprsk", quietly = TRUE)) {
    library(cmprsk)

    # Prepare data
    pbc_fg <- pbc_dt[, .(time, status, trt, age, sex, bili)]
    pbc_fg <- pbc_fg[complete.cases(pbc_fg)]

    # Create covariate matrix
    cov_mat <- model.matrix(~ factor(trt) + age + factor(sex) + log(bili), data = pbc_fg)[, -1]

    # Fine-Gray for death (event = 2)
    fg_death <- crr(
        ftime = pbc_fg$time,
        fstatus = pbc_fg$status,
        cov1 = cov_mat,
        failcode = 2
    )

    cat("Fine-Gray Model for Death (competing with transplant):\n")
    cat("=======================================================\n\n")
    print(summary(fg_death))
} else {
    cat("Install 'cmprsk' package for Fine-Gray models.\n")
}
```

---

## 9.24 Special Survival Designs

### 9.24.1 Interval Censoring

**Prose and Intuition**

Sometimes we only know that an event occurred between two observation times:
- Patient was disease-free at month 6 visit, had disease at month 12 visit
- The event happened somewhere in that interval

This is **interval censoring**. Standard methods assume exact event times (or right censoring only).

```{r interval_censoring}
# Simulate interval-censored data
set.seed(42)
n <- 100

# True event times (unknown)
true_times <- rweibull(n, shape = 2, scale = 20)

# Observation times: visits at months 6, 12, 18, 24
visit_times <- c(0, 6, 12, 18, 24, Inf)

# Find interval containing true event
interval_dt <- data.table(
    id = 1:n,
    true_time = true_times
)

interval_dt[, left := sapply(true_time, function(t) max(visit_times[visit_times < t]))]
interval_dt[, right := sapply(true_time, function(t) min(visit_times[visit_times >= t]))]

cat("Interval Censoring Example:\n")
cat("===========================\n\n")
cat("True times are unknown; we only know event occurred in interval [left, right]\n\n")
print(head(interval_dt, 10))

# Fit interval-censored model (requires survival package update)
# Create Surv object for interval censoring
surv_int <- Surv(time = interval_dt$left, time2 = interval_dt$right, type = "interval2")

cat("\nInterval-Censored Kaplan-Meier (Turnbull estimator):\n")
fit_int <- survfit(surv_int ~ 1)
print(fit_int)
```

### 9.24.2 Left Truncation (Delayed Entry)

**Prose and Intuition**

**Left truncation** occurs when patients only enter the study after surviving to a certain point:
- Study recruits only patients who survived initial surgery
- Registry-based studies where patients must survive to be registered

Without adjustment, we overestimate survival.

```{r left_truncation}
# Example: Study of patients who survived initial diagnosis
# Entry time = time from diagnosis to study enrollment

set.seed(42)
n <- 100

# Time from diagnosis to enrollment (everyone survived this)
entry_time <- runif(n, 0, 100)

# Additional follow-up time
followup <- rexp(n, rate = 0.01)
exit_time <- entry_time + followup

# Event indicator
event <- rbinom(n, 1, 0.6)

trunc_dt <- data.table(
    id = 1:n,
    entry = entry_time,
    exit = exit_time,
    event = event
)

cat("Left Truncation (Delayed Entry) Example:\n")
cat("========================================\n\n")
cat("Patients entered study at time 'entry' after diagnosis\n")
cat("They are at risk only from 'entry' to 'exit'\n\n")

# Fit model with left truncation
fit_trunc <- survfit(Surv(entry, exit, event) ~ 1, data = trunc_dt)

cat("Survival estimate accounting for left truncation:\n")
print(fit_trunc)
```

### 9.24.3 Recurrent Events

**Prose and Intuition**

Some outcomes can happen multiple times:
- Hospital readmissions
- Seizure episodes
- Infection occurrences

Analysis options:
- **Time to first event**: Simple but ignores information
- **Counting process**: Each event counted separately
- **Frailty models**: Account for within-patient correlation
- **Marginal models**: Population-averaged effects

```{r recurrent_events}
# Simulate recurrent event data
set.seed(42)
n_patients <- 50
max_events <- 5

recurrent_dt <- rbindlist(lapply(1:n_patients, function(id) {
    # Patient-specific frailty
    frailty <- rgamma(1, shape = 1, rate = 1)

    events <- list()
    current_time <- 0

    for (k in 1:max_events) {
        gap <- rexp(1, rate = 0.1 * frailty)
        current_time <- current_time + gap

        if (current_time > 100) break

        events[[k]] <- data.table(
            id = id,
            event_num = k,
            tstart = current_time - gap,
            tstop = current_time,
            event = 1,
            frailty = frailty
        )
    }

    # Add final censored period
    if (length(events) > 0) {
        last_event <- events[[length(events)]]$tstop
    } else {
        last_event <- 0
    }

    events[[length(events) + 1]] <- data.table(
        id = id,
        event_num = length(events) + 1,
        tstart = last_event,
        tstop = 100,
        event = 0,
        frailty = frailty
    )

    rbindlist(events)
}))

cat("Recurrent Event Data Structure:\n")
cat("===============================\n\n")
print(head(recurrent_dt[order(id, tstart)], 15))

# Fit Andersen-Gill model (counting process)
cox_recurrent <- coxph(Surv(tstart, tstop, event) ~ 1 + cluster(id),
                        data = recurrent_dt)

cat("\nAndersen-Gill Model for Recurrent Events:\n")
print(summary(cox_recurrent))
```

---

## 9.25 Restricted Mean Survival Time

### 9.25.1 An Alternative to Hazard Ratios

**Prose and Intuition**

The **restricted mean survival time (RMST)** offers an intuitive alternative to hazard ratios:
$$\text{RMST}(\tau) = \int_0^{\tau} S(t) \, dt$$

This is the average survival time up to a time horizon $\tau$, or equivalently, the area under the survival curve up to $\tau$.

**Advantages**:
- Doesn't require proportional hazards
- Directly interpretable (years/months of life)
- Valid even when curves cross

```{r rmst, fig.cap="Restricted mean survival time is the area under the survival curve"}
# Calculate RMST
km_fit <- survfit(Surv(time, event) ~ sex, data = lung_dt)

# Set restriction time (e.g., 365 days)
tau <- 365

# Calculate RMST
if (requireNamespace("survRM2", quietly = TRUE)) {
    library(survRM2)

    rmst_result <- rmst2(time = lung_dt$time, status = lung_dt$event,
                          arm = as.numeric(lung_dt$sex) - 1, tau = tau)

    cat("Restricted Mean Survival Time Analysis (τ =", tau, "days):\n")
    cat("========================================================\n\n")
    print(rmst_result)
} else {
    cat("Install 'survRM2' package for RMST analysis.\n")

    # Manual approximation
    cat("\nManual RMST approximation:\n")
    for (s in c("Male", "Female")) {
        km_s <- survfit(Surv(time, event) ~ 1, data = lung_dt[sex == s])
        # Approximate integral
        times <- c(0, km_s$time[km_s$time <= tau], tau)
        surv <- c(1, km_s$surv[km_s$time <= tau],
                  km_s$surv[max(which(km_s$time <= tau))])
        rmst_approx <- sum(diff(times) * head(surv, -1))
        cat(sprintf("  %s: RMST = %.1f days\n", s, rmst_approx))
    }
}
```

---

## 9.26 Presenting Survival Analysis in Publications

### 9.26.1 STROBE and CONSORT Guidelines

**Prose and Intuition**

For observational studies (STROBE) and clinical trials (CONSORT), survival analyses should include:

**Essential Elements**:
1. Flow diagram showing patient disposition
2. Baseline characteristics table
3. Number at risk over time
4. Median follow-up time
5. Number of events and censored observations
6. Kaplan-Meier curves with confidence bands
7. Median survival with 95% CI (or other quantiles)
8. Hazard ratios with 95% CI and p-values
9. Model diagnostics (especially PH assumption testing)

### 9.26.2 Example Publication-Quality Table

```{r publication_table}
# Create publication-quality results table
lung_complete <- lung_dt[!is.na(ph.ecog)]

# Cox model
cox_final <- coxph(Surv(time, event) ~ age + sex + factor(ph.ecog),
                    data = lung_complete)

# Extract results
cox_summ <- summary(cox_final)
hr_table <- data.table(
    Variable = c("Age (per year)", "Sex: Female vs Male",
                 "ECOG 1 vs 0", "ECOG 2 vs 0", "ECOG 3 vs 0"),
    N = nrow(lung_complete),
    Events = sum(lung_complete$event),
    HR = sprintf("%.2f", cox_summ$conf.int[, "exp(coef)"]),
    CI = sprintf("(%.2f-%.2f)",
                 cox_summ$conf.int[, "lower .95"],
                 cox_summ$conf.int[, "upper .95"]),
    p = format.pval(cox_summ$coefficients[, "Pr(>|z|)"], digits = 3)
)

cat("Table 2: Cox Proportional Hazards Analysis of Survival\n")
cat("======================================================\n\n")
print(hr_table)

cat("\nModel fit: Concordance =", round(concordance(cox_final)$concordance, 3), "\n")
cat("PH assumption: Global test p =", round(cox.zph(cox_final)$table["GLOBAL", "p"], 3), "\n")
```

### 9.26.3 Publication-Quality Figures

```{r publication_figure, fig.cap="Publication-quality Kaplan-Meier curves following reporting guidelines"}
# Create publication-quality KM plot
km_sex <- survfit(Surv(time, event) ~ sex, data = lung_dt)

# Extract data
km_data <- data.table(
    time = rep(c(0, km_sex$time), 2),
    survival = c(1, km_sex$surv[1:km_sex$strata[1]],
                 1, km_sex$surv[(km_sex$strata[1]+1):sum(km_sex$strata)]),
    lower = c(1, km_sex$lower[1:km_sex$strata[1]],
              1, km_sex$lower[(km_sex$strata[1]+1):sum(km_sex$strata)]),
    upper = c(1, km_sex$upper[1:km_sex$strata[1]],
              1, km_sex$upper[(km_sex$strata[1]+1):sum(km_sex$strata)]),
    sex = rep(c("Male", "Female"), c(km_sex$strata[1]+1, km_sex$strata[2]+1))
)

# Risk table at specific times
risk_times <- c(0, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000)
risk_table <- rbindlist(lapply(c("Male", "Female"), function(s) {
    km_s <- survfit(Surv(time, event) ~ 1, data = lung_dt[sex == s])
    data.table(
        time = risk_times,
        n_risk = sapply(risk_times, function(t) sum(lung_dt[sex == s, time >= t])),
        sex = s
    )
}))

# Main plot
p_main <- ggplot2$ggplot(km_data, ggplot2$aes(x = time, colour = sex, fill = sex)) +
    ggplot2$geom_ribbon(ggplot2$aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
    ggplot2$geom_step(ggplot2$aes(y = survival), linewidth = 1) +
    ggplot2$scale_colour_manual(values = c("Male" = "#0072B2", "Female" = "#D55E00")) +
    ggplot2$scale_fill_manual(values = c("Male" = "#0072B2", "Female" = "#D55E00")) +
    ggplot2$scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    ggplot2$scale_x_continuous(breaks = seq(0, 1000, 200)) +
    ggplot2$annotate("text", x = 700, y = 0.85,
                     label = sprintf("Log-rank p = %.3f",
                                     1-pchisq(survdiff(Surv(time, event) ~ sex, lung_dt)$chisq, 1)),
                     size = 4) +
    ggplot2$labs(
        title = "Overall Survival by Sex in Advanced Lung Cancer",
        x = "Time (days)",
        y = "Survival Probability",
        colour = "", fill = ""
    ) +
    ggplot2$theme_minimal() +
    ggplot2$theme(
        legend.position = c(0.85, 0.85),
        plot.margin = ggplot2$margin(10, 10, 5, 10)
    )

print(p_main)

cat("\nNumber at Risk Table:\n")
print(dcast(risk_table, sex ~ time, value.var = "n_risk"))
```

---

## 9.27 Summary and Key Takeaways

**Advanced Survival Methods Overview**

| Method | When to Use | R Function |
|--------|-------------|------------|
| Parametric models | Specific distribution assumed | `survreg()` |
| Time-dependent covariates | Predictors change over time | `coxph()` with `(start, stop)` |
| Competing risks (CIF) | Multiple event types | `survfit()` with factor status |
| Fine-Gray model | Regression with competing risks | `crr()` (cmprsk) |
| Interval censoring | Event in interval, not exact | `Surv(type="interval2")` |
| Left truncation | Delayed entry | `Surv(entry, exit, event)` |
| Recurrent events | Multiple events per person | `cluster()` or frailty |
| RMST | Non-PH or crossing curves | `rmst2()` (survRM2) |

**Publication Checklist**

1. ☑ Report median follow-up time
2. ☑ State number of events and censoring
3. ☑ Present Kaplan-Meier curves with CI
4. ☑ Include number at risk table
5. ☑ Report hazard ratios with 95% CI
6. ☑ Test and report PH assumption
7. ☑ State model discrimination (C-statistic)
8. ☑ Consider sensitivity analyses

**R Packages for Survival Analysis**

| Package | Purpose |
|---------|---------|
| `survival` | Core survival functions |
| `survminer` | Publication-quality plots |
| `cmprsk` | Competing risks |
| `survRM2` | RMST analysis |
| `frailtypack` | Frailty models |
| `flexsurv` | Parametric models |

---

## 9.28 Exercises

1. **Parametric model selection**: Compare exponential, Weibull, and log-normal models for the lung cancer data. Which fits best?

2. **Competing risks**: Using the PBC data, estimate the 5-year cumulative incidence of death and transplant.

3. **RMST analysis**: Calculate the difference in RMST between males and females at 1 year. Is this clinically meaningful?

4. **Publication table**: Create a complete Table 1 and Table 2 for a survival analysis of your choice.

5. **Time-dependent analysis**: Design a study where treatment status changes over time. How would you structure the data?

---

## 9.29 Course Summary: Survival Analysis

Across three chapters, we have covered:

**Part 1: Fundamentals**
- Survival and hazard functions
- Censoring and its implications
- Kaplan-Meier estimation
- Log-rank tests

**Part 2: Cox Model**
- Proportional hazards regression
- Hazard ratio interpretation
- PH assumption testing
- Model diagnostics

**Part 3: Advanced Topics**
- Parametric models
- Time-dependent covariates
- Competing risks
- Special designs (interval censoring, left truncation, recurrent events)
- RMST as an alternative measure
- Publication guidelines

Survival analysis is essential for any researcher working with time-to-event data in medicine and public health.

---

## 9.30 References

- Therneau, T. M., & Grambsch, P. M. (2000). *Modeling Survival Data: Extending the Cox Model*. Springer.
- Collett, D. (2015). *Modelling Survival Data in Medical Research* (3rd ed.). CRC Press.
- Fine, J. P., & Gray, R. J. (1999). A proportional hazards model for the subdistribution of a competing risk. *JASA*, 94(446), 496-509.
- Uno, H., et al. (2014). Moving beyond the hazard ratio in quantifying the between-group difference in survival analysis. *JCO*, 32(22), 2380-2385.
- Royston, P., & Parmar, M. K. (2013). Restricted mean survival time: an alternative to the hazard ratio for the design and analysis of randomized trials. *BMC Medical Research Methodology*, 13, 152.
