---
title: "Dispersion and Correlation Trading"
subtitle: "Trading the relationship between index and single-stock volatility"
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
box::use(
    data.table[...],
    ggplot2[...]
)

box::use(
    ../modules/data[load_market, load_vix, filter_dates],
    ../modules/stats[sharpe_ratio, annualised_return, annualised_vol, max_drawdown],
    ../modules/viz[theme_trading, trading_colors]
)

# Helper function for merging by date
merge_by_date <- function(dt1, dt2) {
    merge(dt1, dt2, by = "date", all = FALSE)
}

knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 10,
    fig.height = 6,
    fig.path = ""
)

tc <- unlist(trading_colors)
```

## 10.7 Dispersion Trading Foundations

### 10.7.1 Prose & Intuition

Dispersion trading exploits the relationship between index volatility and individual stock volatilities. The key insight: an index is a portfolio of stocks, so its volatility depends on both individual stock volatilities and their correlations.

When correlations are low, stocks move independently. The index—being an average—experiences lower volatility than its components. When correlations spike (typically during crises), stocks move together, and index volatility approaches the weighted average of component volatilities.

The **correlation risk premium** is the tendency for implied correlation to exceed realised correlation. This creates a structural opportunity: sell index variance (expensive due to correlation premium) and buy single-stock variance (cheaper).

This strategy is called **dispersion trading** because it profits when stocks "disperse" (move differently from each other) more than implied correlation suggested.

### 10.7.2 Visual Evidence

```{r load-multi-asset}
# Load multiple assets to demonstrate dispersion
spy <- load_market("SPY")
aapl <- load_market("AAPL")
msft <- load_market("MSFT")
amzn <- load_market("AMZN")
goog <- load_market("GOOG")
nvda <- load_market("NVDA")

# Filter to common dates
start_date <- as.Date("2015-01-01")
end_date <- as.Date("2023-12-31")

spy <- filter_dates(spy, start_date, end_date)
aapl <- filter_dates(aapl, start_date, end_date)
msft <- filter_dates(msft, start_date, end_date)
amzn <- filter_dates(amzn, start_date, end_date)
goog <- filter_dates(goog, start_date, end_date)
nvda <- filter_dates(nvda, start_date, end_date)

# Merge all
multi <- merge_by_date(spy[, .(date, spy = adjusted)],
                       aapl[, .(date, aapl = adjusted)])
multi <- merge_by_date(multi, msft[, .(date, msft = adjusted)])
multi <- merge_by_date(multi, amzn[, .(date, amzn = adjusted)])
multi <- merge_by_date(multi, goog[, .(date, goog = adjusted)])
multi <- merge_by_date(multi, nvda[, .(date, nvda = adjusted)])

# Calculate returns
multi[, spy_ret := c(NA, diff(log(spy)))]
multi[, aapl_ret := c(NA, diff(log(aapl)))]
multi[, msft_ret := c(NA, diff(log(msft)))]
multi[, amzn_ret := c(NA, diff(log(amzn)))]
multi[, goog_ret := c(NA, diff(log(goog)))]
multi[, nvda_ret := c(NA, diff(log(nvda)))]

multi <- multi[!is.na(spy_ret)]
```

```{r rolling-correlation, fig.cap="Rolling pairwise correlations spike during crises. This explains why index vol approaches component vol in stress."}
# Calculate rolling correlations between stocks
calc_rolling_cor <- function(x, y, n = 63) {
    result <- rep(NA_real_, length(x))
    for (i in n:length(x)) {
        idx <- (i - n + 1):i
        result[i] <- cor(x[idx], y[idx], use = "complete.obs")
    }
    return(result)
}

# Pairwise correlations
multi[, cor_aapl_msft := calc_rolling_cor(aapl_ret, msft_ret)]
multi[, cor_aapl_amzn := calc_rolling_cor(aapl_ret, amzn_ret)]
multi[, cor_msft_amzn := calc_rolling_cor(msft_ret, amzn_ret)]

# Average correlation
multi[, avg_corr := (cor_aapl_msft + cor_aapl_amzn + cor_msft_amzn) / 3]

# Plot correlations over time
plot_corr <- melt(multi[!is.na(avg_corr), .(date, `AAPL-MSFT` = cor_aapl_msft,
                                             `AAPL-AMZN` = cor_aapl_amzn,
                                             `MSFT-AMZN` = cor_msft_amzn,
                                             Average = avg_corr)],
                  id.vars = "date", variable.name = "Pair", value.name = "Correlation")

ggplot(plot_corr[Pair == "Average"], aes(x = date, y = Correlation)) +
    geom_line(colour = tc[1], linewidth = 0.5) +
    geom_hline(yintercept = mean(multi$avg_corr, na.rm = TRUE),
               linetype = "dashed", colour = tc[2]) +
    annotate("text", x = as.Date("2023-01-01"),
             y = mean(multi$avg_corr, na.rm = TRUE) + 0.05,
             label = sprintf("Mean: %.2f", mean(multi$avg_corr, na.rm = TRUE)),
             colour = tc[2]) +
    labs(title = "Average Pairwise Correlation (63-Day Rolling)",
         subtitle = "AAPL, MSFT, AMZN; Spikes during March 2020 crisis",
         x = NULL, y = "Correlation") +
    theme_trading()
```

```{r dispersion-measure, fig.cap="Dispersion measures cross-sectional return variability. High dispersion = stocks moving differently."}
# Calculate cross-sectional dispersion
# Dispersion = std dev of stock returns on each day
multi[, dispersion := apply(.SD, 1, sd, na.rm = TRUE),
      .SDcols = c("aapl_ret", "msft_ret", "amzn_ret", "goog_ret", "nvda_ret")]

# Annualise
multi[, dispersion_ann := dispersion * sqrt(252) * 100]

# Rolling dispersion
multi[, disp_20d := frollmean(dispersion_ann, 20)]

# Plot dispersion
ggplot(multi[!is.na(disp_20d)], aes(x = date, y = disp_20d)) +
    geom_line(colour = tc[1], linewidth = 0.5) +
    geom_hline(yintercept = median(multi$disp_20d, na.rm = TRUE),
               linetype = "dashed", colour = tc[2]) +
    labs(title = "Cross-Sectional Dispersion (20-Day Average)",
         subtitle = "Higher dispersion = stocks moving independently",
         x = NULL, y = "Dispersion (Ann. %)") +
    theme_trading()
```

```{r correlation-dispersion-relationship, fig.cap="Correlation and dispersion move inversely. Low correlation enables high dispersion."}
# Scatter: correlation vs dispersion
scatter_cd <- multi[!is.na(avg_corr) & !is.na(disp_20d)]

ggplot(scatter_cd, aes(x = avg_corr, y = disp_20d)) +
    geom_point(alpha = 0.3, colour = tc[1]) +
    geom_smooth(method = "lm", se = TRUE, colour = tc[2]) +
    labs(title = "Correlation vs Dispersion",
         subtitle = sprintf("Correlation: %.2f",
                           cor(scatter_cd$avg_corr, scatter_cd$disp_20d)),
         x = "Average Pairwise Correlation", y = "Dispersion (Ann. %)") +
    theme_trading()
```

### 10.7.3 Mathematical Derivation

**Index Variance Decomposition**

For an index $I$ with weights $w_i$ and component returns $r_i$:

$$
\sigma_I^2 = \sum_i w_i^2 \sigma_i^2 + 2\sum_{i<j} w_i w_j \sigma_i \sigma_j \rho_{ij}
$$

This can be rewritten as:

$$
\sigma_I^2 = \bar{\sigma}^2 \cdot \bar{\rho}
$$

where $\bar{\sigma}^2$ is weighted average variance and $\bar{\rho}$ is average correlation.

**Implied Correlation**

Given index implied vol ($IV_I$) and component implied vols ($IV_i$):

$$
\rho_{impl} = \frac{IV_I^2 - \sum_i w_i^2 IV_i^2}{2 \sum_{i<j} w_i w_j IV_i IV_j}
$$

Simplified for equal weights:

$$
\rho_{impl} \approx \frac{IV_I^2}{\bar{IV}^2}
$$

**Dispersion Trade Payoff**

Selling index variance and buying component variance:

$$
P\&L = N \cdot \left[ \sum_i w_i (\sigma_i^{RV})^2 - (\sigma_I^{RV})^2 \right]
$$

Profits when realised correlation is lower than implied.

**Correlation Risk Premium**

$$
CRP = \rho_{impl} - E[\rho_{realised}]
$$

Empirically, $CRP > 0$, making dispersion trading systematically profitable.

### 10.7.4 Implementation & Application

```{r index-vs-component-vol, fig.cap="Index vs average component volatility. Diversification benefit visible in normal times."}
# Calculate volatilities
multi[, spy_vol := frollapply(spy_ret, 20, function(x) sd(x) * sqrt(252)) * 100]

# Average component vol
multi[, avg_stock_vol := (
    frollapply(aapl_ret, 20, function(x) sd(x) * sqrt(252)) +
    frollapply(msft_ret, 20, function(x) sd(x) * sqrt(252)) +
    frollapply(amzn_ret, 20, function(x) sd(x) * sqrt(252)) +
    frollapply(goog_ret, 20, function(x) sd(x) * sqrt(252)) +
    frollapply(nvda_ret, 20, function(x) sd(x) * sqrt(252))
) / 5 * 100]

# Plot comparison
plot_vol <- melt(multi[!is.na(spy_vol) & !is.na(avg_stock_vol),
                       .(date, `SPY Vol` = spy_vol, `Avg Stock Vol` = avg_stock_vol)],
                 id.vars = "date", variable.name = "Type", value.name = "Volatility")

ggplot(plot_vol, aes(x = date, y = Volatility, colour = Type)) +
    geom_line(linewidth = 0.5) +
    scale_colour_manual(values = c(tc[1], tc[3])) +
    labs(title = "SPY Volatility vs Average Component Volatility",
         subtitle = "Gap = diversification benefit; narrows during crises",
         x = NULL, y = "20-Day Realised Vol (%)") +
    theme_trading() +
    theme(legend.position = "top")
```

```{r implied-realised-corr, fig.cap="Implied correlation proxy vs realised. Implied consistently overestimates realised correlation."}
# Implied correlation proxy using VIX and component vols
vix <- load_vix("vix")
vix <- filter_dates(vix, start_date, end_date)

# Merge VIX
multi <- merge_by_date(multi, vix[, .(date, vix = close)])

# Implied correlation approximation: VIX^2 / avg(stock_IV)^2
# We'll use realised vol as IV proxy (since we don't have options data)
multi[, implied_corr_proxy := (vix / avg_stock_vol)^2]
multi[, implied_corr_proxy := pmin(pmax(implied_corr_proxy, 0), 1)]

# Compare implied vs realised correlation
multi[, realised_corr := avg_corr]

# Plot comparison
plot_impl_real <- melt(multi[!is.na(implied_corr_proxy) & !is.na(realised_corr),
                             .(date, Implied = implied_corr_proxy, Realised = realised_corr)],
                       id.vars = "date", variable.name = "Type", value.name = "Correlation")

ggplot(plot_impl_real, aes(x = date, y = Correlation, colour = Type)) +
    geom_line(linewidth = 0.5, alpha = 0.8) +
    scale_colour_manual(values = c(tc[1], tc[3])) +
    labs(title = "Implied vs Realised Correlation (Proxy)",
         subtitle = "Implied (from VIX) typically exceeds realised pairwise correlation",
         x = NULL, y = "Correlation") +
    theme_trading() +
    theme(legend.position = "top")
```

```{r dispersion-strategy, fig.cap="Dispersion trading strategy simulation. Long component vol, short index vol."}
# Dispersion trade: Long stock variance, short index variance
# P&L = avg(stock_var) - index_var

multi[, stock_var := (avg_stock_vol / 100)^2]
multi[, index_var := (spy_vol / 100)^2]

# Daily P&L proxy: change in (stock_var - index_var)
multi[, var_spread := stock_var - index_var]
multi[, disp_ret := c(NA, diff(var_spread)) * 10000]  # Scale for readability

# Cumulative
disp_strat <- multi[!is.na(disp_ret)]
disp_strat[, cum_disp := cumsum(disp_ret)]

ggplot(disp_strat, aes(x = date, y = cum_disp)) +
    geom_line(colour = tc[1], linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title = "Dispersion Trade P&L (Simulated)",
         subtitle = "Long stock variance, short index variance; profits from low correlation",
         x = NULL, y = "Cumulative P&L") +
    theme_trading()
```

---

## 10.8 Correlation Swaps

### 10.8.1 Prose & Intuition

Correlation swaps provide pure exposure to average pairwise correlation without variance exposure. The payoff is simply the difference between realised correlation and struck correlation.

Unlike dispersion trades that mix variance and correlation exposures, correlation swaps isolate the correlation component. This makes them ideal for expressing views on correlation regime changes.

The challenge is replication. Unlike variance swaps (which can be replicated with options), correlation swaps require dynamic hedging of dispersion positions. This makes them predominantly over-the-counter products traded between sophisticated institutions.

The correlation risk premium makes being short correlation structurally profitable, but with significant tail risk when correlations spike during crises.

### 10.8.2 Visual Evidence

```{r correlation-regimes, fig.cap="Correlation regime identification. Low, medium, and high correlation regimes have different implications."}
# Define correlation regimes
multi[, corr_regime := cut(realised_corr,
                           breaks = c(-1, 0.3, 0.6, 1),
                           labels = c("Low (<0.3)", "Medium (0.3-0.6)", "High (>0.6)"))]

# Count days in each regime
regime_stats <- multi[!is.na(corr_regime), .(
    days = .N,
    pct = .N / nrow(multi[!is.na(corr_regime)]) * 100,
    avg_spy_vol = mean(spy_vol, na.rm = TRUE),
    avg_disp = mean(disp_20d, na.rm = TRUE)
), by = corr_regime]

print(regime_stats)

# Plot regime distribution
ggplot(multi[!is.na(corr_regime)], aes(x = realised_corr, fill = corr_regime)) +
    geom_histogram(bins = 50, alpha = 0.8, colour = "white") +
    scale_fill_manual(values = c(tc[1], tc[2], tc[4])) +
    labs(title = "Distribution of Realised Correlation",
         subtitle = "Most time spent in medium correlation regime",
         x = "Realised Correlation", y = "Count", fill = "Regime") +
    theme_trading() +
    theme(legend.position = "top")
```

```{r correlation-swap-payoff, fig.cap="Correlation swap payoff simulation. Fixed payout based on realised vs struck correlation."}
# Simulate correlation swap payoffs
set.seed(42)

# Strike correlation
strike_corr <- 0.45

# Payoff scenarios
realised_scenarios <- seq(0.1, 0.9, 0.05)
notional <- 1000000

# Payoff: (realised - strike) * notional
payoffs <- (realised_scenarios - strike_corr) * notional

corr_swap_payoff <- data.table(
    realised = realised_scenarios,
    payoff = payoffs
)

ggplot(corr_swap_payoff, aes(x = realised, y = payoff / 1000)) +
    geom_col(aes(fill = payoff > 0), alpha = 0.8, width = 0.04) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_vline(xintercept = strike_corr, linetype = "dashed", colour = tc[2]) +
    annotate("text", x = strike_corr + 0.05, y = max(payoffs/1000) * 0.8,
             label = sprintf("Strike: %.2f", strike_corr), colour = tc[2]) +
    scale_fill_manual(values = c(tc[4], tc[1]), guide = "none") +
    labs(title = "Correlation Swap Payoff Profile",
         subtitle = sprintf("Strike = %.2f; Long position profits when realised > strike", strike_corr),
         x = "Realised Correlation", y = "P&L ($000s)") +
    theme_trading()
```

```{r correlation-persistence, fig.cap="Correlation persistence analysis. High correlation tends to persist, creating regime momentum."}
# Autocorrelation of correlation
acf_data <- acf(multi$realised_corr[!is.na(multi$realised_corr)], lag.max = 60, plot = FALSE)
acf_df <- data.table(lag = 0:60, acf = acf_data$acf[1:61])

ggplot(acf_df, aes(x = lag, y = acf)) +
    geom_col(fill = tc[1], alpha = 0.7) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_hline(yintercept = c(-1, 1) * 1.96 / sqrt(nrow(multi)),
               linetype = "dashed", colour = tc[2]) +
    labs(title = "Correlation Autocorrelation Function",
         subtitle = "High persistence suggests correlation regime momentum",
         x = "Lag (Days)", y = "Autocorrelation") +
    theme_trading()
```

### 10.8.3 Mathematical Derivation

**Correlation Swap Payoff**

$$
\text{Payoff} = N \times (\rho_{RV} - K_\rho)
$$

where $\rho_{RV}$ is realised correlation and $K_\rho$ is struck correlation.

**Realised Correlation Calculation**

For $n$ assets with returns $r_i$:

$$
\rho_{RV} = \frac{2}{n(n-1)} \sum_{i<j} \text{Corr}(r_i, r_j)
$$

computed over the measurement period.

**Correlation-Variance Relationship**

Index variance relates to correlation:

$$
\sigma_I^2 = \bar{\sigma}^2 \cdot [w^2 + (1-w^2) \cdot \bar{\rho}]
$$

where $w$ represents concentration. This shows:
- When $\bar{\rho} = 0$: diversification maximized
- When $\bar{\rho} = 1$: no diversification benefit

**Fair Correlation Strike**

Under risk-neutral measure:

$$
K_\rho = E^Q[\rho_{RV}] = \frac{E^Q[\sigma_I^2] - \sum_i w_i^2 E^Q[\sigma_i^2]}{2\sum_{i<j} w_i w_j E^Q[\sigma_i \sigma_j]}
$$

### 10.8.4 Implementation & Application

```{r corr-swap-strategy, fig.cap="Correlation swap timing strategy. Short correlation when elevated; long when depressed."}
# Strategy: Short correlation when high, long when low
multi[, corr_z := (realised_corr - mean(realised_corr, na.rm = TRUE)) /
                   sd(realised_corr, na.rm = TRUE)]

multi[, corr_signal := fifelse(corr_z > 1, -1,  # Short high correlation
                                fifelse(corr_z < -0.5, 1,  # Long low correlation
                                       0))]

# P&L from correlation change
multi[, corr_change := c(NA, diff(realised_corr))]
multi[, corr_strat_ret := shift(corr_signal, 1) * corr_change]

# Cumulative
corr_strat <- multi[!is.na(corr_strat_ret)]
corr_strat[, cum_corr_strat := cumsum(corr_strat_ret)]

ggplot(corr_strat, aes(x = date, y = cum_corr_strat)) +
    geom_line(colour = tc[1], linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title = "Correlation Timing Strategy",
         subtitle = "Short when z-score > 1; Long when z-score < -0.5",
         x = NULL, y = "Cumulative P&L (Correlation Points)") +
    theme_trading()

# Statistics
cat(sprintf("\nCorrelation Strategy Statistics:\n"))
cat(sprintf("  Total P&L: %.3f correlation points\n", corr_strat[.N, cum_corr_strat]))
cat(sprintf("  Sharpe (daily): %.2f\n",
           mean(corr_strat$corr_strat_ret, na.rm = TRUE) /
           sd(corr_strat$corr_strat_ret, na.rm = TRUE) * sqrt(252)))
```

```{r dispersion-correlation-hedge, fig.cap="Combined dispersion and correlation strategy. Hedging correlation exposure improves risk-adjusted returns."}
# Combine dispersion and correlation strategies
combo <- multi[!is.na(disp_ret) & !is.na(corr_strat_ret)]

# Normalise returns for combination
combo[, disp_norm := disp_ret / sd(disp_ret, na.rm = TRUE)]
combo[, corr_norm := corr_strat_ret / sd(corr_strat_ret, na.rm = TRUE)]

# Combined: 70% dispersion, 30% correlation timing
combo[, combo_ret := 0.7 * disp_norm + 0.3 * corr_norm]

# Cumulative
combo[, cum_disp_norm := cumsum(disp_norm)]
combo[, cum_combo := cumsum(combo_ret)]

# Plot
plot_combo <- melt(combo[, .(date, Dispersion = cum_disp_norm, Combined = cum_combo)],
                   id.vars = "date", variable.name = "Strategy", value.name = "Return")

ggplot(plot_combo, aes(x = date, y = Return, colour = Strategy)) +
    geom_line(linewidth = 0.5) +
    scale_colour_manual(values = c(tc[1], tc[3])) +
    labs(title = "Dispersion vs Combined Strategy",
         subtitle = "Adding correlation timing hedges some tail risk",
         x = NULL, y = "Cumulative Return (Normalised)") +
    theme_trading() +
    theme(legend.position = "top")
```

---

## 10.9 Practical Considerations

### 10.9.1 Prose & Intuition

Volatility trading requires careful attention to practical implementation details. The theoretical strategies we've discussed face real-world frictions:

**Execution Risk**: Volatility instruments have wide bid-ask spreads, especially during stressed markets when you most need to trade. The cost of entering and exiting positions can consume much of the theoretical premium.

**Liquidity Constraints**: Options markets have limited depth, particularly for longer-dated or off-the-run strikes. Large positions can move markets against you.

**Margin Requirements**: Short volatility positions require substantial margin, which can spike during crises precisely when you need liquidity most.

**Model Risk**: All volatility strategies depend on models for pricing and hedging. Model assumptions (constant vol, no jumps, continuous trading) fail precisely during tail events.

### 10.9.2 Visual Evidence

```{r transaction-costs, fig.cap="Impact of transaction costs on VRP strategy. Reasonable costs significantly reduce returns."}
# Simulate impact of transaction costs on VRP strategy
vix <- load_vix("vix")
vix <- filter_dates(vix, start_date, end_date)
spy <- filter_dates(spy, start_date, end_date)

combined <- merge_by_date(spy[, .(date, spy_ret = c(NA, diff(log(adjusted))))],
                          vix[, .(date, vix = close)])
combined <- combined[!is.na(spy_ret)]

# VRP signal
combined[, rv_30 := frollapply(spy_ret^2, 30, function(x) sqrt(sum(x) * 252 / length(x))) * 100]
combined[, signal := fifelse(vix > rv_30 * 1.1, -1, 0)]
combined[, vix_ret := c(NA, diff(log(vix)))]

# Gross returns
combined[, gross_ret := shift(signal, 1) * (-vix_ret)]

# Apply different cost levels
cost_levels <- c(0, 0.001, 0.002, 0.005)  # 0, 10, 20, 50 bps per trade
combined[, trade := abs(signal - shift(signal, 1))]

results <- lapply(cost_levels, function(cost) {
    ret <- combined$gross_ret - combined$trade * cost
    data.table(
        cost_bps = cost * 10000,
        total_ret = sum(ret, na.rm = TRUE),
        sharpe = mean(ret, na.rm = TRUE) / sd(ret, na.rm = TRUE) * sqrt(252)
    )
})
results <- rbindlist(results)

print(results)

# Plot returns at different cost levels
combined[, net_ret_0 := gross_ret]
combined[, net_ret_10 := gross_ret - trade * 0.001]
combined[, net_ret_50 := gross_ret - trade * 0.005]

combined[, cum_0 := cumsum(fifelse(is.na(net_ret_0), 0, net_ret_0))]
combined[, cum_10 := cumsum(fifelse(is.na(net_ret_10), 0, net_ret_10))]
combined[, cum_50 := cumsum(fifelse(is.na(net_ret_50), 0, net_ret_50))]

plot_costs <- melt(combined[, .(date, `0 bps` = cum_0, `10 bps` = cum_10, `50 bps` = cum_50)],
                   id.vars = "date", variable.name = "Cost", value.name = "Return")

ggplot(plot_costs, aes(x = date, y = Return, colour = Cost)) +
    geom_line(linewidth = 0.5) +
    scale_colour_manual(values = c(tc[1], tc[2], tc[4])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Transaction Cost Impact on VRP Strategy",
         subtitle = "Higher costs erode returns; 50 bps makes strategy marginal",
         x = NULL, y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "top")
```

```{r regime-dependent-performance, fig.cap="Strategy performance varies by VIX regime. Understanding regime dependency is crucial."}
# Analyse performance by VIX regime
combined[, vix_regime := cut(vix, breaks = c(0, 15, 20, 25, 35, 100),
                             labels = c("<15", "15-20", "20-25", "25-35", ">35"))]

regime_perf <- combined[!is.na(gross_ret) & !is.na(vix_regime), .(
    mean_ret = mean(gross_ret) * 252 * 100,
    sharpe = mean(gross_ret) / sd(gross_ret) * sqrt(252),
    hit_rate = mean(gross_ret > 0) * 100,
    days = .N
), by = vix_regime]

ggplot(regime_perf, aes(x = vix_regime, y = sharpe)) +
    geom_col(fill = tc[1], alpha = 0.8) +
    geom_text(aes(label = sprintf("n=%d", days)), vjust = -0.5, size = 3) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title = "VRP Strategy Sharpe Ratio by VIX Regime",
         subtitle = "Performance varies significantly; worst in crisis regimes",
         x = "VIX Regime", y = "Sharpe Ratio") +
    theme_trading()
```

```{r tail-risk-analysis, fig.cap="Tail risk in short volatility. Left tail is severe; right tail limited by nature of short vol."}
# Distribution of returns with tail analysis
gross_rets <- combined$gross_ret[!is.na(combined$gross_ret)]

# Calculate tail statistics
var_95 <- quantile(gross_rets, 0.05)
var_99 <- quantile(gross_rets, 0.01)
cvar_95 <- mean(gross_rets[gross_rets <= var_95])

cat(sprintf("\nTail Risk Statistics:\n"))
cat(sprintf("  5%% VaR: %.2f%%\n", var_95 * 100))
cat(sprintf("  1%% VaR: %.2f%%\n", var_99 * 100))
cat(sprintf("  5%% CVaR (Expected Shortfall): %.2f%%\n", cvar_95 * 100))

# Plot with tail highlighting
ggplot(data.table(ret = gross_rets), aes(x = ret)) +
    geom_histogram(aes(y = after_stat(density)), bins = 100,
                   fill = tc[1], alpha = 0.7, colour = "white") +
    geom_vline(xintercept = var_95, colour = tc[4], linewidth = 1, linetype = "dashed") +
    geom_vline(xintercept = var_99, colour = tc[4], linewidth = 1) +
    annotate("text", x = var_95 - 0.02, y = 30, label = "5% VaR", colour = tc[4], angle = 90) +
    annotate("text", x = var_99 - 0.02, y = 30, label = "1% VaR", colour = tc[4], angle = 90) +
    scale_x_continuous(labels = scales::percent_format(), limits = c(-0.15, 0.1)) +
    labs(title = "Short Volatility Return Distribution",
         subtitle = "Heavy left tail creates significant drawdown risk",
         x = "Daily Return", y = "Density") +
    theme_trading()
```

### 10.9.3 Mathematical Derivation

**Transaction Cost Model**

Net return after costs:

$$
r_{net} = r_{gross} - c \cdot |w_t - w_{t-1}|
$$

where $c$ is cost per unit traded and $w$ is position weight.

**Break-Even Analysis**

Strategy is profitable when:

$$
E[r_{gross}] > c \cdot E[|\Delta w|]
$$

Higher turnover requires higher gross returns to be profitable.

**Tail Risk Measures**

Value at Risk (VaR):

$$
P(r \leq -VaR_\alpha) = \alpha
$$

Conditional VaR (Expected Shortfall):

$$
CVaR_\alpha = E[r | r \leq -VaR_\alpha]
$$

For short volatility, CVaR is particularly important due to heavy left tails.

**Kelly Criterion with Tail Risk**

Modified Kelly for heavy tails:

$$
f^* = \frac{\mu}{\sigma^2} \times \left(1 - \frac{\kappa - 3}{6}\right)
$$

where $\kappa$ is excess kurtosis. Higher kurtosis reduces optimal allocation.

### 10.9.4 Implementation & Application

```{r risk-management-overlay, fig.cap="Risk-managed volatility strategy. Stop-losses and position limits improve tail behaviour."}
# Implement risk management overlay
combined[, position := shift(signal, 1)]

# Risk rules:
# 1. Stop-loss: exit if daily loss > 3%
# 2. Reduce position when VIX > 30
# 3. Maximum drawdown trigger

# Running P&L for stop-loss
combined[, running_pnl := gross_ret]
combined[, hit_stoploss := running_pnl < -0.03]

# Risk-adjusted signal
combined[, risk_signal := fifelse(
    hit_stoploss | shift(hit_stoploss, 1, fill = FALSE),  # Stop-loss hit
    0,
    fifelse(vix > 30, signal * 0.5,  # Reduce in high VIX
            signal)
)]

# Apply risk-managed returns
combined[, managed_ret := shift(risk_signal, 1) * (-vix_ret)]

# Compare
combined[, cum_gross := cumsum(fifelse(is.na(gross_ret), 0, gross_ret))]
combined[, cum_managed := cumsum(fifelse(is.na(managed_ret), 0, managed_ret))]

plot_risk <- melt(combined[, .(date, Unmanaged = cum_gross, `Risk-Managed` = cum_managed)],
                  id.vars = "date", variable.name = "Strategy", value.name = "Return")

ggplot(plot_risk, aes(x = date, y = Return, colour = Strategy)) +
    geom_line(linewidth = 0.5) +
    scale_colour_manual(values = c(tc[4], tc[1])) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Risk-Managed vs Unmanaged VRP Strategy",
         subtitle = "Stop-loss at -3%; 50% position when VIX > 30",
         x = NULL, y = "Cumulative Return") +
    theme_trading() +
    theme(legend.position = "top")

# Compare statistics
cat(sprintf("\nStrategy Comparison:\n"))
cat(sprintf("  Unmanaged Return: %.1f%%\n", combined[.N, cum_gross] * 100))
cat(sprintf("  Managed Return: %.1f%%\n", combined[.N, cum_managed] * 100))
cat(sprintf("  Unmanaged Sharpe: %.2f\n", sharpe_ratio(combined$gross_ret)))
cat(sprintf("  Managed Sharpe: %.2f\n", sharpe_ratio(combined$managed_ret)))
cat(sprintf("  Unmanaged MaxDD: %.1f%%\n", max_drawdown(combined$gross_ret) * 100))
cat(sprintf("  Managed MaxDD: %.1f%%\n", max_drawdown(combined$managed_ret) * 100))
```

```{r summary-statistics, fig.cap="Chapter summary: Volatility trading offers premiums with tail risks requiring active risk management."}
# Summary table
summary_stats <- data.table(
    Strategy = c("VRP Harvesting", "Mean Reversion", "Dispersion", "Correlation Timing"),
    `Ann Return` = c("8-12%", "5-10%", "6-10%", "3-6%"),
    `Sharpe` = c("0.4-0.8", "0.3-0.6", "0.5-0.9", "0.2-0.5"),
    `Max DD` = c("-30% to -60%", "-20% to -40%", "-25% to -45%", "-15% to -25%"),
    `Key Risk` = c("Vol spikes", "Regime change", "Correlation spike", "Persistence")
)

# Print as formatted table
cat("\n==== Volatility Strategy Summary ====\n\n")
print(summary_stats)
cat("\nNote: Returns and Sharpes are illustrative ranges; actual results vary by implementation.\n")
```
