---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 21: Poisson and Count Models"
part: "Part 1: Poisson Regression"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, poisson-regression, count-data, glm, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

```{r packages, include=FALSE}
box::use(ggplot2)
box::use(data.table)
```

```{r dt-symbols, include=FALSE}
.N <- data.table::.N
.SD <- data.table::.SD
```

# Chapter 21 -- Poisson and Count Models

## Part 1: Poisson Regression

Count data are everywhere: the number of insurance claims filed per year, the number of incidents on a ship, the number of publications by a scientist. These outcomes are non-negative integers, often right-skewed, and frequently have variance that increases with the mean. Ordinary linear regression is the wrong tool -- it predicts negative counts and assumes constant variance. **Poisson regression** is the natural starting point for modelling count data.

In this chapter we derive Poisson regression from the exponential family framework developed in Chapter 19, interpret coefficients as **rate ratios**, handle exposure via offsets, and fit models to real data using `glm(..., family = poisson)`.

---

## 21.1 The Log Link as Canonical for Poisson

### 21.1.1 Review

Recall from Chapter 19 that the Poisson PMF in exponential family form is:

$$P(Y = y \mid \lambda) = \frac{1}{y!} \exp\bigl[y \log \lambda - \lambda\bigr]$$

The canonical parameter is $\theta = \log \lambda$, so the **canonical link** for the Poisson family is the **log link**:

$$g(\mu) = \log \mu$$

The inverse link is:

$$\mu = g^{-1}(\eta) = e^{\eta}$$

### 21.1.2 The Model

The Poisson regression model is:

$$Y_i \sim \text{Poisson}(\mu_i), \qquad \log \mu_i = \mathbf{x}_i^\top \boldsymbol{\beta}$$

or equivalently:

$$\mu_i = \exp(\mathbf{x}_i^\top \boldsymbol{\beta}) = \exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip})$$

Because $\mu_i = E[Y_i]$ and $\text{Var}(Y_i) = \mu_i$ for the Poisson distribution, the mean-variance relationship is fixed: **the variance equals the mean**. This is both a strength (parsimony) and a weakness (rigidity) that we address in Part 2.

---

## 21.2 Deriving the Log-Likelihood

For independent observations $y_1, \ldots, y_n$:

$$\ell(\boldsymbol{\beta}) = \sum_{i=1}^n \bigl[y_i \log \mu_i - \mu_i - \log(y_i!)\bigr]$$

Substituting $\log \mu_i = \mathbf{x}_i^\top \boldsymbol{\beta}$:

$$\ell(\boldsymbol{\beta}) = \sum_{i=1}^n \bigl[y_i \mathbf{x}_i^\top \boldsymbol{\beta} - \exp(\mathbf{x}_i^\top \boldsymbol{\beta})\bigr] + \text{const}$$

The score equations are:

$$\frac{\partial \ell}{\partial \beta_j} = \sum_{i=1}^n (y_i - \mu_i) x_{ij} = 0$$

These are the same score equations we derived for the GLM in Chapter 19, confirming that Poisson regression is a special case of the GLM framework.

---

## 21.3 Interpreting Coefficients as Rate Ratios

### 21.3.1 The Multiplicative Structure

Because $\mu_i = e^{\beta_0 + \beta_1 x_{i1} + \cdots}$, a one-unit increase in $x_j$ multiplies the mean by $e^{\beta_j}$:

$$\frac{\mu(x_j + 1)}{\mu(x_j)} = e^{\beta_j}$$

This is the **rate ratio** (or **incidence rate ratio**) -- the Poisson analogue of the odds ratio in logistic regression.

- $e^{\beta_j} > 1$: predictor increases the rate.
- $e^{\beta_j} = 1$: no effect.
- $e^{\beta_j} < 1$: predictor decreases the rate.

### 21.3.2 Example: Insurance Claims

```{r load-insurance}
data_dir <- "../../../data"
claims <- data.table$fread(file.path(data_dir, "count/insurance_claims.csv"))
str(claims)
```

```{r claims-eda, fig.cap="Distribution of insurance claims. The data are right-skewed with many zeros, as is typical for count data."}
ggplot2$ggplot(claims, ggplot2$aes(x = claims)) +
  ggplot2$geom_histogram(
    binwidth = 1, fill = "#0072B2", colour = "white", alpha = 0.8
  ) +
  ggplot2$labs(
    x = "Number of claims",
    y = "Frequency",
    title = "Distribution of insurance claims"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

```{r fit-poisson-claims}
fit_claims <- glm(claims ~ age + gender + region, data = claims, family = poisson)
summary(fit_claims)
```

```{r rate-ratios-claims}
coefs <- coef(fit_claims)
ci <- confint.default(fit_claims)

rr_table <- data.frame(
  Predictor = names(coefs),
  Coefficient = round(coefs, 4),
  Rate_Ratio = round(exp(coefs), 4),
  RR_Lower = round(exp(ci[, 1]), 4),
  RR_Upper = round(exp(ci[, 2]), 4)
)
rownames(rr_table) <- NULL
knitr::kable(rr_table, caption = "Poisson regression coefficients and rate ratios for insurance claims.")
```

---

## 21.4 Offsets: Accounting for Exposure

### 21.4.1 The Problem

Often the count depends on the amount of "exposure" -- time at risk, area surveyed, number of employees, etc. If person A was observed for 2 years and person B for 6 months, it is unfair to compare their raw counts directly. We need to model the **rate** (count per unit exposure) rather than the raw count.

### 21.4.2 The Solution: Offset

If $t_i$ is the exposure for observation $i$, we model:

$$\log \mu_i = \log t_i + \mathbf{x}_i^\top \boldsymbol{\beta}$$

The term $\log t_i$ is called an **offset** -- it enters the linear predictor with a fixed coefficient of 1. Equivalently:

$$\mu_i = t_i \cdot \exp(\mathbf{x}_i^\top \boldsymbol{\beta})$$

so $\exp(\mathbf{x}_i^\top \boldsymbol{\beta})$ is the **rate per unit exposure**.

### 21.4.3 Ships Damage Data

The classic ships damage dataset records the number of damage incidents by ship type, year of construction, and period of operation. The exposure is the aggregate months of service.

```{r ships-data}
ships <- data.table$fread(file.path(data_dir, "count/ships_damage.csv"))
str(ships)
```

```{r ships-eda, fig.cap="Incidents per ship type, showing both the raw counts and the rates per 1000 service months."}
ships_summary <- ships[service > 0, .(
  total_incidents = sum(incidents),
  total_service = sum(service),
  rate_per_1000 = sum(incidents) / sum(service) * 1000
), by = type]

ggplot2$ggplot(ships_summary, ggplot2$aes(x = reorder(type, -rate_per_1000), y = rate_per_1000)) +
  ggplot2$geom_col(fill = "#0072B2", alpha = 0.8) +
  ggplot2$labs(
    x = "Ship type",
    y = "Incidents per 1000 service months",
    title = "Ship damage rates by type"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

```{r ships-poisson}
# Only use rows with positive service
ships_pos <- ships[service > 0]

fit_ships <- glm(
  incidents ~ type + factor(year) + factor(period),
  data = ships_pos,
  family = poisson,
  offset = log(service)
)
summary(fit_ships)
```

```{r ships-rate-ratios}
coefs_ships <- coef(fit_ships)
ci_ships <- confint.default(fit_ships)

rr_ships <- data.frame(
  Predictor = names(coefs_ships),
  Rate_Ratio = round(exp(coefs_ships), 3),
  Lower = round(exp(ci_ships[, 1]), 3),
  Upper = round(exp(ci_ships[, 2]), 3)
)
rownames(rr_ships) <- NULL
knitr::kable(rr_ships, caption = "Rate ratios from the ships damage Poisson regression with offset.")
```

The offset ensures we are comparing rates, not raw counts. A rate ratio of 2.5 for ship type B means that type B ships have 2.5 times the incident rate of the reference type, after adjusting for year, period, and exposure.

---

## 21.5 Poisson Regression from Scratch

Let us implement the log-likelihood, gradient, and Newton-Raphson fitting for Poisson regression without using `glm()`.

```{r poisson-scratch}
poisson_loglik <- function(beta, X, y, offset = rep(0, length(y))) {
  eta <- X %*% beta + offset
  mu <- exp(eta)
  sum(y * eta - mu - lgamma(y + 1))
}

poisson_score <- function(beta, X, y, offset = rep(0, length(y))) {
  eta <- X %*% beta + offset
  mu <- exp(eta)
  t(X) %*% (y - mu)
}

poisson_fit <- function(X, y, offset = rep(0, length(y)), tol = 1e-8, max_iter = 50) {
  beta <- rep(0, ncol(X))

  for (iter in seq_len(max_iter)) {
    eta <- X %*% beta + offset
    mu <- as.vector(exp(eta))

    # Working weights (for Poisson with log link: w = mu)
    W <- diag(mu)

    # Working response
    z <- eta + (y - mu) / mu - offset
    z <- z + offset  # add back for WLS

    # IRLS update (equivalent to Newton-Raphson for canonical link)
    beta_new <- as.vector(solve(t(X) %*% W %*% X) %*% t(X) %*% (W %*% (eta + (y - mu) / mu)))

    if (max(abs(beta_new - beta)) < tol) {
      beta <- beta_new
      break
    }
    beta <- beta_new
  }

  eta_final <- X %*% beta + offset
  mu_final <- exp(eta_final)
  deviance <- 2 * sum(ifelse(y > 0, y * log(y / mu_final), 0) - (y - mu_final))

  list(
    coefficients = beta,
    iterations = iter,
    deviance = as.numeric(deviance),
    fitted_values = as.vector(mu_final)
  )
}

# Fit to insurance claims data
X_claims <- model.matrix(~ age + gender + region, data = claims)
y_claims <- claims$claims

fit_scratch <- poisson_fit(X_claims, y_claims)

comp_df <- data.frame(
  Parameter = colnames(X_claims),
  Scratch = round(fit_scratch$coefficients, 6),
  glm = round(coef(fit_claims), 6)
)
knitr::kable(comp_df, caption = "Poisson regression: from-scratch vs glm() coefficients.")
```

---

## 21.6 Deviance for Poisson Models

The **Poisson deviance** is:

$$D = 2 \sum_{i=1}^n \left[y_i \log\frac{y_i}{\hat{\mu}_i} - (y_i - \hat{\mu}_i)\right]$$

with the convention $0 \log 0 = 0$.

```{r deviance-check}
mu_hat <- fitted(fit_claims)
dev_terms <- ifelse(y_claims > 0, y_claims * log(y_claims / mu_hat), 0) - (y_claims - mu_hat)
dev_scratch <- 2 * sum(dev_terms)

c(
  "Deviance (scratch)" = round(dev_scratch, 4),
  "Deviance (glm)" = round(deviance(fit_claims), 4)
)
```

### 21.6.1 Goodness of Fit

For Poisson data with large $\mu_i$, the deviance is approximately $\chi^2$ with $n - p$ degrees of freedom under the true model. A deviance much larger than the degrees of freedom suggests **overdispersion** (see Part 2).

```{r gof-test}
gof_df <- data.frame(
  Measure = c("Residual deviance", "Degrees of freedom", "Ratio (deviance/df)"),
  Value = round(c(
    deviance(fit_claims),
    fit_claims$df.residual,
    deviance(fit_claims) / fit_claims$df.residual
  ), 3)
)
knitr::kable(gof_df, caption = "Goodness-of-fit summary. A ratio much greater than 1 suggests overdispersion.")
```

---

## 21.7 Diagnostic Plots

```{r poisson-diagnostics, fig.cap="Diagnostic plots for the Poisson regression. Left: residuals vs fitted (looking for patterns). Right: observed vs fitted counts."}
claims[, fitted_val := fitted(fit_claims)]
claims[, pearson_resid := residuals(fit_claims, type = "pearson")]

p1 <- ggplot2$ggplot(claims, ggplot2$aes(x = fitted_val, y = pearson_resid)) +
  ggplot2$geom_point(alpha = 0.3, colour = "#0072B2", size = 1.5) +
  ggplot2$geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2$geom_smooth(method = "loess", se = FALSE, colour = "#D55E00", linewidth = 0.8) +
  ggplot2$labs(
    x = "Fitted values",
    y = "Pearson residuals",
    title = "Residuals vs fitted"
  ) +
  ggplot2$theme_minimal(base_size = 12)

p2 <- ggplot2$ggplot(claims, ggplot2$aes(x = fitted_val, y = claims)) +
  ggplot2$geom_jitter(alpha = 0.3, colour = "#0072B2", size = 1.5, width = 0.02, Height = 0.2) +
  ggplot2$geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "#D55E00") +
  ggplot2$labs(
    x = "Fitted values",
    y = "Observed counts",
    title = "Observed vs fitted"
  ) +
  ggplot2$theme_minimal(base_size = 12)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

---

## 21.8 Predicted Rates

```{r predicted-rates, fig.cap="Predicted claim rate as a function of age, by gender."}
pred_grid <- data.table$data.table(
  Age = rep(seq(18, 70, length.out = 200), 2),
  Gender = rep(c("male", "female"), each = 200),
  region = "northeast"
)
pred_grid[, pred_rate := predict(fit_claims, newdata = pred_grid, type = "response")]

ggplot2$ggplot(pred_grid, ggplot2$aes(x = Age, y = pred_rate, colour = gender)) +
  ggplot2$geom_line(linewidth = 1.1) +
  ggplot2$labs(
    x = "Age",
    y = "Expected number of claims",
    colour = "Gender",
    title = "Predicted claim rates by age and gender"
  ) +
  ggplot2$theme_minimal(base_size = 13) +
  ggplot2$theme(legend.position = "bottom")
```

---

## 21.9 Model Comparison

```{r model-comparison-poisson}
fit_null <- glm(claims ~ 1, data = claims, family = poisson)
fit_age <- glm(claims ~ Age, data = claims, family = poisson)
fit_full <- glm(claims ~ age + gender + region, data = claims, family = poisson)

comp_models <- data.frame(
  Model = c("Null", "Age only", "Age + gender + region"),
  Deviance = round(c(deviance(fit_null), deviance(fit_age), deviance(fit_full)), 2),
  AIC = round(c(AIC(fit_null), AIC(fit_age), AIC(fit_full)), 2),
  df = c(fit_null$df.residual, fit_age$df.residual, fit_full$df.residual)
)
knitr::kable(comp_models, caption = "Poisson regression model comparison.")
```

```{r lrt-poisson}
anova(fit_age, fit_full, test = "LRT")
```

---

## 21.10 The Mean-Variance Relationship

A fundamental property of the Poisson distribution is $\text{Var}(Y) = E[Y]$. Let us check this in the data by computing the mean and variance of claims within groups.

```{r mean-var-check, fig.cap="Mean vs variance of claims by group. Under the Poisson assumption these should fall on the identity line. Points above the line indicate overdispersion."}
mv_dt <- claims[, .(
  mean_claims = mean(claims),
  var_claims = var(claims)
), by = .(age_group = cut(Age, breaks = 5), region)]

ggplot2$ggplot(mv_dt, ggplot2$aes(x = mean_claims, y = var_claims)) +
  ggplot2$geom_point(colour = "#0072B2", size = 3) +
  ggplot2$geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "#D55E00") +
  ggplot2$labs(
    x = "Group mean",
    y = "Group variance",
    title = "Mean-variance relationship (Poisson expects points on the line)"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

If the variance substantially exceeds the mean, we have **overdispersion** -- a signal that the Poisson model is too restrictive. We address this in Part 2.

---

## Communicating to Stakeholders

When presenting Poisson regression results:

- **Use rate ratios.** Say "customers aged 50-60 file claims at 1.3 times the rate of those aged 30-40" rather than quoting log-scale coefficients.
- **Explain offsets.** "We accounted for the fact that some policyholders were observed for longer than others by modelling the rate (claims per year) rather than the raw count."
- **Flag overdispersion.** If the deviance is much larger than the degrees of freedom, warn stakeholders that the standard errors may be too small and the model's confidence intervals too narrow.
- **Visualise predicted rates.** A curve showing how the expected count changes with a predictor is far more accessible than a table of coefficients.

---

## Quick Reference

| Concept | Details |
|:--------|:--------|
| **Model** | $\log \mu_i = \mathbf{x}_i^\top \boldsymbol{\beta}$ |
| **Inverse link** | $\mu_i = e^{\eta_i}$ |
| **Rate ratio** | $\text{RR}_j = e^{\beta_j}$ |
| **Offset** | $\log \mu_i = \log t_i + \mathbf{x}_i^\top \boldsymbol{\beta}$ |
| **Mean-variance** | $\text{Var}(Y) = \mu$ (Poisson assumption) |
| **Deviance** | $2\sum[y_i \log(y_i / \hat{\mu}_i) - (y_i - \hat{\mu}_i)]$ |
| **GOF check** | Deviance / df $\approx 1$ if model is adequate |
| **R syntax** | `glm(y ~ x, family = poisson)` |
| **With offset** | `glm(y ~ x, family = poisson, offset = log(t))` |

---

## Exercises

1. **Derivation.** Derive the score equations for Poisson regression from the log-likelihood. Show that they reduce to $\mathbf{X}^\top \mathbf{y} = \mathbf{X}^\top \hat{\boldsymbol{\mu}}$ (the canonical link property).

2. **Rate ratio interpretation.** In a Poisson regression for hospital admissions, the coefficient for a binary treatment variable is $\hat{\beta} = -0.25$. What is the rate ratio? Interpret it in plain language.

3. **Offset vs predictor.** Fit the ships data two ways: (a) with `log(service)` as an offset and (b) with `log(service)` as a regular predictor. Compare the coefficients. Under what condition are they equivalent?

4. **Deviance residuals.** For the insurance claims model, plot the deviance residuals against the fitted values. Compare with the Pearson residual plot. Which appears more symmetric?

5. **Prediction.** Using the ships model, predict the expected number of incidents for a type A ship built in 1970, operating in the period 1975-1979, with 2000 months of service. Provide a 95% confidence interval for the prediction.

6. **Simulation.** Generate 500 observations from a Poisson regression with $\log \mu = 1 + 0.5 x$ where $x \sim U(0, 5)$. Fit the model and verify that the estimated coefficients are close to the true values. Repeat 1000 times and examine the distribution of $\hat{\beta}_1$.
