---
title: "Mathematical Statistics with R I: Foundations"
chapter: "Chapter 8: Point Estimation"
part: "Part 3: Fisher Information, Cramér-Rao, and Sufficiency"
coverImage: 13
author: "Dereck Mezquita"
date: "`r Sys.Date()`"
tags: [statistics, mathematics, estimation, fisher-information, sufficiency, R]
published: true
comments: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
if (knitr::is_html_output()) knitr::knit_hooks$set(
  plot = function(x, options) {
    cap <- options$fig.cap
    as.character(htmltools::tag(
      "Figure", list(src = x, alt = cap, paste("\n\t", cap, "\n", sep = ""))
    ))
  }
)

knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 10,
  fig.Height = 7,
  comment = "",
  warning = FALSE,
  collapse = FALSE,
  results = 'hold'
)
```

```{r packages, include=FALSE}
box::use(ggplot2)
box::use(data.table)
```

```{r dt-symbols, include=FALSE}
.N <- data.table::.N
.SD <- data.table::.SD
```

```{r data-path, include=FALSE}
data_dir <- "../../../data"
```

# Chapter 8 -- Point Estimation

## Part 3: Fisher Information, Cramér--Rao, and Sufficiency

Parts 1 and 2 gave us *methods* for producing estimators (MOM and MLE) and *criteria* for judging them (bias, consistency, efficiency). But a fundamental question remains: **how good can an estimator possibly be?** Is there a floor below which no unbiased estimator's variance can sink?

The answer is yes, and it is given by the **Cramér--Rao lower bound (CRLB)**. The key quantity in this bound is the **Fisher information**, which measures how much a sample tells us about the unknown parameter. We then study **sufficiency** -- the idea that certain statistics capture *all* the information in the sample -- and the **Rao--Blackwell theorem**, which shows how sufficient statistics can improve estimators.

This part is more theoretical than the previous two, but the rewards are substantial: you will understand *why* MLE is efficient and *what* makes a good statistic.

---

## 8.9 Fisher Information

### 8.9.1 The Score Function

Let $f(x;\theta)$ be the density (or PMF) of a single observation. The **score function** is:

$$s(x;\theta) = \frac{\partial}{\partial\theta}\log f(x;\theta).$$

The score measures the sensitivity of the log-density to changes in $\theta$. Under regularity conditions (we can interchange differentiation and integration):

$$E[s(X;\theta)] = \int \frac{\partial}{\partial\theta}\log f(x;\theta)\cdot f(x;\theta)\,dx = \int \frac{f'(x;\theta)}{f(x;\theta)}\cdot f(x;\theta)\,dx = \frac{\partial}{\partial\theta}\int f(x;\theta)\,dx = 0.$$

The score has mean zero. Its *variance* is what matters.

### 8.9.2 Definition of Fisher Information

The **Fisher information** in a single observation is:

$$I(\theta) = \text{Var}_\theta[s(X;\theta)] = E_\theta\!\left[\left(\frac{\partial}{\partial\theta}\log f(X;\theta)\right)^{\!2}\right].$$

Under an additional regularity condition (interchange of differentiation and integration for the second derivative):

$$I(\theta) = -E_\theta\!\left[\frac{\partial^2}{\partial\theta^2}\log f(X;\theta)\right].$$

This second form is often easier to compute. It says: Fisher information equals the expected *curvature* of the log-likelihood. A sharply curved log-likelihood means the data are very informative about $\theta$.

For a sample of $n$ iid observations, the total Fisher information is $nI(\theta)$.

### 8.9.3 Computing Fisher Information: Normal($\mu$, $\sigma^2$ known)

Let $X \sim N(\mu, \sigma^2)$ with $\sigma^2$ known. The log-density is:

$$\log f(x;\mu) = -\frac{1}{2}\log(2\pi\sigma^2) - \frac{(x-\mu)^2}{2\sigma^2}.$$

**Score:** $s(x;\mu) = \frac{x - \mu}{\sigma^2}$.

**Second derivative:** $\frac{\partial^2}{\partial\mu^2}\log f = -\frac{1}{\sigma^2}$.

Hence $I(\mu) = \frac{1}{\sigma^2}$. The Fisher information does not depend on $\mu$ here. More spread (larger $\sigma^2$) means less information per observation.

### 8.9.4 Computing Fisher Information: Exponential($\lambda$)

Let $X \sim \text{Exp}(\lambda)$ with density $f(x;\lambda) = \lambda e^{-\lambda x}$.

**Log-density:** $\log f = \log\lambda - \lambda x$.

**Score:** $s(x;\lambda) = \frac{1}{\lambda} - x$.

**Second derivative:** $\frac{\partial^2}{\partial\lambda^2}\log f = -\frac{1}{\lambda^2}$.

Hence $I(\lambda) = \frac{1}{\lambda^2}$.

### 8.9.5 Computing Fisher Information: Poisson($\lambda$)

Let $X \sim \text{Poisson}(\lambda)$ with PMF $f(x;\lambda) = e^{-\lambda}\lambda^x/x!$.

**Log-PMF:** $\log f = -\lambda + x\log\lambda - \log(x!)$.

**Second derivative:** $\frac{\partial^2}{\partial\lambda^2}\log f = -\frac{x}{\lambda^2}$.

Hence $I(\lambda) = E[X/\lambda^2] = \lambda/\lambda^2 = \frac{1}{\lambda}$.

### 8.9.6 Summary Table

| Distribution | Parameter | $I(\theta)$ |
|:-------------|:----------|:------------|
| Normal($\mu$, $\sigma^2$ known) | $\mu$ | $1/\sigma^2$ |
| Normal($\mu$ known, $\sigma^2$) | $\sigma^2$ | $1/(2\sigma^4)$ |
| Exponential($\lambda$) | $\lambda$ | $1/\lambda^2$ |
| Poisson($\lambda$) | $\lambda$ | $1/\lambda$ |
| Bernoulli($p$) | $p$ | $1/[p(1-p)]$ |

---

## 8.10 The Cramér--Rao Lower Bound

### 8.10.1 Statement

**Theorem (Cramér--Rao).** Let $X_1, \ldots, X_n$ be iid with density $f(x;\theta)$ satisfying regularity conditions. If $\hat\theta$ is any unbiased estimator of $\theta$, then:

$$\text{Var}(\hat\theta) \ge \frac{1}{nI(\theta)}.$$

The right-hand side is the **Cramér--Rao lower bound (CRLB)**. An unbiased estimator that achieves this bound is called **efficient** (or a **minimum-variance unbiased estimator**, MVUE).

### 8.10.2 Proof of the Cramér--Rao Lower Bound

*Proof.* Let $T = \hat\theta(X_1, \ldots, X_n)$ be unbiased, so $E[T] = \theta$ for all $\theta$. Let $S = \sum_{i=1}^n s(X_i; \theta)$ denote the total score. We established that $E[S] = 0$ and $\text{Var}(S) = nI(\theta)$.

**Step 1.** Differentiate $E[T] = \theta$ with respect to $\theta$:

$$\frac{d}{d\theta}E[T] = \frac{d}{d\theta}\int \cdots \int T(x_1,\ldots,x_n)\prod_{i=1}^n f(x_i;\theta)\,dx_1\cdots dx_n = 1.$$

Interchanging differentiation and integration (regularity condition), the right side becomes:

$$\int \cdots \int T \cdot \frac{\partial}{\partial\theta}\prod_{i=1}^n f(x_i;\theta)\,dx_1\cdots dx_n.$$

Now observe:

$$\frac{\partial}{\partial\theta}\prod_{i=1}^n f(x_i;\theta) = \prod_{i=1}^n f(x_i;\theta) \cdot \sum_{i=1}^n \frac{f'(x_i;\theta)}{f(x_i;\theta)} = \prod_{i=1}^n f(x_i;\theta) \cdot S.$$

Hence:

$$1 = E[T \cdot S].$$

Since $E[S] = 0$ and $E[T] = \theta$, this gives $\text{Cov}(T, S) = E[TS] - E[T]E[S] = 1$.

**Step 2.** Apply the **Cauchy--Schwarz inequality**:

$$[\text{Cov}(T, S)]^2 \le \text{Var}(T) \cdot \text{Var}(S).$$

Substituting:

$$1 \le \text{Var}(T) \cdot nI(\theta).$$

Therefore:

$$\text{Var}(T) \ge \frac{1}{nI(\theta)}. \quad \square$$

### 8.10.3 When Is the Bound Achieved?

The Cauchy--Schwarz inequality becomes an equality when $T$ and $S$ are linearly related: $T = a + bS$ for constants $a$ and $b$. This happens precisely when the score can be written as:

$$S = nI(\theta)(T - \theta),$$

which occurs for exponential-family distributions. For example, $\bar X$ achieves the CRLB for the Normal mean.

### 8.10.4 Simulation: Comparing Estimator Variance to the CRLB

```{r crlb-simulation, fig.cap="Variance of the sample mean as an estimator of the Normal mean, compared to the Cramer-Rao lower bound. The sample mean achieves the bound exactly, confirming it is efficient."}
set.seed(42)
true_mu <- 10
true_sigma <- 4
sample_sizes <- seq(5, 200, by = 5)
n_reps <- 10000

observed_var <- sapply(sample_sizes, function(n) {
  means <- replicate(n_reps, mean(rnorm(n, true_mu, true_sigma)))
  var(means)
})

crlb <- true_sigma^2 / sample_sizes

crlb_dt <- data.table$data.table(
  n = rep(sample_sizes, 2),
  variance = c(observed_var, crlb),
  type = rep(c("Observed Var(Xbar)", "CRLB = sigma^2/n"), each = length(sample_sizes))
)

ggplot2$ggplot(crlb_dt, ggplot2$aes(x = n, y = variance, colour = type)) +
  ggplot2$geom_line(linewidth = 1) +
  ggplot2$geom_point(size = 1.5) +
  ggplot2$scale_colour_manual(values = c("#D55E00", "#0072B2")) +
  ggplot2$labs(
    x = "Sample size n",
    y = "Variance",
    colour = "",
    title = "Sample mean variance vs. Cramer-Rao lower bound (Normal mean)"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

The two lines coincide -- the sample mean is an *efficient* estimator of the Normal mean.

### 8.10.5 Example: An Inefficient Estimator

The sample median is unbiased for $\mu$ under Normality but its asymptotic variance is $\pi\sigma^2/(2n)$, which exceeds $\sigma^2/n$ by a factor of $\pi/2 \approx 1.57$. It does *not* achieve the CRLB.

```{r crlb-median, fig.cap="Variance of the sample mean and sample median compared to the CRLB. The median's variance exceeds the bound by a factor of approximately pi/2."}
observed_var_median <- sapply(sample_sizes, function(n) {
  medians <- replicate(n_reps, median(rnorm(n, true_mu, true_sigma)))
  var(medians)
})

crlb_med_dt <- data.table$data.table(
  n = rep(sample_sizes, 3),
  variance = c(observed_var, observed_var_median, crlb),
  type = rep(c("Var(Xbar)", "Var(Median)", "CRLB"), each = length(sample_sizes))
)

ggplot2$ggplot(crlb_med_dt, ggplot2$aes(x = n, y = variance, colour = type)) +
  ggplot2$geom_line(linewidth = 1) +
  ggplot2$scale_colour_manual(values = c("#D55E00", "#0072B2", "#009E73")) +
  ggplot2$labs(
    x = "Sample size n",
    y = "Variance",
    colour = "",
    title = "Mean, median, and CRLB for the Normal mean"
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

---

## 8.11 Sufficiency

> *This section is optional on a first reading, but the ideas are important for deeper study.*

### 8.11.1 Motivation

Consider estimating $\mu$ from $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$. Once we compute $\bar X$, the individual observations $X_1, \ldots, X_n$ contain no further information about $\mu$. The statistic $\bar X$ **summarises** the data without losing anything relevant to the parameter. Such a statistic is called **sufficient**.

### 8.11.2 Definition

A statistic $T = T(X_1, \ldots, X_n)$ is **sufficient** for $\theta$ if the conditional distribution of the sample $(X_1, \ldots, X_n)$ given $T = t$ does not depend on $\theta$.

Intuitively: once you know $T$, learning the individual data values gives you no additional information about $\theta$.

### 8.11.3 The Factorisation Theorem (Neyman--Fisher)

**Theorem.** $T(X_1, \ldots, X_n)$ is sufficient for $\theta$ if and only if the joint density (or PMF) can be written as:

$$f(x_1, \ldots, x_n; \theta) = g(T(x_1, \ldots, x_n), \theta) \cdot h(x_1, \ldots, x_n),$$

where $g$ depends on the data only through $T$, and $h$ does not depend on $\theta$.

### 8.11.4 Proof of the Factorisation Theorem (Discrete Case)

*Proof.* We prove both directions for the discrete case.

**($\Rightarrow$)** Suppose $T$ is sufficient, so $P(X = x \mid T = t)$ does not depend on $\theta$. Write:

$$P(X = x; \theta) = P(X = x \mid T(x) = t) \cdot P(T = t; \theta).$$

Here $P(T = t; \theta)$ depends on $x$ only through $T(x)$, so set $g(t, \theta) = P(T = t; \theta)$ and $h(x) = P(X = x \mid T(x) = T(x))$. The factorisation holds.

**($\Leftarrow$)** Suppose $f(x;\theta) = g(T(x), \theta)\cdot h(x)$. Then for any $x$ with $T(x) = t$:

$$P(X = x \mid T = t) = \frac{P(X = x; \theta)}{P(T = t; \theta)} = \frac{g(t, \theta)\cdot h(x)}{\sum_{y: T(y) = t} g(t, \theta)\cdot h(y)} = \frac{h(x)}{\sum_{y: T(y) = t} h(y)},$$

which does not depend on $\theta$. Hence $T$ is sufficient. $\square$

### 8.11.5 Examples of Sufficient Statistics

**Normal:** For $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$ with $\sigma^2$ known, the joint density is:

$$\prod_{i=1}^n \frac{1}{\sqrt{2\pi}\sigma}\exp\!\left(-\frac{(x_i - \mu)^2}{2\sigma^2}\right) = \underbrace{\frac{1}{(2\pi\sigma^2)^{n/2}}\exp\!\left(-\frac{n\bar x^2 - 2\mu n\bar x + n\mu^2}{2\sigma^2}\right)}_{g(\bar x, \mu)} \cdot \underbrace{\exp\!\left(-\frac{\sum x_i^2 - n\bar x^2}{2\sigma^2}\right)}_{h(x_1,\ldots,x_n)}.$$

So $\bar X$ is sufficient for $\mu$.

**Poisson:** For $X_1, \ldots, X_n \sim \text{Poisson}(\lambda)$:

$$\prod_{i=1}^n \frac{e^{-\lambda}\lambda^{x_i}}{x_i!} = \underbrace{e^{-n\lambda}\lambda^{\sum x_i}}_{g(\sum x_i, \lambda)} \cdot \underbrace{\frac{1}{\prod x_i!}}_{h(x_1,\ldots,x_n)}.$$

So $\sum X_i$ (equivalently $\bar X$) is sufficient for $\lambda$.

**Bernoulli:** For $X_1, \ldots, X_n \sim \text{Bernoulli}(p)$:

$$\prod_{i=1}^n p^{x_i}(1-p)^{1-x_i} = p^{\sum x_i}(1-p)^{n - \sum x_i} \cdot 1.$$

So $\sum X_i$ is sufficient for $p$.

---

## 8.12 The Rao--Blackwell Theorem

### 8.12.1 Statement

**Theorem (Rao--Blackwell).** Let $\hat\theta$ be any unbiased estimator of $\theta$ and let $T$ be a sufficient statistic. Define:

$$\tilde\theta = E[\hat\theta \mid T].$$

Then:

1. $\tilde\theta$ is unbiased: $E[\tilde\theta] = \theta$.
2. $\tilde\theta$ has variance no greater than $\hat\theta$: $\text{Var}(\tilde\theta) \le \text{Var}(\hat\theta)$.
3. $\tilde\theta$ is a function of $T$ alone (it does not depend on $\theta$, by sufficiency).

### 8.12.2 Proof

*Proof.*

**(1) Unbiasedness.** By the tower property of conditional expectation:

$$E[\tilde\theta] = E\big[E[\hat\theta \mid T]\big] = E[\hat\theta] = \theta.$$

**(2) Variance reduction.** By the law of total variance:

$$\text{Var}(\hat\theta) = E[\text{Var}(\hat\theta \mid T)] + \text{Var}(E[\hat\theta \mid T]) = E[\text{Var}(\hat\theta \mid T)] + \text{Var}(\tilde\theta).$$

Since $E[\text{Var}(\hat\theta \mid T)] \ge 0$:

$$\text{Var}(\hat\theta) \ge \text{Var}(\tilde\theta),$$

with equality iff $\text{Var}(\hat\theta \mid T) = 0$ a.s., i.e., $\hat\theta$ is already a function of $T$. $\square$

### 8.12.3 Interpretation

The Rao--Blackwell theorem tells us: if you have any unbiased estimator, you can always improve it (or at least not make it worse) by conditioning on a sufficient statistic. The improved estimator $\tilde\theta$ uses the data more efficiently because the sufficient statistic has already captured all the relevant information.

### 8.12.4 Example: Improving an Estimator for $p$

Suppose $X_1, \ldots, X_n \sim \text{Bernoulli}(p)$. A trivial unbiased estimator of $p$ is $\hat p = X_1$ (just look at the first observation). This is unbiased but terrible -- it ignores $n-1$ observations.

The sufficient statistic is $T = \sum_{i=1}^n X_i$. By Rao--Blackwell:

$$\tilde p = E[X_1 \mid T] = \frac{T}{n} = \bar X.$$

(Since $X_1, \ldots, X_n$ are exchangeable given $T$, each $X_i$ has conditional expectation $T/n$.)

Thus $\bar X$ is the Rao--Blackwellised version of $X_1$, and it has much smaller variance.

### 8.12.5 Simulation: Rao--Blackwell in Action

```{r rao-blackwell, fig.cap="Sampling distributions of X1 (naive estimator) and Xbar (Rao-Blackwellised estimator) for estimating p = 0.3. Both are unbiased, but Xbar concentrates far more tightly around the truth."}
set.seed(42)
n <- 30
true_p <- 0.3
n_reps <- 10000

naive <- numeric(n_reps)
rb <- numeric(n_reps)

for (i in seq_len(n_reps)) {
  x <- rbinom(n, 1, true_p)
  naive[i] <- x[1]
  rb[i] <- mean(x)
}

rb_dt <- data.table$data.table(
  value = c(naive, rb),
  estimator = rep(c("X1 (naive)", "Xbar (Rao-Blackwell)"), each = n_reps)
)

ggplot2$ggplot(rb_dt, ggplot2$aes(x = value, fill = estimator)) +
  ggplot2$geom_histogram(ggplot2$aes(y = ggplot2::after_stat(density)),
    bins = 40, alpha = 0.5, position = "identity") +
  ggplot2$geom_vline(xintercept = true_p, linetype = "dashed",
    colour = "red", linewidth = 0.8) +
  ggplot2$labs(
    x = "Estimate of p",
    y = "Density",
    fill = "Estimator",
    title = sprintf("Naive vs. Rao-Blackwell estimator (p = %.1f, n = %d)", true_p, n)
  ) +
  ggplot2$theme_minimal(base_size = 13)
```

```{r rao-blackwell-summary}
data.frame(
  estimator = c("X1 (naive)", "Xbar (Rao-Blackwell)"),
  mean = round(c(mean(naive), mean(rb)), 4),
  variance = round(c(var(naive), var(rb)), 5),
  theoretical_var = round(c(true_p * (1 - true_p), true_p * (1 - true_p) / n), 5)
)
```

The variance reduction is dramatic: from $p(1-p) = 0.21$ to $p(1-p)/n = 0.007$, a 30-fold improvement.

---

## 8.13 Putting It All Together

The ideas in this part form a coherent story:

1. **Fisher information** quantifies how much data tell us about the parameter.
2. The **Cramér--Rao lower bound** sets a floor for the variance of any unbiased estimator.
3. **Sufficient statistics** capture all the information without loss.
4. The **Rao--Blackwell theorem** shows that conditioning on a sufficient statistic can only improve an estimator.
5. The **MLE** (from Part 2) achieves the CRLB asymptotically, which is why it is the dominant estimation method.

```{r big-picture, fig.cap="Conceptual map of estimation theory. Fisher information underpins the CRLB, which sets the efficiency benchmark. Sufficient statistics and the Rao-Blackwell theorem provide tools for constructing efficient estimators. The MLE sits at the intersection of all these ideas.", echo=FALSE}
concepts <- data.table$data.table(
  x = c(0, -1.5, 1.5, 0, 0),
  y = c(2, 0, 0, -1.5, -3),
  label = c(
    "Fisher Information\nI(theta)",
    "Cramer-Rao\nLower Bound",
    "Sufficient\nStatistics",
    "Rao-Blackwell\nTheorem",
    "MLE\n(asymptotically efficient)"
  )
)

arrows <- data.table$data.table(
  x = c(0, 0, -1.5, 1.5, -1.5, 1.5),
  y = c(2, 2, 0, 0, 0, 0),
  xend = c(-1.5, 1.5, 0, 0, 0, 0),
  yend = c(0, 0, -1.5, -1.5, -3, -3)
)

ggplot2$ggplot() +
  ggplot2$geom_segment(
    data = arrows,
    ggplot2$aes(x = x, y = y, xend = xend, yend = yend),
    arrow = ggplot2$arrow(length = ggplot2$unit(0.15, "inches")),
    colour = "#555555", linewidth = 0.8
  ) +
  ggplot2$geom_label(
    data = concepts,
    ggplot2$aes(x = x, y = y, label = label),
    size = 4, fill = "#E8F0FE", label.size = 0.3
  ) +
  ggplot2$coord_cartesian(xlim = c(-3, 3), ylim = c(-4, 3)) +
  ggplot2$theme_void(base_size = 13) +
  ggplot2$labs(title = "Conceptual map of estimation theory")
```

---

## Communicating to Stakeholders

- Replace "Fisher information" with "the precision of our measurement method." More informative data means more precise estimates.
- Explain the Cramér--Rao bound as: "There is a theoretical limit to how precise any method can be. Our chosen method is close to that limit, so we can be confident we are not leaving precision on the table."
- Sufficiency can be communicated as: "We have distilled the entire data set into a single summary number without losing anything that matters for our question."
- The Rao--Blackwell theorem translates to: "By using a smarter summary of the data, we can improve the precision of our estimate without collecting any new data."

---

## Quick Reference

| Concept | Formula / Description |
|:--------|:----------------------|
| **Score** | $s(x;\theta) = \frac{\partial}{\partial\theta}\log f(x;\theta)$, has $E[s] = 0$ |
| **Fisher information** | $I(\theta) = E[s^2] = -E[\ell''(\theta)]$ |
| **CRLB** | $\text{Var}(\hat\theta) \ge \frac{1}{nI(\theta)}$ for unbiased $\hat\theta$ |
| **Efficient estimator** | Achieves the CRLB |
| **Sufficient statistic** | $T$ such that $f(x;\theta) = g(T(x),\theta)\cdot h(x)$ |
| **Rao--Blackwell** | $\tilde\theta = E[\hat\theta \mid T]$ has $\text{Var}(\tilde\theta) \le \text{Var}(\hat\theta)$ |

| Distribution | Sufficient statistic | Fisher information |
|:-------------|:--------------------|:-------------------|
| Normal($\mu$, $\sigma^2$ known) | $\bar X$ | $1/\sigma^2$ |
| Exponential($\lambda$) | $\sum X_i$ | $1/\lambda^2$ |
| Poisson($\lambda$) | $\sum X_i$ | $1/\lambda$ |
| Bernoulli($p$) | $\sum X_i$ | $1/[p(1-p)]$ |

---

## Exercises

**Conceptual**

1. In your own words, explain why the Fisher information measures the "curvature" of the log-likelihood. What does high curvature mean for estimation?

2. A flat log-likelihood implies low Fisher information. What does this mean practically for trying to estimate the parameter?

3. Give an intuitive explanation of sufficiency using the bag-of-marbles analogy: you have a bag containing red and blue marbles, and you draw $n$ with replacement.

4. Can a biased estimator have variance below the CRLB? Why or why not? (Careful -- the CRLB applies only to unbiased estimators.)

**Proofs and Derivations**

5. Compute the Fisher information for $p$ in the Bernoulli distribution and verify that $\text{Var}(\bar X) = p(1-p)/n$ equals the CRLB.

6. Show that $\sum X_i^2$ is **not** sufficient for $\mu$ in the Normal distribution (with $\sigma^2$ known). (Hint: attempt the factorisation and show it fails.)

7. Use the Rao--Blackwell theorem to improve the estimator $\hat\lambda = 1/X_1$ for the Exponential rate parameter. What is $E[1/X_1 \mid \sum X_i]$? (This is harder; a direct computation via the conditional distribution is needed.)

8. Prove that the MLE achieves the CRLB for the Bernoulli distribution by showing that $\text{Var}(\bar X) = 1/(nI(p))$.

**Computational**

9. Simulate 10,000 samples of size $n = 25$ from an $\text{Exp}(\lambda = 2)$ distribution. For each sample, compute $\hat\lambda = 1/\bar X$. Compare the observed variance of the MLE to the CRLB ($\lambda^2/n$). Is the MLE efficient for finite $n$?

10. Using the NHANES data, fit a Normal to BMI and compute the observed Fisher information matrix (the Hessian of the log-likelihood at the MLE). Use this to construct approximate 95% confidence intervals for $\mu$ and $\sigma$.

11. Generate Bernoulli(0.4) data with $n = 50$. Compare the variance of $X_1$ (naive estimator) and $\bar X$ (Rao--Blackwellised) across 10,000 replications. Verify the theoretical variance ratio is $n$.
